```{r}
#| include: false
source("_common.R")
```

# Cleaning Data {#sec-clean}

An auditor asks a works division for its statement of payments, and receives a spreadsheet. It is meant for printing, not for analysis. The column names have spaces and symbols. There are title rows at the top, blank rows in between, and a total row at the bottom. Amounts are written as "1,25,000/-". Dates are typed in four different ways. A missing amount is written as "-", "N/A" or "Nil". The same firm appears as "Sharma Traders" and " sharma traders ". If we analyse such a file as it is, totals will be wrong, the same firm will be counted twice, and some payments will drop out without any warning.

So before any audit check, the data must be cleaned. This is often the longest step of the whole analysis. It is also an audit step in its own right. A missing amount, or a firm entered under two names, is a fact about the quality of the auditee's records, and may be a finding by itself.

Several packages in R help with this. In this chapter we will use `janitor`, along with `readr`, `stringr` and `lubridate`. We first learn each tool on a small piece of office data. We then put them together to clean a messy statement of payments received from a division, and record what the cleaning revealed.

**Prerequisites**

```{r}
#| message: false
#| warning: false
library(tidyverse)
library(janitor)
```

## Cleaning column names

Names of objects in R follow certain rules. For example, a name with a space or a special character must be wrapped in backticks \`. Files from Excel often have such "dirty" column names, which we should clean before going further. The function `clean_names()` does this. Let us take the headings of a statement of works, as an office might type them.

```{r}
# Create a data.frame with dirty names
test_df <- as.data.frame(matrix(ncol = 7))

names(test_df) <- c("Work ID", "Name of Work", "Estimate (Rs. lakh)", "% Completion",
                    "REMARKS", "REMARKS", "")
# View this data
test_df
```

Using `clean_names()`, which works well in a pipe, we can clean all the names in one step. The results are in *snake case*, with lower case words joined by underscores.

```{r}
test_df |>
  clean_names()
```

`clean_names()` does the following.

+ Brings letter case and separators to one consistent format.
+ Uses `snake_case` by default, but other cases like `camelCase` are available.
+ Handles special characters and spaces, including transliterating characters like `œ` to `oe`.
+ Adds numbers to duplicated names, as it did for the two `REMARKS` columns.
+ Converts `"%"` to "percent" and `"#"` to "number", to keep their meaning.
+ Gives a name to a blank column, like `x`.

## Handling duplicate records

`janitor` has a ready function, `get_dupes()`, which finds records that are duplicates on the columns we name. It also works well in a pipe. Let us take a few bills of a division, and look for bills with the same firm and the same bill number.

```{r}
bills <- tibble(
  voucher_no = c("PV/24/011", "PV/24/012", "PV/24/013", "PV/24/014", "PV/24/015", "PV/24/016"),
  firm       = c("Sharma Traders", "Kaveri Infra", "Sharma Traders", "Patel Hardware",
                 "Kaveri Infra", "Om Sai Suppliers"),
  bill_no    = c("ST/118", "KI/2024/07", "ST/121", "PH-33", "KI/2024/07", "OS/09"),
  amount     = c(48500, 312000, 26750, 9800, 312000, 61200)
)

bills |>
  get_dupes(firm, bill_no)
```

It returns all copies of the duplicate records, with an extra column `dupe_count`, so that they can be examined together. Here bill `KI/2024/07` of Kaveri Infra has been paid under two vouchers. Duplicate detection is covered in detail in @sec-dup.

## Removing empty rows and columns

Files exported from Excel often have empty rows, used as spacers, and empty columns. `remove_empty()` removes rows and columns which are entirely empty.

```{r}
with_blanks <- tibble(voucher = c("PV/24/001", NA, "PV/24/002"),
                      blank   = c(NA, NA, NA),
                      amount  = c(18500, NA, 42300))
with_blanks

with_blanks |> remove_empty(c("rows", "cols"))
```

## Removing constant columns

`remove_constant()` drops columns which have the same value in every row. Such columns, like a column holding the name of the office in every row, carry no information for analysis.

```{r}
tibble(voucher = 1:3, office = "Amarpur", amount = c(500, 800, 1200)) |>
  remove_constant()
```

## Fixing Excel dates stored as serial numbers

When an Excel file is read into R, a date sometimes appears as a number like `45383` instead. Excel stores dates as the number of days since its starting date, and the format is sometimes lost. `excel_numeric_to_date()` converts such numbers to dates. Let us see this.

```{r}
excel_numeric_to_date(45383)
```

The number `45383` was the first day of the financial year 2024-25.

## Converting a mix of date formats

Similarly, a column with a mix of Excel serial numbers and dates can be converted with `convert_to_date()` or `convert_to_datetime()`. See this example.

```{r}
unsorted_dates <- c('2024-05-31', '45383', 45383)
convert_to_date(unsorted_dates)
```

::: callout-note
The vector above mixes text and a number, but R's coercion rules have already turned all of them into text.
:::

When data is entered on many computers, or comes from many offices, a date column may mix several formats. The function `parse_date_time()` from `lubridate` can read such a column. The argument `orders` lists the possible orders of day, month and year, and the function tries each of them. Let us take the dates of payment reported by six offices.

```{r}
mixed_dates <- c("13-11-2024", "13-Sep-24",
                 "20 August 2024", "15 August 24",
                 "03/31/25", "12-31-2024")

lubridate::parse_date_time(mixed_dates,
                           orders = c("d m y", "d B Y", "m/d/y", "d B y"),
                           locale = "eng")
```

::: callout-warning
## Caution, ambiguous dates

A date like `03/04/2024` can be the 3rd of April (Indian style, day first) or the 4th of March (American style, month first). No function can tell which was meant. If both orders are allowed, `parse_date_time()` simply takes whichever of them it tries first. Before converting, find out how each source writes dates. Where possible, check a few dates in the converted data against the documents. A day of 13 or more is never a month, which helps to identify the style of a source.
:::

## Missing values written in many ways

Missing values are written in many ways, like an empty cell, `NA`, `-`, `N/A`, `Nil` or `0`. R treats only a true `NA` as missing. The others are text, or worse, a real zero. Two tools help.

- While reading a file, the argument `na` of `read_csv()` lists the values to be read as missing, like `na = c("", "NA", "-", "N/A", "Nil")`.
- After reading, `na_if()` from `dplyr` turns a given value into `NA`, like `mutate(remarks = na_if(remarks, "Nil"))`.

```{r}
tibble(status = c("Paid", "-", "Nil", "Paid")) |>
  mutate(status = na_if(status, "-"),
         status = na_if(status, "Nil"))
```

::: callout-warning
## Caution

Be careful with zero. A zero in an amount column may be a real zero, or a missing value entered as zero. Find out which, from the source, before treating zeros as missing.
:::

## Audit use case, cleaning a file received from an office {#sec-clean-audit}

Let us put everything together. An office has sent its statement of payments for April 2024, prepared in a spreadsheet and saved as csv. Here is how it looks, line by line. We write it to a temporary file, as if it had come by email.

```{r}
messy <- 'Office of the Executive Engineer, PWD Division Amarpur
Statement of payments for April 2024

S.No.,Voucher No.,Name of Vendor,Amount (Rs.),Date of Payment,Remarks,Remarks
1,PV/24/001, Sharma Traders ,"1,25,000/-",01/04/2024,,
2,PV/24/002,GUPTA & SONS,"45,600",02-04-2024,,
,,,,,,
3,PV/24/003,Kaveri Infra,N/A,05 Apr 2024,Cancelled,
4,PV/24/004,sharma traders,"98,000/-",2024-04-08,,
5,PV/24/005,Om Sai Suppliers,-,10/04/2024,,
Total,,,,,,'

messy_file <- file.path(tempdir(), "april_payments.csv")
writeLines(messy, messy_file)
```

It has every problem we have discussed. There are two title lines and a blank line above the column names. The column names have spaces and symbols, and two columns have the same name. There is a blank row, and a total row at the bottom. Vendor names are in different cases and have extra spaces. Amounts have commas and "/-", missing amounts are written as "N/A" and "-", and dates are in four formats.

**Step 1, read everything as text, skipping the title lines.** Reading as text ensures that nothing is lost or wrongly converted at this stage.

```{r}
raw <- read_csv(messy_file, skip = 3,
                col_types = cols(.default = col_character()))
raw
```

`read_csv()` has renamed the two `Remarks` columns to keep them apart, and told us so.

**Step 2, clean the structure.** Clean the names, remove empty rows and columns, and drop the total row.

```{r}
step2 <- raw |>
  clean_names() |>
  remove_empty(c("rows", "cols")) |>
  filter_out(s_no == "Total")
step2
```

Notice that we used `filter_out()`, which we learn in @sec-DPLYRR. It drops only the rows which are certainly the total row.

**Step 3, clean the values.**

```{r}
payments <- step2 |>
  mutate(vendor = str_squish(str_to_upper(name_of_vendor)),
         amount = parse_number(amount_rs, na = c("-", "N/A")),
         date   = as.Date(parse_date_time(date_of_payment,
                                          orders = c("dmy", "ymd", "d b Y")))) |>
  select(voucher_no, vendor, amount, date, remarks = remarks_6)
payments
```

All dates are now real dates, all amounts real numbers, and the two forms of *Sharma Traders* have become the same.

**Step 4, record what was found.** The cleaning itself has revealed facts which belong in the working papers.

```{r}
found <- payments |>
  summarise(payments          = n(),
            amount_missing    = sum(is.na(amount)),
            total_amount      = sum(amount, na.rm = TRUE),
            cancelled         = sum(remarks %in% "Cancelled"))
found
```

```{r}
#| include: false
unexplained <- payments |> filter(is.na(amount), !remarks %in% "Cancelled")
```

The statement has `r found$payments` payments. The amount of `r found$amount_missing` of them is not given. One of these is marked cancelled, but the other, voucher `r unexplained$voucher_no`, is not. That payment needs an explanation. And the statement's own total row was empty, so the total of ₹`r round(found$total_amount / 1e5, 2)` lakh must be reconciled with the cash book.

What would the auditor do next? Ask the division for voucher `r unexplained$voucher_no` and its amount. See the cancelled voucher, to confirm that no payment was made against it. Check in the cash book that the payments to Sharma Traders, written in two ways, are to one and the same firm. Then note all of this in the working papers, along with the code that cleaned the file.

::: callout-tip
## Audit tip

Never clean the original file. Keep it unchanged, and do all cleaning in code, as we did here. The code is then a complete record of every change made to the data, which another auditor can read and repeat. Record in the working papers the problems found while cleaning, like missing amounts and inconsistent names. They are often findings in themselves, about the quality of the auditee's records.
:::

## A suggested workflow

For any file received from an office, an auditor may follow these steps.

1.  **Keep the original unchanged.** Do all cleaning in code, so that every change is on record and can be repeated.
2.  **Read everything as text.** Skip the title lines with `skip`, and read all columns as text so that nothing is lost at this stage.
3.  **Clean the structure.** Use `clean_names()`, `remove_empty()` and `remove_constant()`, and drop the total row with `filter_out()`.
4.  **Clean the values.** Standardise names with `str_to_upper()` and `str_squish()`, and convert amounts with `parse_number()`.
5.  **Fix the dates.** Use `parse_date_time()` with the right `orders`, or `excel_numeric_to_date()` for Excel serial numbers. Find out first whether each source writes the day or the month first.
6.  **Treat missing values properly.** Turn placeholders like "-", "N/A" and "Nil" into `NA`. Find out from the source whether a zero is a real zero.
7.  **Look for duplicates.** Use `get_dupes()` on the columns that should identify a record.
8.  **Record what you found.** Note missing amounts, inconsistent names and other problems in the working papers.

::: callout-important
## Remember

Never clean the original file by hand. Clean it in code, so that the code itself is the record of every change. Problems found while cleaning are often findings in themselves, about the quality of the auditee's records.
:::

## Summary of functions used

| Function | Package | What it does |
|-----------------------|----------------|--------------------------------|
| `clean_names()` | janitor | Makes column names consistent and easy to use |
| `get_dupes()` | janitor | Shows all copies of duplicate records |
| `remove_empty()`, `remove_constant()` | janitor | Drop empty rows and columns, and constant columns |
| `excel_numeric_to_date()`, `convert_to_date()` | janitor | Convert Excel serial numbers to dates |
| `parse_date_time()` | lubridate | Read dates written in several formats |
| `na_if()` | dplyr | Turn a placeholder value into `NA` |
| `filter_out()`, `mutate()`, `select()`, `summarise()` | dplyr | Drop rows, add or change columns, keep columns, and summarise |
| `read_csv()`, `parse_number()` | readr | Read a csv file, and take numbers from text, ignoring commas and symbols |
| `str_squish()`, `str_to_upper()` | stringr | Standardise text, like vendor names |
| `writeLines()`, `tempdir()`, `file.path()` | base R | Write text to a file in a temporary folder |

: Functions used in this chapter {#tbl-cleanfunctions}
