```{r}
#| include: false
source("_common.R")
```

# Data Cleaning in R
Data cleansing is one of the important steps in data analysis. Multiple packages are available in r to clean the data sets.  One of such packages is `janitor` which we will be using in this chapter along with few other packages.  

Let's load it
```{r}
library(janitor)
```

## Cleaning Column names.
We know that names of objects in R follow certain conventions like we may not have certain special characters in names.  If a space has been used that is to be quoted under a pair of backticks \`.  But generally when we read data from files in excel, we can have some 'dirty' names, which we should clean before proceeding.  In such `clean_names()` come handy.  E.g.
```{r}
# Create a data.frame with dirty names
test_df <- as.data.frame(matrix(ncol = 6))

names(test_df) <- c("firstName", "ábc@!*", "% successful (2009)",
                    "REPEAT VALUE", "REPEAT VALUE", "")
# View this data
test_df
```

Using `clean_names()` which is also pipe friendly, we can clean names in one step.  (Results will be in snake case)
```{r}
test_df %>% 
  clean_names()
```

It -

+ Parses letter cases and separators to a consistent format.
+ Default is to `snake_case`, but other cases like `camelCase` are available
+ Handles special characters and spaces, including transliterating characters like `œ` to `oe`.
+ Appends numbers to duplicated names
+ Converts `“%”` to “percent” and `“#”` to `“number”` to retain meaning
+ Spacing (or lack thereof) around numbers is preserved

## Handling duplicate records
In `janitor` package, we have a ready to use function `get_dupes()`. It allows us to find “similar” observations in a data set based on certain characteristics.  Syntax is pretty simple, and function is pipe friendly too.  Suppose we have to find out duplicate in `mtcars` dataset on each combination of `wt` and `cyl`.  
```{r}
mtcars %>% 
  get_dupes(wt, cyl)
```
We can see that it returns all duplicate records with an additional column `dupe_count` so that these duplicates can be analysed separately.

## Remove Constant (Redundant) columns
Dropping columns from a `data.frame` that contain only a single constant value throughout is again easy through `janitor::remove_constant()`.

## Remove empty rows and/or columns
While importing messy data from excel files, we may get some empty rows and/or columns.  Sorting out this issue, is easy using `janitor::remove_empty()`.

## Fix excel dates stored as serial numbers
While loading excel files in R, we may have sometimes noticed `41590` instead of having a `date format`.  Sorting out this issue is again easy in `janitor` as we have a function `excel_numeric_to_date()` for this. Example
```{r}
janitor::excel_numeric_to_date(41590)
```

## Convert a mix of date and datetime formats to date
Similar to above, we can also sort out, if we have a column mix of different date formats, using `janitor::convert_to_date()` or `janitor::convert_to_datetime()`.  See Examples-
```{r}
unsorted_dates <- c('2018-05-31', '41590', 41590)
janitor::convert_to_date(unsorted_dates)
```
**Note in above example, we have created a heterogeneous vector, but implicit coercion rules of R have converted all forms to character only.**

In real world examples, where data is entered through multiple machines/data points simultaneously, we may a column mix of date formats.  In that case, we may use `parse_date_time()` function in `lubridate` package.  To allow different formats we have use `order` agument in this function.  Example

```{r}
mixed_dates <- c("13-11-1991", "13-Sep-22", 
                 "20 August 2000", "15 August 87", 
                 "03/31/23", "12-31-2022")

lubridate::parse_date_time(mixed_dates,
                           orders = c("d m y", "d B Y", "m/d/y", "d B y"),
                           locale = "eng")
```


