# Part-III: Probability and Sampling in R {-}

# Probability in R
We will keep short here.  Instead of learning all the concepts of probability, we will see how to calculate probability, densities, quantiles for nearly any type of distribution.  R's powerhorse has four types of functions for each of the distributions associated called `pqdr` functions.  Actually all these are prefixes.  Consider a probability function $P(X=x) = p$ for a variable $x$ and $p$ be the associated probability.  

| Distribution                   | P |  Q | D | R  |
| ------------------------------ | --------- | ------ | ----- | -----|
| Beta                           | pbeta     | qbeta | dbeta | rbeta |
| Binomial                       | pbinom    | qbinom | dbinom | rbinom |
| Cauchy                         | pcauchy   | qcauchy | dcauchy | rcauchy |
| Chi-Square                     | pchisq    | qchisq | dchisq | rchisq |
| Exponential                    | pexp      | qexp | dexp | rexp |
| F                              | pf        | qf | df | rf |
| Gamma                          | pgamma    | qgamma | dgamma | rgamma |
| Geometric                      | pgeom     | qgeom | dgeom | rgeom |
| Hypergeometric                 | phyper    | qhyper | dhyper | rhyper |
| Logistic                       | plogis    | qlogis | dlogis | rlogis |
| Log Normal                     | plnorm    | qlnorm | dlnorm | rlnorm |
| Negative Binomial              | pnbinom   | qnbinom | dnbinom | rnbinom |
| Normal                         | pnorm     | qnorm | dnorm | rnorm |
| Poisson                        | ppois     | qpois | dpois | rpois |
| Student t                      | pt        | qt | dt | rt |
| Studentized Range              | ptukey    | qtukey | dtukey | rtukey |
| Uniform                        | punif     | qunif | dunif | runif |
| Weibull                        | pweibull  | qweibull | dweibull | rweibull |
| Wilcoxon Rank Sum Statistic    | pwilcox   | qwilcox | dwilcox | rwilcox |
| Wilcoxon Signed Rank Statistic | psignrank | qsignrank | dsignrank | rsignrank |

All these functions are vectorised. Let us explore these one by one.


## `p*()` set of functions
These set of functions give the cumulative **p**robability distribution of that probability function.  

Example-1.  What is the probability of a number being less than or equal to `25` in `Normal` distribution with `mean = 50` and `sd = 10`.

```{r}
pnorm(25, mean = 50, sd = 10)
```

On the contrary, the probability of a number being greater than or equal to 25 in the above distribution is-
```{r}
# Either deduct probability from 1 
1 - pnorm(25, mean = 50, sd = 10)
# Or provide FALSE to lower.tail argument
pnorm(25, mean = 50, sd = 10, lower.tail = FALSE)
```

Example-2: What is the probability of one or more heads out of two tosses of a fair coin (binomial distribution with `p = 0.5`).

```{r}
pbinom(1, size = 2, p = 0.5)
```


## `q*()` set of functions
These set of functions, give **q**uantile which is the inverse of cumulative probability function. So if $f$ is cdf (cumulative distribution function) of a given probability distribution then $F$ the quantile is inverse of `f` i.e. $F = f^{-1}$.  These are related by

\begin{equation} 
p = f(x)
(\#eq:s1)
\end{equation} 

\begin{equation} 
x = F(x) = f^{-1}(x)
(\#eq:s2)
\end{equation} 

Example- In the above same normal distribution (`mean = 50` and `sd = 10`) What is number below which 90% of population will be distributed.

```{r}
qnorm(0.9, mean = 50, sd = 10)
```

Similar to `cdf` here we may use `lower.tail` argument to find the number above which a population percent is distributed.
```{r}
qnorm(0.9, mean = 50, sd = 10, lower.tail = FALSE)
```

## `d*()` set of functions
We saw that `p` group denotes `cdf`, `q` group denotes `inverse cdf`, but `d` group actually denotes probability **d**ensity function of a given distribution.  Simply stating, this returns the height of probability distribution function for a given x value.

So what is expected probability of drawing exactly 2 heads out of two tosses of a single fair coin (i.e. from a binomial distribution with probability `p = 0.5`).

```{r}
dbinom(2, 2, prob = 0.5)
```

## `r*()` set of functions
These set of functions are used to generate **r**andom numbers from a Statistical distribution.  So to generate `10` random numbers from Normal distribution with `mean = 50` and `sd = 10`, we can use `rnorm`.

```{r}
rnorm(10, mean = 50, sd = 10)
```

We can actually check this using histogram.
```{r fig.align='center', fig.show='hold', fig.cap="Histogram of Random numbers generated out of Normal distribution", out.width="75%"}
set.seed(1234)
hist(rnorm(10000, 50, 10), breaks = 50)
```



