```{r}
#| include: false
source("_common.R")
```


# Probability Distributions {#sec-probability}

We will keep this short. Instead of learning all the concepts of probability, we will see how to calculate probabilities, densities, quantiles and random numbers for nearly any distribution. Base R has four functions for each common distribution. Their names start with the prefixes `p`, `q`, `d` and `r`, followed by a short name of the distribution, like `norm` for the normal distribution or `binom` for the binomial.

| Distribution                   | P |  Q | D | R  |
| ------------------------------ | --------- | ------ | ----- | -----|
| Beta                           | pbeta     | qbeta | dbeta | rbeta |
| Binomial                       | pbinom    | qbinom | dbinom | rbinom |
| Cauchy                         | pcauchy   | qcauchy | dcauchy | rcauchy |
| Chi-Square                     | pchisq    | qchisq | dchisq | rchisq |
| Exponential                    | pexp      | qexp | dexp | rexp |
| F                              | pf        | qf | df | rf |
| Gamma                          | pgamma    | qgamma | dgamma | rgamma |
| Geometric                      | pgeom     | qgeom | dgeom | rgeom |
| Hypergeometric                 | phyper    | qhyper | dhyper | rhyper |
| Logistic                       | plogis    | qlogis | dlogis | rlogis |
| Log Normal                     | plnorm    | qlnorm | dlnorm | rlnorm |
| Negative Binomial              | pnbinom   | qnbinom | dnbinom | rnbinom |
| Normal                         | pnorm     | qnorm | dnorm | rnorm |
| Poisson                        | ppois     | qpois | dpois | rpois |
| Student t                      | pt        | qt | dt | rt |
| Studentized Range              | ptukey    | qtukey | dtukey | rtukey |
| Uniform                        | punif     | qunif | dunif | runif |
| Weibull                        | pweibull  | qweibull | dweibull | rweibull |
| Wilcoxon Rank Sum Statistic    | pwilcox   | qwilcox | dwilcox | rwilcox |
| Wilcoxon Signed Rank Statistic | psignrank | qsignrank | dsignrank | rsignrank |

: The p, q, d and r functions of common distributions {#tbl-pqdr}

All these functions are vectorised. Let us explore them one by one.

## `p*()` set of functions

These functions give the cumulative **p**robability, that is, the probability that a value is *less than or equal to* a given number.

Example-1. What is the probability of a value being less than or equal to `25` in a normal distribution with `mean = 50` and `sd = 10`?

```{r}
pnorm(25, mean = 50, sd = 10)
```

On the contrary, the probability of a value being greater than 25 in the same distribution is

```{r}
# Either deduct probability from 1
1 - pnorm(25, mean = 50, sd = 10)
# Or provide FALSE to lower.tail argument
pnorm(25, mean = 50, sd = 10, lower.tail = FALSE)
```

Example-2. What is the probability of getting one or more heads in two tosses of a fair coin? The number of heads follows a binomial distribution with `size = 2` and `prob = 0.5`. "One or more" is the same as "more than zero", so we ask for the upper tail above 0.

```{r}
pbinom(0, size = 2, prob = 0.5, lower.tail = FALSE)
```

::: callout-warning
## Caution

Be careful with discrete distributions. `pbinom(1, ...)` would give the probability of *at most* one head, which is a different question. Here it happens to give the same number, 0.75, by coincidence of this small example, but in general the two differ.
:::

## `q*()` set of functions

These functions give the **q**uantile, which is the inverse of the cumulative probability. If $F$ is the cumulative distribution function, which turns a value $x$ into a probability $p$, then the quantile function $F^{-1}$ turns a probability back into a value.

$$
p = F(x)
$$ {#eq-s1}

$$
x = F^{-1}(p)
$$ {#eq-s2}

Example. In the same normal distribution (`mean = 50` and `sd = 10`), below which value do 90% of the values lie?

```{r}
qnorm(0.9, mean = 50, sd = 10)
```

As with the `p` functions, we may use the `lower.tail` argument to find the value *above* which 90% of the values lie.

```{r}
qnorm(0.9, mean = 50, sd = 10, lower.tail = FALSE)
```

## `d*()` set of functions

The `p` functions give cumulative probabilities and the `q` functions their inverse. The `d` functions give the **d**ensity, the height of the distribution at a given value. For a *discrete* distribution like the binomial, this height is simply the probability of exactly that value.

So what is the probability of getting exactly 2 heads in two tosses of a fair coin?

```{r}
dbinom(2, size = 2, prob = 0.5)
```

For a *continuous* distribution like the normal, the density is not a probability itself. Probabilities come from areas under the curve, which the `p` functions give.

## `r*()` set of functions

These functions generate **r**andom numbers from a distribution. So to generate `10` random numbers from a normal distribution with `mean = 50` and `sd = 10`, we use `rnorm()`. Setting a seed with `set.seed()` makes the result reproducible.

```{r}
set.seed(2024)
rnorm(10, mean = 50, sd = 10)
```

We can check the shape of the distribution with a histogram of many such numbers (@fig-rnormhist).

```{r}
#| label: fig-rnormhist
#| fig-cap: "Histogram of random numbers generated from a normal distribution"
#| fig-align: center
#| out-width: "75%"
set.seed(1234)
hist(rnorm(10000, 50, 10), breaks = 50, main = NULL, xlab = "Value")
```

The `r` functions are how we create the simulated data used throughout this book, like the payments and claims with planted irregularities.

## Probability in audit sampling {#sec-prob-audit}

Why should an auditor care about these functions? Because every statistical sample rests on them. Let us see two questions that come up in every sampling plan.

**Question 1. If I find no errors in my sample, how sure can I be?** Suppose the true rate of deviations from a control is 5%, which we would consider too high. We test 60 items and find none. What is the chance of that happening? Each item is a deviation with probability 0.05, so the number of deviations in 60 items follows a binomial distribution.

```{r}
dbinom(0, size = 60, prob = 0.05)
```

```{r}
#| include: false
p0 <- dbinom(0, size = 60, prob = 0.05)
```

There is only a `r round(100 * p0, 1)`% chance of finding no deviation in 60 items if the true rate were 5%. So if we do find none, we can be about `r round(100 * (1 - p0))`% confident that the true rate is below 5%. This is the logic behind sample sizes for tests of controls.

::: callout-tip
## Audit tip

Decide the rate you would consider too high, and the confidence you need, *before* testing the sample. Record both in the working papers. A conclusion is weaker if these are chosen after seeing the results.
:::

For small rates and large populations, the **Poisson** distribution gives nearly the same answer, with only one parameter, the expected number of deviations, here $60 \times 0.05 = 3$.

```{r}
ppois(0, lambda = 3)
```

**Question 2. What is the highest rate consistent with what I found?** The Poisson distribution also tells us how high the true number of errors could be, at a chosen confidence level, given the errors we found. These upper limits are called **reliability factors** or **confidence factors**, and auditors usually read them from printed tables. In R they come from a single function, `qgamma()`, because of a mathematical link between the Poisson and gamma distributions.

```{r}
# Reliability factors for 0 to 5 errors found
tibble::tibble(errors_found = 0:5,
               factor_95    = round(qgamma(0.95, 0:5 + 1), 2),
               factor_90    = round(qgamma(0.90, 0:5 + 1), 2))
```

These are exactly the factors printed in audit sampling tables. The factor 3.00 for zero errors at 95% confidence is the most used number in audit sampling. We will use these factors to calculate sample sizes and evaluate results in the next chapter (@sec-sampling).

::: callout-note
## Why does `qgamma()` give Poisson limits?

If errors occur at random, at an average of $\lambda$ per sample, then the probability of seeing $k$ or fewer errors is `ppois(k, lambda)`. The upper limit at 95% is the value of $\lambda$ for which this probability falls to 5%. That value happens to be the 95th percentile of a gamma distribution with shape $k + 1$, so `qgamma(0.95, k + 1)` gives it directly. We can check it.

```{r}
ppois(0, qgamma(0.95, 1))
ppois(2, qgamma(0.95, 3))
```
:::

## A suggested workflow

1.  **Know the question.** Decide whether you need the chance of *at most* a value, *more than* a value, or *exactly* a value. Each needs a different function or argument.
2.  **Choose the distribution.** The number of deviations in a sample of vouchers follows the binomial distribution. For small rates and large populations, the Poisson distribution gives nearly the same answer, with one parameter.
3.  **Use the right function.** Use `pbinom()` or `ppois()` for cumulative probability, with `lower.tail = FALSE` for the upper tail. Use `dbinom()` for exactly a value, and a `q` function for the value below which a given share lies.
4.  **Test what "no errors" means.** Before concluding that a control works, use `dbinom()` to find the chance of seeing no deviation in your sample if the true rate were too high.
5.  **Get reliability factors in R.** Calculate the factors for the number of errors found with `qgamma()`, instead of reading them from printed tables. They are used for sample sizes and evaluation in the next chapter.
6.  **Make simulations reproducible.** When you generate test data with `rnorm()` or another `r` function, call `set.seed()` first. A reviewer can then get the same numbers.

::: callout-important
## Remember

For a count, *at most one* and *one or more* are different questions. Check which tail you need before using a `p` function. A wrong tail gives a wrong level of assurance in the sampling conclusion.
:::

## Summary of functions used

| Function | Package | What it does |
|-----------------------|----------------|--------------------------------|
| `pnorm()`, `pbinom()`, `ppois()` | base R (stats) | Cumulative probability, use `lower.tail = FALSE` for the upper tail |
| `qnorm()`, `qgamma()` | base R (stats) | Quantiles, the value below which a given share lies |
| `dbinom()` | base R (stats) | Probability of exactly a value, for discrete distributions |
| `rnorm()`, `set.seed()` | base R | Random numbers, reproducibly |

: Functions used in this chapter {#tbl-probfunctions}
