# Time Series Analysis

Time series analysis is an essential technique for forecasting and analyzing trends in data over time. It is commonly used in various fields like finance, economics, and engineering.

So a question arises, what is a time series? A time series is a sequence of data points collected over time, where the observations are recorded in chronological order. Time series analysis involves analyzing and modeling these data points to understand patterns, trends, and make predictions about future values.  It is therefore used in predictive as well as descriptive analysis.

Time Series Analysis plays an important role in audit analytics as well.  A few of the use cases can be-

- **Revenue Analysis:** An auditor may analyze revenue data over time to identify irregularities or suspicious patterns that could indicate potential fraud or misstatement. By examining the revenue time series, the auditor can look for unexpected fluctuations, unusual growth or decline trends, or abnormal seasonality. Deviations from historical patterns or industry benchmarks may indicate fraudulent activities, such as revenue manipulation, fictitious transactions, or irregular recognition practices.
- **Inventory Analysis:** Auditors often analyze inventory data to assess the adequacy of inventory levels, identify potential inventory obsolescence or shrinkage, and evaluate the efficiency of inventory management. Time series analysis can be valuable in understanding inventory patterns and identifying potential risks or anomalies.

There are several time series data which are loaded by default in R; a few of these are listed below.  We will use these datasets to understand the various concepts related to time series analysis.

- `AirPassengers` containing monthly airline passenger numbers from 1949-1960. See figure \@ref(fig:tsex) (a)
- `Nile` contains flow of Nile river data. See figure \@ref(fig:tsex) (b)
- `sunspots` containing monthly sunspot numbers from 1749-1983. See figure \@ref(fig:tsex) (c)
- `JohnsonJohnson` which contains quarterly earnings per Johnson & Johnson share. See figure \@ref(fig:tsex) (d)

```{r tsex, echo=FALSE, fig.cap="Few time series data sets in R", fig.align='center', fig.show='hold'}
par(mfrow = c(2,2))
plot(AirPassengers, main= "(a) Airline Passengers from 1949-1960")
plot(Nile, main = "(b) Nile flow 1871-1970")
plot(sunspots, main = "(c) Sunspots")
plot(JohnsonJohnson, main = "(d) Earnings of Johnson&Jonson")
```

To analyse any time series, we have to understand its different components.

## Components of Time series
A time series can be decomposed into several components:

- **Level:** The baseline value or average of the series over time. It represents the long-term behavior of the series. This is basic component, and is always present in a time series object.  E.g. In a straight horizontal line only level is there which is equal to the value of y intercept.
- **Trend:** The overall direction of the series. It indicates whether the series is increasing, decreasing, or staying relatively constant over time.  E.g. A clear increasing trend can be seen in Johnson & Johnson earnings, in Figure \@ref(fig:tsex) (d).
- **Seasonality:** The repetitive and predictable patterns within the series that occur at regular intervals. Seasonality can be daily, weekly, monthly, quarterly, or yearly, etc. E.g. In figure \@ref(fig:tsex) (d) we may see seasonal patterns in earnings of Johnson&Johnson share.
- **Errors (Residuals):** The random fluctuations or noise in the series that cannot be explained by the level, trend, or seasonality.  E.g. See figure \@ref(fig:tsex) (c). The error component is an important aspect of time series analysis because it provides information about the uncertainty and variability of the data.

## Additive or multiplicative components
In time series analysis, the trend, seasonal, and residual components can be modeled as either additive or multiplicative. The choice of the model depends on how the components combine to create the observed values of the series.

An **additive model** assumes that the components of the time series are added together to create the observed values. In other words, the value of the time series at any point in time is equal to the sum of the trend, seasonal, and residual components at that point in time. This is expressed mathematically as:

$$
y_t = T_t + S_t + e_t
$$

where $y_t$, $T_t$, $S_t$ and $e_t$ are the values of the series, trend component, seasonal component and  residuals, respectively at time $t$.

A **multiplicative model**, on the other hand, assumes that the components of the time series are multiplied together to create the observed values. In other words, the value of the time series at any point in time is equal to the product of the trend, seasonal, and residual components at that point in time. This is expressed mathematically as:

$$
y_t = T_t \times S_t \times e_t
$$

where $y_t$, $T_t$, $S_t$ and $e_t$ are defined as before.  We can convert multiplicative time series into additive time series by taking $log$.  Note that seasonal component in `AirPassengers` time series depicted in \@ref(fig:tsex) (a) is multiplicative. See Figure \@ref(fig:logts).

```{r logts, echo=FALSE, warning=FALSE, message=FALSE, fig.show='hold', fig.align='center', fig.cap="Transforming Time Series", out.width="49%"}
library(ggplot2)
library(ggfortify)
autoplot(AirPassengers) +
  theme_bw() +
  labs(title = "AirPassengers Time series")
autoplot(log(AirPassengers)) +
  theme_bw() +
  labs(title = "Logirithmic transformation")
```

We will learn about decomposing time series components in next sections.

**Damped Trends:** Gardner and Mckenzie, in 1985 [@mcke] observed that most time series methods assume that any trend will continue unabated, regardless of the forecast lead time. They, based on empirical findings, suggested that forecast accuracy can be improved by either damping or ignoring altogether trends which have a low probability of persistence. We will learn about this concept, as well in next sections.

## Practical examples in R
Before analysing time series, in R let's see how time series objects are dealt in R.

### Pre-requisites {-}
Though most of the time we will be using base R, yet `forecast` is a fabulous package in R, developed by **Rob Hyndman**, which we will be using for predicting.  Predicting values help us understand how well we have captured the hidden trends and patterns in any time series. 

```{r message=FALSE, warning=FALSE}
library(forecast)
library(ggplot2)
```

### Creating a time series object in R

R provides us a simple function `ts()` to convert a series of observations (basically a vector) into a specific time series object.  The syntax is -

```
ts(data, 
  start = 1, 
  end = numeric(length(data)), 
  frequency = 1, 
  ...)
```
where:

- `data`: a numeric vector or matrix containing the data for the time series
- `start`: the start time of the time series, represented as either a numeric or a Date or POSIXct object
- `end`: the end time of the time series, represented as either a numeric or a Date or POSIXct object
- `frequency`: the number of observations per unit time for the time series. For example, if the data is recorded monthly, the frequency would be 12
- `...`: additional arguments that can be passed to the function, such as `names`, `delim`, or `tsp` (time series start and end points)

Example-1: Let's create a time series.  

```{r}
# Create a numeric vector representing monthly sales data for a year
sales <- c(10, 20, 30, 25, 35, 40, 45, 50, 55, 60, 65, 70)

# Create a time series object with monthly frequency starting from January
sales_ts <- ts(sales, start = c(2022, 4), frequency = 12)

# Print the time series object
sales_ts

```

We can check its class.
```{r}
class(sales_ts)
```

Let's also check `Nile`, which is already available in base R.
```{r}
Nile
```

### Plotting Time Series Objects
To plot `ts` object we can simply use `plot()` command or alternatively we can use `ggplot2` as well.

```{r tsex2, fig.cap="Example of time series plotting in R; Base R (Left) and ggplot2 (Right)", fig.align='center', fig.show='hold', fig.align='center', out.width="48%", warning=FALSE, message=FALSE}
# Plotting in base R
plot(sales_ts, main = "Sales during FY 2022-23\nDummy Data by author")

forecast::autoplot(AirPassengers) +
  scale_y_log10('Logrithmic values') +
  ggtitle('AirPassengers converted to additive') +
  theme_bw()
```

*Note: Library `forecast` in R also provides us a function `autoplot()` which plots a time series object, similar to plot(), but the plot returned here is a `ggplot2` object which can be modified/fine-tuned using `ggplot2` functions.*

## Time series modelling and forecasting
As we have already discussed, forecasting future plays an important part of time series analysis; an important prerequisite is to model our data.  In fact, the forecasting/modelling techniques can be broadly classified into two categories, data based techniques and model-based techniques.

### Data-based techniques {-}
Data-based techniques, also known as statistical or empirical techniques, focus on analyzing the patterns and characteristics of the observed time series data directly. These techniques do not explicitly assume a specific underlying model structure. Instead, they rely on statistical properties and patterns present in the data. Examples of data-based techniques include:

### Naive method
This method assumes that the future value will be the same as the last observed value. This is the simplest and most basic forecasting method and hence named *naive*. The formula for the naive method is:

$$
\text{Naive method: }\hat{Y}_{T+1} = Y_T
$$

Here, $\hat{Y}_{T+1}$ is the forecast value for the next time period and $Y_T$ is the last observed value i.e. for a series of $T$ observations.  Thus, predicting future values using naive method require no special skill. 

**Problem Statement-1:** Let's predict say 5, future values of Nile flow data.  In `forecast` there is a function `naive` which will do the job, once we give the value of parameter `h` which is short for forecasting **h**orizon. In Figure \@ref(fig:naive), we can see forecast values, in red, which are exactly same as previous/last value.

```{r naive, fig.show='hold', fig.align='center', fig.cap="Naive Forecasting using R", out.width="98%", fig.cap="Naive forecasting"}

forecast::naive(Nile, h = 5) %>% 
  autoplot() +
  theme_bw() +
  labs(title = "Nile flow - forecast by Naive method",
       x = "Year")


```

If we aren't interested in visualising confidence values (by default 80% and 95% confidence values are shown in bands), we can tune parameter `level` accordingly.

### Moving averages
In this technique, the forecast value is computed as the average of the most recent observations within a sliding window of fixed length, say $n$..  It is a technique to smooth out a time series by averaging neighboring values. It helps in reducing noise and revealing underlying trends. 

The formula for the moving average method is:

$$
\hat{Y}_{t+1} = \frac{1}{n}(Y_t + Y_{t-1} + ... + Y_{t-n+1})
$$

Here, $\hat{Y}_{t+1}$ is the forecast value for the next time period and $Y_t$, $Y_{t-1}$, ..., $Y_{t-n+1}$ are the past $n$ observations. 

Smoothing through moving average can either be **centre-weighted** or **tailed-weighted**.  In former, $k$ i.e. `order` of the moving average, should be an odd number because each data point is replaced with the mean of that observation and $(k-1)/2$ observations before and $(k-1)/2$ observations after it.  Smoothed time series `Nile` for different `k` can be seen in Figure \@ref(fig:ma).

```{r ma, echo=FALSE,fig.align='center', fig.cap="Simple Moving Averages", fig.align='center', fig.show='hold'}
library(patchwork)
lims <- c(min(AirPassengers), max(AirPassengers))

p1 <- autoplot(AirPassengers) +
  scale_y_continuous(limits = lims)+
  labs(title = 'AirPassengers - original') +
  theme_bw()

p2 <- autoplot(ma(AirPassengers,7)) +
  scale_y_continuous(limits = lims)+
  labs(title = 'Moving average k=7') +
  theme_bw()

p3 <- autoplot(ma(AirPassengers,15, centre = FALSE)) +
  scale_y_continuous(limits = lims)+
  labs(title = 'Moving average k=15, tailed') +
  theme_bw()

p4 <- autoplot(ma(AirPassengers,15)) +
  scale_y_continuous(limits = lims)+
  labs(title = 'Moving average k=15') +
  theme_bw()

(p1 + p2)/(p4 + p3)
```

**Problem Statement-2:** Let's try to visualise moving average trend in Nile data.  We will use `ma` function from `forecast` for this.  In left side Figure of \@ref(fig:ma1) we have a smoother (with k = 7) time series and 10 forecast  values using last value of moving average.  

We can also use `rollmean` function from `zoo` library, which is another package to analyse time series objects. Refer right-side figure in \@ref(fig:ma1).  In this example, we have used `wineind` data/time series which shows *Australian total wine sales by wine makers in bottles <= 1 litre. Jan 1980 – Aug 1994.*  One of the advantage using this method, is that we can see both original and smoothed time series. 

```{r ma1, warning=FALSE, message=FALSE, echo=FALSE, fig.align='center', fig.cap="Simple Moving averages", fig.show='hold', out.width="48%"}
# USING FORECAST
ma_nile <- ma(Nile, order = 7, centre = FALSE)
future_nile <- forecast(ma_nile, h = 10)
plot(future_nile, main = "Forecasting\nNile Flow")

# Using zoo
library(zoo)
naive(zoo::rollmean(wineind, k = 9, align = "right")) |> 
  plot(ylim=c(min(wineind), max(wineind)), 
       xlim=c(1980, 1995),
       col="red", 
       main = "Forecasting\nAustralian Wine sales")

lines(wineind)
```

### Exponential smoothing 
Exponential smoothing techniques update forecasts based on weighted averages of past observations, giving more weight to recent observations. Examples include simple exponential smoothing, Holt's method, and Holt-Winters' method.

As the name suggests simple exponential smoothing or SES is simplest of these.  **Simple exponential smoothing (SES)** is actually a moving weighted average where recent observations are given more weights while calculating averages.  *This method is suitable for forecasting data with no clear trend or seasonal pattern.* 

$$
\hat{y}_{T+1} = \alpha{y_T} + \alpha(1- \alpha){y_{T-1}} + \alpha(1 - \alpha)^2{y_{T-2}} + ...
$$

The value of $\alpha$, the smoothing parameter, in above equation should follow $0 \le \alpha \le 1$.  Clearly different values of $\alpha$ will give different smoothing and we will have to adopt most suitable one, one such example with $\alpha = 0.1$ is shown in the Figure \@ref(fig:ses1) (Left). 

Holt extended SES by introducing trend component and thus a new smoothing parameter for trend $\beta$. The method is also named after him as **Holt's Linear Method** of exponential smoothing. Now there are are two equations, mathematically.

$$
\begin{aligned}
\hat{y}_{t+h|t} &= l_t + hb_t \\
l_t &= \alpha y_t + (1-\alpha)(l_{t-1} + b_{t-1}) \\
b_t &= \beta(l_t - l_{t-1}) + (1-\beta)b_{t-1}
\end{aligned}
$$

$\alpha$ and $\beta$ (again $0 \le \beta \le 1$) in above equations are smoothing parameters for level and trend components, respectively; and $h$ is forecast horizon. Also, $y_t$ is the actual value, $l_t$ is the level (or intercept), $b_t$ is the trend (or slope) of the time series at time $t$.  Refer \@ref(fig:ses1) (Right).

To apply *SES*, in R, we can use function `ses` and for applying *Holt's* smoothing, we can use `holt`; both from `forecast` library.  Example usages are as follows.

```{r ses1, fig.cap="Simple Exponential Smoothing (Left) Vs. Holt's Linear Method (Right)", fig.align='center', fig.show='hold', out.width="49%"}

lynxses <- ses(lynx, h= 10, alpha = 0.1)
holtfit <- holt(airmiles, h = 10, alpha = 0.3, beta = 0.1)

autoplot(lynxses) +
  autolayer(lynxses$fitted) +
  ggplot2::ggtitle('SES with alpha = 0.1 : Lynx data') +
  ggplot2::theme_bw() +
  ggplot2::theme(legend.position = "bottom") +
  ggplot2::labs(color = "")

autoplot(holtfit) +
  autolayer(holtfit$fitted) +
  ggplot2::ggtitle("Holt's Linear Trend method with beta = 0.1 : AirMiles data") +
  ggplot2::theme_bw() +
  ggplot2::theme(legend.position = "bottom") +
  ggplot2::labs(color = "")
```

As we have discussed earlier, damped trends work better many times.  So based on Mckenzie and gardner work, an additional dampening paramater $\phi$ was introduced.  The equations were modified as -
$$
\begin{aligned}
\hat{y}_{t+h|t} &= l_t + (\phi + \phi^2 + ... + \phi^h)b_t \\
l_t &= \alpha y_t + (1-\alpha)(l_{t-1} + \phi b_{t-1}) \\
b_t &= \beta(l_t - l_{t-1}) + (1-\beta)\phi b_{t-1}
\end{aligned}
$$

Clearly, if $\phi = 1$, the above equations are equivalent to Holt's Linear Method. $\phi$ for values $0 < \phi < 1$, dampens the trend, as shown in figure \@ref(fig:hl2) (Left). A comparison of both can be seen in the Figure \@ref(fig:hl2) (Right).

To extend holt's linear method with *damped trends* in R, we can set parameter `damped = TRUE` in `holt` function discussed above.  Example code can be seen below.

```{r hl2, fig.cap="Linear methos Vs. Damped trends", fig.align='center', fig.show='hold', out.width="49%"}
library(ggplot2)

holtfit <- holt(airmiles, h = 10, alpha = 0.3, beta = 0.1, damped = TRUE, phi = 0.8)
autoplot(holtfit) +
  autolayer(holtfit$fitted) +
  ggplot2::ggtitle('Damped Trend Forecast with phi=0.8') +
  ggplot2::theme_bw() +
  ggplot2::theme(legend.position = "bottom") +
  ggplot2::labs(color = "")


holtfit <- holt(airmiles,
                h = 10,
                alpha = 0.3,
                beta = 0.1)

holtfit2 <- holt(
  airmiles,
  h = 10,
  alpha = 0.3,
  beta = 0.1,
  damped = TRUE,
  phi = 0.8
)

autoplot(airmiles)+
  autolayer(holtfit, series="Holt's method", PI = FALSE) +
  autolayer(holtfit2, series="Holt's method with damped trend", PI = FALSE)+
  ggplot2::ggtitle("Holt's method") + xlab("Year") +
  ylab("Revenue passenger miles") +
  guides(colour=guide_legend(title="Method")) +
  ggplot2::theme_bw() +
  theme(legend.position = "bottom")
```

Now to incorporate `seasonality` (with say $m$ periods) along with trend and level, Holt and Winters suggested to incorporate additional parameter $\gamma$ in above equations.  

$$
\text{Holt-Winters method:}
$$
$$
\begin{aligned}
\hat{y}_{t+h|t} &= l_t + hb_t + s_{t+m-(m-1 \bmod m)} \\
l_t &= \alpha(y_t - s_{t-m}) + (1-\alpha)(l_{t-1} + b_{t-1}) \\
b_t &= \beta(l_t - l_{t-1}) + (1-\beta)b_{t-1} \\
s_t &= \gamma(y_t - l_{t-1} - b_{t-1}) + (1-\gamma)s_{t-m}
\end{aligned}
$$

An example can be seen in figure \@ref(fig:hw1).  To apply *Holt-Winters* method in R, we can use `hw` function with desired values of parameters.  An example code may be as follows, where we have tried to extend `AirPassengers` time series with this method with both `additive` and `multiplicative` trends.

```{r hw1, fig.cap="Holt-Winter's Method of Forecasting", fig.align='center', fig.show='hold', out.width="80%"}
hw1 <- hw(AirPassengers,seasonal="additive")
hw2 <- hw(AirPassengers, seasonal="multiplicative")
autoplot(AirPassengers) +
  autolayer(hw1, series="HW additive forecasts", PI=FALSE) +
  autolayer(hw2, series="HW multiplicative forecasts",
            PI=FALSE) +
  xlab("Year") +
  ylab("Air Passengers") +
  ggtitle("Holt-Winters model") +
  guides(colour=guide_legend(title="Forecast")) +
  ggplot2::theme_bw() +
  theme(legend.position = "bottom") 
```

### Seasonal decomposition methods {-}

Seasonal decomposition techniques decompose a time series into its different components, such as trend, seasonality, and noise. This allows for a better understanding of the individual components and their impact on the overall series.  There are several methods to decompose a time series into its different components.  We will however, discuss two of these.

### Classical seasonal decomposition
First of these methods is **classical decomposition method** which decomposes a time series into its trend, seasonal, and residual components. It assumes that the seasonal component repeats identically from year to year, and the trend component changes linearly over time. The steps involved in the classical decomposition method are as follows:

- **Trend Component:** The trend component is estimated using a moving average or regression method.
- **Seasonal Component:** The seasonal component is estimated by averaging the values across the same seasonal periods in different years.
- **Residual Component:** The residual component is obtained by subtracting the trend and seasonal components from the original time series.

**Problem Statement:** Let's try to decompose a time series say `AirPassengers` to see its components.  To decompose a time series in R, we will use function `decompose` as shown in the following code.  Decomposing time series in R gives us four different plots each for (i) Observed i.e. original values, (ii) trend, (iii) seasonality and (iv) random noise available. 

Case-1: Additive decomposition

```{r}
decomposed_air_passengers <- decompose(AirPassengers)
summary(decomposed_air_passengers)
```

Case-2: Multiplicative decomposition.  In this case we can use argument `type`.

```{r}
decomposed_air_passengers2 <- decompose(AirPassengers, type = "multiplicative")
summary(decomposed_air_passengers2)
```

A plot of decomposed `AirPassengers`, using classical approach is shown in Figure \@ref(fig:stl1).

```{r stl1, fig.cap="Seasonal Decomposition Techniques", fig.align='center', fig.show='hold'}
AirPassengers |>
  decompose() |>
  plot()
```

### Seasonal and Trend decomposition using Loess^[Loess is a method for estimating nonlinear relationships]. 

STL (Seasonal and Trend decomposition using Loess) is a robust and flexible method for decomposing a time series into its trend, seasonal, and residual components. It uses local regression (Loess) to estimate the trend and seasonal components. The STL algorithm is as follows:

- Seasonal Component: The seasonal component is estimated using Loess, which fits a smooth curve to the seasonal patterns.
- Trend Component: The trend component is estimated by removing the estimated seasonal component from the original time series.
- Residual Component: The residual component is obtained by subtracting the estimated trend and seasonal components from the original time series.

To decompose time series in R, using STL, we will use function `stl` as shown below.  A plot of STL decomposed `AirPassengers` is shown in figure \@ref(fig:stl2).

```{r stl2, fig.cap="STL Decomposition", fig.align='center', fig.show='hold'}
stl(AirPassengers, s.window = "periodic") %>% 
  plot()
```

### Model-based techniques{-}

Model-based techniques involve fitting a specific mathematical or statistical model to the observed time series data. These techniques assume a particular structure for the data and estimate model parameters based on that. Model-based techniques typically require more assumptions but can provide a more detailed understanding of the underlying dynamics. Examples of model-based techniques include:

1. **Autoregressive Integrated Moving Average (ARIMA)**: ARIMA models capture the linear dependencies between lagged observations and differences of the time series. They are commonly used for modeling stationary time series.

2. **Seasonal ARIMA (SARIMA)**: SARIMA models extend the ARIMA framework to incorporate seasonality in the data. They are suitable for time series exhibiting both trend and seasonality.

In future version of the book, we will discuss these methods.

## Plotting different elements of time series
In R, we can use functions such as -

- `ggseasonplot()`: to create a seasonal plot
- `ggsubseriesplot()`: to create mini plots for each season and show seasonal means
- `gglagplot()`: Plot the time series against lags of itself
- `ggAcf()`: Plot the autocorrelation function (ACF)

Example-1:
```{r, fig.show='hold', fig.align='center', fig.cap="Seasonalilty"}
theme_set(theme(axis.text.x = element_text(angle = 90, 
                                           vjust = 0.5, 
                                           hjust=1)))
library(patchwork)
g1 <- ggseasonplot(AirPassengers) +
  theme_bw()
g2 <- ggsubseriesplot(AirPassengers) +
  ggtitle('Sub-Series Plot') +
  theme_bw()
(g1 / g2)
```

Example-2: Lag Plot of `AirPassengers` may be seen at Figure \@ref(fig:lagp).  In the figure notice that plot at lag=12 suggest that series has seasonality with 12 periods.
```{r, lagp, fig.show='hold', fig.align='center', fig.cap="Seasonalilty through Lagplots"}
gglagplot(AirPassengers) +
  ggtitle('Lag Plots') +
  theme_bw() +
  theme(legend.position = "bottom") +
  labs(color = NULL) +
  guides(color = guide_legend(nrow = 1))

```

------------------------------


