```{r}
#| include: false
source("_common.R")
```

# Linear Regression {#sec-regression}

Regression models are a class of statistical models that let us explore the relationship between a response variable and one or more explanatory variables. If such a relationship exists and we can find it, we can predict the value of the response variable from the explanatory variables.

As an example, suppose we establish a relationship between the number of employees in an office and its monthly expenditure on salary. We can then predict the monthly salary expenditure of any office if we know its staff strength. That lets us ask useful questions. How much should an office with 10 employees spend on salary every month? And for the auditor, the more interesting question is the reverse one. Which offices are spending *much more* than their staff strength explains? We will answer exactly this question at the end of this chapter, in @sec-reg-audit.

Explanatory variables are also called *regressors*, *independent* variables or *predictors*. The response variable is also called the *dependent* or *outcome* variable. When the response variable is numeric and its relationship with the predictors is linear, the regression is called *linear regression*. There are some other assumptions too, which we will discuss later in the chapter.

## Prerequisites

Besides `tidyverse`, we will use a few packages in this chapter. `broom` turns model results into tidy data frames. `performance` and `see` draw diagnostic plots. `car` calculates the variance inflation factor. `GGally` draws correlation plots and `moderndive` draws parallel slopes. `palmerpenguins` gives us a dataset for the last example.

```{r}
#| message: false
#| warning: false
library(tidyverse)
library(broom)
library(performance)
library(car)
library(GGally)
library(moderndive)
library(palmerpenguins)
```

```{r helpfun, echo=FALSE, warning=FALSE, message=FALSE}
my_chart <- function(data, x_var, y_var){
  # Model
  lmod <- eval(substitute(lm(y_var ~ x_var , data = data)))
  # Visualise
  data %>%
    mutate(pred = predict(lmod),
           res = residuals(lmod)) %>%
    ggplot(aes({{x_var}}, {{y_var}})) +
    geom_smooth(method = 'lm', color = 'lightgrey', se = FALSE, formula = 'y ~ x') +
    geom_segment(aes(xend = {{x_var}},  yend = pred), alpha = 0.2, linewidth = 0.9, color = 'seagreen') +
    geom_point(aes(color = abs(res), size = abs(res))) +
    scale_color_continuous(low = 'yellow', high = 'darkred') +
    guides(color = "none", size = "none") +
    geom_point(aes(y = pred), shape = 1) +
    theme(panel.grid.major = element_blank(),
          panel.grid.minor = element_blank(),
          panel.background = element_blank(),
          axis.line = element_line(colour = "black")
          )

}

# Correlation panel coloured by strength, used by ggpairs() later
my_fn <- function(data, mapping, method = "p", use = "pairwise", ...){
  x <- eval_data_col(data, mapping$x)
  y <- eval_data_col(data, mapping$y)
  corr <- cor(x, y, method = method, use = use)
  # minus one is blue, zero is white and one is red
  colFn <- colorRampPalette(c("blue", "white", "red"), interpolate = 'spline')
  fill <- colFn(100)[findInterval(corr, seq(-1, 1, length = 100))]
  ggally_cor(data = data, mapping = mapping, ...) +
    theme_void() +
    theme(panel.background = element_rect(fill = fill))
}
```

## Basic concepts

Before we build a regression model, it is always good practice to visualise the data. It helps us see the nature of the relationship between the predictors and the response variable, if there is one. To see the relationship between two numeric variables, we use a scatter plot. See the two examples in @fig-concept. In the first plot there is a near perfect linear relationship between GNP and population (correlation `r round(cor(longley$GNP, longley$Population), 2)`). In the second plot there is a strong but negative relationship between mileage and engine displacement (correlation `r round(cor(mtcars$disp, mtcars$mpg), 2)`). The points are more scattered around the line here.

::: {#fig-concept}

```{r echo=FALSE, fig.align='center', out.height="45%", out.width="47%", fig.show='hold'}

longley |>
  ggplot(aes(Population, GNP)) +
  geom_point(color = 'blue') +
  geom_smooth(method = 'lm', se = FALSE, color = 'red', formula = 'y ~ x') +
  labs(title = 'Ex-1: Gross Nation Product Vs.\nPopulation in longley dataset') +
  theme_bw()

mtcars |>
  ggplot(aes(disp, mpg)) +
  geom_point(color = 'blue') +
  geom_smooth(method = 'lm', se = FALSE, color = 'red', formula = 'y ~ x') +
  labs(title = 'Ex-2: Mileage in miles per gallon\nvs. Displacement') +
  theme_bw()


```

Linear Regression - Intuition
:::

When the outcome variable has a linear relationship with the explanatory variable, the best prediction lies on a straight line. So the maths behind the prediction is finding the equation of that line, @eq-lr1.

$$
y = mx + c
$$ {#eq-lr1}

Where -

- `m` is the slope of the line
- `c` is the intercept on the Y-axis (value of $y$ when $x$ is `0`)

Fitting a regression thus means estimating both $m$ and $c$. In statistics, these two are usually written as $\beta_1$ (slope) and $\beta_0$ (intercept). Those familiar with coordinate geometry know that two points are enough to draw a line. But real data has many points, and they rarely lie on one line. So we can draw many different lines through them. Our goal is to find the *best* of these lines. But how?

To answer this, note that in practice the points do not lie on the regression line. See the plots in @fig-concept again. Hardly any point lies exactly on the line. So for each point there is an *actual* value of $y$ and a *fitted* value, the one on the line. The difference between the two is called the *residual*. It is the part of $y$ that the model is not able to explain. If $\hat{y}$ is the fitted value and $y$ is the actual value, the residual $\epsilon$ is

$$
\epsilon = y - \hat{y}
$$ {#eq-lr2}

or, we can say that

$$
y = \beta_0 + \beta_1x + \epsilon
$$ {#eq-lr3}

The best line is the one which leaves the smallest residuals. We cannot simply add up the residuals, because some are positive and some negative, and they cancel out. See @fig-errors. So we either take their absolute values or their squares. The usual method is to minimise the *sum of squares* of residuals. Squaring gives more weight to large residuals than to small ones. It also gives a simple formula for the best line. That is why this technique is called **Ordinary Least Squares** or **OLS** regression.

```{r fig-errors, echo=FALSE, warning=FALSE, fig.align='center', fig.cap="Residuals - Intuition and Concept", out.height="35%"}
my_chart(LifeCycleSavings, sr, pop75)
```

## Simple Linear Regression in R

We do not have to do this minimisation ourselves. Base R has a function `lm()` which fits the best regression line for a given set of variables. The syntax is simple.

```
lm(formula, data, ...)
```

where -

- `formula` is a symbolic description of the model to be fitted. For our example above, we write `y ~ x`, read as "y is explained by x".
- `data` is the data frame which contains the variables.

`lm()` returns a special object. We can print it directly like other R objects. But it is best printed with the `summary()` function. Let us first see an example of *simple* linear regression, which is linear regression with only one independent variable.

**Problem Statement:** Example-1. `iris` is a famous (Fisher's or Anderson's) dataset, loaded by default in R. It gives the measurements in centimetres of sepal length (`Sepal.Length`), sepal width (`Sepal.Width`), petal length (`Petal.Length`) and petal width (`Petal.Width`) for 50 flowers from each of 3 species (`Species`) of iris. The species are *setosa*, *versicolor*, and *virginica*. Let us try to establish a relationship between `Sepal.Length` and `Sepal.Width` of the *setosa* species, which are the first 50 records of `iris`.

As already stated, it is good practice to visualise the data first. So let us make a scatter plot, as in @fig-ex1plot.

```{r fig-ex1plot, fig.align='center', fig.cap="Relationship between sepal widths and lengths in setosa species", out.height="30%"}

iris %>%
  # Extract First 50 records
  head(50) %>%
  ggplot(aes(Sepal.Length, Sepal.Width)) +
  # Scatter plot
  geom_point() +
  # adding a trend line without error bands
  geom_smooth(method = 'lm', se=FALSE, formula = "y~x") +
  # Theme Black and White
  theme_bw()

```

The relationship looks fairly linear (@fig-ex1plot), so let us build the model.

```{r ex1}
lin_reg1 <- lm(formula = Sepal.Width ~ Sepal.Length, data = iris[1:50,])

# Let us print the object directly
lin_reg1

# Try printing it with summary()
summary(lin_reg1)
```

Simply printing the object gives us only the coefficients. Printing it with `summary()` gives us a lot of other information. But how do we read it? Before that, let us understand a few assumptions which we made, silently, while fitting the line.

## Assumptions of Linear Regression

Linear regression makes several assumptions about the data.

- **Linearity.** The relationship between the predictor ($x$) and the outcome ($y$) is assumed to be linear. If we fit a straight line to a curved relationship, our results will be wrong. See @fig-linearity. The relationship there is actually cubic (green dashed line), and the straight line (red) misses it badly. When we use many predictors, as we will shortly, we also assume that their effects *add up*.

```{r fig-linearity, echo=FALSE, fig.align='center', fig.cap="Is the relationship linear?", out.width="50%"}
# Setting seed for reproducibility
set.seed(1234)
x <- 11:110
y <- x^3 + rnorm(100, 0, 80000)
data.frame(x = x, y = y) %>%
  ggplot(aes(x, y)) +
  geom_point() +
  geom_smooth(method = "lm", color = "red", formula = 'y ~ x', linetype = "solid", se = FALSE) +
  geom_smooth(method = "lm", color = "green", formula = 'y ~ I(x^3)', se = FALSE, linetype = 'dashed') +
  theme_bw()
```

- **Normality of residuals.** The residuals are assumed to be normally distributed with a mean of $0$. Note that it is the *residuals* which should be normal, not the $y$ variable itself. If the model has captured the relationship well, what is left should be random noise. This assumption matters mainly for the p-values and confidence intervals. The line itself can be fitted without it.

- **Homogeneity of residual variance.** The residuals are assumed to have the same spread at all values of $x$. Statistically this is called *homoskedasticity*. If the spread changes with $x$, it is called *heteroskedasticity*. See @fig-homod. The residuals in the left plot have equal spread. In the right plot, the spread grows as $x$ grows, which shows heteroskedasticity clearly. In audit data this is very common. Big offices, big contracts and big claims vary more, in rupees, than small ones.

::: {#fig-homod}

```{r echo=FALSE, fig.align='center', out.width="47%", fig.show='hold'}
set.seed(1234)
data.frame(x = 1:100, y = rnorm(100, 5, 10)+(1:100)) %>%
  ggplot(aes(x, y)) +
  geom_point()+
  geom_smooth(method = 'lm', se = FALSE, formula = "y ~ x") +
  labs(title = "Homoskedasticity of Residuals") +
  theme_bw()

data.frame(x = 1:100, y = rnorm(100, 0.5, 0.07)*(1:100)) %>%
  ggplot(aes(x, y)) +
  geom_point()+
  geom_smooth(method = 'lm', se = FALSE, formula = "y ~ x") +
  labs(title = "Heteroskedasticity of Residuals") +
  theme_bw()


```

Homoskedasticity (left) Vs. Heteroskedasticity (right)
Sample data created by author for demonstration only
:::

- **Independence of residuals.** Each observation is assumed to be independent of the others. In our example, the sepal width of one flower does not affect the width of another. Monthly data of the same office, on the other hand, is usually *not* independent, because this month's expenditure is related to last month's. Such data needs time series methods (@sec-ts).

## Interpreting the output of `lm`

Let us discuss each section of the `summary()` output, which we will call the model from here on.

### `Call`

The `Call` section shows the formula used to fit the model. `Sepal.Width` is our response variable and `Sepal.Length` is the predictor. The data is the first 50 rows of `iris`, or `iris[1:50,]`.

### `Residuals`

This is the five point summary of the residuals. As discussed, residuals are the actual values minus the fitted values. We can get the same numbers ourselves.

```{r resid}
summary(iris$Sepal.Width[1:50] - lin_reg1$fitted.values)
```

Ideally, the median of the residuals should be close to `0`, and the first and third quartiles should be about equally far from it. That tells us the residuals are roughly symmetrical, and the model is not consistently over-predicting or under-predicting. In our example the median is `r round(median(residuals(lin_reg1)), 3)`, practically zero. The quartiles are also nearly balanced. The histogram in @fig-ex1hist shows that the residuals look fairly normal, as we assumed.

```{r fig-ex1hist, echo=FALSE, fig.align='center', fig.cap="Histogram of Residuals"}
broom::augment(lin_reg1) %>%
  ggplot(aes(.resid)) +
  geom_histogram(binwidth = 0.03) +
  geom_density(linewidth = 1, lty = 2) +
  theme_bw() +
  labs(x = "Residuals", y = NULL) + theme(panel.grid.major = element_line(linetype = "blank"),
    panel.grid.minor = element_line(linetype = "blank"))
```

### `Coefficients`

These were the goal of our exercise. There is one coefficient for the predictor (the slope, $\beta_1$) and one for the intercept ($\beta_0$). A positive slope means $y$ goes up when $x$ goes up. A negative slope means $y$ goes down when $x$ goes up. In our example,

- when sepal length is `0`, the sepal width would be `r round(coef(lin_reg1)[1], 4)`. This is only a mathematical value, since no flower has zero length or negative width. The intercept often has no practical meaning when $x = 0$ lies far outside the data.
- for every `1` cm increase in sepal length, the width increases by `r round(coef(lin_reg1)[2], 4)` cm, on average.

Thus our regression line is

$$
Sepal.Width = -0.5694 + 0.7985 * Sepal.Length
$$ {#eq-lr4}

We fitted the line on a sample of 50 flowers. Another sample of 50 setosa flowers would give slightly different coefficients. So the coefficients are only *estimates*, and each comes with some uncertainty. We can see a range of likely values for each coefficient, called the confidence interval, using the function `confint()`. By default, it gives 95% confidence intervals.

```{r}
confint(lin_reg1)
```

The other columns of the coefficients table measure this uncertainty.

- The *standard error* (`Std. Error`) tells us how much the estimate would vary from sample to sample. The confidence intervals above are built from it.
- The *t value* is simply the estimate divided by its standard error. A large t value (positive or negative) means the estimate is large compared to its uncertainty.
- The *p-value* (`Pr(>|t|)`) tests the null hypothesis that the true coefficient is zero, that is, $x$ has no effect on $y$. A small p-value, say below 0.05, lets us reject this.

In our example, the p-value of `Sepal.Length` is very small, so sepal length clearly affects width. The p-value of the intercept is `r round(summary(lin_reg1)$coefficients[1, 4], 3)`, so we cannot say the intercept is different from zero. That is fine. We are rarely interested in the intercept.

### `Signif. codes`

These are just legends for the stars shown next to each p-value. Notice the three stars next to `Sepal.Length`. They mean its p-value is below 0.001, and we can safely reject the null hypothesis that $\beta_1$ is $0$. The stars give a quick visual sense of which coefficients matter.

### `Residual standard error`

The residual standard error (RSE) is one of the measures of how well the model fits the data. It is the standard deviation of the residuals, except that we divide by the *degrees of freedom* instead of $n$.

$$
RSE = \sqrt{\frac{1}{(n-2)} \sum_{i=1}^n (y_i - \hat{y_i})^2}
$$ {#eq-lr5}

Here the degrees of freedom are $n-2$, because we estimated two coefficients, one slope and one intercept. We can verify @eq-lr5 by calculation.

```{r rse}
sqrt(sum((iris[1:50, "Sepal.Width"] - lin_reg1$fitted.values)^2)/48)
```

The RSE is in the units of $y$. So we can say that the model's predictions of sepal width are typically off by about `r round(sigma(lin_reg1), 2)` cm.

### `R-Squared`, both Multiple and Adjusted

*Multiple R-squared* is also called the coefficient of determination. It is the most cited measure of how well a model fits. It tells us what share of the variation in the response variable is explained by the model. In our example about `r round(summary(lin_reg1)$r.squared * 100)`% of the variation in sepal width is explained by sepal length. We will discuss it in detail in the next section.

*Adjusted R-squared* is R-squared with a penalty for the number of predictors in the model. R-squared never goes down when we add a predictor, even a useless one. Adjusted R-squared goes up only if the new predictor improves the model more than would be expected by chance. So it is the better measure to compare models with different numbers of predictors.

### `F-Statistic` and `p-value`

Why another p-value? Here the null hypothesis is about the *whole* model. It says that *none* of the predictors has any relationship with the response variable, or that all slopes are zero. The alternative is that at least one slope is not zero. This is tested with the F-statistic. In our example, the p-value is very small. So we reject the null hypothesis and conclude that there is a relationship between `Sepal.Length` and `Sepal.Width`.

With one predictor, this test gives the same p-value as the t-test of the slope. With many predictors, the F-test is useful because it tests them all together. If we test twenty coefficients one by one, one or two may look significant just by chance.

## Model Evaluation Metrics

To judge how accurate a model is, we need evaluation metrics. There are several. We will discuss a few of them here.

### MAE - Mean Absolute Error

As the name suggests, it is the mean of the absolute values of the residuals, @eq-lr6.

$$
{MAE} = \frac{1}{N}\sum_{i = 1}^{N}{\lvert}{y_i - \hat{y_i}}{\rvert}
$$ {#eq-lr6}

It is the average size of the error, in the units of $y$. A larger value means a less accurate model. On its own, MAE does not tell us much. But when we compare several models on the same data, we can choose the one with the lower MAE.

Once we extract `$residuals` from the model, calculating it is easy. In our example,

```{r ex1mae}
# Mean Absolute Error
lin_reg1$residuals |> abs() |> mean()
```

### MSE - Mean Square Error and RMSE - Root Mean Square Error

The mean of the squares of the residuals is the mean square error or MSE, @eq-lr7.

$$
{MSE} = \frac{1}{N}\sum_{i = 1}^{N}({y_i - \hat{y_i}})^2
$$ {#eq-lr7}

MSE punishes large residuals more, by squaring them. Like MAE, we can use it to choose the better of several models. In fact, MSE is exactly what OLS minimises while finding the coefficients.

```{r}
# Mean Square Error
lin_reg1$residuals^2 |> mean()
```

RMSE or root mean square error is the square root of MSE. It brings the error back to the units of $y$.

$$
{RMSE} = \sqrt{\frac{1}{N}\sum_{i = 1}^{N}({y_i - \hat{y_i}})^2}
$$ {#eq-lr8}

In our example,

```{r}
# Root Mean Square Error
lin_reg1$residuals^2 |> mean() |> sqrt()
```

Notice that RMSE is very close to the residual standard error we saw earlier. The only difference is that RMSE divides by $N$, and RSE divides by the degrees of freedom.

### MAPE - Mean Absolute Percentage Error

This metric takes each residual as a percentage of the actual value, @eq-lr9.

$$
{MAPE} = \frac{1}{N}\sum_{i = 1}^{N}{\lvert}\frac{({y_i - \hat{y_i}})}{y_i}\cdot{100}\%{\rvert}
$$ {#eq-lr9}

Since it is a percentage, MAPE does not depend on the scale of the data. We can compare a model on salaries in rupees with a model on salaries in lakh. In our example,

```{r}
# Mean Absolute Percentage Error
{lin_reg1$residuals/iris$Sepal.Width[1:50]} %>%
  {abs(.)*100} %>%
  mean(.) %>%
  sprintf("MAPE is %1.2f%%", .)
```

::: callout-warning
## Caution

MAPE breaks down when some actual values are zero or close to zero, because we divide by them. A ₹10 error on a ₹20 bill is a 50% error. In audit data, which often has zero or tiny amounts, check for these before using MAPE.
:::

### R-Squared and adjusted R-squared

As already discussed, R-squared is the coefficient of determination. Its formula is @eq-lr10.

$$
R^2 = 1 - \frac{\sum_{i=1}^{N}(y_i - \hat{y_i})^2}{\sum_{i=1}^{N}(y_i - \bar{y})^2}
$$ {#eq-lr10}

The numerator in the fraction is called $SSE$ or Sum of Squares of Errors. The denominator is called $TSS$ or Total Sum of Squares, which measures how much $y$ varies around its own mean $\bar{y}$. So $\frac{SSE}{TSS}$ is the share of variation *left unexplained* by the model, and $R^2$ is the share *explained*. As $SSE$ gets closer to $0$, $R^2$ gets closer to $1$, a perfect model.

We can verify the formula with the results we have.

```{r}
## Sum of Squares of Errors
y_bar <- mean(iris[1:50,"Sepal.Width"])
(SSE <- sum(lin_reg1$residuals^2))
(TSS <- sum((iris[1:50,"Sepal.Width"] - y_bar)^2))
(r_sq <- 1 - SSE/TSS)
```

There is one more way to think of $R^2$. It is simply the square of the correlation between the actual and the fitted values. Let us verify.

```{r}
(cor(iris[1:50, 'Sepal.Width'], lin_reg1$fitted.values))^2
```

As we keep adding predictors to a model, $R^2$ keeps going up, even when the new predictors are useless. The model then starts fitting the noise in our sample, which is called *over-fitting*. Adjusted R-squared corrects for this by penalising the number of predictors, @eq-lr11.

$$
{Adjusted}\;{ R^2} = 1 - \frac{(1-R^2)(N - 1)}{(N - p - 1)}
$$ {#eq-lr11}

where $p$ is the number of predictors in the model. The `summary()` function gives both these metrics. We can verify the calculation.

```{r}
1 - ((1-r_sq)*(50-1)/(50-1-1))
```

## Plotting the results and their interpretation

We can pass the output of `lm()` to `plot()` to see six diagnostic plots. We choose them with the `which` argument. By default, R shows plots 1, 2, 3 and 5. The six plots are

1. Residuals Vs. Fitted Values
2. Normal Q-Q
3. Scale-Location
4. Cook's distance
5. Residuals vs. leverage
6. Cook's distance vs. leverage

Let us see these for our example.

```{r fig-plots1, fig.align='center', fig.cap="First two diagnostic plots", out.height="40%"}
par(mfrow = c(1, 2), oma = c(0, 0, 2, 0))
plot(lin_reg1, which = 1:2, sub.caption = "")
```

1. **Residuals Vs. Fitted Values.** This plot checks the first assumption, linearity. If the red line across the plot is roughly flat, the relationship is linear and a straight line is a fair choice. If the red line shows a clear curve, a straight line is missing some pattern. In our example the red line bends a little at the ends, where there are only a few points, but there is no clear curve. So linear regression is appropriate.
2. **Normal Q-Q.** This plot checks whether the residuals are normally distributed. If the points fall roughly along the straight diagonal line, they are. In our example, they do.

Notice also that R labels the three most extreme residuals with their row numbers. Here rows `r paste(head(order(-abs(rstandard(lin_reg1))), 3), collapse = ", ")` are labelled.

```{r fig-plots2, fig.align='center', fig.cap="Next two diagnostic plots", out.height="40%"}
par(mfrow = c(1, 2), oma = c(0, 0, 2, 0))
plot(lin_reg1, which = 3:4, sub.caption = "")
```

3. **Scale-Location.** This plot checks the assumption of equal variance or homoskedasticity. If the red line is roughly flat and the points are spread evenly around it, the assumption is met. A line going up from left to right is the usual sign of heteroskedasticity.
4. **Cook's Distance.** An *influential* point is one which, if removed, would change the regression line noticeably. Not every outlier is influential. A point with a large residual but an ordinary $x$ value often moves the line very little. A point with an extreme $x$ value has more power to pull the line towards itself. This power is called *leverage*. Cook's distance combines the size of the residual and the leverage into one measure of influence. A common rule of thumb is to look closely at any point with Cook's distance above 0.5, or well above the others. In our example, row `r which.max(cooks.distance(lin_reg1))` has the largest Cook's distance, `r round(max(cooks.distance(lin_reg1)), 2)`, which is not alarming.

```{r fig-plots3, fig.align='center', fig.cap="Last two diagnostic plots", out.height="40%"}
par(mfrow = c(1, 2), oma = c(0, 0, 2, 0))
plot(lin_reg1, which = 5:6, sub.caption = "")
```

5. **Residuals vs. leverage.** This plot also finds influential points. Points with high leverage lie to the right. Any point outside the dashed red lines of Cook's distance is influential. In our example, no point is even close to these lines.
6. **Cook's distance vs. leverage.** This shows the same information in another way. Points in the top right corner, with both high leverage and high Cook's distance, need attention.

Base R's `plot()` can draw all these plots. The package `performance` draws all the relevant diagnostic plots together, nicely, and writes under each plot what to look for. See @fig-perf.

```{r fig-perf, fig.align='center', fig.cap="Output from package-performance", fig.height=8.5, fig.width=10}
check_model(lin_reg1)
```

With only one predictor there is no question of *multi-collinearity*, so `check_model()` cannot draw that plot and may give a message about it. Multi-collinearity is about one more assumption, which applies only when there are two or more predictors. The predictors should not be strongly related to each other. We will discuss it with multiple linear regression below.

## Using `lm` for predictions

The output of `lm()` is actually a list. It contains much more information than we saw above. Let us see what it contains.

```{r objs}
names(lin_reg1) |>
  as.data.frame() |>
  setNames('Objects')
```

We may extract any of these as needed. For example,

```{r}
lin_reg1$coefficients
```

The package `broom` gives all this information as a tidy data frame. Its function `augment()` gives the fitted value, residual, leverage (`.hat`), Cook's distance (`.cooksd`) and standardised residual (`.std.resid`) for every row. We will use these columns a lot in the audit example at the end of the chapter.

```{r aug}
augment(lin_reg1)
```

Let us also visualise the predicted values against the actual values and residuals, in @fig-predvsact.

::: {#fig-predvsact}

```{r echo=FALSE, fig.align='center', fig.show='hold', out.width="48%"}
newdata <- iris[1:50, ]
newdata$predictions <- predict(lin_reg1)
ggplot(newdata, aes(predictions, y = Sepal.Width)) +
  geom_point() +
  geom_abline(color = "blue") +
  labs(x = "Predicted Values",
       y = "Actual Values",
       title = "Predicted Values Vs. Actual Values") +
  theme_bw()

newdata$residuals <- lin_reg1$residuals
ggplot(newdata, aes(predictions, y = residuals)) +
  geom_pointrange(aes(ymin = 0, ymax = residuals), alpha = 0.5) +
  geom_hline(yintercept = 0, linetype = 3, color = "blue") +
  labs(x = "Predicted Values",
       y = "Residuals",
       title = "Predicted Values Vs. Residuals") +
  theme_bw()


```

Predicted Vs. Actual Values (Left) and Residuals (Right)
:::

To predict for new data, we use the base R function `predict()`. The new data must be a data frame with a column of the *same name* as the predictor. Let us predict sepal widths for three new flowers.

```{r preds}
new_vals <- data.frame(Sepal.Length = c(4.5, 5.0, 5.5))
predict(lin_reg1, new_vals)
```

A prediction is never exact. We can ask `predict()` for a *prediction interval*, the range in which the width of a new flower is likely to fall, 95 times out of 100.

```{r}
predict(lin_reg1, new_vals, interval = "prediction")
```

So a flower with a sepal length of 5 cm will most likely have a sepal width between about `r round(predict(lin_reg1, data.frame(Sepal.Length = 5), interval = "prediction")[2], 1)` cm and `r round(predict(lin_reg1, data.frame(Sepal.Length = 5), interval = "prediction")[3], 1)` cm.

::: callout-warning
## Caution

Predict only within the range of the data used to build the model. Our setosa flowers have sepal lengths from `r min(iris$Sepal.Length[1:50])` to `r max(iris$Sepal.Length[1:50])` cm. The model knows nothing about a flower with a 7 cm sepal, and its prediction there may be nonsense. Predicting outside the range is called *extrapolation*. In audit, a model built on small offices should not be used to judge a very large one.
:::

## Multiple Linear Regression

In multiple linear regression, several independent variables together explain the dependent variable. The regression equation becomes

$$
y = \beta_0 + \beta_1x_1 + \beta_2x_2 + \beta_3x_3 + ... + \epsilon
$$ {#eq-lr12}

where $x_1$, $x_2$, $x_3$ ..., $x_n$ are the $n$ independent variables and $y$ is the dependent variable as usual.

With two predictors, this equation is no longer a line but a flat *plane* in three dimensions. With more predictors, it is a *hyperplane*, which we cannot draw. So we usually look at the predictors one at a time, or use correlation plots.

As already stated, there is one more assumption now. **The independent variables should not be strongly related to each other, that is, there should be no multi-collinearity.** Let us build an example. We just add the predictors to the formula using the `+` operator.

**Problem Statement:** Example-2. Let us try to explain cars' mileage (`mpg` in the `mtcars` dataset) using engine displacement `disp`, horse power `hp` and weight `wt`. First, let us visualise each relationship separately, in @fig-ex2vis.

::: {#fig-ex2vis}

```{r echo=FALSE, warning=FALSE, message=FALSE, fig.align='center', out.width="31%", fig.show='hold'}

mtcars %>%
  ggplot(aes(disp, mpg)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE , formula = "y ~ x") +
  theme_bw()

mtcars %>%
  ggplot(aes(hp, mpg)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE , formula = "y ~ x") +
  theme_bw()

mtcars %>%
  ggplot(aes(wt, mpg)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE , formula = "y ~ x") +
  theme_bw()


```

Relationship between regressors and outcome variables
:::

Let us also visualise the correlation between all these variables. See @fig-ex2vis2.

```{r fig-ex2vis2, echo=FALSE, warning=FALSE, message=FALSE, fig.align='center', fig.cap="Correlation between the variables in Example-2"}
ggpairs(
  mtcars[, c("mpg", "disp", "hp", "wt")],
  upper = list(continuous = my_fn),
  lower = list(continuous = 'smooth', wrap=c(colour="blue")))
```

Notice that the predictors are strongly correlated with each other too. Displacement and weight, for example, have a correlation of `r round(cor(mtcars$disp, mtcars$wt), 2)`. Heavy cars have big engines. Keep this in mind. Now let us build the model.

```{r mult}
data <- mtcars[, c("mpg", "disp", "hp", "wt")]
lin_reg2 <- lm(mpg ~ .,
               data = data)
summary(lin_reg2)
```

In the formula, we used `.` instead of naming all the variables. The `.` is a shorthand for "all other columns of the data". There is now one slope for each predictor and one intercept. The equation can be written as @eq-lr13.

$$
{mpg} = -0.000937\cdot{disp} - 0.031157\cdot{hp} - 3.800891\cdot{wt} + 37.105505
$$ {#eq-lr13}

**Interpretation:** All the slopes are negative, so mileage drops as weight, displacement and horse power increase. Each slope is read *keeping the other variables constant*. For example, if two cars have the same horse power and weight, the one with 1 cubic inch more displacement gives 0.000937 miles per gallon less. "Keeping other variables constant" is important. If the other variables also change, the effect on mileage is different.

**Results:** The model explains about `r round(summary(lin_reg2)$r.squared * 100)`% of the variation in mileage (Multiple R-squared). Adjusted R-squared is a little lower, at `r round(summary(lin_reg2)$adj.r.squared * 100)`%, after the penalty for three predictors. The F-test p-value is very small, so at least one slope is not zero.

Look at `disp`, however. On its own, displacement is strongly related to mileage (@fig-ex2vis, left). In this model, its slope is tiny and its p-value is `r round(summary(lin_reg2)$coefficients["disp", 4], 2)`, not significant at all. This is because `disp` carries almost the same information as `wt` and `hp`. Once they are in the model, `disp` has little new to say. We could remove `disp` and re-run the model. We will come back to this when we discuss multi-collinearity.

## Including categorical or factor variables in `lm`

So far, both our response variable and our predictors were numeric. But what if a predictor is categorical, like the city class of an office, or the type of a contract?

In such a case, we should make sure that the categorical variable is of type `factor` before building the model. We learnt about factors in @sec-factors.

**Problem Statement:** Example-3. Let us try to explain `Sepal.Width` by `Species` in the `iris` dataset. This time we take the complete dataset. `Species` is already a factor, so we need not convert it. Before moving on, let us visualise the relationship. Box plots are best suited here.

```{r fig-fact1, echo=FALSE, fig.align='center', fig.cap="Species Vs. Sepal Width", warning=FALSE, fig.show='hold', out.height="30%"}
ggplot(iris, aes(Species, Sepal.Width)) +
  geom_boxplot() +
  stat_summary(fun = mean, shape = 15) +
  labs(x = "Species",
       y = "Sepal.Width") +
  theme_bw()
```

Now let us build the model.

```{r}
lm_fact <- lm(Sepal.Width ~ Species, data = iris)
summary(lm_fact)
```

We now got one intercept and *two* slopes for a single predictor `Species`. What happened?

**Interpretation:** R converts a factor into *dummy variables*. A dummy variable is 1 when a row belongs to a category and 0 otherwise. Being numeric, the dummies can be used in linear regression as before. The first level of the factor is taken as the *baseline*. It gets no dummy of its own. `Species` has three levels, so it is encoded as two dummy variables. To see this, we can use the `contrasts()` function.

```{r}
contrasts(iris$Species)
```

We can also see the actual numbers that went into the model with `model.matrix()`. Let us look at one flower of each species.

```{r}
model.matrix(Sepal.Width ~ Species, data = iris)[c(1, 51, 101), ]
```

When the flower is *versicolor*, `Speciesversicolor` is `1` and `Speciesvirginica` is `0`, and the other way round for *virginica*. When both are `0`, the flower is *setosa*, the baseline. That is why there is no separate coefficient for *setosa*. Now we can write the regression equation as @eq-lr14.

$$
{Sepal.Width} = 3.428 + (-0.658)\cdot{Speciesversicolor} + (-0.454)\cdot{Speciesvirginica}
$$ {#eq-lr14}

Reading this equation is now easy. For *versicolor*, the predicted `Sepal.Width` is `3.428 - 0.658`, or `r 3.428 - 0.658`. When both dummy variables are zero, the predicted width is just the intercept. So the intercept is the prediction for the baseline level, *setosa*, and each slope is the *difference* from the baseline. If we remove the intercept by adding `- 1` to the formula, R gives us one coefficient for each level directly.

```{r}
lm_fact0 <- lm(Sepal.Width ~ Species -1, data = iris)
summary(lm_fact0)
```

**Results:** Compare these coefficients with @fig-fact1. They are nothing but the mean sepal width of each species, which is logical. The equation without intercept is @eq-lr15.

$$
{Sepal.Width} = 3.428\cdot{Speciessetosa} + (2.77)\cdot{Speciesversicolor} + (2.974)\cdot{Speciesvirginica}
$$ {#eq-lr15}

Since these coefficients are not slopes in the true sense, we will call them intercepts in the next examples.

::: callout-warning
## Caution

Look at the R-squared of the two models. With an intercept it is `r round(summary(lm_fact)$r.squared, 3)`. Without an intercept it jumps to `r round(summary(lm_fact0)$r.squared, 3)`. The fit has not improved at all, since the predictions are exactly the same. Without an intercept, R measures R-squared against zero instead of against the mean, which makes it look much larger. So use `- 1` to *read* the coefficients easily, but compare models using the version with an intercept.
:::

### Parallel slopes regression

To understand how a categorical predictor works together with a numeric one, let us build a model step by step.

**Problem statement:** Example-4. We take the `mpg` dataset, which comes with the `ggplot2` package. It has *fuel economy data from 1999 to 2008 for 38 popular models of cars*. Let us explain highway mileage `hwy` by engine displacement `displ` and the year of the model `year`. Let us visualise the variables first. From @fig-ex4vis it is clear that highway mileage falls linearly as displacement increases, in both years.

```{r fig-ex4vis, fig.align='center', fig.cap="Highway Mileage Vs. Displacement over cars manufactured in 1999 Vs. 2008", fig.show='hold', out.height="35%"}
ggplot(mpg, aes(displ, hwy)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE, formula = "y ~ x") +
  facet_wrap(.~ year) +
  theme_bw()
```


**Step-1:** Let us first use only the numeric variable `displ` to explain highway mileage `hwy`, and look at the coefficients.

```{r}
par_slop1 <- lm(hwy ~ displ, data = mpg)
coef(par_slop1)
```

We got one intercept and one slope. We can see the line in @fig-parrslop (left).

**Step-2:** Now let us explain mileage by `year` of manufacture only. We need to convert it into a factor, which we can do directly in the formula. *We also add `- 1` to the formula, which removes the common intercept and gives one intercept for each year.* Now see the coefficients.

```{r}
par_slop2 <- lm(hwy ~ factor(year) - 1, data = mpg)
coef(par_slop2)
```

We got one intercept for each year. As @fig-parrslop (right) shows, these are nothing but the mean mileage for each year. The two means are almost equal.

::: {#fig-parrslop}

```{r warning=FALSE, echo=FALSE, fig.align='center', out.width="47%", fig.show='hold'}

ggplot(mpg, aes(displ, hwy)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE, formula = 'y ~ x') +
  labs(x = "Engine Displacement in Litres",
       y = "Highway Mileage in Miles per Gallon") +
  theme_bw()


ggplot(mpg, aes(factor(year), hwy)) +
  geom_boxplot() +
  stat_summary(fun = mean, shape = 15) +
  labs(x = "Year of Manufacture",
       y = "Highway Mileage in Miles per Gallon") +
  theme_bw()


```

Mileage vs. Displacement (Left) and Year (right)
:::

**Step-3:** Now we build the complete model with both variables together, and look at the coefficients.

```{r}
par_slop <- lm(hwy ~ displ + factor(year) - 1, data = mpg)
coef(par_slop)
```

There is one slope for the numeric variable and a separate intercept for each year. If we draw these, we get two *parallel* lines with the same slope. See @fig-parrslop2.

```{r fig-parrslop2, echo=FALSE, fig.align='center', fig.cap="Parallel Slopes", out.width="95%", out.height="35%", fig.show='hold'}
ggplot(mpg, aes(displ, hwy, color = factor(year))) +
  geom_point() +
  geom_parallel_slopes(se = FALSE) +
  labs(x = "Engine Displacement in Litres",
       y = "Highway Mileage in Miles per Gallon",
       color = "Year Manufactured") +
  theme_bw() +
  theme(legend.position = "bottom")
```

This is also clear from the equation, @eq-lr16.

$$
{hwy} = -3.610986\cdot{displ} + 35.275706\cdot{year1999} + 36.677842\cdot{year2008}
$$ {#eq-lr16}

For any car, one of the two dummy variables is 1 and the other is 0. The one with `1` acts as the intercept, but the slope stays the same. In other words, whatever the year, mileage changes with displacement at the same rate.

Notice something interesting. In Step-2, the two years had almost the same mean mileage. But now, for the *same* displacement, a 2008 car gives about `r round(diff(coef(par_slop)[2:3]), 1)` miles per gallon more than a 1999 car. The improvement was hidden because cars in 2008 had bigger engines on average (`r round(mean(mpg$displ[mpg$year == 2008]), 2)` litres against `r round(mean(mpg$displ[mpg$year == 1999]), 2)` litres). This is the power of multiple regression. It separates the effect of one variable from another.

But is it right to force the same slope on both years? Perhaps engines of 2008 lose mileage at a different rate as they grow. How do we allow for that? The answer is `interaction`, discussed next.

### Extending multiple linear regression by including `interactions`

The parallel slopes model forced a common slope on each category. That is not always the best option.

**Problem Statement:** In the same Example-4, we can allow an `interaction` between the two predictors using the operator `:` in the formula.

```{r interact}
new_model <- lm(hwy ~ displ + factor(year) + displ:factor(year) -1 , data = mpg)
coef(new_model)
```

Notice the extra coefficient. It is the *change* in slope when we move from the baseline year, 1999, to 2008. So the model now has a different slope for each year. See @fig-parrslop3.

```{r fig-parrslop3, echo=FALSE, fig.align='center', fig.cap="Changing Slopes with interaction", out.width="95%", fig.show='hold', out.height="35%"}
ggplot(mpg, aes(displ, hwy, color = factor(year))) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE, formula = "y~x") +
  labs(x = "Engine Displacement in Litres",
       y = "Highway Mileage in Miles per Gallon",
       color = "Year Manufactured") +
  theme_bw() +
  theme(legend.position = "bottom")
```

Let us see the summary.

```{r}
summary(new_model)
```

We may write the equation now (@eq-lr17).

$$
{hwy} = -3.7684\cdot{displ} + 35.7922\cdot{year1999} + 36.1367\cdot{year2008} + 0.3052\cdot{displ}\cdot{year2008}
$$ {#eq-lr17}

So for 2008 cars, the slope of `displ` is $-3.7684 + 0.3052$, or `r -3.7684 + 0.3052`. But look at the p-value of the interaction term, `r round(summary(new_model)$coefficients[4, 4], 2)`. It is far above 0.05. So the data does not give us any real evidence that the slopes differ between the two years. The two lines in @fig-parrslop3 are almost parallel too. Here the simpler parallel slopes model is good enough. Always check whether an interaction is significant before reading meaning into it. We will see an interaction which *does* matter in @sec-reg-audit.

We can add as many interactions as we like using `:`. When there are many, we can use another shorthand, `*`. So `x*z` means `x + z + x:z`, and `x*z*w` means `x + z + w + x:z + x:w + z:w + x:z:w`. It does not change the model, but saves a lot of typing.

## Multi-collinearity and Variance Inflation Factor (VIF)

Multi-collinearity means that the predictor variables are strongly related to each other. This creates a problem. When two predictors move together, the model cannot tell which of them is really affecting $y$. The coefficient estimates become unstable. A small change in the data can change them a lot, sometimes even flip their sign. Their standard errors become large, so useful variables may look insignificant. We saw this with `disp` in Example-2.

Note that multi-collinearity does not spoil the *predictions* of the model much. It spoils our ability to *interpret* the individual coefficients. For an auditor, who usually wants to explain a finding, that matters.

There are many ways to detect multi-collinearity. One is to use correlation plots, as in @fig-ex2vis2. Another is the *Variance Inflation Factor* or VIF. To find the VIF of a predictor, we regress it on all the other predictors and note the $R^2$ of that regression. The VIF score thus tells us how well a predictor is explained by the other predictors.

$$
{VIF} = \frac{1}{1-R^2}
$$ {#eq-lr18}

So the closer this $R^2$ is to $1$, the higher the ${VIF}$.

- $VIF$ starts at $1$ and has no upper limit.
- ${VIF} = 1$ means the predictor is not related to the other predictors at all.
- ${VIF}$ above $5$ is commonly taken to indicate a problem, and above $10$ a serious one.

In R, we do not have to calculate ${VIF}$ by hand. The function `vif()` from the package `car` does it. Let us first check our model of Example-2.

```{r}
car::vif(lin_reg2)
```

The VIF of `disp` is `r round(car::vif(lin_reg2)["disp"], 1)`, above 5. That confirms what we suspected. Now let us try a model on `mtcars` where `mpg` is explained by all the other variables. We again use the shorthand `.` for all other variables.

```{r}
mod <- lm(mpg ~ . , data = mtcars)
car::vif(mod)
```

Several predictors here have a VIF above 10. Such a model would be hard to interpret.

## One Complete Example

**Problem Statement:** Example-5. The data pertains to inter-country Life-Cycle Savings 1960-1970. Under the life-cycle savings hypothesis developed by *Franco Modigliani*, the savings ratio `sr` (aggregate personal saving divided by disposable income) is explained by per-capita disposable income `dpi`, the percentage rate of change in per-capita disposable income `ddpi`, and two demographic variables. These are the percentage of population less than 15 years old `pop15`, and the percentage of the population over 75 years old `pop75`. The data are averaged over the decade 1960–1970 to remove the business cycle or other short-term fluctuations.

Let us try linear regression on the `LifeCycleSavings` data.

```{r ex2}
# Visualise first 6 rows
head(LifeCycleSavings)

# Build a model
ex1 <- lm(sr ~ . , data = LifeCycleSavings)
```

In the formula, `.` again means all variables other than `sr`. Let us see the output.

```{r summ1}
summary(ex1)
```

Multiple R-squared is about `r round(summary(ex1)$r.squared * 100)`%, so the model explains about a third of the variation in savings ratios. Let us see some diagnostic plots (@fig-perf2).

```{r fig-perf2, fig.align='center', fig.cap="Diagnostic Plots", fig.height=8.5, fig.width=10}
performance::check_model(ex1)
```

We can notice some multi-collinearity between the two population variables in @fig-perf2. It is clearly visible in @fig-multi2. Countries with many children have few old people, and the other way round.

```{r fig-multi2, fig.align='center', fig.cap="Are pop15 and pop75 correlated?", out.height="30%"}
LifeCycleSavings %>%
  ggplot(aes(pop15, pop75)) +
  geom_point()+
  geom_smooth(method = 'lm', formula = 'y~x') +
  theme_bw()
```

This can also be seen in the correlation plots in @fig-corrp.

```{r fig-corrp, echo=FALSE,fig.align='center', fig.cap="Correlation Plots", out.height="45%", message=FALSE}
ggpairs(
  LifeCycleSavings,
  upper = list(continuous = my_fn),
  lower = list(continuous = 'smooth', wrap=c(colour="blue")))
```

Let us see the VIF of the predictors.

```{r vifs}
car::vif(ex1)
```

Both population variables have a VIF above 5. So let us remove one of them, `pop75`, and re-run the model.

```{r ex22}
ex2 <- lm(sr ~ pop15 + dpi + ddpi, data = LifeCycleSavings)
summary(ex2)
```

Multiple R-squared has dropped a little, to `r round(summary(ex2)$r.squared * 100, 1)`%. That is expected, since R-squared always falls when we remove a variable. But look at what we gained. The standard error of `pop15` has fallen from `r round(summary(ex1)$coefficients["pop15", 2], 3)` to `r round(summary(ex2)$coefficients["pop15", 2], 3)`. Its estimate is now much more precise. Let us also remove `dpi`, whose coefficient is not significant.

```{r ex33}
ex3 <- lm(sr ~ pop15 + ddpi, data = LifeCycleSavings)
summary(ex3)
```

Multiple R-squared falls slightly again, to `r round(summary(ex3)$r.squared * 100, 1)`%. But adjusted R-squared stays almost the same (`r round(summary(ex2)$adj.r.squared, 4)` before, `r round(summary(ex3)$adj.r.squared, 4)` now). So `dpi` was adding nothing useful, and the simpler model with two predictors does as well as the one with three. Both of its coefficients are significant. When two models explain the data about equally well, we prefer the simpler one.

Let us visualise the main relationship through a scatter plot.

```{r fig-ex5scatter, warning=FALSE, fig.align='center', fig.cap="Ascertaining linear relationship", out.height="30%"}
LifeCycleSavings %>%
  ggplot(aes(pop15, sr)) +
  geom_point() +
  geom_smooth(method = 'lm', formula = 'y ~ x', se = FALSE) +
  theme_bw()
```

The relationship between `pop15` and the savings ratio is real but not very strong. The points are widely scattered around the line. That is why the R-squared stays below 30%. Savings depend on many things this data does not have.

## Simpson's paradox

Edward Hugh Simpson, a statistician and former cryptanalyst at Bletchley Park, described this phenomenon in a paper in 1951. It is a classic example of how a regression that leaves out an important variable can mislead. The paradox arises when a trend seen in each subgroup of the data *reverses* or *disappears* when the subgroups are combined.

**Problem Statement:** Example-6. Let us take the `penguins` data from `palmerpenguins`, and try to establish a relationship between the depth and length of penguins' bills, `bill_depth_mm` and `bill_length_mm`. See the plots in @fig-ex3.

::: {#fig-ex3}

```{r echo=FALSE, fig.align='center', out.width="47%", fig.show='hold'}
penguins %>%
  na.omit() %>%
  ggplot(aes(bill_length_mm, bill_depth_mm)) +
  geom_point() +
  geom_smooth(method = 'lm', formula = 'y ~ x') +
  theme_bw()

penguins %>%
  na.omit() %>%
  ggplot(aes(bill_length_mm, bill_depth_mm, color = species)) +
  geom_point() +
  geom_smooth(method = 'lm', formula = 'y ~ x') +
  theme_bw() +
  theme(legend.position = 'bottom')
```

Regression having variable hidden (left) and exposed (right)
:::

In the left plot, the relationship is negative. Longer bills seem to be shallower. But when the hidden variable `species` is exposed (right), we see a *positive* relationship within each species. Let us confirm this with two models.

```{r}
peng <- na.omit(penguins)
coef(lm(bill_depth_mm ~ bill_length_mm, data = peng))
coef(lm(bill_depth_mm ~ bill_length_mm + species, data = peng))
```

Without species, the slope of bill length is `r sprintf('%.2f', coef(lm(bill_depth_mm ~ bill_length_mm, data = peng))[2])`. Once species is in the model, it becomes `r sprintf('%.2f', coef(lm(bill_depth_mm ~ bill_length_mm + species, data = peng))[2])`. The sign has flipped. Species is a *confounding* variable here. So before finalising a model, we must make sure that we are not missing an important variable. To avoid being misled by Simpson's paradox, we should

- **Identify confounding variables.** Think about what else could affect both variables under study.
- **Consider all levels.** Analyse the data at the subgroup level and at the combined level, and compare.
- **Use statistical techniques** such as multiple regression, as we just did, to control for confounding variables.
- **Report transparently.** Clearly state the method, assumptions and limitations of the analysis so that others can evaluate the findings.

::: callout-tip
## Audit tip

Simpson's paradox is common in audit data. A scheme may show a falling rejection rate in every district, yet a rising rate for the state as a whole, because the mix of districts has changed. Before reporting a trend on combined data, always check it within the main subgroups, like districts, years, or types of contract.
:::

## An audit example: offices that spend more than expected {#sec-reg-audit}

Let us now use regression the way an auditor would. Remember the question we started this chapter with. Salary expenditure of an office should depend mainly on its staff strength. If an office spends far more than its staff strength explains, it may have *ghost employees* on its pay roll, or other irregular payments. Regression gives us an *expected* expenditure for every office. The residual is then the *unexplained* part. Large positive residuals are our leads.

### Our sample data

We create the data of 90 sub-divisional offices (SDOs) of an imaginary department. For each office we have its staff strength, the class of the city where it is located, and its monthly salary expenditure in ₹ lakh. House rent allowance (HRA) is 30%, 20% and 10% of pay in X, Y and Z class cities respectively. So, for the same staff, an office in an X class city spends more.

We have planted a few cases.

- Three offices have salary expenditure inflated by 16%, as if ghost employees were added.
- In one office, the staff strength was keyed in with an extra zero.

As in other chapters, new readers may skip the code which creates the data.

```{r}
#| code-fold: true
#| code-summary: "Code to create the sample office data"
set.seed(2026)
n <- 90
offices <- tibble(
  office_id  = sprintf("SDO-%02d", 1:n),
  city_class = factor(sample(c("X", "Y", "Z"), n, replace = TRUE,
                             prob = c(0.2, 0.4, 0.4)),
                      levels = c("Z", "Y", "X")),
  staff      = sample(8:60, n, replace = TRUE)
) |>
  mutate(hra    = c(Z = 0.10, Y = 0.20, X = 0.30)[as.character(city_class)],
         # about ₹0.85 lakh basic cost per employee, plus HRA, plus noise
         salary = round(staff * 0.85 * (1 + hra) + rnorm(n, 0, 1.2), 2)) |>
  select(-hra)

# Planted case 1. Ghost employees, salary inflated by 16%
ghost <- c("SDO-17", "SDO-44", "SDO-63")
offices <- offices |>
  mutate(salary = if_else(office_id %in% ghost, round(salary * 1.16, 2), salary))

# Planted case 2. Staff strength keyed in with an extra zero
offices <- offices |>
  mutate(staff = if_else(office_id == "SDO-81", staff * 10L, staff))
```

```{r}
head(offices)
```

### First model, and a surprise

Let us fit a model of salary on staff strength and city class.

```{r}
off_m1 <- lm(salary ~ staff + city_class, data = offices)
summary(off_m1)
```

Something is wrong. Salary expenditure must surely rise steeply with staff strength. Yet the slope of `staff` is only `r round(coef(off_m1)["staff"], 3)` lakh per employee, and R-squared is only `r round(summary(off_m1)$r.squared * 100)`%. Let us look at the influence of each office, using `augment()` and Cook's distance.

```{r}
augment(off_m1, data = offices) |>
  arrange(desc(.cooksd)) |>
  select(office_id, staff, salary, .fitted, .hat, .cooksd) |>
  head(4)
```

One office, `SDO-81`, has a Cook's distance of `r round(max(cooks.distance(off_m1)))`, where the rest are below 0.1. Its staff strength is shown as `r offices$staff[offices$office_id == "SDO-81"]`, while no other office has more than `r max(offices$staff[offices$office_id != "SDO-81"])`. Its leverage (`.hat`) is close to 1, the maximum possible. This single wrong entry has pulled the whole line towards itself. @fig-reg-lever shows this.

```{r fig-reg-lever, fig.align='center', fig.cap="One wrong entry (red) pulls the whole line", out.height="35%"}
ggplot(offices, aes(staff, salary)) +
  geom_point(aes(color = office_id == "SDO-81"), show.legend = FALSE) +
  geom_smooth(method = "lm", formula = "y ~ x", se = FALSE) +
  scale_color_manual(values = c("grey40", "red")) +
  labs(x = "Staff strength (as recorded)", y = "Monthly salary (₹ lakh)") +
  theme_bw()
```

::: callout-tip
## Audit tip

Always check Cook's distance and leverage *before* reading the residuals. A single data entry error with an extreme value can spoil the whole model. It then makes genuine outliers look normal, and normal offices look odd. Such an entry is itself an audit finding about the quality of the data.
:::

On checking the records, suppose we find that the office actually has `r offices$staff[offices$office_id == "SDO-81"] / 10` staff. We should correct the entry. Here we simply keep it aside and re-fit.

### A better model

HRA is a *percentage* of pay. So each extra employee in an X class city costs more than an extra employee in a Z class city. In other words, the slope of `staff` should be different for each city class. This is exactly what an interaction does. Let us fit both the parallel slopes model and the interaction model on the clean data, and compare them.

```{r}
clean <- offices |> filter(office_id != "SDO-81")

off_m2 <- lm(salary ~ staff + city_class, data = clean)
off_m3 <- lm(salary ~ staff * city_class, data = clean)
summary(off_m3)
```

Both interaction terms are significant. For every extra employee, the expected salary goes up by `r round(coef(off_m3)["staff"], 2)` lakh in a Z class city, and by `r round(coef(off_m3)["staff"] + coef(off_m3)["staff:city_classX"], 2)` lakh in an X class city. R-squared is now `r round(summary(off_m3)$r.squared * 100, 1)`%.

To check formally whether the interaction model is better, we can compare the two models with `anova()`. It tests whether the extra terms reduce the residuals by more than chance would.

```{r}
anova(off_m2, off_m3)
```

The p-value is very small, so the interaction model is clearly better. Its residual standard error is `r round(sigma(off_m3), 2)` lakh, against `r round(sigma(off_m2), 2)` lakh for the parallel slopes model. The more accurate our expected values are, the easier it is to spot offices which do not fit.

### Finding the offices

Now let us look at the residuals. Raw residuals are in lakh. To judge how unusual each one is, we use the *standardised* residual (`.std.resid`). It is the residual divided by its standard error, so it is on a common scale. Under our assumptions, values beyond 3 are very rare, about 3 in 1,000. Let us also calculate the excess expenditure in rupees.

```{r}
flagged <- augment(off_m3, data = clean) |>
  mutate(excess = salary - .fitted) |>
  arrange(desc(.std.resid)) |>
  select(office_id, staff, city_class, salary, .fitted, excess, .std.resid)

head(flagged, 5)
```

```{r}
#| include: false
top2 <- flagged |> filter(.std.resid > 3)
sdo17 <- flagged |> filter(office_id == "SDO-17")
```

Two offices, `r paste(top2$office_id, collapse = " and ")`, have standardised residuals above 3. Together they spend about ₹`r round(sum(top2$excess), 1)` lakh a month, or about ₹`r round(sum(top2$excess) * 12 / 100, 1)` crore a year, more than expected. Both are planted ghost employee cases. They are clearly our first leads.

The third planted office, `SDO-17`, has a standardised residual of only `r round(sdo17$.std.resid, 2)`. Its excess of ₹`r round(sdo17$excess, 2)` lakh is within the normal variation between offices. Regression cannot separate it from genuine offices. Let us see all of this in @fig-reg-flags.

```{r fig-reg-flags, fig.align='center', fig.cap="Standardised residuals of offices. Offices beyond 3 are leads."}
augment(off_m3, data = clean) |>
  ggplot(aes(.fitted, .std.resid)) +
  geom_hline(yintercept = c(-3, 3), linetype = 2, color = "red") +
  geom_hline(yintercept = 0, color = "grey50") +
  geom_point(aes(color = office_id %in% ghost)) +
  geom_text(data = \(d) filter(d, abs(.std.resid) > 3 | office_id %in% ghost),
            aes(label = office_id), vjust = -0.8, size = 3) +
  scale_color_manual(values = c("grey40", "red"),
                     labels = c("Other offices", "Planted cases")) +
  labs(x = "Expected salary (₹ lakh)", y = "Standardised residual", color = NULL) +
  theme_bw() +
  theme(legend.position = "bottom")
```

::: callout-warning
## Caution

Regression finds only those deviations which are large compared to the normal variation between offices. A small office with a few ghost employees, like `SDO-17`, can hide in that variation. So a clean result from regression does not prove that there is no fraud. For individual offices, it must be followed by other tests, like matching the pay roll with the attendance and bank account data (see @sec-dup).
:::

::: callout-note
## What about large negative residuals?

Offices which spend much *less* than expected may also deserve a look. The reason may be vacant posts, salary not paid on time, or expenditure booked under the wrong head. Here no office has a standardised residual below -3.
:::

## Uses of regression in audit

We have seen that regression analysis is a powerful tool for an auditor. It may be used, inter alia, for

- **Detection of anomalies and outliers.** As in our office example, regression gives the *expected* value of an amount based on its drivers. Large residuals point to transactions or units which do not fit, and these may be errors, irregularities or fraud.
- **Analytical procedures.** The auditor can estimate what an expenditure *should* be, say fuel cost from vehicle kilometres, or electricity charges from the connected load, and compare it with what was booked.
- **Risk assessment.** Units with large unexplained deviations can be ranked and taken up first in the audit plan. This helps the auditor use limited time and staff well.
- **Benchmarking.** Regression can compare the performance of an organisation with similar ones, after adjusting for size and other factors. Deviations from the expected relationship highlight areas that deserve a closer look.

## A suggested workflow

For any regression analysis in audit, we may follow these steps.

1. **Understand the drivers.** Decide, from knowledge of the business, which variables should explain the amount under audit. Regression only confirms or rejects relationships which make sense.
2. **Visualise first.** Draw scatter plots and box plots to see whether the relationships look linear, and to spot obvious errors.
3. **Fit a simple model** with `lm()`, and read its `summary()`.
4. **Check influence.** Look at Cook's distance and leverage. Investigate and correct data entry errors before going further.
5. **Check the assumptions** with `plot()` or `check_model()`. Check multi-collinearity with `vif()` when there are several predictors.
6. **Improve the model.** Add factors and interactions that make business sense. Compare models with adjusted R-squared or `anova()`.
7. **Flag the residuals.** Use `augment()` and pick the units with large standardised residuals. Quantify the excess in rupees.
8. **Verify with records.** Examine the documents of the flagged units before drawing any conclusion.

::: callout-important
## Remember

A regression model shows association, not cause. A large residual means only that a unit does not fit the pattern of the others. It is a *lead*, not a *finding*. There may be a genuine reason which the model does not know about. Always examine the records before reporting.
:::

## Summary of functions used

| Function | Package | What it does |
|-----------------------|----------------|--------------------------------|
| `lm()` | base R (stats) | Fits a linear regression model |
| `summary()` | base R | Shows coefficients, p-values, R-squared and F-test of a model |
| `coef()`, `confint()` | base R (stats) | Gives the coefficients and their confidence intervals |
| `residuals()`, `fitted()`, `sigma()` | base R (stats) | Gives the residuals, fitted values and residual standard error |
| `predict()` | base R (stats) | Predicts for new data, optionally with prediction intervals |
| `plot()` on a model | base R | Draws the six diagnostic plots, chosen with `which` |
| `cooks.distance()`, `rstandard()` | base R (stats) | Gives Cook's distance and standardised residuals |
| `contrasts()`, `model.matrix()` | base R (stats) | Shows how factors are converted to dummy variables |
| `anova()` | base R (stats) | Compares two nested models |
| `augment()` | broom | Gives fitted values, residuals, leverage and Cook's distance as a data frame |
| `check_model()` | performance | Draws all diagnostic plots together, with guidance |
| `vif()` | car | Calculates the variance inflation factor of each predictor |
| `ggpairs()` | GGally | Draws a matrix of scatter plots and correlations |
| `geom_parallel_slopes()` | moderndive | Draws a parallel slopes model in ggplot2 |

: Functions used in this chapter {#tbl-regfunctions}
