```{r}
#| include: false
source("_common.R")
```

# Large Data with DuckDB and data.table {#sec-duckdb}

Audit data keeps getting bigger. A year of treasury vouchers of a State, the GST returns of a commissionerate, the billing records of an electricity distribution company, or the transactions of a scheme paid through PFMS easily run into crores of rows. Excel stops at 10,48,576 rows. R does better, but it normally reads the whole data into the computer's memory (RAM) before doing anything. With a few crore rows, that is slow, and on an ordinary office laptop it may not be possible at all.

**DuckDB** solves this problem. It is a database engine that runs *inside* R, with nothing to set up, no server, and no database administrator. It can read CSV and Parquet files directly from the disk, without loading them into R, and it is remarkably fast. Best of all, with the package **dbplyr** we use the same `dplyr` verbs we already know (@sec-DPLYRR). We write `filter()`, `summarise()` and `left_join()` as usual, and dbplyr quietly translates them into SQL, the language of databases, for DuckDB to run.

Later in the chapter we also meet **data.table**, a package that makes work on data in memory very fast, and compare the approaches.

Think of it like this. R is the officer who decides what is to be examined. DuckDB is a very fast assistant in the record room. Instead of having every register carried to the officer's table, the officer sends instructions, the assistant goes through the registers where they lie, and brings back only the summary asked for.

**Prerequisites**

```{r}
#| message: false
#| warning: false
library(tidyverse)
library(duckdb)   # the database engine; also loads DBI
library(dbplyr)   # translates dplyr code into SQL
```

Install them once with `install.packages(c("duckdb", "dbplyr"))`.

## Our data, a year of treasury vouchers {#sec-duckdb-data}

The treasuries of a State have given us the payment vouchers of the financial year 2024-25, as 12 monthly CSV files. Each row is one payment: the voucher number, the drawing and disbursing officer (DDO) who drew it, the date, the type of bill, the payee, the payee's bank account and the amount. There are 24 lakh rows in all, more than twice what Excel can open.

The code below creates the files in a temporary folder. In a real audit the files would arrive from the department, and this step would not exist. Two kinds of irregularity are planted, which we will find later.

```{r}
#| code-fold: true
#| code-summary: "Code to create the 12 monthly voucher files"
set.seed(2025)
n <- 2400000
ddo_master <- tibble(
  ddo_code   = sprintf("DDO%04d", 1:800),
  treasury   = sample(sprintf("TR%02d", 1:20), 800, TRUE),
  department = sample(c("Education", "Health", "Police", "Revenue", "Public Works",
                        "Agriculture", "Rural Development", "Water Resources"), 800, TRUE)
)
# More payments towards the end of the year, especially in March
month_wt <- c(7, 7, 8, 8, 8, 8, 8, 8, 8, 8, 9, 13)    # April to March
dates    <- seq(as.Date("2024-04-01"), as.Date("2025-03-31"), by = "day")
fy_month <- (month(dates) - 4) %% 12 + 1
date_wt  <- month_wt[fy_month] / tabulate(fy_month)[fy_month]

vouchers <- tibble(
  voucher_no   = sprintf("V%08d", 1:n),
  ddo_code     = sample(ddo_master$ddo_code, n, TRUE),
  voucher_date = sample(dates, n, TRUE, prob = date_wt),
  bill_type    = sample(c("Salary", "Contingent", "Works", "Grant", "Pension"), n, TRUE,
                        prob = c(0.45, 0.25, 0.10, 0.10, 0.10)),
  payee_id     = sprintf("P%07d", sample(1:400000, n, TRUE)),
  amount       = round(rlnorm(n, log(25000), 1.1))
) |>
  mutate(payee_account = paste0("AC", substr(payee_id, 2, 8), "0"))

# Planted 1. Contingent and works bills paid twice, under a new voucher number
dup <- vouchers |>
  filter(bill_type %in% c("Contingent", "Works")) |>
  slice_sample(n = 400) |>
  mutate(voucher_no = sprintf("V9%07d", row_number()))

# Planted 2. Salaries of 14 "employees" in three DDOs credited to one bank account
ghost <- vouchers |>
  filter(bill_type == "Salary") |>
  slice_sample(n = 14 * 12) |>
  mutate(payee_id      = rep(sprintf("P8%06d", 1:14), 12),
         ddo_code      = rep(c("DDO0117", "DDO0342", "DDO0655"), length.out = n()),
         payee_account = "AC00451230",
         amount        = round(runif(n(), 38000, 52000)))

vouchers <- bind_rows(vouchers, dup, ghost) |> arrange(voucher_date)

# Write one CSV file per month, as received from the treasuries
data_dir <- file.path(tempdir(), "treasury")
dir.create(data_dir, showWarnings = FALSE)
vouchers |>
  split(format(vouchers$voucher_date, "%Y_%m")) |>
  iwalk(\(d, m) write_csv(d, file.path(data_dir, paste0("vouchers_", m, ".csv"))))

rm(vouchers)   # from here on, the data exist only as files on the disk
```

Let us see what we have received.

```{r}
csv_files <- list.files(data_dir, pattern = "^vouchers_.*csv$", full.names = TRUE)
tibble(file = basename(csv_files), size_mb = round(file.size(csv_files) / 1e6, 1))
```

```{r}
#| include: false
csv_mb <- sum(file.size(csv_files)) / 1e6
```

Twelve files, `r round(csv_mb)` MB in all. Notice that we removed the data from R's memory at the end of the code above. From now on, we work on the files directly.

## Connecting to DuckDB {#sec-duckdb-connect}

First we start DuckDB and *connect* to it. `dbConnect()` comes from the package `DBI`, which works the same way for every database R can talk to.

```{r}
#| message: false
con <- dbConnect(duckdb())
```

This creates a database in memory, which disappears when we disconnect. Now we tell DuckDB where the files are. A single line of SQL creates a *view* named `vouchers`, a name for "all the rows in all the files matching this pattern". The `*` in the file name matches every month.

```{r}
csv_pattern <- file.path(normalizePath(data_dir, winslash = "/"), "vouchers_*.csv")
dbExecute(con, paste0("CREATE VIEW vouchers AS SELECT * FROM read_csv('", csv_pattern, "')"))
```

(`normalizePath(..., winslash = "/")` writes the folder path with forward slashes, which DuckDB prefers, even on Windows.) Finally, `tbl()` gives us an R object pointing at the view, which we can use like a data frame.

```{r}
vouchers <- tbl(con, "vouchers")
vouchers
```

It looks like a tibble, with two differences. The header says `# Source: SQL` and the database, and the number of rows is shown as `??`. DuckDB has only peeked at the first few rows to show us a preview. It has not read the files yet.

## Lazy queries and `collect()` {#sec-duckdb-lazy}

This is the one new idea in this chapter. A table from a database is **lazy**. When we write `dplyr` code on it, nothing is computed. dbplyr only writes down what we want. The work is done when we ask to see the result, and then only the *result* comes into R.

Let us count the rows.

```{r}
row_time <- system.time(
  n_rows <- vouchers |> count() |> pull(n)
)
n_rows
```

DuckDB read all twelve files and counted `r format(n_rows, big.mark = ",")` rows in `r round(row_time["elapsed"], 2)` seconds. Now a real summary, the number and value of payments in each month.

```{r}
monthly <- vouchers |>
  mutate(year = year(voucher_date), month = month(voucher_date)) |>
  summarise(payments = n(), crore = sum(amount, na.rm = TRUE) / 1e7, .by = c(year, month)) |>
  arrange(year, month)

monthly
```

Printing a lazy query shows the first rows of the result, which is enough to check that it is right. To bring the full result into R as an ordinary tibble, for a chart or for further work, we use **`collect()`**.

```{r}
monthly_df <- collect(monthly)
class(monthly_df)
```

```{r}
#| fig-cap: "Payments by month, 2024-25. March stands out."
#| fig-height: 3.5
monthly_df |>
  mutate(month_label = factor(month.abb[month], levels = month.abb[c(4:12, 1:3)])) |>
  ggplot(aes(month_label, crore)) +
  geom_col(fill = "steelblue") +
  labs(x = NULL, y = "₹ crore") +
  theme_minimal()
```

**What does dbplyr send to DuckDB?** `show_query()` prints the SQL it has written for us.

```{r}
show_query(monthly)
```

We never have to write this ourselves. But it is worth a look. Auditors who know SQL, or the "equations" of tools like IDEA and ACL, will recognise `SELECT`, `GROUP BY` and `ORDER BY`. And reading the SQL is a good way to confirm that dbplyr has understood what we meant.

::: callout-note
## Missing values in databases

In R, `sum(c(5, NA))` is `NA`, as we saw in @sec-basics. In SQL, missing values (called `NULL`) are simply *left out* of sums and averages. dbplyr warns about this the first time, unless we write `na.rm = TRUE`, as we did above, to show that we know. Keep the difference in mind. A total from a database never tells you that some amounts were missing, so count the missing values separately with `sum(is.na(amount))`.
:::

## Audit tests on the full data {#sec-duckdb-tests}

Now let us run some real tests, on all 24 lakh rows, exactly as we would on a small data frame.

### Which DDOs rushed their spending into March?

```{r}
march_rush <- vouchers |>
  summarise(total = sum(amount, na.rm = TRUE),
            march = sum(if_else(month(voucher_date) == 3, amount, 0), na.rm = TRUE),
            .by = ddo_code) |>
  mutate(march_share = march / total) |>
  slice_max(march_share, n = 10) |>
  collect()

march_rush |> mutate(across(c(total, march), \(x) round(x / 1e5, 1)),
                     march_share = round(100 * march_share, 1))
```

```{r}
#| include: false
overall_march <- monthly_df |> summarise(s = sum(crore[month == 3]) / sum(crore)) |> pull(s)
```

Across the State, March accounts for `r round(100 * overall_march, 1)`% of the year's payments. These DDOs spent `r round(100 * min(march_rush$march_share))`% or more in March alone, and would be the first to examine for the rush of expenditure.

### Duplicate payments

A *same-same-same* test (@sec-dup): the same bank account, the same amount and the same date, under different voucher numbers. We restrict it to contingent and works bills, where a repeated payment is suspicious. (Salaries and pensions legitimately repeat the same amount every month, though not on the same date.)

```{r}
duplicates <- vouchers |>
  filter(bill_type %in% c("Contingent", "Works")) |>
  summarise(times_paid = n(), .by = c(payee_account, amount, voucher_date)) |>
  filter(times_paid > 1)

duplicates |>
  summarise(groups = n(), excess_paid = sum(amount * (times_paid - 1), na.rm = TRUE)) |>
  collect()
```

```{r}
#| include: false
dup_sum <- duplicates |> summarise(groups = n(), excess = sum(amount * (times_paid - 1), na.rm = TRUE)) |> collect()
```

`r dup_sum$groups` groups of payments were made more than once, with ₹`r round(dup_sum$excess / 1e7, 2)` crore paid in excess. To examine them, we bring only these rows into R. A join with the full table gives the voucher details, and still only the matching rows are collected.

```{r}
dup_vouchers <- vouchers |>
  semi_join(duplicates, by = c("payee_account", "amount", "voucher_date")) |>
  arrange(payee_account, voucher_date) |>
  collect()
nrow(dup_vouchers)
head(dup_vouchers, 6)
```

### One bank account, many employees

Salaries of different employees should go to different bank accounts. An account receiving the salary of several persons may indicate *ghost employees*, whose pay is drawn into one person's account.

```{r}
shared_accounts <- vouchers |>
  filter(bill_type == "Salary") |>
  summarise(employees = n_distinct(payee_id), ddos = n_distinct(ddo_code),
            amount = sum(amount, na.rm = TRUE), .by = payee_account) |>
  filter(employees > 1) |>
  arrange(desc(employees)) |>
  collect()
shared_accounts
```

```{r}
#| include: false
sa <- shared_accounts |> slice(1)
```

One account, `r sa$payee_account`, received the salary of `r sa$employees` different payees, drawn by `r sa$ddos` DDOs, ₹`r round(sa$amount / 1e5, 1)` lakh in the year. The next step is to collect its vouchers, and ask the DDOs for the service records of these "employees".

```{r}
vouchers |>
  filter(payee_account == !!sa$payee_account) |>
  count(ddo_code, payee_id) |>
  arrange(ddo_code, payee_id) |>
  collect() |>
  head(8)
```

The first row is the account's genuine holder, a single payee in one DDO. The other 14 payees, in three other DDOs, all have their salaries credited to this one account. The `!!` in front of `sa$payee_account` tells dbplyr to insert the *value* of this R object into the query, rather than look for a column of that name in the database.

## Joining with data in R {#sec-duckdb-join}

Often a small table lives in R, like the DDO master with each DDO's department, and we want to join it with the large table in the database. `copy_to()` sends the small table to DuckDB, and the join is then done there.

```{r}
ddo_db <- copy_to(con, ddo_master, name = "ddo_master", overwrite = TRUE)

vouchers |>
  left_join(ddo_db, by = "ddo_code") |>
  summarise(crore = sum(amount, na.rm = TRUE) / 1e7, .by = department) |>
  arrange(desc(crore)) |>
  collect()
```

All the join checks of @sec-joins still apply. Compare row counts and totals before and after, and use `anti_join()` to find vouchers whose DDO code is not in the master.

```{r}
vouchers |> anti_join(ddo_db, by = "ddo_code") |> count() |> pull(n)
```

## When dbplyr cannot translate {#sec-duckdb-translate}

dbplyr knows how to translate most `dplyr` verbs, and many common functions, like `if_else()`, `case_when()`, `year()`, `month()`, `n_distinct()`, `str_detect()` and `round()`, into SQL. But it cannot translate a function it does not know, such as one we wrote ourselves. Remember the `fin_year()` function of @sec-lubri-audit.

```{r}
fin_year <- function(date) {
  start <- if_else(month(date) >= 4, year(date), year(date) - 1)
  paste0(start, "-", sprintf("%02d", (start + 1) %% 100))
}
```

```{r}
#| error: true
vouchers |> mutate(fy = fin_year(voucher_date)) |> head()
```

dbplyr does not look inside our function. It passes the name `fin_year` on to DuckDB as it is, and DuckDB has never heard of it. (Even if we copied the code in, `sprintf()` has no translation either.) There are two ways out.

1. **Filter and summarise in the database, then `collect()`, then use any R function** on the much smaller result. This is the usual pattern: let DuckDB cut the data down, and let R do the rest.
2. **Rewrite the logic with functions dbplyr knows.** Here, `if_else()`, `year()` and `month()` are enough to compute the starting year of the financial year inside the database.

```{r}
vouchers |>
  mutate(fy_start = if_else(month(voucher_date) >= 4, year(voucher_date), year(voucher_date) - 1)) |>
  count(fy_start) |>
  collect()
```

::: callout-caution
Rows in a database have no fixed order. Without `arrange()`, the same query may return rows in a different order each time, and `head()` may show different rows. Always `arrange()` before `head()`, or before comparing results.
:::

## Writing SQL directly {#sec-duckdb-sql}

Those who know SQL can send it straight to DuckDB with `dbGetQuery()`, which returns a data frame. This is also how audit queries written for other databases can be reused.

```{r}
dbGetQuery(con, "
  SELECT bill_type, COUNT(*) AS payments, ROUND(SUM(amount) / 1e7, 1) AS crore
  FROM vouchers
  GROUP BY bill_type
  ORDER BY crore DESC
")
```

## Parquet, a better format for large data {#sec-duckdb-parquet}

CSV files are simple, but large and slow to read, because every number is stored as text. **Parquet** is a file format designed for large data. It stores each column together, compressed, with its type. DuckDB can write and read it directly.

```{r}
parquet_file <- file.path(normalizePath(data_dir, winslash = "/"), "vouchers_2024_25.parquet")
dbExecute(con, paste0("COPY (SELECT * FROM vouchers) TO '", parquet_file, "' (FORMAT parquet)"))

parquet_mb <- file.size(parquet_file) / 1e6
c(csv_mb = round(csv_mb), parquet_mb = round(parquet_mb))
```

The same data take `r round(parquet_mb)` MB instead of `r round(csv_mb)` MB. And queries are faster still. `read_parquet()` takes the place of `read_csv()` in the view.

```{r}
dbExecute(con, paste0("CREATE VIEW vouchers_pq AS SELECT * FROM read_parquet('", parquet_file, "')"))
vouchers_pq <- tbl(con, "vouchers_pq")

pq_time <- system.time(
  by_type <- vouchers_pq |> count(bill_type) |> arrange(bill_type) |> collect()
)
by_type
```

This query took `r round(pq_time["elapsed"], 2)` seconds.

::: callout-tip
## Audit tip

When a large data set arrives, keep the original files untouched, record their hashes as in @sec-file-hash, and convert a working copy to Parquet once. All further analysis then runs on the Parquet file, which is smaller and much faster, and the original files remain as received, for evidence.
:::

## data.table, fast in memory {#sec-duckdb-datatable}

DuckDB is one answer to large data. The other well-known answer in R is the package **data.table**. It takes a different approach. It *does* load the data into memory, like `read_csv()`. It reads files somewhat faster, but its real strength is what comes after: summaries over many groups, joins and new columns, computed at remarkable speed. For data that fits in the computer's memory, which with today's laptops means up to a few crore rows, data.table is often the fastest tool in R. Many government and research teams use it for exactly that reason.

```{r}
#| message: false
#| warning: false
library(data.table)
```

::: callout-caution
`data.table` has its own functions named `year()`, `month()`, `quarter()` and a few others, which *mask* the lubridate functions of the same names once it is loaded. They give the same answers for dates, so nothing breaks here, but R prints a message about the conflict. If you need to be sure which one is used, write `lubridate::year()` or `data.table::year()`, as in @sec-doublec.
:::

### Reading files with `fread()`

`fread()`, "fast read", reads a CSV file using all the processor cores of the computer, and works out the column types by itself. To read our 12 files, we read each with `fread()` and stack them with `rbindlist()`, the data.table version of `bind_rows()`.

```{r}
fread_time <- system.time(
  dt <- rbindlist(lapply(csv_files, fread))
)
fread_time["elapsed"]
dim(dt)
```

```{r}
#| include: false
dt_mb <- as.numeric(object.size(dt)) / 1e6
```

All 24 lakh rows were read in `r round(fread_time["elapsed"], 1)` seconds, and take about `r round(dt_mb)` MB of memory. The result is a *data.table*, which is also a data frame, so every function that works on data frames works on it too.

### The `DT[i, j, by]` syntax

data.table has a compact syntax of its own, inside square brackets.

```
DT[i, j, by]
```

Read it as: take the data.table `DT`, keep the rows `i`, compute `j`, grouped `by`. Those who know SQL will see `WHERE`, `SELECT` and `GROUP BY`. Inside the brackets, column names are used directly, without `$`, and `.N` stands for the number of rows. `.()` is a short form of `list()`, used to name the results.

Our monthly summary becomes

```{r}
dt[, .(payments = .N, crore = sum(amount) / 1e7),
   by = .(year = year(voucher_date), month = month(voucher_date))][order(year, month)]
```

The second pair of brackets, `[order(year, month)]`, works on the result of the first. This *chaining* of brackets plays the part of the pipe.

### The same audit tests

The duplicate payments test. `%chin%` is a fast version of `%in%` for text.

```{r}
dt_dups <- dt[bill_type %chin% c("Contingent", "Works"),
              .(times_paid = .N),
              by = .(payee_account, amount, voucher_date)][times_paid > 1]
dt_dups[, .(groups = .N, excess_paid = sum(amount * (times_paid - 1)))]
```

One bank account, many employees. Here we work in two steps. First we keep the distinct pairs of account and payee with `unique()`, and count the payees of each account with `.N`. Then, only for the few accounts with more than one payee, we compute the details. `uniqueN()` counts distinct values, like `n_distinct()`.

```{r}
account_payee <- unique(dt[bill_type == "Salary", .(payee_account, payee_id)])
shared <- account_payee[, .N, by = payee_account][N > 1]

dt[bill_type == "Salary" & payee_account %chin% shared$payee_account,
   .(employees = uniqueN(payee_id), ddos = uniqueN(ddo_code), amount = sum(amount)),
   by = payee_account][order(-employees)]
```

::: callout-tip
Why two steps? data.table computes `sum()`, `mean()` and `.N` within groups with special optimised code, but not `uniqueN()`. With about four lakh accounts, `uniqueN()` would be called four lakh times, which takes close to a minute. Counting the distinct pairs first takes a fraction of a second. When a grouped calculation is unexpectedly slow, see whether the work can be reduced first, like this.
:::

The results are the same as from DuckDB, as they must be. Joining with the DDO master can be done with `merge()`, which works like `left_join()` when `all.x = TRUE`.

```{r}
merge(dt, as.data.table(ddo_master), by = "ddo_code", all.x = TRUE)[
  , .(crore = sum(amount) / 1e7), by = department][order(-crore)]
```

### Adding columns by reference with `:=`

In `dplyr`, `mutate()` returns a *new copy* of the data with the extra column. With 24 lakh rows, every copy costs time and memory. data.table's `:=` operator adds or changes a column *in place*, without copying. Here we add the financial year, using `fifelse()`, data.table's fast `if_else()`.

```{r}
dt[, fy_start := fifelse(month(voucher_date) >= 4, year(voucher_date), year(voucher_date) - 1)]
dt[, .N, by = fy_start]
```

Notice that we did not write `dt <- ...`. The column was added to `dt` itself.

::: callout-caution
Because `:=` changes the data in place, the change also shows up in any other name that points to the same data. `dt2 <- dt` does *not* make a copy in data.table. Use `dt2 <- copy(dt)` when you need a separate copy to modify.
:::

### dplyr and data.table side by side

| Task | dplyr | data.table |
|----------------|------------------------|------------------------|
| Keep rows | `filter(df, amount > 1e5)` | `dt[amount > 1e5]` |
| Choose columns | `select(df, ddo_code, amount)` | `dt[, .(ddo_code, amount)]` |
| Summarise by group | `summarise(df, total = sum(amount), .by = ddo_code)` | `dt[, .(total = sum(amount)), by = ddo_code]` |
| Count rows by group | `count(df, bill_type)` | `dt[, .N, by = bill_type]` |
| Add a column | `mutate(df, lakh = amount / 1e5)` | `dt[, lakh := amount / 1e5]` |
| Sort | `arrange(df, desc(amount))` | `dt[order(-amount)]` |
| Join | `left_join(df, ddo, by = "ddo_code")` | `merge(dt, ddo, by = "ddo_code", all.x = TRUE)` |

: The same tasks in dplyr and data.table {#tbl-dt-dplyr}

If you prefer `dplyr`'s verbs but want data.table's speed, the package `dtplyr` translates `dplyr` code into data.table code, much as dbplyr translates it into SQL.

## How much faster? {#sec-duckdb-speed}

Let us compare the approaches on two jobs: reading the 12 CSV files, and then running the duplicate payments test, which groups 24 lakh rows by three columns.

```{r}
dup_test_dplyr <- function(df) {
  df |>
    filter(bill_type %in% c("Contingent", "Works")) |>
    count(payee_account, amount, voucher_date) |>
    filter(n > 1)
}

# readr and dplyr
read_readr <- system.time(all_rows <- read_csv(csv_files, show_col_types = FALSE))
dup_dplyr  <- system.time(dup_test_dplyr(all_rows))
in_memory_mb <- as.numeric(object.size(all_rows)) / 1e6
rm(all_rows)

# data.table
read_dt <- system.time(dt <- rbindlist(lapply(csv_files, fread)))
dup_dt  <- system.time(
  dt[bill_type %chin% c("Contingent", "Works"), .N, by = .(payee_account, amount, voucher_date)][N > 1]
)
rm(dt)

# DuckDB, on the CSV files and on the Parquet file
dup_duck <- system.time(vouchers |> dup_test_dplyr() |> collect())
dup_pq   <- system.time(vouchers_pq |> dup_test_dplyr() |> collect())
```

Notice that the same function, `dup_test_dplyr()`, ran on a tibble in memory and on the DuckDB tables. That is the convenience of dbplyr.

| Approach | Reading the files (seconds) | Duplicate test (seconds) | Memory used by R |
|----------|---------------|---------------|------------------|
| readr and dplyr | `r round(read_readr["elapsed"], 1)` | `r round(dup_dplyr["elapsed"], 2)` | about `r round(in_memory_mb)` MB |
| data.table | `r round(read_dt["elapsed"], 1)` | `r round(dup_dt["elapsed"], 2)` | about `r round(dt_mb)` MB |
| DuckDB on the CSV files | not needed | `r round(dup_duck["elapsed"], 2)` | only the result |
| DuckDB on the Parquet file | not needed | `r round(dup_pq["elapsed"], 2)` | only the result |

: Reading 24 lakh rows and testing them for duplicate payments {#tbl-duckdb-speed}

The times will differ on your computer, but the pattern will not. Reading is only somewhat faster with `fread()`, because most of the time goes in turning text into columns. The grouping is where data.table shines, many times faster than dplyr on the same data in memory. DuckDB needs no reading step at all, since it works on the files where they lie, and it is fastest of all on Parquet. As the data grows to several crore rows, `read_csv()` and even `fread()` may fail for lack of memory, while DuckDB works through the files on disk without trouble.

**Which one to use?**

| Situation | Good choice |
|------------------------|------------------------|
| Data fits comfortably in memory, and you like dplyr | dplyr, as in the rest of this book |
| Data fits in memory, but dplyr feels slow, or the same heavy job is run often | data.table |
| Data is larger than memory, or arrives as many large files | DuckDB with dbplyr |
| Data will be queried repeatedly over weeks | DuckDB, with the data in Parquet or a database file |

: Choosing a tool for large data {#tbl-large-choice}

## A database file for repeated work {#sec-duckdb-file}

So far, DuckDB lived in memory and read the files afresh for every query. For data we will query again and again, over days, we can keep a DuckDB *database file*. Give `dbConnect()` a file name, and load the data into a table once.

```{r}
#| eval: false
con <- dbConnect(duckdb(), dbdir = "treasury_2024_25.duckdb")
dbExecute(con, "CREATE TABLE vouchers AS SELECT * FROM read_parquet('vouchers_2024_25.parquet')")
# ... next day, in a new R session
con <- dbConnect(duckdb(), dbdir = "treasury_2024_25.duckdb")
vouchers <- tbl(con, "vouchers")
```

Whichever way we connect, we disconnect at the end, which also frees the memory.

```{r}
dbDisconnect(con)
```

## A suggested workflow

1. **Keep the original files untouched**, and record their hashes.
2. **Connect** with `dbConnect(duckdb())`, and create a view over the files with `read_csv()`, using a `*` pattern for many files.
3. **Point `tbl()` at the view**, and check a preview and the row count against the control totals received with the data.
4. **Convert to Parquet** if the data will be used more than once, and query that instead.
5. **Write the analysis in `dplyr`**, as usual. Use `show_query()` when in doubt.
6. **Filter and summarise in DuckDB**, and `collect()` only the small result, or the exceptions.
7. **Use R-only functions after `collect()`**, or rewrite them with translatable functions.
8. **Disconnect** at the end.

::: callout-important
## Remember

- DuckDB runs inside R and queries large CSV and Parquet files on disk, without loading them into memory.
- With dbplyr, we use the same `dplyr` verbs; they are translated into SQL.
- Database tables are lazy. Nothing happens until we print or `collect()` the result.
- Let DuckDB cut the data down, and bring only the result into R.
- Missing values are silently left out of SQL sums; check for them separately.
- Parquet files are smaller and faster than CSV; keep the originals for evidence.
- data.table is the fastest choice when the data fits in memory; DuckDB when it does not.
:::

## Summary of functions used

| Function | Package | What it does |
|-----------------------|----------------|--------------------------------|
| `dbConnect(duckdb())`, `dbDisconnect()` | DBI, duckdb | Start and close a DuckDB database, in memory or in a file |
| `dbExecute()` | DBI | Run an SQL command, like creating a view or writing Parquet |
| `dbGetQuery()` | DBI | Run an SQL query and get a data frame |
| `read_csv()`, `read_parquet()` (inside SQL) | DuckDB | Read CSV or Parquet files, including many files with a `*` pattern |
| `tbl()` | dplyr | Point at a table or view in the database |
| `collect()` | dplyr | Run the query and bring the result into R |
| `show_query()` | dplyr | Show the SQL that dbplyr has written |
| `copy_to()` | dplyr | Send a small R table to the database, for joins |
| `!!` | rlang | Insert the value of an R object into a database query |
| `fread()`, `rbindlist()` | data.table | Read CSV files fast, and stack a list of tables |
| `DT[i, j, by]`, `.N`, `.()`, `uniqueN()`, `%chin%` | data.table | Filter, compute and group; count rows and distinct values |
| `:=`, `fifelse()`, `copy()` | data.table | Add columns in place, fast `if_else()`, and make a true copy |
| `merge()` | base R, data.table | Join two tables |
| `normalizePath()` | base R | Write a folder path with forward slashes for DuckDB |
| `system.time()` | base R | Time a piece of code |

: Functions used in this chapter {#tbl-duckdbfunctions}
