```{r}
#| include: false
source("_common.R")
```

# Time Series Analysis {#sec-ts}

Every office keeps figures month by month. A treasury records the expenditure of each department every month. A sub-registrar office records the stamp duty it collects. A commercial tax circle records its tax collections every quarter. An auditor who looks at only one month, or only one year, sees a single number. An auditor who looks at the whole run of months sees a *pattern*. And a pattern tells us what to expect.

That is the idea behind this chapter. If receipts have grown by about 8% every year, and always peak in March, then we know roughly what this year's receipts should be. If the actual receipts fall well short, we have a lead. Perhaps some revenue has leaked. In the same way, if a department always spends much more in March, we can measure that *March rush*, and spot a March which is unusual even for March.

A sequence of values recorded over time, in order, is called a **time series**. Monthly expenditure, daily cash balance, quarterly GST collections and yearly excise revenue are all time series. Time series analysis means studying such data to understand its patterns, and to predict its future values. So it serves for description as well as prediction.

A few uses in audit are these.

- **Revenue analysis.** An auditor may examine revenue over time to find irregular patterns. Unexpected falls, unusual growth, or a change in the usual seasonal pattern may point to misstatement, leakage or fraud. We will see a complete example at the end of the chapter, in @sec-ts-audit.
- **Expenditure analysis.** Government offices are known to spend much more in March, to use up the budget before the financial year ends. Time series methods let us measure this March rush, and find months which are unusual even after allowing for it. The anomaly detection chapter takes this further, in @sec-ano-ts.
- **Inventory analysis.** Auditors examine stock levels to judge whether they are adequate, and to find obsolescence or shrinkage. Time series methods help us understand the usual pattern of stock, and see when it changes.

In this chapter we will learn how to turn monthly figures into a time series in R, and how to plot it. We will separate a series into its trend and its seasonal pattern. We will forecast it with simple and advanced methods, and test which method forecasts best. Finally, we will use a forecast to judge whether a year's stamp duty receipts are believable.

## Our data

All the series in this chapter are made up. They are realistic, but no real office or state is meant. We use five of them.

- `spend`, the monthly expenditure of a state's rural development department, in ₹ crore, from April 2014 to March 2024. It grows by about 7% a year, and has a strong March rush. We have planted one irregularity in it. In March 2021 the department spent far more than even its usual March rush. This is the kind of year-end spending that an auditor would want to examine, to see whether funds were parked or drawn only to avoid lapse.
- `fines`, the yearly penalties collected by a district transport office, in ₹ lakh, from 1985 to 2024. It goes up and down from year to year, around a steady level.
- `royalty`, the yearly mineral royalty collected by a district mining office, in ₹ lakh, from 1975 to 2024. It rises and falls with booms and slumps in construction.
- `ctax`, the quarterly commercial tax collections of a state, in ₹ crore, from the April to June quarter of 2005 to the January to March quarter of 2024.
- `excise`, the yearly state excise revenue of a state, in ₹ crore, from 1995 to 2023.

New readers may skip the code which creates the data.

```{r}
#| code-fold: true
#| code-summary: "Code to create the five series"
set.seed(2025)

# Monthly expenditure of the rural development department, Rs crore
# April 2014 to March 2024. Growth of 7% a year and a March rush.
n <- 120
spend_season <- c(0.55, 0.70, 0.80, 0.75, 0.80, 0.90,   # April to September
                  0.95, 1.00, 1.05, 1.10, 1.25, 2.15)   # October to March
x <- 40 * 1.07^((0:(n - 1)) / 12) * rep(spend_season, 10) * exp(rnorm(n, 0, 0.06))
# Planted case. March 2021 (month 84) is 45% above its usual March rush
x[84] <- x[84] * 1.45
spend <- ts(round(x, 1), start = c(2014, 4), frequency = 12)

# Yearly penalties of a district transport office, Rs lakh, 1985 to 2024
fines <- ts(round(120 + 6 * arima.sim(list(ar = 0.5), 40), 1), start = 1985)

# Yearly mineral royalty of a district mining office, Rs lakh, 1975 to 2024
# Booms and slumps of uneven length, 6 to 10 years each
phase   <- cumsum(2 * pi / runif(50, 6, 10))
royalty <- ts(round(200 + 60 * sin(phase) + rnorm(50, 0, 12), 1), start = 1975)

# Quarterly commercial tax of a state, Rs crore, April-June 2005 to Jan-March 2024
ctax_season <- c(0.92, 0.95, 1.00, 1.13)   # Q1 (Apr-Jun) to Q4 (Jan-Mar)
ctax <- ts(round(1500 * 1.09^((0:75) / 4) * rep(ctax_season, 19) * exp(rnorm(76, 0, 0.03))),
           start = c(2005, 2), frequency = 4)

# Yearly state excise revenue, Rs crore, 1995 to 2023
excise <- ts(round(800 * 1.10^(0:28) * exp(rnorm(29, 0, 0.04))), start = 1995)
```

Let us look at the first of them. The monthly expenditure is printed in a calendar layout, one row for each calendar year.

```{r}
spend
```

@fig-tsex plots the first four series. Each shows a different kind of pattern.

```{r fig-tsex, echo=FALSE, fig.cap="Four audit time series", fig.align='center', fig.show='hold'}
par(mfrow = c(2,2))
plot(spend, main = "(a) Monthly expenditure, Rs crore", xlab = NULL, ylab = NULL)
plot(fines, main = "(b) Yearly transport penalties, Rs lakh", xlab = NULL, ylab = NULL)
plot(royalty, main = "(c) Yearly mineral royalty, Rs lakh", xlab = NULL, ylab = NULL)
plot(ctax, main = "(d) Quarterly commercial tax, Rs crore", xlab = NULL, ylab = NULL)
```

To analyse any time series, we have to understand its different components.

## Components of a time series

A time series can be broken down into several components.

- **Level.** The baseline value or average of the series. If a series were a straight horizontal line, it would have only a level. The transport penalties in @fig-tsex (b) wander up and down, but always come back to about the same level. They have no lasting rise or fall.
- **Trend.** The long-term direction of the series. It shows whether the series is rising, falling or staying about constant over time. For example, a clear rising trend can be seen in the commercial tax collections in @fig-tsex (d).
- **Seasonality.** A pattern which repeats at a *fixed and known* interval, like every year, every quarter or every week. In @fig-tsex (d), commercial tax rises and falls in the same way every year, with the January to March quarter on top. In @fig-tsex (a), expenditure peaks every March.
- **Cycle.** Rises and falls which are *not* of a fixed length. The mineral royalty in @fig-tsex (c) rises in construction booms and falls in slumps. Each boom and slump lasts 6 to 10 years, and the length varies from one to the next. That is why they form a cycle and not a season.
- **Errors (residuals).** The random ups and downs that cannot be explained by the level, trend, seasonality or cycle. The year-to-year changes in the transport penalties in @fig-tsex (b) are mostly of this kind. The error component tells us how variable the data is, and so how much we can trust a forecast.

## Additive or multiplicative components

The trend, seasonal and error components can combine with each other in two ways, additive or multiplicative. Which one fits depends on the data.

An **additive model** assumes that the components are *added* together. The value of the series at any time is the sum of the trend, seasonal and error components at that time.

$$
y_t = T_t + S_t + e_t
$$ {#eq-ts-add}

where $y_t$, $T_t$, $S_t$ and $e_t$ are the values of the series, trend component, seasonal component and residuals, respectively, at time $t$. In an additive series, the seasonal ups and downs stay about the same size, whatever the level of the series.

A **multiplicative model** assumes that the components are *multiplied* together.

$$
y_t = T_t \times S_t \times e_t
$$ {#eq-ts-mult}

where $y_t$, $T_t$, $S_t$ and $e_t$ are defined as before. In a multiplicative series, the seasonal ups and downs grow as the series grows. The seasonal component is then a *percentage* of the trend, like "March is 40% above an average month". Look at the expenditure in @fig-tsex (a). The March peaks rise higher every year, because the March rush is a share of a growing budget. So its seasonality is multiplicative.

We can convert a multiplicative series into an additive one by taking its logarithm, because $\log(T \times S \times e) = \log T + \log S + \log e$. See @fig-logts. On the log scale, the March peaks of `spend` stand about the same height above the other months every year.

::: {#fig-logts}

```{r echo=FALSE, warning=FALSE, message=FALSE, fig.show='hold', fig.align='center', out.width="49%"}
library(ggplot2)
library(forecast)
autoplot(spend) +
  theme_bw() +
  labs(title = "Monthly expenditure (Rs crore)", x = NULL, y = NULL)
autoplot(log(spend)) +
  theme_bw() +
  labs(title = "Logarithmic transformation", x = NULL, y = NULL)
```

Transforming a time series
:::

We will learn how to separate these components in later sections.

**Damped trends.** Gardner and McKenzie [@mcke] observed in 1985 that most forecasting methods assume a trend will continue forever, however far ahead we forecast. Based on their findings, they suggested that forecasts become more accurate if trends are *damped*, that is, made to flatten out gradually. We will learn about this idea as well in later sections.

## Working with time series in R

Let us first see how time series objects are handled in R.

### Prerequisites {.unnumbered}

Most of the time we will use base R. For forecasting, we will use `forecast`, a fabulous package developed by **Rob Hyndman**. Forecasting future values also tells us how well we have captured the trends and patterns in a series. We will also use `zoo` for rolling averages, `patchwork` for combining plots, and `tidyverse` for the audit use case.

```{r message=FALSE, warning=FALSE}
library(forecast)
library(zoo)
library(patchwork)
library(tidyverse)
```

### Creating a time series object in R

R provides a simple function `ts()` to convert a series of observations (basically a vector) into a time series object. The main arguments are

```
ts(data,
  start = 1,
  end = numeric(),
  frequency = 1,
  ...)
```

where

- `data` is a numeric vector or matrix containing the observations.
- `start` is the time of the first observation. It is a single number, like `1985` for yearly data, or a pair of numbers, like `c(2022, 4)` for April 2022 in monthly data.
- `end` is the time of the last observation, written the same way. We usually give only `start`, and R works out `end` from the length of the data.
- `frequency` is the number of observations per unit of time. Monthly data has a frequency of 12, quarterly data 4, and yearly data 1.
- `...` are other, rarely needed arguments, like `names` for the columns of a matrix of series.

Let us create a small time series. A block office spent the following amounts, in ₹ lakh, in each month of the financial year 2022-23.

```{r}
# Monthly expenditure of a block office in FY 2022-23, Rs lakh
block_exp <- c(12, 15, 18, 16, 17, 20, 22, 21, 24, 26, 30, 58)

# A monthly time series starting from April 2022
block_ts <- ts(block_exp, start = c(2022, 4), frequency = 12)

# Print the time series object
block_ts
```

Notice that R has placed the 12 values from April 2022 to March 2023 in a calendar layout. The March rush is already visible. We can check the class of the object.

```{r}
class(block_ts)
```

Let us also print `fines`. It is a yearly series, so it has a frequency of 1.

```{r}
fines
```

We can take out a part of a time series with `window()`. For example, the penalties from 2015 onwards are

```{r}
window(fines, start = 2015)
```

For a monthly series, `start` and `end` are again pairs. The expenditure of the financial year 2023-24 alone is `window(spend, start = c(2023, 4), end = c(2024, 3))`.

### Plotting time series objects

To plot a `ts` object we can simply use `plot()`. Alternatively we can use `autoplot()`, which draws the series with `ggplot2`.

```{r fig-tsex2, fig.cap="Plotting a time series in base R (left) and ggplot2 (right)", fig.align='center', fig.show='hold', out.width="48%", warning=FALSE, message=FALSE}
# Plotting in base R
plot(block_ts, main = "Block office expenditure, FY 2022-23",
     ylab = "Rs lakh")

# Plotting with ggplot2, on a log scale
autoplot(spend) +
  scale_y_log10('Rs crore, log scale') +
  ggtitle('Department expenditure on a log scale') +
  theme_bw()
```

`autoplot()` is a function of `ggplot2`. The package `forecast` teaches it how to draw time series and forecasts. The plot is a `ggplot2` object, so it can be fine-tuned with the usual `ggplot2` functions, as we did with `scale_y_log10()` above.

## Time series modelling and forecasting

Forecasting the future is an important part of time series analysis. For an auditor, a forecast is an *expectation*. It says what the figures should have been if the office behaved as it always did. To forecast, we first need to *model* the data, that is, describe its pattern. Forecasting techniques can be broadly divided into two groups. *Data-based techniques*, like averaging, smoothing and decomposition, work directly on the observed values. *Model-based techniques*, like ARIMA, assume a particular statistical structure for the data and estimate its parameters. In practice the line between the two is not sharp. We will start with the simplest ones.

### Naive method

This method assumes that the future value will be the same as the last observed value. It is the simplest and most basic forecasting method, and hence named *naive*. The formula is

$$
\text{Naive method: }\hat{Y}_{T+1} = Y_T
$$ {#eq-ts-naive}

Here, $\hat{Y}_{T+1}$ is the forecast for the next period and $Y_T$ is the last observed value, in a series of $T$ observations. So predicting with the naive method needs no special skill. Yet it is useful. Any better method should at least beat the naive forecast.

Let us forecast the transport penalties for 5 future years. The package `forecast` has a function `naive()` which does this, once we give the value of the parameter `h`, short for forecasting **h**orizon. In @fig-naive, the forecast values (the blue line) are exactly the same as the last observed value. The shaded bands show the range in which the actual values are likely to fall.

```{r fig-naive, fig.show='hold', fig.align='center', fig.cap="Naive forecasting", out.width="98%"}
forecast::naive(fines, h = 5) %>%
  autoplot() +
  theme_bw() +
  labs(title = "Transport penalties, forecast by the naive method",
       x = "Year", y = "Rs lakh")
```

By default, 80% and 95% prediction intervals are shown as bands. We can choose other levels with the parameter `level`, for example `level = 90`. If we do not want the bands at all, we can use `autoplot(..., PI = FALSE)`.

For seasonal data, there is a variant called the *seasonal naive* method, `snaive()`. It forecasts each month as the value of the same month last year. Auditors use this idea all the time, when they compare this March with last March.

::: callout-tip
## Audit tip

For a seasonal series, never compare a month with the month just before it. March expenditure will always look alarming next to February. Compare it with the same month of earlier years, as `snaive()` does, or with a forecast that allows for the season.
:::

### Moving averages

In this technique, the forecast is the average of the most recent $n$ observations. As new data comes in, the window of $n$ observations slides forward. A moving average smooths a time series by averaging neighbouring values. It reduces noise and reveals the underlying trend.

The formula for forecasting with a moving average is

$$
\hat{Y}_{t+1} = \frac{1}{n}(Y_t + Y_{t-1} + ... + Y_{t-n+1})
$$ {#eq-ts-ma}

Here, $\hat{Y}_{t+1}$ is the forecast for the next period and $Y_t$, $Y_{t-1}$, ..., $Y_{t-n+1}$ are the past $n$ observations.

A moving average can be *centred* or *trailing*.

- In a **centred** moving average of order $k$, each value is replaced by the mean of itself, the $(k-1)/2$ values before it and the $(k-1)/2$ values after it. So $k$ is naturally an odd number. For an even $k$, like 12 for monthly data, `ma()` from `forecast` takes care of centring automatically. Centred averages are used to *see* the trend in past data.
- In a **trailing** moving average, each value is replaced by the mean of itself and the $k-1$ values *before* it. It uses only past data, so it is the one we use for *forecasting*. The function `rollmeanr()` from `zoo` calculates it.

Smoothed versions of the monthly expenditure for different $k$ can be seen in @fig-ma. A 7-month average softens the March peaks but does not remove them. A 12-month average removes them almost fully, because each average contains exactly one March. Notice also that the trailing average lags behind the data, since it uses only the past.

```{r fig-ma, echo=FALSE, fig.align='center', fig.cap="Simple moving averages of monthly expenditure", fig.show='hold', warning=FALSE}
lims <- c(min(spend), max(spend))

p1 <- autoplot(spend) +
  scale_y_continuous(limits = lims) +
  labs(title = 'Expenditure, original', x = NULL, y = 'Rs crore') +
  theme_bw()

p2 <- autoplot(ma(spend, 7)) +
  scale_y_continuous(limits = lims) +
  labs(title = 'Moving average k=7, centred', x = NULL, y = 'Rs crore') +
  theme_bw()

p3 <- autoplot(rollmeanr(spend, 12, fill = NA)) +
  scale_y_continuous(limits = lims) +
  labs(title = 'Moving average k=12, trailing', x = NULL, y = 'Rs crore') +
  theme_bw()

p4 <- autoplot(ma(spend, 12)) +
  scale_y_continuous(limits = lims) +
  labs(title = 'Moving average k=12, centred', x = NULL, y = 'Rs crore') +
  theme_bw()

(p1 + p2)/(p4 + p3)
```

Let us now forecast with a moving average. For the transport penalties, we take a trailing moving average of the last 7 years with `rollmeanr()`. The forecast for every future year is then simply the last value of this moving average. Applying `naive()` to the moving average gives exactly that, as in the left side of @fig-ma1.

In the right side of @fig-ma1, we have done the same with the monthly expenditure. This series has a strong seasonal pattern, so we have used a 12-month moving average. An average over exactly one full year contains every month once, so the seasonal ups and downs cancel out and only the trend remains. We have also drawn the original series in black, so we can see both the original and the smoothed series.

::: {#fig-ma1}

```{r warning=FALSE, message=FALSE, fig.align='center', fig.show='hold', out.width="48%"}
# Transport penalties, 7 year trailing moving average
fines_ma <- rollmeanr(fines, k = 7)
naive(fines_ma, h = 10) |>
  plot(ylim = range(fines), col = "red",
       main = "Forecasting\ntransport penalties")
lines(fines)

# Monthly expenditure, 12 month trailing moving average
naive(rollmeanr(spend, k = 12), h = 12) |>
  plot(ylim = range(spend),
       xlim = c(2014, 2025.5),
       col = "red",
       main = "Forecasting\nmonthly expenditure")
lines(spend)
```

Forecasting with simple moving averages
:::

::: callout-warning
## Caution

A moving average forecast is flat. It cannot follow a trend or a seasonal pattern into the future. Look at the expenditure forecast in @fig-ma1. It gives the average month, not the March rush that we can clearly expect. An auditor who used it would find every March "unusual". For such data we need the methods discussed next.
:::

### Exponential smoothing

Exponential smoothing methods forecast with weighted averages of past observations, giving more weight to recent observations. Examples are simple exponential smoothing, Holt's method and the Holt-Winters method.

**Simple exponential smoothing (SES)** is the simplest of these. It is a weighted moving average in which the weights fall *exponentially* as the observations get older. *This method is suitable for data with no clear trend or seasonal pattern.*

$$
\hat{y}_{T+1} = \alpha{y_T} + \alpha(1- \alpha){y_{T-1}} + \alpha(1 - \alpha)^2{y_{T-2}} + ...
$$ {#eq-ts-ses}

The smoothing parameter $\alpha$ lies between $0$ and $1$. A value close to 1 gives almost all the weight to the latest observation, so the forecast follows the data closely. A value close to 0 spreads the weight over many past observations, so the forecast is very smooth. An example with $\alpha = 0.1$, on the transport penalties, is shown in @fig-ses1 (left).

Holt extended SES by adding a trend component, with its own smoothing parameter $\beta$. The method is named after him as **Holt's linear method**. Mathematically, there are now three equations.

$$
\begin{aligned}
\hat{y}_{t+h|t} &= l_t + hb_t \\
l_t &= \alpha y_t + (1-\alpha)(l_{t-1} + b_{t-1}) \\
b_t &= \beta(l_t - l_{t-1}) + (1-\beta)b_{t-1}
\end{aligned}
$$ {#eq-ts-holt}

Here $\alpha$ and $\beta$ (again $0 \le \beta \le 1$) are the smoothing parameters for level and trend respectively, and $h$ is the forecast horizon. Also, $y_t$ is the actual value, $l_t$ is the level (or intercept) and $b_t$ is the trend (or slope) of the series at time $t$. The first equation says that the forecast $h$ periods ahead is the latest level plus $h$ times the latest slope. We apply it to the yearly excise revenue, which has a clear upward trend. See @fig-ses1 (right).

To apply *SES* in R, we use the function `ses()`, and for *Holt's* method, `holt()`, both from `forecast`. If we do not give the values of $\alpha$ and $\beta$, R chooses the best values itself. Here we have fixed them, to see their effect.

::: {#fig-ses1}

```{r fig.align='center', fig.show='hold', out.width="49%"}
fines_ses <- ses(fines, h = 10, alpha = 0.1)
holtfit <- holt(excise, h = 10, alpha = 0.3, beta = 0.1)

autoplot(fines_ses) +
  autolayer(fines_ses$fitted) +
  ggplot2::ggtitle('SES with alpha = 0.1, transport penalties') +
  ggplot2::theme_bw() +
  ggplot2::theme(legend.position = "bottom") +
  ggplot2::labs(color = "", y = "Rs lakh")

autoplot(holtfit) +
  autolayer(holtfit$fitted) +
  ggplot2::ggtitle("Holt's linear method with beta = 0.1, excise revenue") +
  ggplot2::theme_bw() +
  ggplot2::theme(legend.position = "bottom") +
  ggplot2::labs(color = "", y = "Rs crore")
```

Simple exponential smoothing (left) and Holt's linear method (right)
:::

As discussed earlier, damped trends often work better. Based on the work of Gardner and McKenzie, a damping parameter $\phi$ was added. The equations become

$$
\begin{aligned}
\hat{y}_{t+h|t} &= l_t + (\phi + \phi^2 + ... + \phi^h)b_t \\
l_t &= \alpha y_t + (1-\alpha)(l_{t-1} + \phi b_{t-1}) \\
b_t &= \beta(l_t - l_{t-1}) + (1-\beta)\phi b_{t-1}
\end{aligned}
$$ {#eq-ts-damped}

If $\phi = 1$, these are the same as Holt's linear method. A value $0 < \phi < 1$ damps the trend, so the forecast line gradually flattens, as shown in @fig-hl2 (left). The two methods are compared in @fig-hl2 (right).

To use *damped trends* in R, we set `damped = TRUE` in the `holt()` function. Example code is below.

::: {#fig-hl2}

```{r fig.align='center', fig.show='hold', out.width="49%"}
holtfit <- holt(excise, h = 10, alpha = 0.3, beta = 0.1, damped = TRUE, phi = 0.8)
autoplot(holtfit) +
  autolayer(holtfit$fitted) +
  ggplot2::ggtitle('Damped trend forecast with phi = 0.8') +
  ggplot2::theme_bw() +
  ggplot2::theme(legend.position = "bottom") +
  ggplot2::labs(color = "", y = "Rs crore")

holtfit <- holt(excise,
                h = 10,
                alpha = 0.3,
                beta = 0.1)

holtfit2 <- holt(
  excise,
  h = 10,
  alpha = 0.3,
  beta = 0.1,
  damped = TRUE,
  phi = 0.8
)

autoplot(excise) +
  autolayer(holtfit, series = "Holt's method", PI = FALSE) +
  autolayer(holtfit2, series = "Holt's method with damped trend", PI = FALSE) +
  ggplot2::ggtitle("Holt's method") + xlab("Year") +
  ylab("Excise revenue (Rs crore)") +
  guides(colour = guide_legend(title = "Method")) +
  ggplot2::theme_bw() +
  theme(legend.position = "bottom")
```

Linear method and damped trend
:::

Which one should an auditor prefer? If we expect revenue to keep growing at the same pace, the linear trend is right. If growth is likely to slow down, for example because a tax base is nearing saturation, the damped trend is safer. It is a judgement, and it should be recorded in the working papers.

To include *seasonality* with $m$ periods in a year (12 for monthly data) along with trend and level, Holt and Winters added a seasonal component $s_t$ with its own smoothing parameter $\gamma$. The equations of the **Holt-Winters additive method** are

$$
\begin{aligned}
\hat{y}_{t+h|t} &= l_t + hb_t + s_{t+h-m(k+1)} \\
l_t &= \alpha(y_t - s_{t-m}) + (1-\alpha)(l_{t-1} + b_{t-1}) \\
b_t &= \beta(l_t - l_{t-1}) + (1-\beta)b_{t-1} \\
s_t &= \gamma(y_t - l_{t-1} - b_{t-1}) + (1-\gamma)s_{t-m}
\end{aligned}
$$ {#eq-ts-hw}

where $k$ is the whole number part of $(h-1)/m$. The term $s_{t+h-m(k+1)}$ looks complicated, but it only means "the seasonal value of the same month in the latest year of data". So a forecast for next March uses the seasonal effect estimated for the latest March. In the *multiplicative* version, the seasonal effect multiplies the level and trend instead of being added.

To apply the *Holt-Winters* method in R, we use the function `hw()`. In the code below, we forecast the monthly expenditure with both `additive` and `multiplicative` seasonality. See @fig-hw1.

::: {#fig-hw1}

```{r fig.align='center', fig.show='hold', out.width="48%"}
hw1 <- hw(spend, seasonal = "additive")
hw2 <- hw(spend, seasonal = "multiplicative")

autoplot(hw1) +
  ggtitle("Additive method") +
  labs(x = NULL, y = "Rs crore") +
  theme_bw()

autoplot(hw2) +
  ggtitle("Multiplicative method") +
  labs(x = NULL, y = "Rs crore") +
  theme_bw()
```

Holt-Winters method of forecasting
:::

The multiplicative forecast (right) lets the March peaks keep growing with the budget, as in the data. The additive one (left) keeps the seasonal swings at a fixed size in rupees. As we saw, the seasonality of `spend` is multiplicative, so the right one is more believable.

We do not always have to choose between SES, Holt and Holt-Winters ourselves. The function `ets()` (short for Error, Trend, Seasonal) tries all these methods, with and without damping, additive and multiplicative. It picks the best one automatically. We will use it in the audit use case.

### Seasonal decomposition methods {.unnumbered}

Seasonal decomposition separates a time series into its components, like trend, seasonality and noise. This lets us study each component and its effect on the series separately. There are several methods. We will discuss two of them.

### Classical seasonal decomposition

The first is the **classical decomposition method**. It assumes that the seasonal pattern repeats *identically* every year. Its steps are

- **Trend component.** The trend is estimated with a centred moving average over one full season, a 12-month average for monthly data. As we saw in the moving averages section, such an average cancels out the seasonal ups and downs.
- **Seasonal component.** The trend is removed from the series. What is left is averaged for each month (or quarter) across all the years. This gives one seasonal value for each month.
- **Residual component.** The trend and seasonal components are removed from the original series. What remains is the residual.

Let us decompose the monthly expenditure to see its components. In R, we use the function `decompose()`. Since we know the seasonality of `spend` is multiplicative, we set `type = "multiplicative"` (the default is additive).

```{r}
spend_dec <- decompose(spend, type = "multiplicative")
```

The result is a list. Its element `figure` gives the seasonal value of each month. In a multiplicative decomposition, these are seasonal *indices*. An index of 1.10 means that the month is usually 10% above the trend. The values are in the order of the months of the series. Our series starts in April, so we name them from April to March.

```{r}
spend_idx <- round(spend_dec$figure, 3) |>
  setNames(month.abb[c(4:12, 1:3)])
spend_idx
```

March has an index of `r spend_idx["Mar"]`. So the department usually spends about `r round(spend_idx["Mar"], 1)` times as much in March as in an average month of the year. The lowest month is `r month.name[match(names(which.min(spend_idx)), month.abb)]`, the first month of the financial year, at about `r round(100 * (1 - min(spend_idx)))`% below the trend. Such an index is a good way to report the March rush. It is one number, it allows for the growth of the budget, and it can be compared across departments.

::: callout-tip
## Audit tip

The financial rules of most governments discourage a rush of expenditure in the last months of the year. A March index above 2 means that the March of an average year carries twice the spending of an average month. Compute this index for each department, and take up the departments with the highest index for detailed checks of March vouchers.
:::

A plot of the decomposed expenditure is shown in @fig-stl1. It has four panels, (i) the observed, that is, original values, (ii) the trend, (iii) the seasonal pattern and (iv) the random noise.

```{r fig-stl1, fig.cap="Classical decomposition of monthly expenditure (multiplicative)", fig.align='center', fig.show='hold'}
spend_dec |>
  plot()
```

Notice that the trend is missing for the first six and last six months. A centred 12-month average needs six months on each side, which are not available at the ends of the series. Notice also the tall spike in the random panel early in 2021. We will find out which month it is in the next section.

### Seasonal and trend decomposition using Loess^[Loess is a method for fitting a smooth curve through data, by fitting many small regressions on nearby points.]

STL (Seasonal and Trend decomposition using Loess) is a robust and flexible method for decomposing a time series into trend, seasonal and remainder components. Instead of simple averages, it uses Loess, a method of fitting smooth curves, to estimate both the trend and the seasonal component. It repeats the two steps a few times, each time improving both estimates. It has some advantages over classical decomposition.

- It gives a trend for the whole series, including the first and last months.
- The seasonal pattern may be allowed to change slowly over the years.
- It is *robust*, meaning a few unusual values do not spoil the estimates much.

STL works only with additive components. So for a multiplicative series like `spend`, we first take its logarithm. In R, we use the function `stl()`. Setting `s.window = "periodic"` tells it to keep the seasonal pattern the same every year. Setting `robust = TRUE` asks it to resist unusual months. A plot of the STL decomposition is shown in @fig-stl2.

```{r fig-stl2, fig.cap="STL decomposition of log(monthly expenditure)", fig.align='center', fig.show='hold'}
spend_stl <- stl(log(spend), s.window = "periodic", robust = TRUE)
spend_stl |>
  plot()
```

The element `time.series` of the result holds the three components. The remainder is what is left after removing the trend and the March rush. On the log scale, a remainder of 0.1 means that the month was about 10% above what its trend and season would suggest. Let us list the months with the largest remainders.

```{r}
spend_rem <- tibble(month     = as.yearmon(time(spend)),
                    spend     = as.numeric(spend),
                    remainder = round(as.numeric(spend_stl$time.series[, "remainder"]), 3)) |>
  arrange(desc(remainder))
head(spend_rem, 5)
```

```{r}
#| include: false
top_rem <- spend_rem[1, ]
```

`r format(top_rem$month, "%B %Y")` stands out, with a remainder of `r top_rem$remainder`. So the department spent ₹`r top_rem$spend` crore that month, about `r round(100 * (exp(top_rem$remainder) - 1))`% more than even a normal March would explain. This is our planted case. An auditor would examine the vouchers of that March. Were the funds drawn and kept in bank accounts, or released to agencies only to avoid lapse of the budget? The anomaly detection chapter shows how to flag such months with a formal rule, in @sec-ano-ts.

### Model-based techniques {.unnumbered}

Model-based techniques fit a specific statistical model to the time series, and estimate its parameters from the data. They need more assumptions, but can give a more detailed understanding of the series. The most widely used are

1. **Autoregressive Integrated Moving Average (ARIMA)** models. These explain each value from the previous values of the series (the *autoregressive* part, of order $p$) and from the previous forecast errors (the *moving average* part, of order $q$). Many series, like those with a trend, do not have a constant level. ARIMA handles them by working on the *differences* between consecutive values (the *integrated* part, of order $d$). An ARIMA model is written as ARIMA($p$,$d$,$q$).
2. **Seasonal ARIMA (SARIMA)** models. These extend ARIMA to include seasonality, by also relating each value to the values of the same month in earlier years. They are written as ARIMA($p$,$d$,$q$)($P$,$D$,$Q$)$_m$, where the second set of orders is for the seasonal part, and $m$ is the number of periods in a season.

Choosing $p$, $d$, $q$ and their seasonal versions by hand needs experience. Fortunately, the function `auto.arima()` from `forecast` tries many combinations and picks the best one. The argument `lambda = 0` asks it to work on the logarithm of the series, as `spend` is multiplicative. The forecasts are converted back to the original scale automatically.

```{r}
spend_arima <- auto.arima(spend, lambda = 0)
spend_arima
```

`auto.arima()` has chosen the model `r as.character(spend_arima)`. The seasonal difference ($D = 1$) means that the model works on the change from the same month of last year. This is the auditor's familiar "this March compared with last March", built into a model. Its forecast for the next two years is shown in @fig-arima.

```{r fig-arima, fig.cap="Forecast of monthly expenditure with a seasonal ARIMA model", fig.align='center'}
forecast(spend_arima, h = 24) |>
  autoplot() +
  labs(x = NULL, y = "Rs crore") +
  theme_bw()
```

### Which method to choose?

We have seen many methods. Which one forecasts best? The honest way to find out is to *test* them. We hide the last part of the data, forecast it with each method using only the earlier data, and compare the forecasts with what actually happened. The earlier part is called the *training* data, and the hidden part the *test* data.

Let us keep April 2014 to March 2022 of `spend` for training, and the last two financial years, 2022-23 and 2023-24, for testing. The function `accuracy()` compares a forecast with the actual values. It gives the error measures we met in the regression chapter (@sec-regression), like RMSE, MAE and MAPE.

```{r}
train <- window(spend, end = c(2022, 3))
test  <- window(spend, start = c(2022, 4))

forecasts <- list(
  "Naive"                 = naive(train, h = 24),
  "Seasonal naive"        = snaive(train, h = 24),
  "Holt-Winters (mult.)"  = hw(train, h = 24, seasonal = "multiplicative"),
  "ETS (automatic)"       = forecast(ets(train), h = 24),
  "ARIMA (automatic)"     = forecast(auto.arima(train, lambda = 0), h = 24)
)

sapply(forecasts, \(f) accuracy(f, test)["Test set", c("RMSE", "MAE", "MAPE")]) |>
  t() |>
  round(1)
```

```{r}
#| include: false
acc <- sapply(forecasts, \(f) accuracy(f, test)["Test set", "MAPE"])
```

The naive forecast is off by `r round(acc["Naive"], 1)`% on average. It is badly wrong because the last month of the training data is a March. It repeats that March peak for every month of the next two years. The seasonal naive forecast does far better, at `r round(acc["Seasonal naive"], 1)`%. The best here is `r names(which.min(acc))`, at `r round(min(acc), 1)`%. The other two methods that allow for trend and season are close, between `r round(sort(acc)[2], 1)`% and `r round(sort(acc)[3], 1)`%.

::: callout-note
## No method is always best

The automatic choices of `ets()` and `auto.arima()` are made on the *training* data. They are good starting points, but they do not always win on the test data. Here the automatic ETS model came a little behind the multiplicative Holt-Winters method. On another series, the order may change. So it is good practice to try a few methods and compare them on test data, as we did here.
:::

## Plotting different elements of time series

`forecast` has some more plots which help us see the elements of a time series.

- `ggseasonplot()` draws each year as a separate line against the months. It shows the seasonal pattern and how it changes over the years.
- `ggsubseriesplot()` draws a mini plot for each month across the years, with the mean of that month as a horizontal line.
- `gglagplot()` plots the series against *lagged* (earlier) values of itself.
- `ggAcf()` plots the *autocorrelation function* (ACF), that is, the correlation of the series with its own lagged values.

The seasonal plot and sub-series plot of the monthly expenditure are in @fig-seasplots. Both show that March is far above every other month, every year. The seasonal plot also shows that each year lies above the previous one, which is the trend. In the sub-series plot, the highest point on the March line is March 2021. It is again our planted case.

```{r fig-seasplots, fig.show='hold', fig.align='center', fig.cap="Seasonality through seasonal and sub-series plots", fig.height=8}
g1 <- ggseasonplot(spend) +
  labs(y = "Rs crore") +
  theme_bw()
g2 <- ggsubseriesplot(spend) +
  ggtitle('Sub-series plot') +
  labs(y = "Rs crore") +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))
(g1 / g2)
```

Note that the seasonal plot runs from January to December, the calendar year. So each financial year is split across two lines. This does not matter for seeing the pattern.

The lag plots of the monthly expenditure can be seen in @fig-lagp. Each panel plots the series against itself a certain number of months earlier. Look at the panel for lag 12. The points lie close to the diagonal line, and the March points sit together at the top right. So each month is very similar to the same month a year earlier. This tells us that the series has a seasonality of 12 periods.

```{r fig-lagp, fig.show='hold', fig.align='center', fig.cap="Seasonality through lag plots"}
gglagplot(spend) +
  ggtitle('Lag plots') +
  theme_bw() +
  theme(legend.position = "bottom") +
  labs(color = NULL) +
  guides(color = guide_legend(nrow = 1))
```

The same information appears as numbers in the ACF plot of @fig-acf. Each bar is the correlation of the series with itself at that lag. The bars peak at lags 12, 24 and 36, because of the March rush every twelve months. Bars crossing the blue dashed lines are significantly different from zero.

```{r fig-acf, fig.show='hold', fig.align='center', fig.cap="Autocorrelation of monthly expenditure", out.height="35%"}
ggAcf(spend, lag.max = 36) +
  ggtitle('Autocorrelation function') +
  theme_bw()
```

## Audit use case, is this year's revenue believable? {#sec-ts-audit}

Let us now use forecasting the way an auditor would. Revenue collected by a government office usually follows a steady pattern of growth and seasonality. If we forecast this year's revenue from the past years, and the actual revenue falls well below the forecast, we have a lead. There may be leakage of revenue, like undervaluation, short levy, or receipts not accounted for.

### Our sample data

We create monthly stamp duty receipts of an imaginary sub-registrar office, in ₹ lakh, from April 2017 to March 2024. Receipts grow by about 8% every year. They dip in the monsoon months, rise in the festive months of October and November, and peak in March. We have built two events into the data.

- In April and May 2020, receipts nearly stopped because of the COVID-19 lockdown. This is a genuine, known event.
- From September 2023, receipts are about 22% lower than they should be. This is our planted case of leakage.

As usual, new readers may skip the code which creates the data.

```{r}
#| code-fold: true
#| code-summary: "Code to create the stamp duty receipts"
set.seed(2024)
n <- 84  # April 2017 to March 2024

# Seasonal pattern from April to March
season <- c(0.95, 0.95, 0.85, 0.80, 0.85, 0.90, 1.10, 1.15, 1.00, 1.00, 1.05, 1.40)

receipts <- 250 * 1.08^((0:(n - 1)) / 12) * rep(season, 7) * exp(rnorm(n, 0, 0.05))

# Lockdown in April and May 2020
receipts[37:38] <- receipts[37:38] * c(0.08, 0.35)

# Planted case. Receipts about 22% lower from September 2023
receipts[78:84] <- receipts[78:84] * 0.78

stamp <- ts(round(receipts, 1), start = c(2017, 4), frequency = 12)
```

```{r}
stamp
```

Let us plot the data in @fig-stamp. The lockdown dip in 2020 is obvious. The fall after September 2023 is harder to see by eye.

```{r fig-stamp, fig.cap="Monthly stamp duty receipts of the sub-registrar office", fig.align='center', out.height="35%"}
autoplot(stamp) +
  labs(x = NULL, y = "Receipts (₹ lakh)") +
  theme_bw()
```

### Forecasting the year under audit

The year under audit is 2023-24. So we take the data up to March 2023 as training data, and keep 2023-24 aside to compare with the forecast.

```{r}
train_stamp  <- window(stamp, end = c(2023, 3))
actual_stamp <- window(stamp, start = c(2023, 4))
```

We fit a model with `ets()` on the training data and forecast the 12 months of 2023-24, with a 95% prediction interval. Let us also write a small function to put the forecast and actual values side by side.

```{r}
compare_with_actual <- function(fc, actual) {
  tibble(month    = as.yearmon(time(actual)),
         actual   = as.numeric(actual),
         forecast = as.numeric(fc$mean),
         lower    = as.numeric(fc$lower),
         upper    = as.numeric(fc$upper)) |>
    mutate(gap   = actual - forecast,
           below = actual < lower)
}

fc1 <- forecast(ets(train_stamp), h = 12, level = 95)
compare_with_actual(fc1, actual_stamp)
```

```{r}
#| include: false
cmp1 <- compare_with_actual(fc1, actual_stamp)
w1 <- mean(cmp1$upper - cmp1$lower)
```

Not a single month is below the lower limit. The intervals are very wide. On average, the range from the lower to the upper limit is about ₹`r round(w1)` lakh, almost as large as a month's receipts. With such wide limits, almost nothing can look unusual. Why?

The reason is the lockdown. The model has seen receipts fall to nearly zero in April 2020, and so it believes that such falls can happen any time. A single extraordinary event in the past has made the model very unsure about the future.

### Cleaning a known event

The lockdown was a known, one-time event. It is not part of the normal behaviour of this office. So it is fair to replace those two months with values the series would normally have had. The function `tsclean()` from `forecast` finds such outliers and replaces them with values estimated from the rest of the series.

```{r}
train_clean <- tsclean(train_stamp)

# Compare the original and cleaned values for March to June 2020
cbind(original = train_stamp, cleaned = train_clean) |>
  window(start = c(2020, 3), end = c(2020, 6))
```

Only April and May 2020 have changed. Now let us forecast again from the cleaned data.

```{r}
fc2 <- forecast(ets(train_clean), h = 12, level = 95)
cmp2 <- compare_with_actual(fc2, actual_stamp)
cmp2
```

```{r}
#| include: false
w2 <- mean(cmp2$upper - cmp2$lower)
short <- cmp2 |> filter(below)
```

The intervals are now much narrower, about ₹`r round(w2)` lakh on average. And the picture is clear. From April to August 2023, the actual receipts are close to the forecast. From `r format(min(short$month), "%B %Y")` onwards, *every* month is below the lower limit. Together, the receipts in these `r nrow(short)` months are ₹`r round(-sum(short$gap))` lakh, or about ₹`r round(-sum(short$gap) / 100, 1)` crore, less than forecast. @fig-stampfc shows this.

```{r fig-stampfc, fig.cap="Forecast (blue, with 95% interval) and actual receipts (red) for 2023-24", fig.align='center'}
autoplot(fc2) +
  autolayer(actual_stamp, series = "Actual", linewidth = 0.8) +
  scale_colour_manual(values = c(Actual = "red")) +
  coord_cartesian(xlim = c(2021, 2024.25)) +
  labs(x = NULL, y = "Receipts (₹ lakh)", title = NULL, colour = NULL) +
  theme_bw() +
  theme(legend.position = "bottom")
```

What would the auditor do next? The shortfall is a lead, so it goes into the working papers with the forecast, the interval and the gap in rupees. The next step is to ask the sub-registrar office what changed from September 2023, and to test-check the documents registered after that point.

::: callout-tip
## Audit tip

A *sustained* shortfall, month after month from a clear starting point, is a much stronger lead than a single bad month. Find out what changed at that point. It could be a new officer, a change in the guideline values of property, new software, or a change in the procedure of registration. Then examine the documents registered after that point, for example for undervaluation of property.
:::

::: callout-warning
## Caution

Clean only those past events which are known, genuine and one-time, like a lockdown, a strike or a system failure. Record what was cleaned and why. Never clean the period under audit itself. That would hide the very problem we are looking for.
:::

## A suggested workflow

For any time series analysis in audit, we may follow these steps.

1. **Create the time series.** Arrange the data by month or quarter, without gaps, and convert it with `ts()`, giving the correct `start` and `frequency`. A financial year series starts at `c(year, 4)`.
2. **Plot it.** Look at the trend, the seasonality and any unusual events with `autoplot()`, `ggseasonplot()` and `ggsubseriesplot()`.
3. **Decide additive or multiplicative.** If the seasonal swings grow with the level of the series, take the logarithm, or use multiplicative methods.
4. **Decompose** the series with `decompose()` or `stl()`, to measure the seasonal pattern (like the March index) and see the trend clearly. Look at the largest remainders for unusual months.
5. **Clean known one-time events** in the past with `tsclean()`, and record what was cleaned.
6. **Forecast** the period under audit from the earlier data, with `ets()`, `hw()` or `auto.arima()`. Where possible, compare methods on test data with `accuracy()`.
7. **Compare the actual values with the forecast** and its prediction interval. Quantify the gap in rupees.
8. **Verify with records.** Find out what changed, and examine the documents of the period.

::: callout-important
## Remember

A forecast is the auditor's *expectation*, not the truth. A gap between the actual and the forecast is a *lead*, not a *finding*. There may be a genuine reason, like a change in tax rates, a court order or a slowdown in the market. Always find out the reason and examine the records before reporting.
:::

## Where next

This chapter looked at the pattern of a series as a whole. The anomaly detection chapter picks out single months which are unusual after allowing for trend and season (@sec-ano-ts). It also finds the point at which a series shifted to a new level and stayed there (@sec-ano-changepoint), as the stamp duty receipts did in September 2023.

## Summary of functions used

| Function | Package | What it does |
|-----------------------|----------------|--------------------------------|
| `ts()` | base R (stats) | Creates a time series object |
| `window()` | base R (stats) | Takes out a part of a time series between two times |
| `plot()`, `autoplot()` | base R, ggplot2 with forecast | Plots a time series or a forecast |
| `naive()`, `snaive()` | forecast | Naive and seasonal naive forecasts |
| `ma()` | forecast | Centred moving average |
| `rollmeanr()` | zoo | Trailing moving average |
| `ses()`, `holt()`, `hw()` | forecast | Simple exponential smoothing, Holt's method and Holt-Winters method |
| `ets()` | forecast | Chooses and fits the best exponential smoothing model automatically |
| `auto.arima()` | forecast | Chooses and fits the best ARIMA model automatically |
| `forecast()` | forecast | Forecasts from a fitted model, with prediction intervals |
| `accuracy()` | forecast | Compares forecasts with actual values (RMSE, MAE, MAPE and others) |
| `decompose()`, `stl()` | base R (stats) | Classical and STL decomposition into trend, season and remainder |
| `tsclean()` | forecast | Replaces outliers and missing values in a time series |
| `ggseasonplot()`, `ggsubseriesplot()` | forecast | Seasonal and sub-series plots |
| `gglagplot()`, `ggAcf()` | forecast | Lag plots and autocorrelation plot |
| `as.yearmon()` | zoo | Turns the time of a monthly series into a readable month, like Mar 2021 |

: Functions used in this chapter {#tbl-tsfunctions}
