```{r}
#| include: false
source("_common.R")
```


# Networks and Graph Theory {#sec-graph}

Network analysis has become a powerful tool in audit analytics. It lets auditors uncover relationships and patterns hidden in large datasets. Which beneficiaries share a bank account? Which firms always bid together? Does money go round in a circle between a few companies? These are questions about *connections*, and network analysis is built to answer them. Using R, auditors can construct, visualise and analyse such networks with ease, detect anomalies, and identify the key players.

In this chapter we will learn the concepts of graph theory and the tools to work with networks in R. In the next chapter (@sec-netaudit), we will apply them to real audit problems. So let us dive in.

## An introduction to Graph theory

Network analysis applies the concepts of *graph theory*, a branch of mathematics, to understand complex systems like social networks, transport networks and financial transactions. Graph theory studies the properties of graphs, such as paths, cycles and connectivity.

### Definition of a graph

A graph is a mathematical structure consisting of a set of *vertices* (also called nodes or actors) and a set of *edges* (also called connections or links) between these vertices. Mathematically, a graph $G = (V, E)$ consists of a set $V$ of vertices and a set $E$ of edges, as illustrated in @fig-gra1. A vertex may represent a real-world object, like a person, a computer, a firm or a bank account. An edge represents the *relationship* between the two vertices it connects, like friendship between two persons, a cable between two computers, or a payment from one firm to another.

```{r fig-gra1, echo=FALSE, message=FALSE, warning=FALSE, fig.align='center', fig.show='hold', fig.cap="An Example Graph", out.width="100%"}
library(igraph)
library(igraphdata)
data("kite")
kite <- upgrade_graph(kite)
plot(kite, vertex.size = 25)
```

### Types of graph {.unnumbered}

The graph in @fig-gra1 is an **undirected graph**, in which the edges have no direction. A cable connecting two computers is an example. If the edges have a direction, from one vertex to another, the graph is called a **directed graph**. For example, on X (formerly Twitter) a user may follow another user without being followed back. A payment from one firm to another is also directed. See @fig-gra2.

```{r fig-gra2, echo=FALSE, message=FALSE, warning=FALSE, fig.align='center', fig.show='hold', fig.cap="An Example Directed Graph", out.width="100%", out.height="33%"}
library(tidyverse)
set.seed(123)
igraph::as_data_frame(kite) %>%
  slice(1:9) %>%
  graph_from_data_frame() %>%
  plot(vertex.size = 25)
```

Sometimes an edge connects a vertex with itself. Such an edge is called a **self-loop**. Graphs which allow self-loops are sometimes called **pseudographs**. For a real-world example, think of transactions between firms. A firm usually transacts with other firms, but it may also transfer money between two of its own accounts. In @fig-gra3 (left), vertex `E` is connected with itself.

Further, there may be more than one edge connecting the same pair of vertices (in the same direction). These are called **multi-edges**. In @fig-gra3 (right) there are two edges from `B` to `E`. In the real world, think of flights by different airlines between the same pair of airports, or many invoices from the same seller to the same buyer.

::: {#fig-gra3}

```{r echo=FALSE, message=FALSE, warning=FALSE, fig.align='center', fig.show='hold', out.width="48%", out.height="33%"}
set.seed(123)
igraph::as_data_frame(kite) %>%
  slice(c(1, 3, 5:6)) %>%
  add_row(from = "E", to = "E") %>%
  graph_from_data_frame() %>%
  plot(vertex.size = 45)
igraph::as_data_frame(kite) %>%
  slice(c(1, 3, 5:6)) %>%
  add_row(from = "B", to = "E") %>%
  graph_from_data_frame() %>%
  plot(vertex.size = 45)
```

Examples of Self Edge and Multi Edge
:::

**Complete graphs** are graphs in which every pair of vertices is connected by an edge. Examples are shown in @fig-gra4. Complete graphs are rare in the real world. But a few vertices of a bigger graph may all be connected with each other. Such a group, which is a complete graph in itself, is called a **clique**. For example, the vertices `B, D, E and G` in @fig-gra1 form a clique. Every one of them is connected with the other three. In audit, a clique of firms which all trade with each other, or of bidders who all bid together, deserves attention.

::: {#fig-gra4}

```{r echo=FALSE, message=FALSE, warning=FALSE, fig.align='center', fig.show='hold', out.width="49%", out.height="31%"}
plot(make_full_graph(4), vertex.size = 45)
plot(make_full_graph(6), vertex.size = 45)
```

Examples of Complete Graphs
:::

Let us learn about two more types of graph. The first is the **bipartite graph**. Its vertices are divided into two separate sets, and edges connect only vertices of *different* sets, never two vertices of the same set. The graph in @fig-gra5 (left) is bipartite. Think of officials of two departments, where an edge shows correspondence between two officials of different departments. In audit, bipartite graphs are very common. Beneficiaries and their bank accounts, or bidders and tenders, form two sets with edges only between the sets. The idea can be extended to *k-partite* graphs, with `k` such sets.

The second is the **tree**, which represents a hierarchy. A graph is a tree if there is exactly one path between any pair of vertices (ignoring direction). The bipartite graph in @fig-gra5 (left) is also a tree, since it meets this condition. In @fig-gra5 (right) the same graph has been redrawn as a tree.

::: {#fig-gra5}

```{r echo=FALSE, message=FALSE, warning=FALSE, fig.align='center', fig.show='hold', out.width="49%", out.height="40%"}
# Officials of department A (A1 to A4) and department B (B1 to B5)
g <- graph_from_literal(A1-B1, A1-B2, A2-B2, A2-B3, A3-B3, A3-B4, A4-B4, A4-B5)
V(g)$type <- startsWith(V(g)$name, "B")

# define color and shape mappings.
col <- c("steelblue", "orange")
shape <- c("circle", "square")

plot(g,
     vertex.color = col[V(g)$type + 1],
     vertex.shape = shape[V(g)$type + 1],
     vertex.size = 35,
     layout = layout_as_bipartite
)

plot(g,
     vertex.color = "pink",
     vertex.shape = "square",
     vertex.size = 35,
     layout = layout_as_tree
)
```

Bipartite and Tree Graph example
:::

### Vertex and Edge attributes

Recall that a graph consists of two sets $(V, E)$, where $V$ is a set of vertices and $E$ is a set of edges. In their simplest form, these sets are just lists. But both can carry additional *attributes*. For example, if a graph consists of airports and flights between them, each airport in $V$ could have attributes like (i) its name, (ii) whether it is domestic or international, (iii) its geographical coordinates, and (iv) the State in which it is located. Similarly, each edge in $E$ could have attributes like (i) the number of flights between the two airports, (ii) the airlines operating them, and (iii) the aerial distance. An example may be seen in @fig-gra6. In audit data, typical edge attributes are the amount and date of a transaction.

```{r fig-gra6, echo=FALSE, message=FALSE, warning=FALSE, fig.align='center', fig.show='hold', fig.cap="Examples of Vertex and Edge properties", out.width="98%", out.height="32%"}
library(ggraph)
airtrans <- data.frame(
  from = c("A", "A", "A", "B", "C"),
  to = c("B", "C", "D", "C", "D")
) |> as.matrix()

airtransgraph <- graph_from_edgelist(airtrans, directed = FALSE)
V(airtransgraph)$loc <- c("New Delhi", "Bangalore", "Mumbai", "Pune")
V(airtransgraph)$country <- rep("India", 4)
E(airtransgraph)$dist <- c(1750, 1150, 1200, 850, 120)
E(airtransgraph)$airline <- c(rep("Air India", 3), "IndiGo", "Akasa Air")

ggraph(airtransgraph, layout = 'kk') +
  geom_edge_fan(aes(label = paste0("Operating Airlines:", airline, ", Distance:", dist, " Kms")),
                angle_calc = "along",
                label_dodge = unit(2.5, 'mm'),
                label_size = 4
  ) +
  geom_node_point(size = 20, color = "#D4E7C5") +
  geom_node_text(aes(label = paste0(name, " \n(Airport:", loc, ")\n", country)), size = 3) +
  scale_x_continuous(expand = expansion(mult = 0.1)) +
  theme_void()
```


## Practical approach for creating graphs in R

### Packages to use

One of the best packages for creating and analysing graphs in R is `igraph`. It was originally developed by **Gábor Csárdi** and **Tamás Nepusz**, and its core is written in the C programming language for good performance. So it can handle networks with millions of edges. Let us load it.

```{r}
#| message: false
#| warning: false
library(igraph)
library(tidyverse)
```

`igraph` is a complete package for network analysis in itself. We will also use `visNetwork` and `ggraph`. The former creates interactive network charts, and the latter creates `ggplot2` compatible plots, which can be customised further in a familiar way.

::: callout-note
## Function names in igraph

Version 2 of `igraph` renamed many functions to a consistent `snake_case` style. For example, `as.undirected()` is now `as_undirected()` and `graph.data.frame()` is now `graph_from_data_frame()`. The old names still work for now, with a warning. You may find old names in books and websites, but use the new ones in your code.
:::

Let us now learn how to create graph objects in R. Graphs can be created in many ways. We will learn three methods, which will serve our purpose most of the time.

### Creating a graph from data frame

In data analysis, data frames are our primary tool. So the most common and practical way to create a graph is from a `data.frame`. The data frame should have at least two columns, where each row is one edge. The first column is taken as the `'from'` vertex and the second as the `'to'` vertex, whatever their column names. We use the function `graph_from_data_frame()` from `igraph`. Its syntax is

```
graph_from_data_frame(
  d =         ,  # data.frame of edges
  directed = TRUE,
  vertices = NULL # optional data.frame of vertices
)
```

Example-1: Let us construct a graph from a data frame containing four edges.

```{r gra7}
# Example data frame of edges
df <- data.frame(
  from = c(1, 1, 3, 3),
  to = c(2, 3, 5, 4)
)
# Creating graph object
gr1 <- graph_from_data_frame(df)
# Let's print it
gr1
```

A graph object named `gr1` has been created. Printing it gives us useful information about it.

+ The first line always begins with `IGRAPH`, followed by seven characters. These are the first characters of a unique graph ID.
+ Next comes a four-letter code. In this example it is `DN--`. The letters mean
  - The first letter tells whether the graph is directed (`D`) or undirected (`U`).
  - The second letter, `N`, means a *named* graph, that is, the vertices have a `name` attribute.
  - The third letter, `W`, means a *weighted* graph, that is, the edges have a `weight` attribute.
  - The fourth letter, `B`, means a *bipartite* graph, that is, the vertices have a `type` attribute.
+ Next come the number of vertices and edges, separated by two dashes.
+ The second line lists the attributes of the graph. For each attribute, its kind, graph (`g`), vertex (`v`) or edge (`e`), and its data type, character (`c`), numeric (`n`), logical (`l`) or other (`x`), are shown in brackets.
+ The last lines show the edges.

All four edges of our `df` are in `gr1`. Notice that the vertex names are characters, `"1"`, `"2"` and so on, even though our data had numbers. `igraph` always stores vertex names as characters.

::: callout-warning
## Caution
In audit, vertices are often account numbers, PANs or vendor codes. Read such columns as text, not numbers. If an account number like `00123` is read as a number, it becomes `123`. It may then fail to match the same account in another table, or two different accounts may end up as one vertex.
:::

The argument `directed` is `TRUE` by default, so the graph is directed, as confirmed by the first letter `D`. Let us create an undirected graph.

```{r gra8}
# Creating graph object
gr2 <- graph_from_data_frame(df, directed = FALSE)
# Let's print it
gr2
```

The first letter is now `U`, and the edges are printed with `--` instead of an arrow `->`. An existing directed graph can also be converted into an undirected one using the function `as_undirected()` (and the other way round with `as_directed()`).

```{r gra9}
as_undirected(gr1)
```

Now what about the last argument, `vertices`? In most cases, the edge data is enough. But a graph may contain *isolates*, that is, vertices not connected to any edge. These do not appear in the edge data, so to include them we use the `vertices` argument. Its first column must contain the vertex IDs used in the edge data. Any other columns become vertex attributes. Similarly, any columns of the edge data frame after the first two become edge attributes.

Example-2: Let us add an isolated vertex `"6"` to our graph, along with a `label` for each vertex.

```{r gra10}
df_v <- data.frame(
  id = c(1, 2, 3, 4, 5, 6),
  label = c("Ram", "Shyam", "Alex", "Bob", "Charlie", "Kumar")
)
# Creating graph object
gr3 <- graph_from_data_frame(df, directed = FALSE, vertices = df_v)
# Let's print it
gr3
```

The number of vertices is now 6, though there are still 4 edges. The vertex attribute `label` has also been added, shown as `label (v/c)`.

Readers may try adding more columns to the edge data frame, like an amount or a date, and see how they become edge attributes.

### Creating a graph from Edge list

Creating a graph from an edge list is nearly the same as creating it from a data frame. The differences are that the input must be a two column *matrix*, character or numeric, and that there is no `vertices` argument. The syntax is

```
graph_from_edgelist(el, directed = TRUE)
```

```{r gra11}
edges <- data.frame(
  origin = c("Ram", "Ram", "Alex", "Alex"),
  dest = c("Shyam", "Alex", "Charlie", "Bob")
)
edges <- as.matrix(edges)
# Now let's create a graph
gr4 <- graph_from_edgelist(edges, directed = FALSE)
# And print it
gr4
```

We have created an undirected graph with `5` vertices and `4` edges.

### Creating a graph from adjacency matrix

An adjacency matrix is a square $n \times n$ matrix, where both the rows and the columns stand for the $n$ vertices. The $(i, j)$-th entry of this matrix means

*   In a graph without weights, `1` means there is an edge from vertex $v_{i}$ to $v_j$, and `0` means there is none.
*   In a weighted graph, the entry is the weight of the edge from vertex $v_{i}$ to $v_j$, and `0` means there is no edge.

Example: Let us create the graph of @fig-gra6 through its adjacency matrix. First, without weights.

```{r gra12}
# Let's create the matrix
adj_mat <- structure(c(0, 1, 1, 1, 1, 0, 1, 0, 1, 1, 0, 1, 1, 0, 1, 0), dim = c(4L,
4L), dimnames = list(c("DEL", "BNG", "BOM", "PNQ"), c("DEL",
"BNG", "BOM", "PNQ")))
adj_mat
# Let's create the graph
gr5 <- graph_from_adjacency_matrix(adj_mat)
# Let's print it
gr5
```

A directed graph (the default) has been created, with 10 edges. Since our matrix is symmetric, each route appears twice, once in each direction, like `DEL->BNG` and `BNG->DEL`. For a symmetric matrix, we would usually set `mode = "undirected"` to get one edge per route. Now let us create a weighted graph of the same data, where the weights are the distances between the airports.

```{r gra13}
# Let's create the matrix
adj_mat2 <- structure(c(0, 1750, 1150, 1200, 1750, 0, 850, 0, 1150, 850,
0, 120, 1200, 0, 120, 0), dim = c(4L, 4L), dimnames = list(c("DEL",
"BNG", "BOM", "PNQ"), c("DEL", "BNG", "BOM", "PNQ")))
adj_mat2
# Let's create the graph
gr6 <- graph_from_adjacency_matrix(adj_mat2, weighted = TRUE)
# Let's print it
gr6
```

This time an edge attribute named `weight`, of `numeric` type, has been added to the graph.

### Getting data back from graph objects

Once a graph has been created, getting the data back as a `data.frame` or an adjacency `matrix` is easy. We use the `igraph` functions `as_data_frame()` and `as_adjacency_matrix()`.

1. `as_data_frame()`: Since older versions of the package `tibble` also had a function of the same name, it is safest to write `igraph::as_data_frame()`. It takes a graph and returns a `data.frame`. Let us get the data back from `gr6`, which we created from an adjacency matrix.

```{r gra14}
igraph::as_data_frame(gr6)
```

The function worked correctly. Since the graph was directed, each route appears in both directions. Let us also see the output for an undirected graph.

```{r gra15}
igraph::as_data_frame(as_undirected(gr6))
```

There is one row per route now. But notice that the weights have *doubled*, for example DEL to BNG is now 3,500. When two directed edges are merged into one, `as_undirected()` by default *adds up* their weights. We can choose another way to combine them using the argument `edge.attr.comb`, which takes a function (or its name) for each attribute. Here, the mean is right.

```{r gra16}
gr7 <- as_undirected(gr6, edge.attr.comb = list(weight = "mean"))
igraph::as_data_frame(gr7)
```

This function has another argument, `what`. It can be `"edges"` (the default), `"vertices"`, or `"both"`, in which case the two data frames are returned in a list.

```{r gra17}
# Vertices data only
igraph::as_data_frame(gr7, what = "vertices")
```

2. `as_adjacency_matrix()`: This function takes a graph and returns its adjacency `matrix`.

```{r gra18}
as_adjacency_matrix(gr7)
```

The adjacency matrix has been returned, but without weights. The dots are zeros. R stores the matrix in a *sparse* form, which saves memory when most entries are zero, as in most real networks. To get the weights, we use the `attr` argument, which is `NULL` by default.

```{r gra19}
as_adjacency_matrix(gr7, attr = "weight")
```

## Adding vertex and/or edge attributes to an existing `igraph` object

We have seen how vertex and edge attributes can be included while creating a graph. But often we need to add attributes to a graph which already exists. Let us first see how to get the existing vertices, edges and their attributes.

1. Getting the vertices of a graph using `V()`

```{r gra20}
V(gr7)
```

2. Getting the edges of a graph using `E()`

```{r gra21}
E(gr7)
```

3. Extracting an edge or vertex attribute using `$`

```{r gra22}
# Vertex attribute "name"
V(gr7)$name
# Edge attribute "weight"
E(gr7)$weight
```

4. Adding a vertex or edge attribute, also using `$`

```{r gra23}
# adding vertex attribute say "airport_name"
V(gr7)$airport_name <- c("New Delhi", "Bengaluru", "Mumbai", "Pune")
# A single value will be repeated for all vertices
V(gr7)$country <- "India"

## Let's check it
igraph::as_data_frame(gr7, what = "vertices")
```

Let us also add some edge attributes. The values are given in the same order as the edges, which we saw with `E(gr7)` above.

```{r gra24}
# Adding edge attribute "carrier"
E(gr7)$carrier <- c("Air India", "Air India", "IndiGo", "Air India", "Akasa Air")
# Adding another edge attribute on the basis of a condition
E(gr7)$route <- case_when(
  E(gr7)$weight <= 400 ~ "Short",
  E(gr7)$weight <= 1000 ~ "Medium",
  TRUE ~ "Long"
)

## Let's check it
igraph::as_data_frame(gr7)
```

## Visualising graphs

After learning what graphs are and how to store them, let us see how to visualise them. Visualisation is crucial to convey the essence of a network. Apart from looks, the most important choice is the *layout*, that is, where each vertex is placed. The same graph can look very different in two layouts. Compare the two plots in @fig-gra25. Both show the same graph as @fig-gra2.

::: {#fig-gra25}

```{r echo=FALSE, message=FALSE, warning=FALSE, fig.align='center', fig.show='hold', out.height="34%", out.width="48%"}
set.seed(123)
igraph::as_data_frame(kite) %>%
  slice(1:9) %>%
  graph_from_data_frame() %>%
  plot(vertex.size = 25, layout = layout_as_star)

igraph::as_data_frame(kite) %>%
  slice(1:9) %>%
  graph_from_data_frame() %>%
  plot(vertex.size = 25, layout = layout_as_tree)

```

Two layouts of a same graph
:::

Many packages in R can plot graphs. We will learn two of them here, (i) `igraph` itself, which draws static plots, and (ii) `visNetwork`, which draws interactive plots. We used `ggraph` above for @fig-gra6. Visualising networks on a map is discussed in the geospatial chapters.

## Plotting using `plot()` in `igraph`

Plotting a graph in R is very simple. We just use `plot()`, which calls the `plot.igraph()` method with default values for all its other arguments. As we have seen, the layout is important, and we can set it with the argument `layout`.

In the following example (@fig-gra26), we can see two random layouts of the same graph as in @fig-gra25.

::: {#fig-gra26}

```{r message=FALSE, warning=FALSE, fig.align='center', fig.show='hold', out.height="34%", out.width="48%"}
my_gr <- igraph::as_data_frame(kite) %>%
  slice(1:9) %>%
  graph_from_data_frame()

set.seed(12345)
plot(my_gr, vertex.size = 25, layout = layout_randomly)

set.seed(54321)
plot(my_gr, vertex.size = 25, layout = layout_randomly)
```

Two Other layouts of a same graph
:::

Many layouts involve randomness, so the same layout may place the vertices differently every time. By setting a random number seed with `set.seed()`, we can fix the layout and reproduce it.

Let us now load a famous network, the *Zachary karate club*, and learn to plot networks effectively. Wikipedia describes it thus. *A social network of a karate club was studied by Wayne W. Zachary for a period of three years from 1970 to 1972. The network captures 34 members of a karate club, documenting links between pairs of members who interacted outside the club. During the study a conflict arose between the administrator "John A" and instructor "Mr. Hi" (pseudonyms), which led to the split of the club into two. Half of the members formed a new club around Mr. Hi; members from the other part found a new instructor or gave up karate. Based on collected data Zachary correctly assigned all but one member of the club to the groups they actually joined after the split.*

This graph is available in the package `igraphdata`, from which we can load it.

```{r gra27, warning=FALSE}
library(igraphdata)
data("karate")
karate <- upgrade_graph(karate)
karate
karate_data <- igraph::as_data_frame(karate)
head(karate_data)
plot(karate, layout = layout_nicely)
```

The vertex attribute `Faction` tells us which group each member actually joined after the split. The edge attribute `weight` is the number of different situations in which the two members interacted.

### Layouts

Network layouts are algorithms which return coordinates for each vertex. `igraph` offers several built-in layouts. The algorithms behind them are outside the scope of this chapter.

- `layout_as_bipartite()` Minimises edge crossings in a simple two-row (or column) layout, for bipartite graphs.
- `layout_as_star()` Places one vertex in the centre of a circle and the rest equally spaced on the circle.
- `layout_as_tree()` The **Reingold-Tilford** tree-like layout. Perfect for trees, and acceptable for graphs with not too many cycles.
- `layout_in_circle()` Places vertices on a circle, in the order of their vertex ids.
- `layout_nicely()` Tries to choose a suitable layout automatically, based on the graph.
- `layout_on_grid()` Places vertices on a rectangular grid, in two or three dimensions.
- `layout_on_sphere()` Places vertices on a sphere, approximately uniformly, in the order of their vertex ids.
- `layout_randomly()` Places the vertices randomly, in two or three dimensions.
- `layout_with_dh()` Places vertices by the simulated annealing algorithm of **Davidson and Harel**.
- `layout_with_fr()` Places vertices by the force-directed algorithm of **Fruchterman and Reingold**. Connected vertices pull each other close, and all vertices push each other away.
- `layout_with_gem()` Places vertices by the GEM force-directed algorithm.
- `layout_with_kk()` The **Kamada-Kawai** algorithm, based on a physical model of springs.
- `layout_with_sugiyama()` The **Sugiyama** algorithm for layered directed acyclic graphs. It minimises edge crossings.

Readers are advised to try these layouts to get a fair understanding of them. Let us see a few useful ones on the `karate` data.

- Circular layout: @fig-gra28 Left.
- Fruchterman Reingold: @fig-gra28 Right.
- Kamada Kawai: @fig-gra29 Left.
- Sugiyama: @fig-gra29 Right.
- Tree layouts (two representations): @fig-gra30.

```{r fig-gra28, message=FALSE, warning=FALSE, fig.align='center', fig.show='hold', fig.cap="Two layouts of Zachary Karate Club Network", out.height="34%", out.width="48%"}
igraph_options(vertex.size = 18)
par(mfrow = c(1, 2))

plot(karate, layout = layout_in_circle)
title("Circular layout")

plot(karate, layout = layout_with_fr)
title("Fruchterman - Reingold")
```

```{r fig-gra29, message=FALSE, warning=FALSE, fig.align='center', fig.show='hold', fig.cap="Two Other layouts of Zachary Karate Club Network", out.height="34%", out.width="48%"}
par(mfrow = c(1, 2))

plot(karate, layout = layout_with_kk)
title("Kamada Kawai")

plot(karate, layout = layout_with_sugiyama)
title("Sugiyama")
```

```{r fig-gra30, message=FALSE, warning=FALSE, fig.align='center', fig.show='hold', fig.cap="Two Tree layouts of Zachary Karate Club Network", out.height="34%", out.width="48%"}
par(mfrow = c(1, 2))

plot(karate, layout = layout_as_tree)
title("Default Tree layout")

plot(karate, layout = layout_as_tree(karate, circular = TRUE))
title("circular Tree")
```

For a social network like this, the force-directed layouts (Fruchterman-Reingold and Kamada-Kawai) show the structure best. The two groups around Mr Hi and John A can be seen clearly. The circular and tree layouts hide it.

### Displaying Vertex/Edge properties

Besides the layout, we can show vertex and edge attributes, like their categories or amounts, through colour, size, width and shape. Our `karate` graph is already coloured. How? The vertices of a graph are coloured by its vertex attribute `color`. Let us see it.

```{r gra31}
V(karate)$color
```

It contains the integers 1 and 2, which `plot()` maps to its default colours. Let us replace them with colours of our choice.

```{r fig-gra32, message=FALSE, warning=FALSE, fig.align='center', fig.show='hold', fig.cap="Vertex coloring in Network", out.height="34%", out.width="98%"}
V(karate)$color <- c("red", "dodgerblue")[V(karate)$color]
head(V(karate)$color)
plot(karate, layout = layout_nicely)
```

Similarly, we can use the following attributes to display properties of vertices and edges in an `igraph` plot.

+ **Vertex Properties**
  - `size` - Size of the vertex. Default is 15
  - `color` - Fill colour of the vertex
  - `frame.color` - Border colour of the vertex
  - `shape` - Shape of the vertex, like `"circle"`, `"square"`, `"rectangle"` or `"none"`
  - `label` - A character vector used to label the vertices
  - `label.family` - Font family of the label. Default is `serif`
  - `label.font` - Font of the label. `1` means plain (default), `2` bold, `3` italic, `4` bold italic and `5` symbol
  - `label.cex` - Font size of the label

+ **Edge properties**
  - `color` - Colour of the edge
  - `width` - Edge width
  - `lty` - Line type for edges (`0` or `"blank"`; `1` or `"solid"`; `2` or `"dashed"`; `3` or `"dotted"`; `4` or `"dotdash"`; `6` or `"twodash"`)
  - `label` - Label of the edge (with `label.family`, `label.font` and `label.cex` for font family, font type and font size respectively)
  - `curved` - Edge curvature, from `0` to `1`
  - `arrow.size` - Arrow size (default is 1), for directed graphs
  - `arrow.width` - Arrow width, default is `1`
  - `arrow.mode` - Arrow mode (`0` means no arrow, `1` backward, `2` forward and `3` both)

Instead of storing these as attributes in the graph, we can also give them as arguments to `plot()`, by adding `vertex.` or `edge.` before the property name, like `vertex.size = 18`.

Let us see some of these through the following example.

```{r fig-gra33, message=FALSE, warning=FALSE, fig.align='center', fig.show='hold', fig.cap="Edge and Vertex properties in Network", out.width="98%"}
# set reproducible layout
set.seed(123)
l_s <- layout_nicely(karate)

# Add vertex labels
V(karate)$label <- stringr::str_remove(V(karate)$name, "Actor ")

# Change shapes of two prominent actors
V(karate)$shape <- ifelse(V(karate)$name %in% c("Mr Hi", "John A"), "rectangle", "circle")

# Change Edge width as per "weight"
E(karate)$width <- E(karate)$weight

plot(karate, layout = l_s, edge.label.cex = 0.7)
```

## Using `visNetwork` for interactive visualizations

Static plots are fine for small graphs and for printed reports. But while *exploring* a network, we want to zoom in, drag vertices around, and click on a vertex to see its details. The package `visNetwork` does exactly this. It draws interactive networks in a web browser, in RStudio's viewer, and in HTML documents like this book.

`visNetwork()` takes two data frames.

- `nodes` must have a column `id`. Optional columns control the look, like `label` (text shown), `group` (for colouring), `title` (tooltip shown on hover), `value` (size), `shape` and `color`.
- `edges` must have the columns `from` and `to`. Optional columns include `value` or `width` (thickness), `label`, `title`, `arrows` and `color`.

Let us draw the karate network interactively. We colour the members by the faction they joined, and size them by their number of connections. Try hovering over a member, clicking one to highlight its neighbours, choosing a member from the drop-down list, and dragging a member around.

```{r}
#| label: fig-vis1
#| fig-cap: "Interactive karate club network with visNetwork"
library(visNetwork)

kar_nodes <- data.frame(
  id    = V(karate)$name,
  label = V(karate)$label,
  group = ifelse(V(karate)$Faction == 1, "Mr Hi's group", "John A's group"),
  value = degree(karate),
  title = paste0("<b>", V(karate)$name, "</b><br>Connections: ", degree(karate))
)

kar_edges <- igraph::as_data_frame(karate) |>
  select(from, to, weight) |>
  mutate(value = weight, title = paste("Interactions:", weight))

visNetwork(kar_nodes, kar_edges, width = "100%", height = "500px") |>
  visOptions(highlightNearest = TRUE, nodesIdSelection = TRUE) |>
  visLegend()
```

A few more useful functions of `visNetwork` are

- `visIgraph(g)` draws an `igraph` object directly, using its attributes.
- `toVisNetworkData(g)` converts an `igraph` object into the `nodes` and `edges` data frames which `visNetwork()` needs. We can then add or change columns before drawing.
- `visIgraphLayout()` calculates the layout in R with `igraph`, instead of in the browser. The network then appears instantly and does not keep moving. This is very useful for bigger networks.
- `visEdges(arrows = "to")` draws arrows on directed edges.
- `visSave(network, file = "network.html")` saves the interactive network as a standalone HTML file, which can be shared with colleagues who do not use R.

::: callout-tip
## Tip

The browser has to draw every vertex and edge of a `visNetwork`. A few hundred vertices are fine. With many thousands, the page becomes slow, and the picture is a hairball that nobody can read anyway. So first reduce the network in R, as explained in the next section and in the next chapter, and then draw only the part that matters.
:::

## Subgraphs, adding or deleting edges/vertices

Real networks are big. Usually we are interested in only a part of them. Let us see how to take out parts of a graph, and how to add or delete vertices and edges.

**Selecting vertices and edges.** We can select vertices and edges by a condition on their attributes, inside `V()` and `E()`.

```{r}
# Members with more than 10 connections
V(karate)[degree(karate) > 10]

# Edges with 5 or more interactions
E(karate)[weight >= 5]
```

**Neighbours.** `neighbors()` gives the vertices directly connected to a vertex.

```{r}
neighbors(karate, "John A")
```

**Subgraphs.** `induced_subgraph()` keeps the chosen vertices and all the edges *between* them. `subgraph_from_edges()` keeps the chosen edges and their vertices.

```{r}
# The graph of Mr Hi's faction only
mrhi_group <- induced_subgraph(karate, V(karate)[Faction == 1])
mrhi_group
```

**Ego networks.** The *ego network* of a vertex is the vertex itself, its neighbours, and the edges between them. The argument `order` sets how many steps away we go. `make_ego_graph()` returns a list of such graphs, one for each vertex we ask for. In audit, the ego network of a suspicious entity is often the first thing we look at.

```{r fig-ego, fig.cap="Ego network of Actor 20, up to 2 steps", fig.align='center', out.width="80%"}
ego20 <- make_ego_graph(karate, order = 2, nodes = "Actor 20")[[1]]
plot(ego20, layout = layout_with_fr)
```

**Deleting.** `delete_vertices()` and `delete_edges()` remove vertices and edges. When a vertex is deleted, all its edges go with it.

```{r}
# Remove the two leaders and see whether the club stays connected
without_leaders <- delete_vertices(karate, c("Mr Hi", "John A"))
components(without_leaders)$csize
```

Without its two leaders, the club falls apart into `r components(without_leaders)$no` separate pieces. Those leaders were holding it together.

**Adding.** `add_vertices()` and `add_edges()` add new vertices and edges. A convenient alternative is the `+` operator with `vertices()` and `edges()`.

```{r}
small <- make_graph(c("A", "B", "B", "C"), directed = FALSE)
small <- small + vertices("D") + edges("C", "D", "D", "A")
small
```

## Describing network/graph for edges and nodes therein

Once we have a network, we want to describe it. How big and how dense is it? Which vertices are the most important? Are there groups within it? Let us answer these questions for the karate club.

### Measures for the whole network

```{r}
# Number of vertices and edges
vcount(karate)
ecount(karate)

# Density, the share of all possible edges which exist
edge_density(karate)

# Diameter, the longest of all shortest paths (counting steps)
diameter(karate, weights = NA)

# Average number of steps between two members
mean_distance(karate, weights = NA)

# Transitivity, the chance that two friends of a member are friends too
transitivity(karate)
```

Only `r round(edge_density(karate) * 100)`% of all possible pairs of members are connected. Yet any member can reach any other in at most `r diameter(karate, weights = NA)` steps, and on average in `r round(mean_distance(karate, weights = NA), 1)` steps. Most real networks, from friendships to company shareholdings, are like this, sparse but closely knit.

::: callout-note
## Weights and distances

`igraph` treats the `weight` attribute as a *distance* in path-based measures like `diameter()`, `distances()`, `betweenness()` and `closeness()`. In the karate data, a larger weight means a *stronger* relation, which should mean a *shorter* distance. So here we use `weights = NA` to ignore the weights and simply count steps. Always check what the weights in your data mean before using them in such measures.
:::

### Paths

`shortest_paths()` gives the shortest path between two vertices, and `distances()` gives the number of steps between every pair. In audit, the path tells us *how* two entities are connected, for example how a vendor is linked to an employee.

```{r}
shortest_paths(karate, from = "Actor 17", to = "John A", weights = NA)$vpath
```

### Importance of vertices (centrality)

Which members are the most important? There are several ways to measure this, called *centrality* measures.

- **Degree** is the number of edges of a vertex. For a directed graph, *in-degree* and *out-degree* count incoming and outgoing edges separately. A bank account receiving money from 40 beneficiaries has a high in-degree.
- **Strength** is the sum of the weights of the edges of a vertex, for example the total amount paid or received.
- **Betweenness** counts how often a vertex lies on the shortest paths between other vertices. A vertex with high betweenness is a *bridge* or a middleman between groups.
- **Closeness** measures how close a vertex is to all others. A vertex with high closeness can reach everyone quickly.
- **Eigenvector centrality** and **PageRank** give high scores to vertices connected to other important vertices. Being connected to the boss matters more than being connected to many juniors.

Let us calculate these for the karate club and look at the top members.

```{r}
centrality <- tibble(
  member      = V(karate)$name,
  degree      = degree(karate),
  strength    = strength(karate),
  betweenness = round(betweenness(karate, weights = NA), 1),
  closeness   = round(closeness(karate, weights = NA), 3),
  eigen       = round(eigen_centrality(karate)$vector, 2)
) |>
  arrange(desc(betweenness))

head(centrality)
```

Mr Hi and John A, the two leaders, are at or near the top of every measure. That is what we would expect. But look at members like Actor 33, Actor 3 and Actor 32 too. They are not leaders, but they connect many others. Actor 3 even ranks above Mr Hi in eigenvector centrality, because it is connected to well-connected members of both groups, like Mr Hi, Actor 2 and Actor 33. In an investigation, such vertices are worth a look.

### Communities

A *community* is a group of vertices which are densely connected with each other, and only loosely connected with the rest. Finding communities is called *community detection* or *clustering*. `igraph` has several algorithms for it. The Louvain method, `cluster_louvain()`, is fast and works well for most networks.

```{r}
set.seed(42)
kar_comm <- cluster_louvain(karate, weights = NA)
sizes(kar_comm)
```

Let us compare the communities found with the faction each member actually joined after the split.

```{r}
table(community = membership(kar_comm), faction = V(karate)$Faction)
```

Louvain has found `r length(kar_comm)` communities. Each of them lies almost entirely within one faction. Only `r sum(apply(table(membership(kar_comm), V(karate)$Faction), 1, min))` member is placed in a community dominated by the other faction. Just from the pattern of friendships, the algorithm has predicted the split of the club, as Zachary did in 1977. @fig-comm shows the communities.

```{r fig-comm, fig.cap="Communities found by the Louvain method", fig.align='center', out.width="90%"}
plot(kar_comm, karate, layout = l_s)
```

In audit, communities point to groups like bidders who bid together, firms which trade mostly among themselves, or beneficiaries who share agents. We will see this in the next chapter.

## A suggested workflow

1.  **Build an edge list.** From the payment register or beneficiary list, make a data frame in which each row is one link, like a beneficiary and a bank account, or a payment from one firm to another. Keep the amount and the date as further columns, so they become edge attributes.
2.  **Create the graph.** Use `graph_from_data_frame()`. Keep it directed for payments, and use `directed = FALSE` for shared details like a common account. Give a `vertices` table to keep entities with no links, and their attributes.
3.  **Check the graph object.** Print it and confirm the number of vertices and edges, and the attributes. Get the data back with `igraph::as_data_frame()` whenever you need to inspect it.
4.  **Describe the whole network.** Use `vcount()`, `ecount()`, `edge_density()` and `components()` to see how big the network is and whether it falls into separate pieces.
5.  **Find the important entities.** Calculate `degree()`, `strength()` and `betweenness()`. A bank account receiving money from many beneficiaries, or a firm acting as a bridge between groups, deserves a look.
6.  **Find the groups.** Run `cluster_louvain()` to find communities, such as bidders who bid together or firms which trade mostly among themselves.
7.  **Zoom into doubtful cases.** Take the ego network of a doubtful entity with `make_ego_graph()`, or a part of the graph with `induced_subgraph()`. Use `shortest_paths()` to see how two entities are linked.
8.  **Show only what matters.** Draw the reduced network with `plot()` or `visNetwork()`. Save the interactive version with `visSave()` to share with colleagues who do not use R.

::: callout-important
## Remember

A connection in a network is a lead, not a finding. Two beneficiaries may share an account for a genuine reason, so examine the records before reporting. Also check what the weights in your data mean, because `igraph` treats `weight` as a distance in measures like `betweenness()`.
:::

## Summary of functions used

| Function | Package | What it does |
|-----------------------|----------------|--------------------------------|
| `graph_from_data_frame()` | igraph | Creates a graph from a data frame of edges (and vertices) |
| `graph_from_edgelist()`, `graph_from_adjacency_matrix()` | igraph | Create a graph from an edge matrix or adjacency matrix |
| `as_undirected()`, `as_directed()` | igraph | Convert between directed and undirected graphs |
| `igraph::as_data_frame()`, `as_adjacency_matrix()` | igraph | Get the data back from a graph |
| `V()`, `E()` | igraph | Vertices and edges of a graph, with their attributes via `$` |
| `plot()`, `layout_*()` | igraph | Static plot of a graph, with a chosen layout |
| `visNetwork()`, `visOptions()`, `visIgraphLayout()`, `visSave()` | visNetwork | Interactive network plots |
| `neighbors()`, `induced_subgraph()`, `make_ego_graph()` | igraph | Parts of a graph around chosen vertices |
| `delete_vertices()`, `delete_edges()`, `add_vertices()`, `add_edges()` | igraph | Remove or add vertices and edges |
| `components()` | igraph | Splits a graph into connected pieces |
| `edge_density()`, `diameter()`, `mean_distance()`, `transitivity()` | igraph | Measures of the whole network |
| `shortest_paths()`, `distances()` | igraph | Paths and number of steps between vertices |
| `degree()`, `strength()`, `betweenness()`, `closeness()`, `eigen_centrality()` | igraph | Importance (centrality) of each vertex |
| `cluster_louvain()`, `membership()` | igraph | Finds communities in a network |

: Functions used in this chapter {#tbl-graphfunctions}
