```{r}
#| include: false
source("_common.R")
```


# Networks and Graph Theory {#sec-graph}

Suppose we audit the works contracts of a Public Works Division. Sixty contractors are registered with it. On paper they are sixty separate firms, and every tender gets three or four bids. But some of these firms may share a partner, an office address, a mobile number or a bank account. If such linked firms bid against each other, the competition is only on paper. A table of sixty firms will not show this. A picture of who is linked to whom shows it at once.

Many audit questions are like this. Which beneficiaries of a scheme share one bank account? Which firms always bid together? Does money go round in a circle between a few companies? Is a vendor linked to the clerk who passes its bills? These are questions about *connections*. Network analysis is built to answer them. With R, we can build such networks from ordinary audit data, draw them, find the key players, and find the hidden groups.

In this chapter we learn the ideas of graph theory and the tools to work with networks in R. We start with small examples to learn the words, like vertex, edge, directed and bipartite. We then learn to create graphs, add details to them, and draw them. Finally we build a network of linked contractors of an imaginary Division. We find its key firms and its hidden groups, and we check the Division's tenders for bids from linked firms. In @sec-netaudit, we apply the same tools to more audit problems.

## An introduction to graph theory

Network analysis applies the concepts of *graph theory*, a branch of mathematics, to understand complex systems like social networks, transport networks and financial transactions. Graph theory studies the properties of graphs, such as paths, cycles and connectivity.

### Definition of a graph

A graph is a mathematical structure consisting of a set of *vertices* (also called nodes or actors) and a set of *edges* (also called connections or links) between these vertices. Mathematically, a graph $G = (V, E)$ consists of a set $V$ of vertices and a set $E$ of edges, as illustrated in @fig-gra1. A vertex may represent a real-world object, like a person, a computer, a firm or a bank account. An edge represents the *relationship* between the two vertices it connects, like friendship between two persons, a cable between two computers, or a payment from one firm to another.

```{r fig-gra1, echo=FALSE, message=FALSE, warning=FALSE, fig.align='center', fig.show='hold', fig.cap="An example graph", out.width="100%"}
library(igraph)
library(igraphdata)
data("kite")
kite <- upgrade_graph(kite)
set.seed(123)
plot(kite, vertex.size = 25)
```

In audit terms, think of the ten vertices of @fig-gra1 as ten firms, and each edge as a detail two firms share, like a partner or an address.

### Types of graph {.unnumbered}

The graph in @fig-gra1 is an **undirected graph**, in which the edges have no direction. A shared address is an example. If firm A shares an address with firm B, then B shares it with A. If the edges have a direction, from one vertex to another, the graph is called a **directed graph**. For example, on X (formerly Twitter) a user may follow another user without being followed back. A payment from one firm to another is also directed. See @fig-gra2.

```{r fig-gra2, echo=FALSE, message=FALSE, warning=FALSE, fig.align='center', fig.show='hold', fig.cap="An example directed graph", out.width="100%", out.height="33%"}
library(tidyverse)
set.seed(123)
igraph::as_data_frame(kite) %>%
  slice(1:9) %>%
  graph_from_data_frame() %>%
  plot(vertex.size = 25)
```

Sometimes an edge connects a vertex with itself. Such an edge is called a **self-loop**. Graphs which allow self-loops are sometimes called **pseudographs**. For a real-world example, think of transactions between firms. A firm usually transacts with other firms, but it may also transfer money between two of its own accounts. In @fig-gra3 (left), vertex `E` is connected with itself.

Further, there may be more than one edge connecting the same pair of vertices (in the same direction). These are called **multi-edges**. In @fig-gra3 (right) there are two edges from `B` to `E`. In audit data, multi-edges are common. One seller raises many invoices on the same buyer, and one Division makes many payments to the same contractor.

::: {#fig-gra3}

```{r echo=FALSE, message=FALSE, warning=FALSE, fig.align='center', fig.show='hold', out.width="48%", out.height="33%"}
set.seed(123)
igraph::as_data_frame(kite) %>%
  slice(c(1, 3, 5:6)) %>%
  add_row(from = "E", to = "E") %>%
  graph_from_data_frame() %>%
  plot(vertex.size = 45)
igraph::as_data_frame(kite) %>%
  slice(c(1, 3, 5:6)) %>%
  add_row(from = "B", to = "E") %>%
  graph_from_data_frame() %>%
  plot(vertex.size = 45)
```

Examples of a self-loop (left) and a multi-edge (right)
:::

**Complete graphs** are graphs in which every pair of vertices is connected by an edge. Examples are shown in @fig-gra4. Complete graphs are rare in the real world. But a few vertices of a bigger graph may all be connected with each other. Such a group, which is a complete graph in itself, is called a **clique**. For example, the vertices `B, D, E and G` in @fig-gra1 form a clique. Every one of them is connected with the other three. In audit, a clique of firms which all trade with each other, or of bidders who all bid together, deserves attention.

::: {#fig-gra4}

```{r echo=FALSE, message=FALSE, warning=FALSE, fig.align='center', fig.show='hold', out.width="49%", out.height="31%"}
set.seed(123)
plot(make_full_graph(4), vertex.size = 45)
plot(make_full_graph(6), vertex.size = 45)
```

Examples of complete graphs
:::

Let us learn about two more types of graph. The first is the **bipartite graph**. Its vertices are divided into two separate sets, and edges connect only vertices of *different* sets, never two vertices of the same set. The graph in @fig-gra5 (left) is bipartite. Think of four firms (A1 to A4) and five tenders (B1 to B5). An edge shows that a firm bid in a tender. A firm never "bids" in another firm, and a tender never bids in a tender. In audit, bipartite graphs are very common. Beneficiaries and their bank accounts, or employees and the bills they passed, also form two sets with edges only between the sets. The idea can be extended to *k-partite* graphs, with `k` such sets.

The second is the **tree**, which represents a hierarchy. The offices of a department form a tree, with the head office at the top, then the circles, then the divisions. A graph is a tree if there is exactly one path between any pair of vertices (ignoring direction). The bipartite graph in @fig-gra5 (left) is also a tree, since it meets this condition. In @fig-gra5 (right) the same graph has been redrawn as a tree.

::: {#fig-gra5}

```{r echo=FALSE, message=FALSE, warning=FALSE, fig.align='center', fig.show='hold', out.width="49%", out.height="40%"}
# Firms (A1 to A4) and the tenders they bid in (B1 to B5)
g <- graph_from_literal(A1-B1, A1-B2, A2-B2, A2-B3, A3-B3, A3-B4, A4-B4, A4-B5)
V(g)$type <- startsWith(V(g)$name, "B")

# define color and shape mappings.
col <- c("steelblue", "orange")
shape <- c("circle", "square")

plot(g,
     vertex.color = col[V(g)$type + 1],
     vertex.shape = shape[V(g)$type + 1],
     vertex.size = 35,
     layout = layout_as_bipartite
)

plot(g,
     vertex.color = "pink",
     vertex.shape = "square",
     vertex.size = 35,
     layout = layout_as_tree
)
```

A bipartite graph (left) and a tree (right)
:::

### Vertex and edge attributes

Recall that a graph consists of two sets $(V, E)$, where $V$ is a set of vertices and $E$ is a set of edges. In their simplest form, these sets are just lists. But both can carry additional *attributes*.

Let us take an example from the audit of travelling allowance (TA). Officers of a department travel between four cities on tour, and claim TA for the distance travelled. An auditor may keep the routes as a graph. Each city is a vertex, with attributes like (i) its name, (ii) its airport, and (iii) its country. Each route is an edge, with attributes like (i) the airlines operating it, and (ii) the distance in kilometres. A claim for more than the distance of the route can then be spotted. The graph is shown in @fig-gra6. In other audit data, typical edge attributes are the amount and the date of a payment.

```{r fig-gra6, echo=FALSE, message=FALSE, warning=FALSE, fig.align='center', fig.show='hold', fig.cap="Vertex and edge attributes in a graph of tour routes", out.width="98%", out.height="32%"}
library(ggraph)
airtrans <- data.frame(
  from = c("A", "A", "A", "B", "C"),
  to = c("B", "C", "D", "C", "D")
) |> as.matrix()

airtransgraph <- graph_from_edgelist(airtrans, directed = FALSE)
V(airtransgraph)$loc <- c("New Delhi", "Bangalore", "Mumbai", "Pune")
V(airtransgraph)$country <- rep("India", 4)
E(airtransgraph)$dist <- c(1750, 1150, 1200, 850, 120)
E(airtransgraph)$airline <- c(rep("Air India", 3), "IndiGo", "Akasa Air")

set.seed(123)
ggraph(airtransgraph, layout = 'kk') +
  geom_edge_fan(aes(label = paste0("Airline: ", airline, ", Distance: ", dist, " km")),
                angle_calc = "along",
                label_dodge = unit(2.5, 'mm'),
                label_size = 4
  ) +
  geom_node_point(size = 20, color = "#D4E7C5") +
  geom_node_text(aes(label = paste0(name, " \n(Airport: ", loc, ")\n", country)), size = 3) +
  scale_x_continuous(expand = expansion(mult = 0.1)) +
  theme_void()
```


## Practical approach for creating graphs in R

### Packages to use

One of the best packages for creating and analysing graphs in R is `igraph`. It was originally developed by **Gábor Csárdi** and **Tamás Nepusz**, and its core is written in the C programming language for good performance. So it can handle networks with millions of edges. Let us load it.

```{r}
#| message: false
#| warning: false
library(igraph)
library(tidyverse)
```

`igraph` is a complete package for network analysis in itself. We will also use `visNetwork` and `ggraph`. The former creates interactive network charts, and the latter creates `ggplot2` compatible plots, which can be customised further in a familiar way.

::: callout-note
## Function names in igraph

Version 2 of `igraph` renamed many functions to a consistent `snake_case` style. For example, `as.undirected()` is now `as_undirected()` and `graph.data.frame()` is now `graph_from_data_frame()`. The old names still work for now, with a warning. You may find old names in books and websites, but use the new ones in your code.
:::

Let us now learn how to create graph objects in R. Graphs can be created in many ways. We will learn three methods, which will serve our purpose most of the time.

### Creating a graph from a data frame

In data analysis, data frames are our primary tool. So the most common and practical way to create a graph is from a `data.frame`. The data frame should have at least two columns, where each row is one edge. The first column is taken as the `'from'` vertex and the second as the `'to'` vertex, whatever their column names. We use the function `graph_from_data_frame()` from `igraph`. Its syntax is

```
graph_from_data_frame(
  d =         ,  # data.frame of edges
  directed = TRUE,
  vertices = NULL # optional data.frame of vertices
)
```

Let us construct a graph from a data frame of four edges. Think of the numbers as codes of five bank accounts. Each row is a transfer from one account to another.

```{r gra7}
# Example data frame of edges, one row per transfer
df <- data.frame(
  from = c(1, 1, 3, 3),
  to = c(2, 3, 5, 4)
)
# Creating graph object
gr1 <- graph_from_data_frame(df)
# Let us print it
gr1
```

A graph object named `gr1` has been created. Printing it gives us useful information about it.

+ The first line always begins with `IGRAPH`, followed by seven characters. These are the first characters of a unique graph ID.
+ Next comes a four-letter code. In this example it is `DN--`. The letters mean
  - The first letter tells whether the graph is directed (`D`) or undirected (`U`).
  - The second letter, `N`, means a *named* graph, that is, the vertices have a `name` attribute.
  - The third letter, `W`, means a *weighted* graph, that is, the edges have a `weight` attribute.
  - The fourth letter, `B`, means a *bipartite* graph, that is, the vertices have a `type` attribute.
+ Next come the number of vertices and edges, separated by two dashes.
+ The second line lists the attributes of the graph. For each attribute, its kind, graph (`g`), vertex (`v`) or edge (`e`), and its data type, character (`c`), numeric (`n`), logical (`l`) or other (`x`), are shown in brackets.
+ The last lines show the edges.

All four edges of our `df` are in `gr1`. Notice that the vertex names are characters, `"1"`, `"2"` and so on, even though our data had numbers. `igraph` always stores vertex names as characters.

::: callout-warning
## Caution
In audit, vertices are often account numbers, PANs or vendor codes. Read such columns as text, not numbers. If an account number like `00123` is read as a number, it becomes `123`. It may then fail to match the same account in another table, or two different accounts may end up as one vertex.
:::

The argument `directed` is `TRUE` by default, so the graph is directed, as confirmed by the first letter `D`. For transfers, this is right, because money goes from one account to the other. But if we only want to know which accounts have dealt with each other, the direction does not matter. Let us create an undirected graph.

```{r gra8}
# Creating graph object
gr2 <- graph_from_data_frame(df, directed = FALSE)
# Let us print it
gr2
```

The first letter is now `U`, and the edges are printed with `--` instead of an arrow `->`. An existing directed graph can also be converted into an undirected one using the function `as_undirected()` (and the other way round with `as_directed()`).

```{r gra9}
as_undirected(gr1)
```

Now what about the last argument, `vertices`? In most cases, the edge data is enough. But a graph may contain *isolates*, that is, vertices not connected to any edge. These do not appear in the edge data, so to include them we use the `vertices` argument. Its first column must contain the vertex IDs used in the edge data. Any other columns become vertex attributes. Similarly, any columns of the edge data frame after the first two become edge attributes.

Let us add a sixth account, `"6"`, which made no transfer during the period. It is an isolated vertex. Let us also add the name of each account holder as a `label`.

```{r gra10}
df_v <- data.frame(
  id = c(1, 2, 3, 4, 5, 6),
  label = c("Ramesh", "Suresh", "Anil", "Kavita", "Bimla", "Mohan")
)
# Creating graph object
gr3 <- graph_from_data_frame(df, directed = FALSE, vertices = df_v)
# Let us print it
gr3
```

The number of vertices is now 6, though there are still 4 edges. The vertex attribute `label` has also been added, shown as `label (v/c)`. In audit, isolated vertices matter. A contractor who shares nothing with anyone is still a contractor, and must stay in the count.

Readers may try adding more columns to the edge data frame, like an amount or a date, and see how they become edge attributes.

### Creating a graph from an edge list

Creating a graph from an edge list is nearly the same as creating it from a data frame. The differences are that the input must be a two column *matrix*, character or numeric, and that there is no `vertices` argument. The syntax is

```
graph_from_edgelist(el, directed = TRUE)
```

Here each edge joins two claimants of a scheme who stood as witness for each other.

```{r gra11}
edges <- data.frame(
  origin = c("Ramesh", "Ramesh", "Anil", "Anil"),
  dest = c("Suresh", "Anil", "Kavita", "Bimla")
)
edges <- as.matrix(edges)
# Now let us create a graph
gr4 <- graph_from_edgelist(edges, directed = FALSE)
# And print it
gr4
```

We have created an undirected graph with `5` vertices and `4` edges.

### Creating a graph from an adjacency matrix

An adjacency matrix is a square $n \times n$ matrix, where both the rows and the columns stand for the $n$ vertices. The $(i, j)$-th entry of this matrix means

*   In a graph without weights, `1` means there is an edge from vertex $v_{i}$ to $v_j$, and `0` means there is none.
*   In a weighted graph, the entry is the weight of the edge from vertex $v_{i}$ to $v_j$, and `0` means there is no edge.

Let us create the graph of tour routes in @fig-gra6 through its adjacency matrix. First, without weights.

```{r gra12}
# Let us create the matrix
adj_mat <- structure(c(0, 1, 1, 1, 1, 0, 1, 0, 1, 1, 0, 1, 1, 0, 1, 0), dim = c(4L,
4L), dimnames = list(c("DEL", "BNG", "BOM", "PNQ"), c("DEL",
"BNG", "BOM", "PNQ")))
adj_mat
# Let us create the graph
gr5 <- graph_from_adjacency_matrix(adj_mat)
# Let us print it
gr5
```

A directed graph (the default) has been created, with 10 edges. Since our matrix is symmetric, each route appears twice, once in each direction, like `DEL->BNG` and `BNG->DEL`. For a symmetric matrix, we would usually set `mode = "undirected"` to get one edge per route. Now let us create a weighted graph of the same data, where the weights are the distances between the cities, in kilometres.

```{r gra13}
# Let us create the matrix
adj_mat2 <- structure(c(0, 1750, 1150, 1200, 1750, 0, 850, 0, 1150, 850,
0, 120, 1200, 0, 120, 0), dim = c(4L, 4L), dimnames = list(c("DEL",
"BNG", "BOM", "PNQ"), c("DEL", "BNG", "BOM", "PNQ")))
adj_mat2
# Let us create the graph
gr6 <- graph_from_adjacency_matrix(adj_mat2, weighted = TRUE)
# Let us print it
gr6
```

This time an edge attribute named `weight`, of `numeric` type, has been added to the graph.

### Getting data back from graph objects

Once a graph has been created, getting the data back as a `data.frame` or an adjacency `matrix` is easy. We use the `igraph` functions `as_data_frame()` and `as_adjacency_matrix()`.

1. `as_data_frame()`. Since older versions of the package `tibble` also had a function of the same name, it is safest to write `igraph::as_data_frame()`. It takes a graph and returns a `data.frame`. Let us get the data back from `gr6`, which we created from an adjacency matrix.

```{r gra14}
igraph::as_data_frame(gr6)
```

The function worked correctly. Since the graph was directed, each route appears in both directions. Let us also see the output for an undirected graph.

```{r gra15}
igraph::as_data_frame(as_undirected(gr6))
```

There is one row per route now. But notice that the weights have *doubled*, for example DEL to BNG is now 3,500. When two directed edges are merged into one, `as_undirected()` by default *adds up* their weights. We can choose another way to combine them using the argument `edge.attr.comb`, which takes a function (or its name) for each attribute. Here, the mean is right.

```{r gra16}
gr7 <- as_undirected(gr6, edge.attr.comb = list(weight = "mean"))
igraph::as_data_frame(gr7)
```

This function has another argument, `what`. It can be `"edges"` (the default), `"vertices"`, or `"both"`, in which case the two data frames are returned in a list.

```{r gra17}
# Vertices data only
igraph::as_data_frame(gr7, what = "vertices")
```

2. `as_adjacency_matrix()`. This function takes a graph and returns its adjacency `matrix`.

```{r gra18}
as_adjacency_matrix(gr7)
```

The adjacency matrix has been returned, but without weights. The dots are zeros. R stores the matrix in a *sparse* form, which saves memory when most entries are zero, as in most real networks. To get the weights, we use the `attr` argument, which is `NULL` by default.

```{r gra19}
as_adjacency_matrix(gr7, attr = "weight")
```

## Adding attributes to an existing graph

We have seen how vertex and edge attributes can be included while creating a graph. But often we need to add attributes to a graph which already exists. Let us first see how to get the existing vertices, edges and their attributes.

1. Getting the vertices of a graph using `V()`

```{r gra20}
V(gr7)
```

2. Getting the edges of a graph using `E()`

```{r gra21}
E(gr7)
```

3. Extracting an edge or vertex attribute using `$`

```{r gra22}
# Vertex attribute "name"
V(gr7)$name
# Edge attribute "weight"
E(gr7)$weight
```

4. Adding a vertex or edge attribute, also using `$`

```{r gra23}
# Adding the vertex attribute "airport_name"
V(gr7)$airport_name <- c("New Delhi", "Bengaluru", "Mumbai", "Pune")
# A single value will be repeated for all vertices
V(gr7)$country <- "India"

# Let us check it
igraph::as_data_frame(gr7, what = "vertices")
```

Let us also add some edge attributes. The values are given in the same order as the edges, which we saw with `E(gr7)` above.

```{r gra24}
# Adding edge attribute "carrier"
E(gr7)$carrier <- c("Air India", "Air India", "IndiGo", "Air India", "Akasa Air")
# Adding another edge attribute on the basis of a condition
E(gr7)$route <- case_when(
  E(gr7)$weight <= 400 ~ "Short",
  E(gr7)$weight <= 1000 ~ "Medium",
  TRUE ~ "Long"
)

# Let us check it
igraph::as_data_frame(gr7)
```

With the distance stored on each route, a TA claim can be joined to its route by the two cities. Any claim for more kilometres than the `weight` of the route then stands out.

## Visualising graphs

After learning what graphs are and how to store them, let us see how to visualise them. Visualisation is crucial to convey the essence of a network. Apart from looks, the most important choice is the *layout*, that is, where each vertex is placed. The same graph can look very different in two layouts. Compare the two plots in @fig-gra25. Both show the same graph as @fig-gra2.

::: {#fig-gra25}

```{r echo=FALSE, message=FALSE, warning=FALSE, fig.align='center', fig.show='hold', out.height="34%", out.width="48%"}
set.seed(123)
igraph::as_data_frame(kite) %>%
  slice(1:9) %>%
  graph_from_data_frame() %>%
  plot(vertex.size = 25, layout = layout_as_star)

igraph::as_data_frame(kite) %>%
  slice(1:9) %>%
  graph_from_data_frame() %>%
  plot(vertex.size = 25, layout = layout_as_tree)

```

Two layouts of the same graph
:::

Many packages in R can plot graphs. We will learn two of them here, (i) `igraph` itself, which draws static plots, and (ii) `visNetwork`, which draws interactive plots. We used `ggraph` above for @fig-gra6. Visualising networks on a map is discussed in the geospatial chapters.

## Plotting with `plot()` in igraph

Plotting a graph in R is very simple. We just use `plot()`, which calls the `plot.igraph()` method with default values for all its other arguments. As we have seen, the layout is important, and we can set it with the argument `layout`.

In the following example (@fig-gra26), we can see two random layouts of the same graph as in @fig-gra25.

::: {#fig-gra26}

```{r message=FALSE, warning=FALSE, fig.align='center', fig.show='hold', out.height="34%", out.width="48%"}
my_gr <- igraph::as_data_frame(kite) %>%
  slice(1:9) %>%
  graph_from_data_frame()

set.seed(12345)
plot(my_gr, vertex.size = 25, layout = layout_randomly)

set.seed(54321)
plot(my_gr, vertex.size = 25, layout = layout_randomly)
```

Two random layouts of the same graph
:::

Many layouts involve randomness, so the same layout may place the vertices differently every time. By setting a random number seed with `set.seed()`, we can fix the layout and reproduce it.

To learn the plotting options, let us use a small and famous network, the *Zachary karate club*. It is used in almost every book on networks, so it is worth knowing. Wayne W. Zachary studied a karate club of 34 members from 1970 to 1972. He recorded which pairs of members met outside the club. During the study, the administrator ("John A") and the instructor ("Mr Hi") fell out, and the club split into two. Each member joined one side or the other. Later in this chapter we will use the same tools on audit data, a network of contractors.

This graph is available in the package `igraphdata`, from which we can load it.

```{r gra27, warning=FALSE}
library(igraphdata)
data("karate")
karate <- upgrade_graph(karate)
karate
karate_data <- igraph::as_data_frame(karate)
head(karate_data)
set.seed(123)
plot(karate, layout = layout_nicely)
```

The vertex attribute `Faction` tells us which group each member actually joined after the split. The edge attribute `weight` is the number of different situations in which the two members interacted.

### Layouts

Network layouts are algorithms which return coordinates for each vertex. `igraph` offers several built-in layouts. The algorithms behind them are outside the scope of this chapter.

- `layout_as_bipartite()` Minimises edge crossings in a simple two-row (or column) layout, for bipartite graphs.
- `layout_as_star()` Places one vertex in the centre of a circle and the rest equally spaced on the circle.
- `layout_as_tree()` The **Reingold-Tilford** tree-like layout. Perfect for trees, and acceptable for graphs with not too many cycles.
- `layout_in_circle()` Places vertices on a circle, in the order of their vertex ids.
- `layout_nicely()` Tries to choose a suitable layout automatically, based on the graph.
- `layout_on_grid()` Places vertices on a rectangular grid, in two or three dimensions.
- `layout_on_sphere()` Places vertices on a sphere, approximately uniformly, in the order of their vertex ids.
- `layout_randomly()` Places the vertices randomly, in two or three dimensions.
- `layout_with_dh()` Places vertices by the simulated annealing algorithm of **Davidson and Harel**.
- `layout_with_fr()` Places vertices by the force-directed algorithm of **Fruchterman and Reingold**. Connected vertices pull each other close, and all vertices push each other away.
- `layout_with_gem()` Places vertices by the GEM force-directed algorithm.
- `layout_with_kk()` The **Kamada-Kawai** algorithm, based on a physical model of springs.
- `layout_with_sugiyama()` The **Sugiyama** algorithm for layered directed acyclic graphs. It minimises edge crossings.

Readers are advised to try these layouts to get a fair understanding of them. Let us see a few useful ones on the `karate` data.

- Circular layout, @fig-gra28 left.
- Fruchterman-Reingold, @fig-gra28 right.
- Kamada-Kawai, @fig-gra29 left.
- Sugiyama, @fig-gra29 right.
- Tree layouts (two representations), @fig-gra30.

Some of these layouts start from random positions. So we set a seed before each plot, to get the same picture every time.

```{r fig-gra28, message=FALSE, warning=FALSE, fig.align='center', fig.show='hold', fig.cap="Two layouts of the karate club network", out.height="34%", out.width="48%"}
igraph_options(vertex.size = 18)
par(mfrow = c(1, 2))
set.seed(123)

plot(karate, layout = layout_in_circle)
title("Circular layout")

plot(karate, layout = layout_with_fr)
title("Fruchterman-Reingold")
```

```{r fig-gra29, message=FALSE, warning=FALSE, fig.align='center', fig.show='hold', fig.cap="Two more layouts of the karate club network", out.height="34%", out.width="48%"}
par(mfrow = c(1, 2))
set.seed(123)

plot(karate, layout = layout_with_kk)
title("Kamada-Kawai")

plot(karate, layout = layout_with_sugiyama)
title("Sugiyama")
```

```{r fig-gra30, message=FALSE, warning=FALSE, fig.align='center', fig.show='hold', fig.cap="Two tree layouts of the karate club network", out.height="34%", out.width="48%"}
par(mfrow = c(1, 2))

plot(karate, layout = layout_as_tree)
title("Default tree layout")

plot(karate, layout = layout_as_tree(karate, circular = TRUE))
title("Circular tree")
```

For a social network like this, the force-directed layouts (Fruchterman-Reingold and Kamada-Kawai) show the structure best. The two groups around Mr Hi and John A can be seen clearly. The circular and tree layouts hide it.

### Showing vertex and edge properties

Besides the layout, we can show vertex and edge attributes, like their categories or amounts, through colour, size, width and shape. Our `karate` graph is already coloured. How? The vertices of a graph are coloured by its vertex attribute `color`. Let us see it.

```{r gra31}
V(karate)$color
```

It contains the integers 1 and 2, which `plot()` maps to its default colours. Let us replace them with colours of our choice.

```{r fig-gra32, message=FALSE, warning=FALSE, fig.align='center', fig.show='hold', fig.cap="Vertices coloured by our choice of colours", out.height="34%", out.width="98%"}
V(karate)$color <- c("red", "dodgerblue")[V(karate)$color]
head(V(karate)$color)
set.seed(123)
plot(karate, layout = layout_nicely)
```

Similarly, we can use the following attributes to display properties of vertices and edges in an `igraph` plot.

+ **Vertex properties**
  - `size`. Size of the vertex. Default is 15.
  - `color`. Fill colour of the vertex.
  - `frame.color`. Border colour of the vertex.
  - `shape`. Shape of the vertex, like `"circle"`, `"square"`, `"rectangle"` or `"none"`.
  - `label`. A character vector used to label the vertices.
  - `label.family`. Font family of the label. Default is `serif`.
  - `label.font`. Font of the label. `1` means plain (default), `2` bold, `3` italic, `4` bold italic and `5` symbol.
  - `label.cex`. Font size of the label.

+ **Edge properties**
  - `color`. Colour of the edge.
  - `width`. Edge width.
  - `lty`. Line type for edges (`0` or `"blank"`; `1` or `"solid"`; `2` or `"dashed"`; `3` or `"dotted"`; `4` or `"dotdash"`; `6` or `"twodash"`).
  - `label`. Label of the edge (with `label.family`, `label.font` and `label.cex` for font family, font type and font size respectively).
  - `curved`. Edge curvature, from `0` to `1`.
  - `arrow.size`. Arrow size (default is 1), for directed graphs.
  - `arrow.width`. Arrow width, default is `1`.
  - `arrow.mode`. Arrow mode (`0` means no arrow, `1` backward, `2` forward and `3` both).

Instead of storing these as attributes in the graph, we can also give them as arguments to `plot()`, by adding `vertex.` or `edge.` before the property name, like `vertex.size = 18`.

In audit networks, we may size vendors by the amount paid, colour them by type, and draw thicker edges for larger payments. Let us see some of these properties through the following example.

```{r fig-gra33, message=FALSE, warning=FALSE, fig.align='center', fig.show='hold', fig.cap="Vertex shapes, labels and edge widths", out.width="98%"}
# set reproducible layout
set.seed(123)
l_s <- layout_nicely(karate)

# Add vertex labels
V(karate)$label <- stringr::str_remove(V(karate)$name, "Actor ")

# Change shapes of two prominent actors
V(karate)$shape <- ifelse(V(karate)$name %in% c("Mr Hi", "John A"), "rectangle", "circle")

# Change edge width as per "weight"
E(karate)$width <- E(karate)$weight

plot(karate, layout = l_s, edge.label.cex = 0.7)
```

## Our audit data, linked contractors {#sec-graph-contractors}

The karate club taught us how to draw a network. For the rest of the chapter, let us work on audit data. We audit the works contracts of an imaginary Public Works Division for the financial year 2024-25. The Division has 60 registered contractors, `C01` to `C60`. From the registration papers, we have listed every pair of contractors who share some detail, a partner, an office address, a mobile number, an email ID or a bank account. Each such pair is one row of a table `links`. Its column `weight` is the number of details the two firms share.

Most firms share nothing with anyone. A few pairs share one detail for a genuine reason, like two family firms or two firms in the same market building. But we have planted two hidden groups.

- **Group A.** Eight firms around `C07`, which share partners, addresses and accounts among themselves.
- **Group B.** Six firms around `C33`, linked in the same way.
- **A bridge.** Firm `C21` shares one detail with `C07` and one with `C33`, and nothing else.

The code to create the data is folded.

```{r}
#| code-fold: true
#| code-summary: "Code to create the contractor master and the links"
set.seed(2024)
prefix <- c("Shree Balaji", "Maa Durga", "Om Sai", "Jai Ambe", "Ganpati", "Kaveri",
            "Narmada", "Shiv Shakti", "Laxmi", "Sai Krupa", "Annapurna", "Bharat",
            "Vishwakarma", "Gayatri", "Sri Venkatesh")
suffix <- c("Constructions", "Builders", "Infra", "Enterprises", "Engineers", "Contractors")

# Contractor master
firms <- tibble(
  firm_id   = sprintf("C%02d", 1:60),
  firm_name = sample(as.vector(outer(prefix, suffix, paste)), 60),
  class     = sample(c("Class I", "Class II", "Class III"), 60, TRUE, prob = c(0.2, 0.3, 0.5))
)

# A link between two firms, sharing k details
details <- c("partner", "address", "mobile", "bank account", "email")
link <- function(a, b, k) {
  tibble(firm_a = a, firm_b = b,
         shared = map_chr(k, \(i) paste(sort(sample(details, i)), collapse = ", ")),
         weight = k)
}

# A hidden group. The first firm is linked to all, the others to some of each other
group_links <- function(g, p) {
  hub <- link(g[1], g[-1], sample(1:3, length(g) - 1, TRUE))
  pairs <- t(combn(g[-1], 2))
  keep <- runif(nrow(pairs)) < p
  bind_rows(hub, link(pairs[keep, 1], pairs[keep, 2], sample(1:2, sum(keep), TRUE)))
}

# Planted. Two hidden groups and a bridge between them
group_a <- c("C07", "C12", "C15", "C19", "C26", "C40", "C44", "C52")
group_b <- c("C33", "C03", "C09", "C28", "C47", "C55")
bridge  <- "C21"

# Genuine. A few pairs of other firms share one detail
others  <- setdiff(firms$firm_id, c(group_a, group_b, bridge))
genuine <- t(replicate(22, sample(others, 2)))

links <- bind_rows(
  group_links(group_a, 0.35),
  group_links(group_b, 0.35),
  link(bridge, c("C07", "C33"), c(1, 1)),
  link(genuine[, 1], genuine[, 2], rep(1, nrow(genuine)))
) |>
  distinct(firm_a, firm_b, .keep_all = TRUE)
```

Let us look at the two tables.

```{r}
glimpse(firms)
head(links)
```

Let us build the graph. Sharing a detail has no direction, so the graph is undirected. We give the contractor master as `vertices`, so that firms with no links stay in the graph, and the firm name and class become vertex attributes.

```{r}
cg <- graph_from_data_frame(links, directed = FALSE, vertices = firms)
cg
```

The graph has `r vcount(cg)` vertices, one for each registered firm, and `r ecount(cg)` edges, one for each linked pair.

## Interactive networks with visNetwork

Static plots are fine for small graphs and for printed reports. But while *exploring* a network, we want to zoom in, drag vertices around, and click on a vertex to see its details. The package `visNetwork` does exactly this. It draws interactive networks in a web browser, in RStudio's viewer, and in HTML documents like this book.

`visNetwork()` takes two data frames.

- `nodes` must have a column `id`. Optional columns control the look, like `label` (text shown), `group` (for colouring), `title` (tooltip shown on hover), `value` (size), `shape` and `color`.
- `edges` must have the columns `from` and `to`. Optional columns include `value` or `width` (thickness), `label`, `title`, `arrows` and `color`.

Let us draw the contractor network interactively. We colour the firms by their class of registration, and size them by their number of links. The tooltip of a firm shows its name, and the tooltip of an edge shows what the two firms share. Try hovering over a firm, clicking one to highlight its neighbours, choosing a firm from the drop-down list, and dragging a firm around.

```{r}
#| label: fig-vis1
#| fig-cap: "Interactive network of the Division's contractors"
library(visNetwork)

cg_nodes <- tibble(
  id    = V(cg)$name,
  label = V(cg)$name,
  group = V(cg)$class,
  value = degree(cg) + 1,
  title = paste0("<b>", V(cg)$name, "</b><br>", V(cg)$firm_name,
                 "<br>Links: ", degree(cg))
)

cg_edges <- links |>
  transmute(from = firm_a, to = firm_b, value = weight,
            title = paste("Shared:", shared))

visNetwork(cg_nodes, cg_edges, width = "100%", height = "450px") |>
  visIgraphLayout(layout = "layout_with_fr", randomSeed = 11) |>
  visOptions(highlightNearest = TRUE, nodesIdSelection = TRUE) |>
  visLegend()
```

Even at a glance, two tight clusters of firms stand out from the scattered pairs. We will find them with R shortly.

A few more useful functions of `visNetwork` are

- `visIgraph(g)` draws an `igraph` object directly, using its attributes.
- `toVisNetworkData(g)` converts an `igraph` object into the `nodes` and `edges` data frames which `visNetwork()` needs. We can then add or change columns before drawing.
- `visIgraphLayout()` calculates the layout in R with `igraph`, instead of in the browser. The network then appears instantly and does not keep moving. This is very useful for bigger networks. Its argument `randomSeed` fixes the layout, like `set.seed()`.
- `visEdges(arrows = "to")` draws arrows on directed edges.
- `visSave(network, file = "network.html")` saves the interactive network as a standalone HTML file, which can be shared with colleagues who do not use R.

::: callout-tip
## Keep interactive networks small

The browser has to draw every vertex and edge of a `visNetwork`. A few hundred vertices are fine. With many thousands, the page becomes slow, and the picture is a hairball that nobody can read anyway. So first reduce the network in R, as explained in the next section and in @sec-netaudit. Then draw only the part that matters.
:::

## Parts of a graph, and adding or deleting vertices and edges

Real networks are big. Usually we are interested in only a part of them. Let us see how to take out parts of a graph, and how to add or delete vertices and edges.

**Selecting vertices and edges.** We can select vertices and edges by a condition on their attributes, inside `V()` and `E()`.

```{r}
# Firms linked with more than 4 other firms
V(cg)[degree(cg) > 4]

# Pairs of firms sharing 3 or more details
E(cg)[weight >= 3]
```

Only `r sum(degree(cg) > 4)` firms are linked with more than four others. And `r sum(E(cg)$weight >= 3)` pairs share three or more details. Sharing so much is hard to explain for two independent firms.

**Neighbours.** `neighbors()` gives the vertices directly connected to a vertex.

```{r}
neighbors(cg, "C07")
```

**Subgraphs.** `induced_subgraph()` keeps the chosen vertices and all the edges *between* them. `subgraph_from_edges()` keeps the chosen edges and their vertices. Let us keep only the firms which share something with at least one other firm.

```{r}
linked <- induced_subgraph(cg, V(cg)[degree(cg) > 0])
linked
```

`r vcount(cg) - vcount(linked)` firms share nothing with anyone, and have dropped out.

**Ego networks.** The *ego network* of a vertex is the vertex itself, its neighbours, and the edges between them. The argument `order` sets how many steps away we go. `make_ego_graph()` returns a list of such graphs, one for each vertex we ask for. In audit, the ego network of a doubtful entity is often the first thing we look at. Let us look at firm `C21`, up to two steps away (@fig-ego).

```{r fig-ego, fig.cap="Ego network of firm C21, up to 2 steps", fig.align='center', out.width="80%"}
ego21 <- make_ego_graph(cg, order = 2, nodes = "C21")[[1]]
set.seed(123)
plot(ego21, layout = layout_with_fr, vertex.size = 20, vertex.color = "orange")
```

`C21` has only two links. But two steps away, it reaches `r vcount(ego21) - 1` firms, on both sides.

**Deleting.** `delete_vertices()` and `delete_edges()` remove vertices and edges. When a vertex is deleted, all its edges go with it. Let us remove `C21`, and see whether `C07` and `C33` are still connected.

```{r}
without_c21 <- delete_vertices(cg, "C21")
components(cg)$membership[c("C07", "C33")]
components(without_c21)$membership[c("C07", "C33")]
```

With `C21`, the two firms fall in the same connected piece (component) of the network. Without it, they fall in different pieces. So `C21` is the only link between the two groups. We will meet this idea again as *betweenness*.

**Adding.** `add_vertices()` and `add_edges()` add new vertices and edges. A convenient alternative is the `+` operator with `vertices()` and `edges()`. Suppose a new firm, `C61`, registers with the Division, and we find that it gives the same mobile number as `C07`.

```{r}
cg_new <- cg + vertices("C61") + edges("C61", "C07", shared = "mobile", weight = 1)
neighbors(cg_new, "C61")
```

## Describing a network

Once we have a network, we want to describe it. How big and how dense is it? Which vertices are the most important? Are there groups within it? Let us answer these questions for the contractors.

### Measures for the whole network

```{r}
# Number of vertices and edges
vcount(cg)
ecount(cg)

# Density, the share of all possible edges which exist
edge_density(cg)

# Connected pieces, and their sizes
components(cg)$no
table(components(cg)$csize)

# Diameter, the longest of all shortest paths (counting steps)
diameter(cg, weights = NA)

# Average number of steps between two firms, where a path exists
mean_distance(cg, weights = NA)

# Transitivity, the chance that two firms linked to a firm are linked to each other too
transitivity(cg)
```

```{r}
#| include: false
comp <- components(cg)
```

Only `r round(edge_density(cg) * 100, 1)`% of all possible pairs of firms are linked. The network falls into `r comp$no` pieces. Of these, `r sum(comp$csize == 1)` are single firms with no links. The biggest piece has `r max(comp$csize)` firms. Within a piece, any firm can reach any other in at most `r diameter(cg, weights = NA)` steps. A transitivity of `r round(transitivity(cg), 2)` is high for such a sparse network. It means that linked firms tend to come in closely knit groups.

::: callout-note
## Weights and distances

`igraph` treats the `weight` attribute as a *distance* in path-based measures like `diameter()`, `distances()`, `betweenness()` and `closeness()`. In our data, a larger weight means more shared details, that is, a *stronger* link. A stronger link should mean a *shorter* distance, not a longer one. So here we use `weights = NA` to ignore the weights and simply count steps. Always check what the weights in your data mean before using them in such measures.
:::

### Paths

`shortest_paths()` gives the shortest path between two vertices, and `distances()` gives the number of steps between pairs of vertices. In audit, the path tells us *how* two entities are connected. Let us see how firm `C40` of one group is linked with firm `C55` of the other.

```{r}
shortest_paths(cg, from = "C40", to = "C55", weights = NA)$vpath
distances(cg, v = c("C40", "C07"), to = c("C33", "C55"), weights = NA)
```

The path runs through `C07`, `C21` and `C33`. Firms with no path between them are shown at a distance of `Inf`, that is, infinite.

### Importance of vertices (centrality)

Which firms are the most important? There are several ways to measure this, called *centrality* measures.

- **Degree** is the number of edges of a vertex. For a directed graph, *in-degree* and *out-degree* count incoming and outgoing edges separately. A bank account receiving money from 40 beneficiaries has a high in-degree.
- **Strength** is the sum of the weights of the edges of a vertex, for example the total amount paid or received. Here, it is the total number of details a firm shares.
- **Betweenness** counts how often a vertex lies on the shortest paths between other vertices. A vertex with high betweenness is a *bridge* or a middleman between groups.
- **Closeness** measures how close a vertex is to all others. A vertex with high closeness can reach everyone quickly.
- **Eigenvector centrality** and **PageRank** give high scores to vertices connected to other important vertices. Being connected to the boss matters more than being connected to many juniors.

Let us calculate these for the contractors and look at the top firms.

```{r}
centrality <- tibble(
  firm        = V(cg)$name,
  degree      = degree(cg),
  strength    = strength(cg),
  betweenness = round(betweenness(cg, weights = NA), 1),
  closeness   = round(closeness(cg, weights = NA), 3),
  eigen       = round(eigen_centrality(cg)$vector, 2)
) |>
  arrange(desc(betweenness))

head(centrality, 8)
```

```{r}
#| include: false
c21 <- centrality |> filter(firm == "C21")
```

`C07` and `C33`, the centres of the two groups, have the most links and the highest betweenness. That is what we would expect. Now look at `C21`. It has only `r c21$degree` links, as few as many ordinary firms. Yet its betweenness is `r c21$betweenness`, close to that of the two group leaders. Every path between the two groups passes through it. A degree count alone would never have picked it out.

Look also at the closeness of firms like `C02` and `C30`. It is higher than that of `C07`. This is because `igraph` measures closeness only over the firms a vertex can reach. In a network of many separate pieces, firms in a small piece look "close" to everything. So use closeness within one connected piece, not across the whole network.

::: callout-tip
## Audit tip

Sort by betweenness, not only by degree. A firm, account or person with few links but high betweenness is a go-between. In real cases, this is often a common accountant, an agent, or a relative who connects firms that claim to be independent. Find out who is behind that vertex.
:::

### Communities

A *community* is a group of vertices which are densely connected with each other, and only loosely connected with the rest. Finding communities is called *community detection* or *clustering*. `igraph` has several algorithms for it. The Louvain method, `cluster_louvain()`, is fast and works well for most networks. It starts from a random order, so we set a seed.

```{r}
set.seed(42)
cg_comm <- cluster_louvain(cg, weights = NA)
sizes(cg_comm)
```

```{r}
#| include: false
n_single <- sum(sizes(cg_comm) == 1)
```

Louvain has found `r length(cg_comm)` communities. Of these, `r n_single` are single firms with no links. Let us see the communities of `C07` and `C33`.

```{r}
cg_comm[[membership(cg_comm)["C07"]]]
cg_comm[[membership(cg_comm)["C33"]]]
```

```{r}
#| include: false
comm_a <- cg_comm[[membership(cg_comm)["C07"]]]
comm_b <- cg_comm[[membership(cg_comm)["C33"]]]
```

The community of `C07` has all `r sum(group_a %in% comm_a)` firms of Group A, along with the bridge firm `C21`. The community of `C33` has all `r sum(group_b %in% comm_b)` firms of Group B. Just from the pattern of shared details, the algorithm has found both hidden groups. @fig-comm shows the linked firms, coloured by community.

```{r}
#| label: fig-comm
#| fig-cap: "Linked contractors coloured by the community found by the Louvain method"
comm_nodes <- tibble(
  id    = V(linked)$name,
  label = V(linked)$name,
  group = paste("Community", membership(cg_comm)[V(linked)$name]),
  title = V(linked)$firm_name
)

visNetwork(comm_nodes, cg_edges, width = "100%", height = "450px") |>
  visIgraphLayout(layout = "layout_with_fr", randomSeed = 11) |>
  visOptions(highlightNearest = TRUE)
```

Communities point to groups like bidders who bid together, firms which trade mostly among themselves, or beneficiaries who share agents. Let us now use them on the Division's tenders.

## Audit use case, linked firms bidding against each other {#sec-graph-tenders}

A tender is competitive only if the bidders are independent. If every bid in a tender comes from one hidden group, the group can decide the winner and the rate among themselves. Let us check the Division's 40 tenders of 2024-25. The table `bids` has one row for each bid, and the table `tenders` gives the estimated cost of each work. We have planted seven tenders in which all the bidders come from one of the two hidden groups.

```{r}
#| code-fold: true
#| code-summary: "Code to create the tenders and bids"
set.seed(7)
n_t <- 40
bids <- map_dfr(1:n_t, function(t) {
  b <- sample(firms$firm_id, sample(3:5, 1))
  # Planted. All bidders from Group A, or all from Group B
  if (t %in% c(4, 11, 18, 26, 33)) b <- sample(group_a, sample(3:4, 1))
  if (t %in% c(8, 29)) b <- sample(group_b, 3)
  tibble(tender_no = sprintf("PWD/%02d/2024-25", t), firm_id = b)
})
tenders <- tibble(
  tender_no = sprintf("PWD/%02d/2024-25", 1:n_t),
  est_cost  = round(rlnorm(n_t, log(60e5), 0.6), -3)
)
```

```{r}
head(bids)
head(tenders)
```

We attach the community of each bidder, and count, for each tender, the bidders and the different communities they come from.

```{r}
firm_comm <- tibble(firm_id   = V(cg)$name,
                    community = as.integer(membership(cg_comm)))

tender_check <- bids |>
  left_join(firm_comm, by = "firm_id") |>
  group_by(tender_no) |>
  summarise(bidders     = n(),
            communities = n_distinct(community),
            firms       = paste(sort(firm_id), collapse = ", ")) |>
  left_join(tenders, by = "tender_no")

flagged <- tender_check |>
  filter(communities == 1)
flagged
```

```{r}
#| include: false
partly <- tender_check |> filter(communities > 1, communities < bidders)
```

In `r nrow(flagged)` tenders, every bidder comes from the same community. These are exactly the seven tenders we planted. Their estimated cost adds up to ₹`r round(sum(flagged$est_cost) / 1e7, 2)` crore. In these tenders, the "competition" was among firms that share partners, addresses or bank accounts. In `r nrow(partly)` more tenders, two or more of the bidders come from the same community, but at least one bidder is from outside. These are weaker leads, and may be looked at after the first seven.

What should the auditor do next? For each flagged tender, the auditor would

1. pull the tender file, and see who won, at what rate, and how far the losing bids were above the winning bid;
2. ask the Division for the registration papers of the linked firms, and confirm the shared partner, address or bank account from the documents;
3. check whether the Division had any declaration from the bidders that they were not related;
4. record the linked firms, the tenders and the amounts in the working papers, and raise an audit query to the Division.

::: callout-warning
## Caution

A community is a group found by an algorithm, not a fact. Louvain may put a loosely linked firm into a group, as it did with `C21` here, and a different seed may move such a firm. So before raising a query, look at the actual links between the bidders of a tender, with `induced_subgraph()` or `shortest_paths()`. Report what the firms share, not the community number.
:::

## A suggested workflow

1.  **Build an edge list.** From the payment register, the beneficiary list or the registration papers, make a data frame in which each row is one link. Examples are a beneficiary and a bank account, a payment from one firm to another, or two firms sharing a partner. Keep the amount, the date or the shared detail as further columns, so they become edge attributes.
2.  **Create the graph.** Use `graph_from_data_frame()`. Keep it directed for payments, and use `directed = FALSE` for shared details like a common account. Give a `vertices` table to keep entities with no links, and their attributes.
3.  **Check the graph object.** Print it and confirm the number of vertices and edges, and the attributes. Get the data back with `igraph::as_data_frame()` whenever you need to inspect it.
4.  **Describe the whole network.** Use `vcount()`, `ecount()`, `edge_density()` and `components()` to see how big the network is and whether it falls into separate pieces.
5.  **Find the important entities.** Calculate `degree()`, `strength()` and `betweenness()`. A bank account receiving money from many beneficiaries, or a firm acting as a bridge between groups, deserves a look.
6.  **Find the groups.** Set a seed and run `cluster_louvain()` to find communities, such as firms that share partners or bidders who bid together.
7.  **Zoom into doubtful cases.** Take the ego network of a doubtful entity with `make_ego_graph()`, or a part of the graph with `induced_subgraph()`. Use `shortest_paths()` to see how two entities are linked.
8.  **Join the findings back to the transactions.** Attach the community or centrality of each entity to the tenders, bills or payments, and value the doubtful ones in rupees.
9.  **Show only what matters.** Draw the reduced network with `plot()` or `visNetwork()`. Set a seed before any random layout. Save the interactive version with `visSave()` to share with colleagues who do not use R.

::: callout-important
## Remember

A connection in a network is a lead, not a finding. Two firms may share an address because they sit in the same market building, so examine the records before reporting. Set a seed before any layout or community detection that has randomness in it, so that the result can be reproduced. Also check what the weights in your data mean, because `igraph` treats `weight` as a distance in measures like `betweenness()`.
:::

## Where next

In @sec-netaudit, we apply these tools to more audit problems. We find one person behind many registrations of a scheme, a cartel of bidders, circular trading between firms, and vendors linked with employees.

## Summary of functions used

| Function | Package | What it does |
|-----------------------|----------------|--------------------------------|
| `graph_from_data_frame()` | igraph | Creates a graph from a data frame of edges (and vertices) |
| `graph_from_edgelist()`, `graph_from_adjacency_matrix()` | igraph | Create a graph from an edge matrix or adjacency matrix |
| `as_undirected()`, `as_directed()` | igraph | Convert between directed and undirected graphs |
| `igraph::as_data_frame()`, `as_adjacency_matrix()` | igraph | Get the data back from a graph |
| `V()`, `E()` | igraph | Vertices and edges of a graph, with their attributes via `$` |
| `plot()`, `layout_*()` | igraph | Static plot of a graph, with a chosen layout |
| `ggraph()`, `geom_edge_fan()`, `geom_node_point()`, `geom_node_text()` | ggraph | Plots a graph in the style of `ggplot2` |
| `set.seed()` | base | Fixes random layouts and community detection, so that results can be reproduced |
| `visNetwork()`, `visOptions()`, `visLegend()`, `visIgraphLayout()`, `visSave()` | visNetwork | Interactive network plots |
| `neighbors()`, `induced_subgraph()`, `subgraph_from_edges()`, `make_ego_graph()` | igraph | Parts of a graph around chosen vertices or edges |
| `delete_vertices()`, `delete_edges()`, `add_vertices()`, `add_edges()`, `vertices()`, `edges()` | igraph | Remove or add vertices and edges |
| `components()` | igraph | Splits a graph into connected pieces |
| `vcount()`, `ecount()`, `edge_density()`, `diameter()`, `mean_distance()`, `transitivity()` | igraph | Measures of the whole network |
| `shortest_paths()`, `distances()` | igraph | Paths and number of steps between vertices |
| `degree()`, `strength()`, `betweenness()`, `closeness()`, `eigen_centrality()` | igraph | Importance (centrality) of each vertex |
| `cluster_louvain()`, `sizes()`, `membership()` | igraph | Finds communities in a network, their sizes and members |
| `glimpse()`, `left_join()`, `n_distinct()` | dplyr | Look at the data, and attach communities to the bids |

: Functions used in this chapter {#tbl-graphfunctions}
