```{r}
#| include: false
source("_common.R")
```

# Readable Regular Expressions with rebus {#sec-rebus}

An audit team checks the vendor master of a department. Every PAN must have five letters, four digits and a letter. Every IFSC must have four letters, a zero and six letters or digits. Voucher dates must be written day first. In @sec-regex we wrote regular expressions for such checks. They work, but they are hard to read. A pattern like `^[A-Z]{5}[0-9]{4}[A-Z]$` means little to a reviewer, and often even to the person who wrote it, a few months later. Yet the reviewer must be sure that the check tests the right rule.

The package `rebus` solves this problem. It lets us build a regular expression from functions whose names say what they do. Instead of writing `[a-zA-Z0-9._%+-]+@[a-zA-Z0-9.-]+\.[a-zA-Z]{2,4}`, and later trying to decipher it, we can write this.

```
one_or_more(char_class(ASCII_ALNUM %R% "._%+-")) %R%
  "@" %R%
  one_or_more(char_class(ASCII_ALNUM %R% ".-")) %R%
  DOT %R%
  ascii_alpha(2, 4)
```

Many of us could have guessed by now that both are patterns to detect email addresses. `rebus` contains functions like `char_class()` and `one_or_more()`. The result is an ordinary regular expression, which can be used in any `stringr` function.

In this chapter we will learn the `rebus` functions in the same order as the regular expressions of @sec-regex. Our examples come from audit records, such as payee names, voucher narrations, bill numbers and amounts. At the end, we will use `rebus` to write readable checks for PAN, IFSC and GSTIN, check a vendor master with them, and check dates in a voucher register.

Let us first load the packages. Along with `rebus`, we load `stringr`, so that we can use its function `str_view()` to see the matches.

::: callout-warning
## Caution

`rebus` also has a function named `regex()`, which conflicts with `stringr::regex()`. To avoid confusion, we use the package `conflicted` to state which one we prefer.
:::

```{r}
#| message: false
#| warning: false
library(rebus)
library(stringr)
library(conflicted)
conflicts_prefer(stringr::regex, rebus::or)
```

## Operators for concatenation and alternation

To join two regular expressions one after the other, we may use either of the operators `%R%` or `%c%`. Like a pipe, the operator joins its left and right sides into one regular expression.

For alternation, we may use either the operator `%|%` or the function `or()`. Both separate two regular expressions with a `|`, which means "either this or that".

A payee name is often typed in more than one way. Let us find a trader whose name is spelt *Traders* in some vouchers and *Tradars* in others.

```{r}
payees <- c("Gupta Traders", "Gupta Tradars", "Gupta Trading Co", "Kumar Cars")
str_view(payees, "Trad" %R% or("e", "a") %R% "rs")
# Note operator precedence
# This actually means either "Trade" or "ars"
str_view(payees, "Trad" %R% "e" %|% "a" %R% "rs")
```

Printing a `rebus` expression shows the regular expression it builds. This is the best way to check that it means what we intended.

```{r}
"Trad" %R% or("e", "a") %R% "rs"
"Trad" %R% "e" %|% "a" %R% "rs"
```

The first is `Trad(?:e|a)rs`. It means "Trad, then e or a, then rs", as intended. The second is `Trade|ars`.

::: callout-warning
## Caution

The second pattern matched both spellings of Gupta Traders, but for the wrong reasons. It found *Trade* in one and *ars* in the other. It also matched *Kumar Cars*, which is not a trader at all. A pattern can give the right answer on a few test values and still be wrong. Use `or()` to keep alternatives together.
:::

## Literal characters, special characters and case sensitivity

Literal characters are written in the pattern just as they are, as we learnt earlier. Literals are case sensitive. So if we want a match that ignores case, we wrap the pattern in the function `case_insensitive()`.

```{r}
narration <- "Paid to M/s Gupta Traders against Bill No. 45 dated 12.04.2024."
# Match the literal "gupta"
str_view(narration, "gupta", match = NA)
str_view(narration, case_insensitive("gupta"), match = NA)
```

Some characters have a special meaning in regular expressions. To match them as ordinary characters, `rebus` has constants. In `rebus`, the constants are usually written in `UPPER CASE`, and equivalent functions are written in `lower case`. The following special character constants are available.

```{r}
BACKSLASH
CARET
DOLLAR
DOT
PIPE
QUESTION
STAR
PLUS
OPEN_PAREN
CLOSE_PAREN
OPEN_BRACKET
CLOSE_BRACKET
OPEN_BRACE
```

Let us find every dot in the narration.

```{r}
str_view(narration, DOT)
```

## Metacharacters

### Character classes

To match any one of a group of characters, we put them in a class with the function `char_class()`. Bill numbers are often written with different separators. Let us find the separators in a few bill numbers.

```{r}
bill_nos <- c("BL/2024/001", "BL-2024-002", "BL.2024.003", "BL 2024 004")
# matches a slash, a dot, a space or a hyphen. A hyphen goes last, or it would mean a range
str_view(bill_nos, char_class("/. -"))
```

To **match a range of characters or numbers**, we separate the two ends with a hyphen inside the class. So, `char_class("A-D")` matches any one of `A`, `B`, `C` or `D`. Suppose cheque books of series A to D were issued to an office, and the cheque numbers start with the series letter.

```{r}
cheques <- c("A 500301", "b 500302", "E 500303", "D 500304")
# Match a series letter from A to D, in upper case
str_view(cheques, char_class("A-D"), match = NA)
# The same, irrespective of case
str_view(cheques, case_insensitive(char_class("A-D")), match = NA)
```

The first pattern misses `b`, because it is in lower case. The cheque of series `E` matches neither, so it is not from the books issued to the office.

For negated character classes, there is an equally clear function, `negated_char_class()`. It matches any character *not* in the class. Amounts typed into a spreadsheet often carry extra characters. Let us find every character which is not a digit.

```{r}
amounts_text <- c("25000", "25,000", "Rs 25000", "25000/-")
# Match everything except digits
str_view(amounts_text, negated_char_class("0-9"), match = NA)
```

These are the characters to remove before the amounts can be converted to numbers.

### Built-in character classes

We can also use the **ready-made character classes** of `rebus`, listed below.

```{r}
ALPHA
ALNUM
BLANK
DIGIT
LOWER
PRINT
PUNCT
SPACE
UPPER
HEX_DIGIT
ANY_CHAR
GRAPHEME
NEWLINE
DGT
WRD
SPC
NOT_DGT
NOT_WRD
NOT_SPC  # Equivalent to "\\S"
ASCII_DIGIT
ASCII_LOWER
ASCII_UPPER
ASCII_ALPHA
ASCII_ALNUM
```

Let us use two of them on our narration.

```{r}
# Match TAB or SPACE characters
str_view(narration, BLANK)
# Match all UPPER CASE characters
str_view(narration, UPPER)
```

Besides these constants, there are `lower case` functions for the same character classes. They are more useful for building patterns, because they also take a repetition.

```
alnum(lo, hi, char_class = TRUE)
alpha(lo, hi, char_class = TRUE)
blank(lo, hi, char_class = TRUE)
cntrl(lo, hi, char_class = TRUE)
digit(lo, hi, char_class = TRUE)
graph(lo, hi, char_class = TRUE)
lower(lo, hi, char_class = TRUE)
printable(lo, hi, char_class = TRUE)
punct(lo, hi, char_class = TRUE)
space(lo, hi, char_class = TRUE)
upper(lo, hi, char_class = TRUE)
hex_digit(lo, hi, char_class = TRUE)
any_char(lo, hi)
grapheme(lo, hi)
newline(lo, hi)
dgt(lo, hi, char_class = FALSE)
wrd(lo, hi, char_class = FALSE)
spc(lo, hi, char_class = FALSE)
not_dgt(lo, hi, char_class = FALSE)
not_wrd(lo, hi, char_class = FALSE)
not_spc(lo, hi, char_class = FALSE)
ascii_digit(lo, hi, char_class = TRUE)
ascii_lower(lo, hi, char_class = TRUE)
ascii_upper(lo, hi, char_class = TRUE)
ascii_alpha(lo, hi, char_class = TRUE)
ascii_alnum(lo, hi, char_class = TRUE)
char_range(lo, hi, char_class = lo < hi)
number_range(lo, hi, allow_leading_zeroes = FALSE, capture = FALSE)
```

In these functions, `lo` and `hi` take positive whole numbers. They give the smallest and largest number of repetitions. The argument `char_class` takes `TRUE` or `FALSE`. Let us try them on the address line of a vendor.

```{r}
address <- "Shop 12, Station Road, Kota  PIN 324001"
# one to three digits at a time
str_view(address, digit(1, 3))
# exactly six digits, the PIN code
str_view(address, digit(6))
# five characters which are not digits
str_view(address, not_dgt(5))
# two spaces together
str_view(address, space(2), match = NA)
```

The last pattern finds the double space before `PIN`. Such stray spaces are a common reason why two copies of the same address fail to match.

### Word boundaries

To match a word boundary, or its negation, we have `BOUNDARY` and `NOT_BOUNDARY` in `rebus`. The function `whole_word(x)`, on the other hand, wraps the pattern in boundaries, so that it matches only a whole word.

Suppose we want to find the vouchers paid in cash, from their narrations.

```{r}
narrations <- c("Paid in cash to the contractor", "Cashier advance for petty expenses",
                "Encashment of leave", "Cash withdrawn for wages")
# Note the match inside "Encashment"
str_view(narrations, "cash", match = NA)
# Only the whole word, in lower case
str_view(narrations, whole_word("cash"), match = NA)
str_view(narrations, case_insensitive(whole_word("cash")), match = NA)
# Words that begin with "cash", in any case
str_view(narrations, BOUNDARY %R% case_insensitive("cash"), match = NA)
# "cash" inside a word
str_view(narrations, NOT_BOUNDARY %R% "cash", match = NA)
```

The plain pattern `"cash"` picks up *Encashment of leave*, which is not a cash payment at all. `case_insensitive(whole_word("cash"))` finds exactly the two cash payments.

## Quantifiers

We learnt the following quantifiers in regular expressions.

- `+` **1 or more** occurrences
- `*` **0 or more**
- `?` **0 or 1**
- `{}` specified numbers
  + `{n}` exactly n
  + `{n,}` n or more
  + `{n,m}` between n and m

`rebus` has a clearly named function for each of them.

- `one_or_more(x, char_class = NA)`
- `zero_or_more(x, char_class = NA)`
- `optional(x, char_class = NA)`
- `repeated(x, lo, hi, lazy = FALSE, char_class = NA)`
  + where `lo` and `hi` stand for `n` and `m`.

Each of them has an argument `char_class = NA`. It takes `TRUE` when `x` should be wrapped in a character class.

Round amounts are a well known sign of estimates or made-up figures. We may find amounts ending in three or more zeros with `repeated("0", 3, Inf) %R% END`. We may also pull out the quantities of material from bill descriptions.

```{r}
amounts_text <- c("125000", "48750", "200000", "3500")
# three or more zeros at the end
str_view(amounts_text, repeated("0", 3, Inf) %R% END, match = NA)

items <- c("10 bags cement", "5bags sand", "25 tins paint")
# a number followed by "bags", with an optional space
str_view(items, one_or_more(DIGIT) %R% optional(BLANK) %R% "bags", match = NA)
```

## Anchors

Anchors match a position in the string, not a character. In `rebus`, the constants `START` and `END` match the beginning and the end of the string.

```{r}
narrations2 <- c("Paid to Gupta Traders", "Advance paid to Gupta Traders")
str_view(narrations2, START %R% case_insensitive("paid"), match = NA)
```

Only the first narration matches. In the second, *paid* is not at the start. There is one more function, `exactly(x)`, which makes the pattern match the whole string, from start to end. It is a shortcut for `START %R% x %R% END`.

```{r}
modes <- c("Cash", "Cheque", "Cash/Cheque", "Petty Cash", "RTGS")
str_view(modes, exactly("Cash"), match = NA)
```

Only the value which is exactly `Cash` matches. This is how we must check codes and identifiers. A pattern without anchors would also pass `Cash/Cheque` and `Petty Cash`.

## Capture groups

We have seen that a capture group keeps a part of the match as a separate piece. This is useful when we want to extract or replace a specific part of a string. In regular expressions, capture groups are written in parentheses `()`. We also learnt that a captured piece can be referred back to with a *back reference*. `\1` refers to the match in the first pair of parentheses, `\2` in the second, and so on.

In `rebus`, there are two functions for this, `capture(x)` and `group(x)`. `capture()` makes a capturing group, `(x)`, whose match can be referred back to. `group()` makes a *non-capturing* group, `(?:x)`, which only keeps its contents together, as we need with alternations. It does not remember its match.

Let us remove the separators from our bill numbers, and pull out the mode of payment from a narration.

```{r}
## capture(x)
str_remove_all(bill_nos, capture(char_class("/. -")))

## group()
pay_modes <- group("cheque" %|% "RTGS" %|% "NEFT")
narration3 <- "Part payment by cheque, balance by RTGS after deduction of TDS."
str_extract_all(narration3, pattern = pay_modes)
```

Back references for replacement are written with the constants `REF1` to `REF9`. We can also use them inside the pattern. Account and mobile numbers with the same digit repeated four times in a row are often dummy entries. Let us find them.

```{r}
account_nos <- c("30114567892", "99999012345", "20188880123", "50100234567")
# a digit, followed by the same digit three more times
str_view(account_nos, capture(DGT) %R% REF1 %R% REF1 %R% REF1, match = NA)
```

As in @sec-regex, we can use capture groups to rearrange a value. Let us turn names written as *last name, first name* into *first name last name*.

```{r}
# We have names in last_name, first_name format
staff_names <- c("Sharma, Ramesh", "Verma, Sunita", "Iyer, Lakshmi")
# Pattern to capture the last name and the first name
pat <- capture(whole_word(one_or_more(ALPHA))) %R% "," %R% optional(SPACE) %R%
  capture(whole_word(one_or_more(ALPHA)))
repl <- REF2 %R% " " %R% REF1
# Using this regex, we can convert these to first_name last_name format
str_replace_all(staff_names, pat, repl)
```

## Lookarounds

As we learnt, **lookaheads** and **lookbehinds** are zero-width assertions. They match a pattern only if it is followed or preceded by another pattern. The pattern inside the lookahead or lookbehind is not part of the match.

`rebus` has clearly named functions for all four lookarounds.

```
lookahead(x)
negative_lookahead(x)
lookbehind(x)
negative_lookbehind(x)
```

Let us find invoice numbers where `INV` is *not* followed by a hyphen.

```{r}
invoices <- c("INV-101", "INV102", "INV")
# With lookahead
str_view(invoices, "INV" %R% negative_lookahead("-"), match = NA)
# Without lookahead. Notice the difference
str_view(invoices, "INV" %R% negated_char_class("-"), match = NA)
```

The negative lookahead finds both `INV102` and the bare `INV`. The negated class needs one more character after `INV`, so it misses the bare `INV`. It also takes that character into the match.

Now let us use all four lookarounds on the remarks column of an advance register. Advances which are not adjusted are a regular audit point.

```{r}
remarks <- c("advance adjusted", "advance not adjusted", "advance")

# Lookahead
str_view(remarks, "advance" %R% lookahead(optional(SPACE) %R% "adjusted"), match = NA)
# Note that "adjusted" is not included in the match

# Lookbehind
str_view(remarks, lookbehind("advance" %R% optional(SPACE)) %R% "adjusted", match = NA)

# Negative lookahead
str_view(remarks, "advance" %R% negative_lookahead(optional(SPACE) %R% "adjusted"), match = NA)

# Negative lookbehind
str_view(remarks, negative_lookbehind("advance" %R% optional(SPACE)) %R% "adjusted", match = NA)
```

The negative lookahead picks out exactly the two advances that are not shown as adjusted.

A lookahead is also handy for pulling out numbers that belong to a particular word.

```{r}
supply <- "Supplied 10 bags cement, 6 bags sand and 5 tins paint"

# lookahead to match the number of bags
bags_pattern <- one_or_more(DIGIT) %R% lookahead(optional(SPACE) %R% "bags")
str_view(string = supply, pattern = bags_pattern, match = NA)
```

## Ready patterns for dates and times

`rebus` has ready patterns for the parts of dates and times, and for whole dates and times.

```{r}
# Individual date-time components
DTSEP             # optional selected punctuation or space
CENTURY           # exactly two digits
YEAR              # one to four digits
YEAR2             # exactly two digits
YEAR4             # exactly four digits
MONTH             # number from 1 to 12, leading zero
WEEK_OF_YEAR      # number from 0 to 53, leading zero
DAY               # number from 1 to 31, leading zero
DAY_SINGLE        # leading space
HOUR24            # 24 hour clock, leading zero
HOUR12            # 12 hour clock, leading zero
HOUR24_SINGLE     # 24 hour clock, leading space
HOUR12_SINGLE     # 12 hour clock, leading space
MINUTE            # number from 0 to 59, leading zero
SECOND            # number from 0 to 61 (leap seconds), leading zero
FRACTIONAL_SECOND # a second optional decimal point and up to 6 digits
AM_PM             # AM or PM, any case
TIMEZONE_OFFSET   # optional plus or minus, then four digits
TIMEZONE          # Any value returned by OlsonNames()
# ISO 8601 formats
ISO_DATE          # %Y-%m-%d
ISO_TIME          # %H:%M:%S
ISO_DATETIME      # ISO_DATE followed by ISO_TIME, separated by space or "T".
# Compound forms, separated by DTSEP
YMD
YDM
MYD
MDY
DYM
DMY
HMS
HM
MS
```

Let us check some dates and times exported from a system in ISO format.

```{r}
# Some dates, valid and invalid
some_dates <- c("2000-13-01", "2025-08-09","2000-01-32", "2000-00-01", "2000-01-00", "2020-05-20")

str_view(some_dates, ISO_DATE)

# Similarly some time formats
some_times <- c("24:00:00", "23:60:59", "23:59:62", "23 59 59", "23:55:55", "00:00:00")
str_view(some_times, ISO_TIME)
```

Only the dates with a possible month and day match. Among the times, only `23:55:55` and `00:00:00` match. In `24:00:00`, `23:60:59` and `23:59:62`, the hour, the minute or the second is out of range. And `23 59 59` uses spaces in place of colons.

## Roman numerals

Roman numerals appear in government records more often than we may think. Schedules of an Act, classes of posts and parts of a report are often numbered this way. To match Roman numerals, `rebus` has a constant `ROMAN` and a function `roman(lo, hi)`.

```{r}
# Schedule numbers typed in a register, valid and invalid
schedule_nos <- c("VII", "IV", "XIL", "IIII", "XL", "MCD")
# Find valid Roman numerals
str_view(schedule_nos, exactly(roman()), match = NA)
```

`XIL` and `IIII` are not valid Roman numerals, so they do not match.

## Audit use case, readable validation patterns {#sec-rebus-audit}

In @sec-regex-validate we wrote patterns for Indian identifiers. Written with `rebus`, they can be read almost like the official description of the format. Let us build three of them, and print the regular expressions they produce.

```{r}
# PAN. Five letters, four digits, one letter
pan_rx  <- exactly(ascii_upper(5) %R% ascii_digit(4) %R% ascii_upper(1))

# IFSC. Four letters for the bank, a zero, six letters or digits for the branch
ifsc_rx <- exactly(ascii_upper(4) %R% "0" %R% repeated(char_class("A-Z0-9"), 6))

# GSTIN. State code, PAN, entity number, Z, check character
gstin_rx <- exactly(ascii_digit(2) %R%
                    ascii_upper(5) %R% ascii_digit(4) %R% ascii_upper(1) %R%
                    char_class("1-9A-Z") %R% "Z" %R% char_class("0-9A-Z"))

pan_rx
ifsc_rx
gstin_rx
```

These are exactly the patterns we wrote by hand in the previous chapter. Notice that for the IFSC we used `char_class("A-Z0-9")` and not `ascii_alnum()`, because `ascii_alnum()` would also allow lower case letters. Printing the pattern is how we catch such differences.

```{r}
str_detect(c("ABCPK1234D", "ABCP1234D"), pan_rx)
str_detect(c("SBIN0001234", "SBIN1001234"), ifsc_rx)
str_detect(c("27AAPFU0939F1ZV", "27AAPFU0939F1Z"), gstin_rx)
```

A pattern only checks the format. The check character of a GSTIN still needs the function `gstin_valid()` of @sec-checkdigits.

### Checking a vendor master

Let us now run the patterns over a small vendor master. A few errors have been planted in it.

```{r}
#| code-fold: true
#| code-summary: "Code to create the vendor master"
vendor_master <- tibble::tibble(
  vendor_id = sprintf("V%03d", 1:8),
  vendor    = c("Gupta Traders", "Shree Balaji Enterprises", "Narmada Builders",
                "Om Sai Caterers", "Kaveri Suppliers", "Laxmi Hardware",
                "Rajdhani Travels", "Yamuna Cleaning Services"),
  pan       = c("AAKPG4521K", "ABJPS7788Q", "AAFFN1203M", "AGSPO6634",
                "aakfk3390d", "AALPL5521H", "AACFR9087B", "AAAPY1122C"),
  ifsc      = c("SBIN0004521", "PUNB0112200", "CNRBO001234", "UBIN0556789",
                "BARB0KOTAXX", "SBIN0007788", "HDFC0001234", "SBIN1003344")
)
```

```{r}
#| message: false
library(dplyr)
conflicts_prefer(dplyr::filter)

vendor_master |>
  mutate(pan_ok  = str_detect(pan, pan_rx),
         ifsc_ok = str_detect(ifsc, ifsc_rx)) |>
  filter(!pan_ok | !ifsc_ok)
```

```{r}
#| include: false
fails <- vendor_master |>
  mutate(pan_ok = str_detect(pan, pan_rx), ifsc_ok = str_detect(ifsc, ifsc_rx)) |>
  dplyr::filter(!pan_ok | !ifsc_ok)
```

`r nrow(fails)` of the `r nrow(vendor_master)` vendors fail. Om Sai Caterers has a PAN with only nine characters. Kaveri Suppliers has a PAN in lower case. That one is probably correct, only untidily typed. Converting to upper case with `str_to_upper()` before the check, as advised in @sec-regex-validate, would pass it. Narmada Builders has the letter `O` in place of the zero in its IFSC. Yamuna Cleaning Services has `1` where the zero should be.

The auditor's next step is to ask the office for the PAN cards and cancelled cheques of these vendors. The office must also explain how payments were made to bank accounts with an invalid IFSC. Did they fail, or were they paid to some other account?

### Dates in vouchers

Dates in Indian records are usually written day first, like `15/04/2024`. The `rebus` constant `DMY` matches such dates. Let us check a few dates typed in a voucher register.

```{r}
#| message: false
library(lubridate)

voucher_dates <- c("15/04/2024", "31/02/2024", "29/02/2023", "29/02/2024",
                   "4/7/2024", "2024/07/04", "31-12-2024")

tibble::tibble(date_text   = voucher_dates,
               shape_ok    = str_detect(voucher_dates, exactly(DMY)),
               parsed_date = dmy(voucher_dates, quiet = TRUE))
```

The regex checks only the *shape* of a date. It accepts `31/02/2024` and `29/02/2023`, because 31 is a possible day and 02 a possible month. But there is no 31st of February, and 2023 was not a leap year. The function `dmy()` from `lubridate`, which we met in @sec-lubridate, knows the calendar. It returns `NA` for both. On the other hand, `dmy()` reads `4/7/2024` correctly, which the `DMY` pattern rejects for lack of leading zeros.

::: callout-tip
## Audit tip

Use regex to find dates written in the wrong *format*, and a date parser like `dmy()` to find dates that do not *exist*. A voucher dated 31 February, or a bill dated after the date of its payment, is worth a closer look. Impossible dates are sometimes typing errors. Sometimes they are a sign that the document was prepared later, in a hurry.
:::

## A suggested workflow

1.  **Write down the official format.** Start from the rule that defines the identifier, such as PAN, IFSC or GSTIN, in plain words. For a PAN, it is five letters, then four digits, then one letter.
2.  **Build the pattern piece by piece.** Join the parts with `%R%`, using `ascii_upper()`, `ascii_digit()`, `char_class()` and `repeated()`. Keep alternatives together with `or()`.
3.  **Anchor the whole value.** Wrap the pattern in `exactly()`. Otherwise a longer or shorter value may pass just because a part of it matches.
4.  **Print the pattern.** Print the `rebus` object to see the regular expression it builds. Check that it means what you intended, for example that lower case letters are not allowed by mistake.
5.  **Test on known values.** Try the pattern with `str_detect()` on a few valid and invalid samples before running it on the whole beneficiary or vendor list.
6.  **List the failures.** Run the pattern over the column received from the auditee and pick out the records that fail. Each one is a data quality issue to be explained.
7.  **Check dates twice.** Use `DMY` to find dates written in the wrong format, and `dmy()` to find dates that do not exist, like 31 February. Look closely at impossible dates in vouchers and bills.

::: callout-important
## Remember

A pattern checks only the shape of a value, not whether it is true. A date like 31/02/2024 passes `DMY`, and a GSTIN in the right format may have a wrong check character. Use `dmy()` for dates and `gstin_valid()` for GSTINs, and confirm doubtful cases from the records.
:::

## Summary of functions used

| Function | Package | What it does |
|-----------------------|----------------|--------------------------------|
| `%R%`, `or()`, `%|%` | rebus | Join patterns, and alternatives |
| `case_insensitive()` | rebus | Match a pattern in any case |
| `exactly()`, `START`, `END` | rebus | Anchor a pattern to the whole string, or to its start or end |
| `char_class()`, `negated_char_class()` | rebus | Sets of characters to match, or not to match |
| `ascii_upper()`, `ascii_digit()`, `digit()` and others | rebus | Built-in character classes, with optional repetition |
| `one_or_more()`, `zero_or_more()`, `optional()`, `repeated()` | rebus | Quantifiers |
| `whole_word()`, `BOUNDARY` | rebus | Word boundaries |
| `capture()`, `group()`, `REF1` | rebus | Capturing and non-capturing groups, and back references |
| `lookahead()`, `lookbehind()` and their negative forms | rebus | Look-arounds |
| `ISO_DATE`, `DMY`, `roman()` | rebus | Ready patterns for dates and Roman numerals |
| `str_view()`, `str_detect()`, `str_extract_all()`, `str_remove_all()`, `str_replace_all()` | stringr | Use the patterns to view, test, extract, remove and replace |
| `mutate()`, `filter()` | dplyr | Add the check results as columns, and keep the failures |
| `dmy()` | lubridate | Converts day-month-year text to dates, `NA` for impossible dates |

: Functions used in this chapter {#tbl-rebusfunctions}
