# Regex - A quick introduction

A __Regular Expression__, or __regex__ for short, is a powerful tool, which helps us writing code for pattern matching in texts.  Regex, is a pattern that describes a set of strings. It is a sequence of characters that define a search pattern. It is used to search for and manipulate text. Regex can be used in many programming languages, including R.

Regex patterns are made up of a combination of regular characters and special characters. Regular characters include letters, digits, and punctuation marks. Special characters have a specific meaning in regex and are used to represent patterns of characters.

Regex patterns can be used for a variety of purposes, including:

- Searching for specific strings in text
- Extracting specific parts of a string
- Replacing parts of a string with other text
- Validating input from users

In R, we can use the `grep` and `gsub` functions to search for and manipulate text using regex.

## Basic Regex -  Literal Characters

Every _literal character_, in itself is a `regex` that matches itself. Thus, `a` matches third character in text `Charles`.  These literal characters are case sensitive.

Example-1
```{r}
ex_text <- "This is an example text"
# Match literal `x`
str_view(ex_text, "x")
# Match Upper case literal "X"
str_view(ex_text, "X", match = NA)
```

### Case sensitivity
As the literals are case_sensitive and we sometimes are not aware of exact case, to match case insensitive literals, we can make use of `stringr` function `regex` in this case, wherein an argument `ignore_case` (note snake case) is there.  Actually, behind the scenes, all regex expressions in stringr are wrapped in this function with argument defaults as `FALSE`. Thus, the code in above example is actually equivalent to the following-

```{r}
# Match literal `x`
str_view(ex_text, regex("x"))
# Match Upper case literal "X"
str_view(ex_text, regex("X"), match = NA)
```

Thus, to match case insensitive literals (or other regex expressions) we may make use of the argument `ignore_case` like this-

```{r}
# Match literal `x`
str_view(ex_text, regex("X", ignore_case = TRUE))
```

## Metacharacters
### Character sets

It is always not feasible to put every literal characters.  We may also match literal characters from a given set of options. To **match a group of characters** we have to put all these in square brackets.  So, `[abc]` matches either of `a`, `b`, or `c`.  

Example-
```{r}
ex_vec <- c("Apple", "Orange", "Myrrh")
# matches a vowel
str_view(ex_vec, "[aeiou]")
# matches a vowel irrespective of case
str_view(ex_vec, regex("[aeiou]", ignore_case = TRUE))
```
To **match a range of characters/numbers** we can separate these by hyphen in square brackets. So, `[a-n]` will match a character from range `[abcdefghijklmn]`.

Example-
```{r}
ex_text <- "The quick brown fox jumps over the lazy dog"
# Match a, b or c in lower case
str_view(ex_text, regex("[a-c]"))
```
Example-2
```{r}
ex_colors <- c("grey", "black", "gray")
str_view(ex_colors, "gr[ae]y")
```

We can also use __pre-built character classes__ listed below.

  + `[:punct:]` punctuation.
  + `[:alpha:]` letters.
  + `[:lower:]` lowercase letters.
  + `[:upper:]` uppercase letters.
  + `[:digit:]` digits.
  + `[:xdigit:]` hex digits.
  + `[:alnum:]` letters and numbers.
  + `[:cntrl:]` control characters.
  + `[:graph:]` letters, numbers, and punctuation.
  + `[:print:]` letters, numbers, punctuation, and white-space.
  + `[:space:]` space characters (basically equivalent to `\\s`).
  + `[:blank:]` space and tab.

Example-
```{r}
ex_vec2 <- c("One apple", "2 Oranges")
str_view(ex_vec2, "[:digit:]", match = NA)
```

### Negation of character sets/classes
Typing a caret [`^`] after the opening square bracket negates the character class.  Example-
```{r}
ex_colors <- c("grey", "black", "gray")
str_view(ex_colors, "gr[^c-f]y")
```

So, in this case `gr` followed by character except from `c` to `f` and further followed by `y` will only be matched, resulting in matching `gray` but not matching `grey`.  So, putting a `^` caret character inside the character classes before putting matching characters/classes will match everything except those characters/classes.

### Non-printable characters/ Meta characters (short-hand character classes)

We can use special character sequences to put non-printable characters in our regular expression(s). E.g. `\t` matches a tab character.  __But since `\` is an escape character in R, we need to escape it too.__  So to match a tab character we have to put `\\t` in our regex sequence.  Regex for that matches new line (line feed) is `\\n`. `Regex` for other meta characters is listed below-

  + `\\s` matches a white-space character.  Moreover, its complement `\\S` matches any character except a white-space.
  + `\\w` matches any alphanumeric character. Similarly, its complement is `\\W` which matches any character except alphanumeric characters.
  + `\\d` matches any digit.  Similarly, its complement is `\\D` which matches any character except digits.
  + `\\b` matches any word boundary.  Thus, `\\B` matches any character except a word boundary.
  + `.` matches any character. To match a literal dot `.` we have to escape that; and thus `\\.` matches a dot character.
  
See these examples-
```{r}
ex_vec3 <- c("One apple", "2 oranges & 3 bananas.")
# match word character
str_view(ex_vec3, "\\w", match = NA)
# match any character followed by a dot character
str_view(ex_vec3, ".\\.", match = NA)
# Note both character and dot will be matched
```

## Quantifiers

What if we want to match more than one literal/character through `regex`? Let's say if we want to check whether the given string or string vector contain two consecutive vowels.  One method may be to use character classes two times i.e. using `[aeiou][aeiou]`. But this method is against the principles of __DRY__^[Dont repeat yourself] which is one of the common principle of programming.  To solve these issues, we have quantifiers.

- `+` __1 or more__ occurrences
- `*` __0 or more__
- `?` __0 or 1__
- `{}` specified numbers
  + `{n}` exactly n
  + `{n,}` n or more
  + `{n,m}` between n and m

Thus, we may match two consecutive vowels using `[aeiou]{2}`.  See this example

```{r}
ex_vec <- c("Apple", "Banana", "pineapple")
str_view(ex_vec, "[aeiou]{2}", match = NA)
```
## Alternation

Alternation in regular expressions allows you to match one pattern or another, depending on which one appears first in the input string. The pipe symbol `|` is used to separate the alternative patterns. 

##### Basic Alternation
Let's start with a basic example to illustrate how alternation works:

```{r}
string <- "I have an apple and a banana"
pattern <- "apple|banana"

str_extract(string, pattern)
```
##### Order of Precedence

When using alternation, it's important to keep in mind the order of precedence rules. In general, the first pattern that matches the input string will be selected, and subsequent patterns will not be considered. Here's an example to illustrate this:
```{r}
string <- "I have a pineapple and an apple"
str_extract(string, pattern = "apple|pineapple")
```
In this example, we have a string `string` that contains the words "apple" and "pineapple". We want to extract the first occurrence of either "apple" or "pineapple" from this text using a regular expression pattern that utilizes alternation. The pattern `apple|pineapple` means "match 'apple' OR 'pineapple'". However, since the input string contains "pineapple" before "apple", the `str_extract()` function selects the first matching string "pineapple".

##### Grouping Alternatives
We can also use parentheses to group alternative patterns together. This can be useful for specifying more complex patterns. Example:

```{r}
string <- "Apple and pineapples are good for health"
pattern <- "(apple|banana|cherry) (and|or) (pineapple|kiwi|mango)"

str_view(string, regex(pattern, ignore_case = TRUE))
```

In above examples, we have used `stringr::regex()` to modify regex flag to ignore cases while matching.

## Anchors

Anchors in regular expressions allow you to match patterns at specific positions within the input string. In R, you can use various anchors in your regular expressions to match the beginning, end, or specific positions within the input text. 

### Beginning and End Anchors

The beginning anchor `^` and end anchor `$` are used to match patterns at the beginning or end of the input string, respectively.  Example
```{r}
string <- "The quick brown fox jumps over the lazy dog. The fox is brown."
pattern <- "^the"
str_view(string, regex(pattern, ignore_case = TRUE))
```
In the above example, we are matching word `the` which is at the beginning of a sentence only.

### Word Boundary Anchors
The word boundary anchor `\\b` is used to match patterns at the beginning or end of a word within the input string.  Example
```{r}
string <- 'Apple and pineapple, both are good for health'
pattern <- '\\bapple\\b'
str_view(string, regex(pattern, ignore_case = TRUE))
```
In the above example, though `apple` string is contained in another word `pineapple` we are limiting our search for whole words only.

## Capture Groups

A capture group is a way to group a part of a regular expression and capture it as a separate sub-string. This can be useful when you want to extract or replace a specific part of a string.  In R, capture groups are denoted by parentheses `()`. Anything inside the parentheses is captured and can be referenced later in the regular expression or in the replacement string.

One use of capturing group is to refer back to it within a match with back reference: `\1` refers to the match contained in the first parenthesis, `\2` in the second parenthesis, and so on. 

Example-1

```{r}
my_fruits <- c('apple', 'banana', 'coconut', 'berry', 'cucumber', 'date')
# search for repeated alphabet
pattern <- '(.)\\1'
str_view(my_fruits, regex(pattern), match = NA)
```
Example-2
```{r}
# search for repeated pair of alphabets
pattern <- '(..)\\1'
str_view(my_fruits, regex(pattern), match = NA)
```

Another way to use capturing group is, when we want to replace the pattern with something else.  It is better to understand this with the following example-
```{r}
# We have names in last_name, first_name format
names <- c('Hanks, Tom', 'Affleck, Ben', 'Damon, Matt')
str_view(names)

# Using this regex, we can convert these to first_name last_name format
str_replace_all(names, '(\\w+),\\s+(\\w+)', '\\2 \\1')

```

## Lookarounds
__Look-ahead__ and __look-behinds__ are zero-width assertions in regex. They are used to match a pattern only if it is followed or preceded by another pattern, respectively. The pattern in the look-ahead or look-behind is not included in the match.

### Lookahead

A look-ahead is used to match a pattern only if it is followed by another pattern. *Positive Lookaheads* are written as `(?=...)`, where `...` is the pattern that must follow the match.

For example, the regex pattern `hello(?= world)` matches "hello" only if it is followed by " world". It matches "hello world" but not "hello there world" or "hello".

Example

```{r}
string <- c("hello world", "hello there world", "hello")
str_view(string, "hello(?= world)", match = NA)
# Note that "world" is not included in the match
```

### Lookbehind
A look-behind is used to match a pattern only if it is preceded by another pattern. Look-behinds are written as `(?<=...)`, where `...` is the pattern that must precede the match.

For example, the regex pattern `(?<=hello )world` matches "world" only if it is preceded by "hello ". It matches "hello world" but not "world hello" or "hello there world".

Example

```{r}
string <- c("hello world", "world hello", "hello there world")
str_view(string, "(?<=hello )world", match = NA)

```

### Negative Lookahead and Lookbehinds

Negative look-ahead and negative look-behinds are used to match a pattern only if it is not followed or preceded by another pattern, respectively. Negative look-ahead and look-behinds are written as `(?!...)` and `(?<!...)`, respectively.

For example, the regex pattern `hello(?! world)` matches "hello" only if it is not followed by " world". It matches "hello there" but not "hello world" or "hello world there".

Example-

```{r}
string <- c("hello there", "hello world", "hello world there")
str_view(string, "hello(?! world)", match = NA)

```

And the regex pattern `(?<!hello )world` matches "world" only if it is not preceded by "hello ". It matches "world hello" and "hello there world" but not "hello world".

```{r}
string <- c("hello world", "world hello", "hello there world")
str_view(string, "(?<!hello )world", match = NA)
```

*While the difference between the look-ahead and look-behind may be subtle, yet these become clear when string/pattern replacement or extraction is required.*  

Examples- 
```{r}
string <- "I have 10 apples, 6 pineapples and 5 bananas"

# look-behind to match "apples" preceded by a digit and a space
pattern1 <- "(?<=\\d\\s)apples"  

# look-ahead to match count of apples
pattern2 <- "\\d+(?=\\sapple)"  

str_view(string = string, pattern = pattern1, match = NA)
str_view(string = string, pattern = pattern2, match = NA)
```

## Comments
### Comments within regex
We can use the # character to add comments within a regex pattern.  Any text following the `#` symbol on a line is ignored by the regex engine and treated as a comment. This can be useful for documenting your regex patterns or temporarily disabling parts of a pattern for testing or debugging.  Example-

```{r}
str_view(c("xyz","abc"), "x(?#this is a comment)", match = NA)
```

### Verbose Mode (multi-line comments)
In regular expressions, verbose mode is a feature that allows you to write more readable and maintainable regex patterns by adding comments and white-space without affecting their behavior. To enable verbose mode, we can use the `(?x)` or `(?verbose)` modifier at the beginning of your regex pattern.

Example - Using this regex we can extract words that contain a vowel at third place.
```{r}
string <- "The quick brown fox jumps over the lazy dog"
pattern <- "(?x)      # Enable verbose mode
            \\b       # Match word boundary
            \\w{2}    # matches first two alphabets
            [aeiou]   # Match a vowel
            \\w*      # Match optional word characters
            \\b       # Match word boundary"
str_view(string, pattern, match = NA)
```


# Regex in human readble format using `rebus`

Regular expressions, as explained in earlier chapter, are very powerful. However they are often difficult to interpret. There is a package called `rebus` in R, which allows us to build complex regular expressions from human readable expressions. So instead of writing `[a-zA-Z0-9._%+-]+@[a-zA-Z0-9.-]+\.[a-zA-Z]{2,4}` and later trying to decipher what this expression actually meant, we can use a more human-readble format -

```
one_or_more(char_class(ASCII_ALNUM %R% "._%+-")) %R%
  "@" %R%
  one_or_more(char_class(ASCII_ALNUM %R% ".-")) %R%
  DOT %R%
  ascii_alpha(2, 4)
```
Many of us could have by now correctly guessed that these both are regular expressions to detect email addresses from the given text strings.  Rebus actually contains functions like `char_class()` or `one_or_more()` to make building regular expressions easier.  So let's dive in to learn the package.

First of all let's load the library.  Alongside let's also load `stringr` so that we can use its function `str_view` to understand the examples. **Readers may note here that this library `rebus` also contains a function `regex` which creates a conflict with `stringr::regex`.  So to avoid that conflict we may use library `conflicted` here which may set our preferential library in case of conflicts.**

```{r}
library(rebus)
library(stringr)
library(conflicted)
conflict_prefer("regex", "stringr")
conflicts_prefer(rebus::or)
```

## Operators for concatenation and alternation
For concatenating two regular expressions we may use either of the operators `%R%` or `%c%` which like a pipe operator concatenates LHS and RHS together as one regular expression.

For alternation we may use either the operator `%|%` or function `or()` from the package which separates two regular expressions through a pipe for alternation.
Example-1:
```{r}
my_colors <- c("red", "grey", "blue", "gray")
str_view(my_colors, "gr" %R% or("a" , "e") %R% "y")
# Note operator precedence
# This actually means either "gra" or "ey"
str_view(my_colors, "gr" %R% "a" %|% "e" %R% "y")
```

## Literal Characters, specvial characters and case sensitivity
The literal characters may, as we learnt earlier, may be given in the string as they mean. Literals are case sensitive too.  So if we want a case insensitive match, there is a function `case_insensitive` here which will take care of the requirement.

Example-2:
```{r}
ex_text <- "This is an example text."
# Match literal `x`
str_view(ex_text, "X", match = NA)
str_view(ex_text, case_insensitive("X"), match = NA)
```

To match special characters which may have special meaning in regular expressions, we have constants here.  In rebus, the constants are usually available in `UPPER CASE` and equivalent functions are available in `lower case`.  The following special character constants are available -

```{r}
BACKSLASH
CARET
DOLLAR
DOT
PIPE
QUESTION
STAR
PLUS
OPEN_PAREN
CLOSE_PAREN
OPEN_BRACKET
CLOSE_BRACKET
OPEN_BRACE
```

Example-3:
```{r}
str_view(ex_text, DOT)
```


## Metacharacters
### Character classes
To group characters together in a class to match any of them, we may use function `char_class()` in rebus package. See example-4:
```{r}
ex_vec <- c("Apple", "Orange", "Myrrh")
# matches a vowel
str_view(ex_vec, char_class("aeiou"))
# matches a vowel irrespective of case
str_view(ex_vec, case_insensitive(char_class("aeiou")))
```

To **match a range of characters/numbers** we separate these by hyphen in square brackets (in normal regex building). So, `char_class("a-d")` will match a character from range `[abcd]`.

Example-5:
```{r}
ex_text <- "The quick brown fox jumps over the lazy dog."
# Match a, b or c in lower case
str_view(ex_text, char_class("a-d"))
```

For negated character classes we again have an intuitively named function `negated_char_class()` in R, which we can use as per our requirement.

Example-6:
```{r}
ex_text <- "The quick brown fox jumps over the lazy dog."
# Match all text except vowels
str_view(ex_text, negated_char_class("aeiou"))
# Note that upper case and dot character have also been matched.
```


### Built-in Character classes
We can also use __pre-built character classes__ available in rebus, as listed below.

```{r}
ALPHA
ALNUM
BLANK
DIGIT
LOWER
PRINT
PUNCT
SPACE
UPPER
HEX_DIGIT
ANY_CHAR
GRAPHEME
NEWLINE
DGT
WRD
SPC
NOT_DGT
NOT_WRD
NOT_SPC  # Equivalent to "\\S"
ASCII_DIGIT
ASCII_LOWER
ASCII_UPPER
ASCII_ALPHA
ASCII_ALNUM
```

See another example.

Example-7:
```{r}
ex_text <- "The quick brown fox jumps over the lazy dog."
# Match TAB or SPACE Characters
str_view(ex_text, BLANK)
# Match all UPPER CASE characters
str_view(ex_text, UPPER)
```

Besides the afore-mentioned class constants, we have `lower case` equivalent functions for these character classes for a more useful regex building.

```
alnum(lo, hi, char_class = TRUE)
alpha(lo, hi, char_class = TRUE)
blank(lo, hi, char_class = TRUE)
cntrl(lo, hi, char_class = TRUE)
digit(lo, hi, char_class = TRUE)
graph(lo, hi, char_class = TRUE)
lower(lo, hi, char_class = TRUE)
printable(lo, hi, char_class = TRUE)
punct(lo, hi, char_class = TRUE)
space(lo, hi, char_class = TRUE)
upper(lo, hi, char_class = TRUE)
hex_digit(lo, hi, char_class = TRUE)
any_char(lo, hi)
grapheme(lo, hi)
newline(lo, hi)
dgt(lo, hi, char_class = FALSE)
wrd(lo, hi, char_class = FALSE)
spc(lo, hi, char_class = FALSE)
not_dgt(lo, hi, char_class = FALSE)
not_wrd(lo, hi, char_class = FALSE)
not_spc(lo, hi, char_class = FALSE)
ascii_digit(lo, hi, char_class = TRUE)
ascii_lower(lo, hi, char_class = TRUE)
ascii_upper(lo, hi, char_class = TRUE)
ascii_alpha(lo, hi, char_class = TRUE)
ascii_alnum(lo, hi, char_class = TRUE)
char_range(lo, hi, char_class = lo < hi)
number_range(lo, hi, allow_leading_zeroes = FALSE, capture = FALSE)
```
In the above functions, both `lo` and `hi` accept positive integers as quantifiers; and `char_class` argument a logical value.  See these examples.

Example-8:
```{r}
ip_add <- "My IP address is 255.1.2.50"
str_view(ip_add, digit(1, 3))
str_view(ip_add, digit(3))
str_view(ip_add, not_dgt(5))
# Note this will match none
str_view(ip_add, space(2))
```

### Word Boundaries

To match a word boundary (or its negation) we have `BOUNDARY` (and `NOT_BOUNDARY`) in rebus. Function `whole_word(x)`, on the other hand, wraps the regex in word boundary tokens to match a whole word.  See following example.

Example-9:
```{r}
ex_text <- "The thermometer, they were searching is placed in a leather box."
# Note three matches
str_view(ex_text, "the")
# There is no match
str_view(ex_text, whole_word("the"), match = NA)
str_view(ex_text, case_insensitive(whole_word("the")), match = NA)
str_view(ex_text, BOUNDARY %R% case_insensitive("the"))
str_view(ex_text, NOT_BOUNDARY %R% "the")
```

## Quantifiers

We learnt following quantifiers in regular expressions.

- `+` __1 or more__ occurrences
- `*` __0 or more__
- `?` __0 or 1__
- `{}` specified numbers
  + `{n}` exactly n
  + `{n,}` n or more
  + `{n,m}` between n and m

We have meaningfully named functions for each of the above quantifiers in rebus.

- `one_or_more(x, char_class = NA)`
- `zero_or_more(x, char_class = NA)`
- `optional(x, char_class = NA)`
- `repeated(x, lo, hi, lazy = FALSE, char_class = NA)`
  + where `lo` and `hi` represent `n` and `m` equivalently.
  
Additionally, we may notice an argument `char_class = NA` which accepts a logical value for case when `x` is required to be wrapped in a character class.

We may match two consecutive vowels using `repeated("aeiou", 2, char_class = TRUE)`.  See this example.

Example-10:
```{r}
ex_vec <- c("1 apple", "2 bananas", "3 pineapples")
# match two consecutive vowels
str_view(ex_vec, repeated("aeiou", 2, char_class = TRUE))
# match a number followed by apple with an optional space
str_view(ex_vec, one_or_more(DIGIT) %R% optional(BLANK) %R% "apple")
```

## Anchors

Anchors in regular expressions allow us to match patterns at specific positions within the input string. In rebus, we have constants `START` and `END` to match the beginning or end positions within the input text, respectively. 

Example-11:
```{r}
string <- "The quick brown fox jumps over the lazy dog. The fox is brown."
str_view(string, START %R% case_insensitive("the"))
```
In the above example, we are matching word `the` which is at the beginning of a sentence only.  There is one more function `exactly(x)` in rebus which makes the regular expression match the whole string, from start to end, in fact a shortcut of `START %R% x %R% END`.

Example-12:
```{r}
ex_vec <- c("apple", "banana", "cherry", "pineapple")
str_view(ex_vec, exactly("apple"))
```

## Capture Groups
We have seen that a capture group is a way to group a part of a regular expression and capture it as a separate substring. This is useful when we want to extract or replace a specific part of a string.  In R, capture groups are denoted by parentheses `()`. Anything inside the parentheses is captured and can be referenced later in the regular expression or in the replacement string.  We also learnt that capturing groups are useful to refer back to it within a match with back reference: `\1` refers to the match contained in the first parenthesis, `\2` in the second parenthesis, and so on. 

In rebus, there are two functions `capture(x)` and `group(x)` for capturing regex.  Former is good with match functions while latter is mostly used with alternations.

Example-13:
```{r}
## capture(x)
my_fruits <- c('apple', 'banana', 'coconut', 'berry', 'cucumber', 'date')
str_remove_all(my_fruits, capture(char_class("aeiou")))

## group()
my_toppings <- group("olive" %|% "mushroom" %|% "tomato")
pizza <- "We have olive, mushroom, chicken and capsicum pizza in our menu."
# Extract my favourite topping from available pizza menu.
str_extract_all(pizza, pattern = my_toppings)

```

Backreferences for replacement operations, are denoted by constants `REF#` where `#` is actually a digit from 1 to 9.

Example-14:

```{r}
my_fruits <- c('apple', 'banana', 'coconut', 'berry', 'cucumber', 'date')
# search for repeated alphabet
str_view(my_fruits, capture(ANY_CHAR) %R% REF1 , match = NA)
```

Similar to previous chapter, we can use capturing group to replace the pattern with something else.  

Example-15:
```{r}
# We have names in last_name, first_name format
names <- c('Hanks, Tom', 'Affleck, Ben', 'Damon, Matt')
str_view(names)
# Pattern to capture first name and last name
pat <- capture(whole_word(one_or_more(ALPHA))) %R% "," %R% optional(SPACE) %R% capture(whole_word(one_or_more(ALPHA)))
repl <- REF2 %R% " " %R% REF1
# Using this regex, we can convert these to first_name last_name format
str_replace_all(names, pat, repl)
```

## Lookarounds
As we had learnt, __lookaheads__ and __lookbehinds__ are zero-width assertions in regex. They are used to match a pattern only if it is followed or preceded by another pattern, respectively. The pattern in the `lookahead` or `lookbehind` is not included in the match.

We have intuitively named functions to deal with all four lookarounds-
```
lookahead(x)
negative_lookahead(x)
lookbehind(x)
negative_lookbehind(x)
```

Example-16: Find character `q` not followed by `u`

```{r}
countries <- c("mozambique", "qatar", "iraq")
# With lookahead
str_view(countries, "q" %R% negative_lookahead("u"))
# Without Lookahaed - Notice the difference
str_view(countries, "q" %R% negated_char_class("u"))
```

Example-17:  
```{r}
# Lookahead
string <- c("hello world", "hello there world", "hello")
str_view(string, "hello" %R% lookahead(optional(SPACE) %R% "world"), match = NA)
# Note that "world" is not included in the match

# Lookbehind
str_view(string, lookbehind("hello" %R% optional(SPACE)) %R% "world", match = NA)

# Negative lookahead
str_view(string, "hello" %R% negative_lookahead(optional(SPACE) %R% "world"), match = NA)

# Negative lookbehind
str_view(string, negative_lookbehind("hello" %R% optional(SPACE)) %R% "world", match = NA)
```

More examples.

Example-18:
```{r}
string <- "I have 10 apples, 6 pineapples and 5 bananas"

# lookahead to match count of apples
pattern1 <- one_or_more(DIGIT) %R% lookahead(optional(SPACE) %R% "apple") 
# How many apples?
str_view(string = string, pattern = pattern1, match = NA)
```

## Some useful regex functions

### Matching valid dates


```{r}
# Individual date-time components
DTSEP             # optional selected punctuation or space
CENTURY           # exactly two digits
YEAR              # one to four digits
YEAR2             # exactly two digits
YEAR4             # exactly four digits
MONTH             # number from 1 to 12, leading zero
WEEK_OF_YEAR      # number from 0 to 53, leading zero
DAY               # number from 1 to 31, leading zero
DAY_SINGLE        # leading space
HOUR24            # 24 hour clock, leading zero
HOUR12            # 12 hour clock, leading zero
HOUR24_SINGLE     # 24 hour clock, leading space
HOUR12_SINGLE     # 12 hour clock, leading space
MINUTE            # number from 0 to 59, leading zero
SECOND            # number from 0 to 61 (leap seconds), leading zero
FRACTIONAL_SECOND # a second optional decimal point and up to 6 digits
AM_PM             # AM or PM, any case
TIMEZONE_OFFSET   # optional plus or minus, then four digits
TIMEZONE          # Any value returned by OlsonNames()
# ISO 8601 formats
ISO_DATE          # %Y-%m-%d
ISO_TIME          # %H:%M:%S
ISO_DATETIME      # ISO_DATE followed by ISO_TIME, separated by space or "T".
# Compound forms, separated by DTSEP
YMD
YDM
MYD
MDY
DYM
DMY
HMS
HM
MS
```

Example-19:
```{r}
# We have some dates - both valid and invalid
some_dates <- c("2000-13-01", "2025-08-09","2000-01-32", "2000-00-01", "2000-01-00", "2020-05-20")

str_view(some_dates, ISO_DATE)

# Similarly some time formats
some_times <- c("24:00:00", "23:60:59", "23:59:62", "23 59 59", "23:55:55", "00:00:00")
str_view(some_times, ISO_TIME)
```

## Roman numerals
To match Roman numerals we have a constant `ROMAN` as well as a function `roman(lo, hi)` in rebus.

Example-20:
```{r}
# Some Roman numerals, both valid and invalid
some_numbers <- c("MMMDCCCXLVIII", "MMMCMDCCCXLVIIV", "MCD", "XIL", "LIX", "XL")
# Find valid roam numerals
str_view(some_numbers, exactly(roman()))
```




