```{r}
#| include: false
source("_common.R")
```

# Readable Regular Expressions with rebus {#sec-rebus}

Regular expressions, as explained in the previous chapter (@sec-regex), are very powerful. However they are often difficult to read, even for the person who wrote them, a few months later. There is a package called `rebus` in R, which allows us to build complex regular expressions from human readable expressions. So instead of writing `[a-zA-Z0-9._%+-]+@[a-zA-Z0-9.-]+\.[a-zA-Z]{2,4}` and later trying to decipher what this expression actually meant, we can use a more human-readable format -

```
one_or_more(char_class(ASCII_ALNUM %R% "._%+-")) %R%
  "@" %R%
  one_or_more(char_class(ASCII_ALNUM %R% ".-")) %R%
  DOT %R%
  ascii_alpha(2, 4)
```
Many of us could have guessed by now that both are regular expressions to detect email addresses. `rebus` contains functions like `char_class()` and `one_or_more()`, with names that say what they do. The result is an ordinary regular expression, which can be used in any `stringr` function. So let's dive in to learn the package.

First of all let's load the library. Alongside, let's also load `stringr`, so that we can use its function `str_view()` to understand the examples.

::: callout-warning
## Caution

`rebus` also has a function named `regex()`, which conflicts with `stringr::regex()`. To avoid confusion, we use the package `conflicted` to state which one we prefer.
:::

```{r}
#| message: false
#| warning: false
library(rebus)
library(stringr)
library(conflicted)
conflicts_prefer(stringr::regex, rebus::or)
```

## Operators for concatenation and alternation
For concatenating two regular expressions we may use either of the operators `%R%` or `%c%` which like a pipe operator concatenates LHS and RHS together as one regular expression.

For alternation we may use either the operator `%|%` or function `or()` from the package which separates two regular expressions through a pipe for alternation.
Example-1:
```{r}
my_colors <- c("red", "grey", "blue", "gray")
str_view(my_colors, "gr" %R% or("a" , "e") %R% "y")
# Note operator precedence
# This actually means either "gra" or "ey"
str_view(my_colors, "gr" %R% "a" %|% "e" %R% "y")
```

Printing a `rebus` expression shows the regular expression it builds. This is the best way to check that it means what we intended.

```{r}
"gr" %R% or("a" , "e") %R% "y"
"gr" %R% "a" %|% "e" %R% "y"
```

The first is `gr(?:a|e)y`, "gr, then a or e, then y", as intended. The second is `gra|ey`.

::: callout-warning
## Caution

The second pattern matched both *grey* and *gray* above, but for the wrong reasons, *ey* in one and *gra* in the other. A pattern can give the right answer on a few test values and still be wrong. Use `or()` to keep alternatives together.
:::

## Literal Characters, special characters and case sensitivity
The literal characters may, as we learnt earlier, may be given in the string as they mean. Literals are case sensitive too.  So if we want a case insensitive match, there is a function `case_insensitive` here which will take care of the requirement.

Example-2:
```{r}
ex_text <- "This is an example text."
# Match literal `x`
str_view(ex_text, "X", match = NA)
str_view(ex_text, case_insensitive("X"), match = NA)
```

To match special characters which may have special meaning in regular expressions, we have constants here.  In rebus, the constants are usually available in `UPPER CASE` and equivalent functions are available in `lower case`.  The following special character constants are available -

```{r}
BACKSLASH
CARET
DOLLAR
DOT
PIPE
QUESTION
STAR
PLUS
OPEN_PAREN
CLOSE_PAREN
OPEN_BRACKET
CLOSE_BRACKET
OPEN_BRACE
```

Example-3:
```{r}
str_view(ex_text, DOT)
```


## Metacharacters
### Character classes
To group characters together in a class to match any of them, we may use function `char_class()` in rebus package. See example-4:
```{r}
ex_vec <- c("Apple", "Orange", "Myrrh")
# matches a vowel
str_view(ex_vec, char_class("aeiou"))
# matches a vowel irrespective of case
str_view(ex_vec, case_insensitive(char_class("aeiou")))
```

To **match a range of characters/numbers** we separate these by hyphen in square brackets (in normal regex building). So, `char_class("a-d")` will match a character from range `[abcd]`.

Example-5:
```{r}
ex_text <- "The quick brown fox jumps over the lazy dog."
# Match a, b, c or d in lower case
str_view(ex_text, char_class("a-d"))
```

For negated character classes we again have an intuitively named function `negated_char_class()` in R, which we can use as per our requirement.

Example-6:
```{r}
ex_text <- "The quick brown fox jumps over the lazy dog."
# Match all text except vowels
str_view(ex_text, negated_char_class("aeiou"))
# Note that upper case and dot character have also been matched.
```


### Built-in Character classes
We can also use __pre-built character classes__ available in rebus, as listed below.

```{r}
ALPHA
ALNUM
BLANK
DIGIT
LOWER
PRINT
PUNCT
SPACE
UPPER
HEX_DIGIT
ANY_CHAR
GRAPHEME
NEWLINE
DGT
WRD
SPC
NOT_DGT
NOT_WRD
NOT_SPC  # Equivalent to "\\S"
ASCII_DIGIT
ASCII_LOWER
ASCII_UPPER
ASCII_ALPHA
ASCII_ALNUM
```

See another example.

Example-7:
```{r}
ex_text <- "The quick brown fox jumps over the lazy dog."
# Match TAB or SPACE Characters
str_view(ex_text, BLANK)
# Match all UPPER CASE characters
str_view(ex_text, UPPER)
```

Besides the afore-mentioned class constants, we have `lower case` equivalent functions for these character classes for a more useful regex building.

```
alnum(lo, hi, char_class = TRUE)
alpha(lo, hi, char_class = TRUE)
blank(lo, hi, char_class = TRUE)
cntrl(lo, hi, char_class = TRUE)
digit(lo, hi, char_class = TRUE)
graph(lo, hi, char_class = TRUE)
lower(lo, hi, char_class = TRUE)
printable(lo, hi, char_class = TRUE)
punct(lo, hi, char_class = TRUE)
space(lo, hi, char_class = TRUE)
upper(lo, hi, char_class = TRUE)
hex_digit(lo, hi, char_class = TRUE)
any_char(lo, hi)
grapheme(lo, hi)
newline(lo, hi)
dgt(lo, hi, char_class = FALSE)
wrd(lo, hi, char_class = FALSE)
spc(lo, hi, char_class = FALSE)
not_dgt(lo, hi, char_class = FALSE)
not_wrd(lo, hi, char_class = FALSE)
not_spc(lo, hi, char_class = FALSE)
ascii_digit(lo, hi, char_class = TRUE)
ascii_lower(lo, hi, char_class = TRUE)
ascii_upper(lo, hi, char_class = TRUE)
ascii_alpha(lo, hi, char_class = TRUE)
ascii_alnum(lo, hi, char_class = TRUE)
char_range(lo, hi, char_class = lo < hi)
number_range(lo, hi, allow_leading_zeroes = FALSE, capture = FALSE)
```
In the above functions, both `lo` and `hi` accept positive integers as quantifiers; and `char_class` argument a logical value.  See these examples.

Example-8:
```{r}
ip_add <- "My IP address is 255.1.2.50"
str_view(ip_add, digit(1, 3))
str_view(ip_add, digit(3))
str_view(ip_add, not_dgt(5))
# Note this will match none
str_view(ip_add, space(2), match = NA)
```

### Word Boundaries

To match a word boundary (or its negation) we have `BOUNDARY` (and `NOT_BOUNDARY`) in rebus. Function `whole_word(x)`, on the other hand, wraps the regex in word boundary tokens to match a whole word.  See following example.

Example-9:
```{r}
ex_text <- "The thermometer, they were searching is placed in a leather box."
# Note three matches
str_view(ex_text, "the")
# There is no match
str_view(ex_text, whole_word("the"), match = NA)
str_view(ex_text, case_insensitive(whole_word("the")), match = NA)
str_view(ex_text, BOUNDARY %R% case_insensitive("the"))
str_view(ex_text, NOT_BOUNDARY %R% "the")
```

## Quantifiers

We learnt following quantifiers in regular expressions.

- `+` __1 or more__ occurrences
- `*` __0 or more__
- `?` __0 or 1__
- `{}` specified numbers
  + `{n}` exactly n
  + `{n,}` n or more
  + `{n,m}` between n and m

We have meaningfully named functions for each of the above quantifiers in rebus.

- `one_or_more(x, char_class = NA)`
- `zero_or_more(x, char_class = NA)`
- `optional(x, char_class = NA)`
- `repeated(x, lo, hi, lazy = FALSE, char_class = NA)`
  + where `lo` and `hi` represent `n` and `m` equivalently.
  
Additionally, we may notice an argument `char_class = NA` which accepts a logical value for case when `x` is required to be wrapped in a character class.

We may match two consecutive vowels using `repeated("aeiou", 2, char_class = TRUE)`.  See this example.

Example-10:
```{r}
ex_vec <- c("1 apple", "2 bananas", "3 pineapples")
# match two consecutive vowels
str_view(ex_vec, repeated("aeiou", 2, char_class = TRUE))
# match a number followed by apple with an optional space
str_view(ex_vec, one_or_more(DIGIT) %R% optional(BLANK) %R% "apple")
```

## Anchors

Anchors in regular expressions allow us to match patterns at specific positions within the input string. In rebus, we have constants `START` and `END` to match the beginning or end positions within the input text, respectively. 

Example-11:
```{r}
string <- "The quick brown fox jumps over the lazy dog. The fox is brown."
str_view(string, START %R% case_insensitive("the"))
```
In the above example, only the `The` at the beginning of the *string* is matched, not the one beginning the second sentence. There is one more function `exactly(x)` in rebus which makes the regular expression match the whole string, from start to end, in fact a shortcut of `START %R% x %R% END`.

Example-12:
```{r}
ex_vec <- c("apple", "banana", "cherry", "pineapple")
str_view(ex_vec, exactly("apple"))
```

## Capture Groups
We have seen that a capture group is a way to group a part of a regular expression and capture it as a separate substring. This is useful when we want to extract or replace a specific part of a string.  In R, capture groups are denoted by parentheses `()`. Anything inside the parentheses is captured and can be referenced later in the regular expression or in the replacement string.  We also learnt that capturing groups are useful to refer back to it within a match with back reference: `\1` refers to the match contained in the first parenthesis, `\2` in the second parenthesis, and so on. 

In rebus, there are two functions for this, `capture(x)` and `group(x)`. `capture()` makes a capturing group, `(x)`, whose match can be referred back to. `group()` makes a *non-capturing* group, `(?:x)`, which only keeps its contents together, as we need with alternations. It does not remember its match.

Example-13:
```{r}
## capture(x)
my_fruits <- c('apple', 'banana', 'coconut', 'berry', 'cucumber', 'date')
str_remove_all(my_fruits, capture(char_class("aeiou")))

## group()
my_toppings <- group("olive" %|% "mushroom" %|% "tomato")
pizza <- "We have olive, mushroom, chicken and capsicum pizza in our menu."
# Extract my favourite topping from available pizza menu.
str_extract_all(pizza, pattern = my_toppings)

```

Backreferences for replacement operations, are denoted by constants `REF#` where `#` is actually a digit from 1 to 9.

Example-14:

```{r}
my_fruits <- c('apple', 'banana', 'coconut', 'berry', 'cucumber', 'date')
# search for repeated alphabet
str_view(my_fruits, capture(ANY_CHAR) %R% REF1 , match = NA)
```

Similar to previous chapter, we can use capturing group to replace the pattern with something else.  

Example-15:
```{r}
# We have names in last_name, first_name format
names <- c('Hanks, Tom', 'Affleck, Ben', 'Damon, Matt')
str_view(names)
# Pattern to capture first name and last name
pat <- capture(whole_word(one_or_more(ALPHA))) %R% "," %R% optional(SPACE) %R% capture(whole_word(one_or_more(ALPHA)))
repl <- REF2 %R% " " %R% REF1
# Using this regex, we can convert these to first_name last_name format
str_replace_all(names, pat, repl)
```

## Lookarounds
As we had learnt, __lookaheads__ and __lookbehinds__ are zero-width assertions in regex. They are used to match a pattern only if it is followed or preceded by another pattern, respectively. The pattern in the `lookahead` or `lookbehind` is not included in the match.

We have intuitively named functions to deal with all four lookarounds-
```
lookahead(x)
negative_lookahead(x)
lookbehind(x)
negative_lookbehind(x)
```

Example-16: Find character `q` not followed by `u`

```{r}
countries <- c("mozambique", "qatar", "iraq")
# With lookahead
str_view(countries, "q" %R% negative_lookahead("u"))
# Without Lookahead - Notice the difference
str_view(countries, "q" %R% negated_char_class("u"))
```

Example-17:  
```{r}
# Lookahead
string <- c("hello world", "hello there world", "hello")
str_view(string, "hello" %R% lookahead(optional(SPACE) %R% "world"), match = NA)
# Note that "world" is not included in the match

# Lookbehind
str_view(string, lookbehind("hello" %R% optional(SPACE)) %R% "world", match = NA)

# Negative lookahead
str_view(string, "hello" %R% negative_lookahead(optional(SPACE) %R% "world"), match = NA)

# Negative lookbehind
str_view(string, negative_lookbehind("hello" %R% optional(SPACE)) %R% "world", match = NA)
```

More examples.

Example-18:
```{r}
string <- "I have 10 apples, 6 pineapples and 5 bananas"

# lookahead to match count of apples
pattern1 <- one_or_more(DIGIT) %R% lookahead(optional(SPACE) %R% "apple") 
# How many apples?
str_view(string = string, pattern = pattern1, match = NA)
```

## Some useful regex functions

### Matching valid dates


```{r}
# Individual date-time components
DTSEP             # optional selected punctuation or space
CENTURY           # exactly two digits
YEAR              # one to four digits
YEAR2             # exactly two digits
YEAR4             # exactly four digits
MONTH             # number from 1 to 12, leading zero
WEEK_OF_YEAR      # number from 0 to 53, leading zero
DAY               # number from 1 to 31, leading zero
DAY_SINGLE        # leading space
HOUR24            # 24 hour clock, leading zero
HOUR12            # 12 hour clock, leading zero
HOUR24_SINGLE     # 24 hour clock, leading space
HOUR12_SINGLE     # 12 hour clock, leading space
MINUTE            # number from 0 to 59, leading zero
SECOND            # number from 0 to 61 (leap seconds), leading zero
FRACTIONAL_SECOND # a second optional decimal point and up to 6 digits
AM_PM             # AM or PM, any case
TIMEZONE_OFFSET   # optional plus or minus, then four digits
TIMEZONE          # Any value returned by OlsonNames()
# ISO 8601 formats
ISO_DATE          # %Y-%m-%d
ISO_TIME          # %H:%M:%S
ISO_DATETIME      # ISO_DATE followed by ISO_TIME, separated by space or "T".
# Compound forms, separated by DTSEP
YMD
YDM
MYD
MDY
DYM
DMY
HMS
HM
MS
```

Example-19:
```{r}
# We have some dates - both valid and invalid
some_dates <- c("2000-13-01", "2025-08-09","2000-01-32", "2000-00-01", "2000-01-00", "2020-05-20")

str_view(some_dates, ISO_DATE)

# Similarly some time formats
some_times <- c("24:00:00", "23:60:59", "23:59:62", "23 59 59", "23:55:55", "00:00:00")
str_view(some_times, ISO_TIME)
```

## Roman numerals
To match Roman numerals we have a constant `ROMAN` as well as a function `roman(lo, hi)` in rebus.

Example-20:
```{r}
# Some Roman numerals, both valid and invalid
some_numbers <- c("MMMDCCCXLVIII", "MMMCMDCCCXLVIIV", "MCD", "XIL", "LIX", "XL")
# Find valid Roman numerals
str_view(some_numbers, exactly(roman()))
```

## Audit use case, readable validation patterns {#sec-rebus-audit}

In @sec-regex-validate we wrote patterns for Indian identifiers. Written with `rebus`, they can be read almost like the official description of the format. Let us build three of them, and print the regular expressions they produce.

```{r}
# PAN: five letters, four digits, one letter
pan_rx  <- exactly(ascii_upper(5) %R% ascii_digit(4) %R% ascii_upper(1))

# IFSC: four letters for the bank, a zero, six letters or digits for the branch
ifsc_rx <- exactly(ascii_upper(4) %R% "0" %R% repeated(char_class("A-Z0-9"), 6))

# GSTIN: State code, PAN, entity number, Z, check character
gstin_rx <- exactly(ascii_digit(2) %R%
                    ascii_upper(5) %R% ascii_digit(4) %R% ascii_upper(1) %R%
                    char_class("1-9A-Z") %R% "Z" %R% char_class("0-9A-Z"))

pan_rx
ifsc_rx
gstin_rx
```

These are exactly the patterns we wrote by hand in the previous chapter. Notice that for the IFSC we used `char_class("A-Z0-9")` and not `ascii_alnum()`, because `ascii_alnum()` would also allow lower case letters. Printing the pattern is how we catch such differences.

```{r}
str_detect(c("ABCPK1234D", "ABCP1234D"), pan_rx)
str_detect(c("SBIN0001234", "SBIN1001234"), ifsc_rx)
str_detect(c("27AAPFU0939F1ZV", "27AAPFU0939F1Z"), gstin_rx)
```

A pattern only checks the format. The check character of a GSTIN still needs the function `gstin_valid()` of @sec-checkdigits.

### Dates in vouchers

Dates in Indian records are usually written day first, like `15/04/2024`. The `rebus` constant `DMY` matches such dates. Let us check a few dates typed in a voucher register.

```{r}
#| message: false
library(lubridate)

voucher_dates <- c("15/04/2024", "31/02/2024", "29/02/2023", "29/02/2024",
                   "4/7/2024", "2024/07/04", "31-12-2024")

tibble::tibble(date_text   = voucher_dates,
               shape_ok    = str_detect(voucher_dates, exactly(DMY)),
               parsed_date = dmy(voucher_dates, quiet = TRUE))
```

The regex checks only the *shape* of a date. It accepts `31/02/2024` and `29/02/2023`, because 31 is a possible day and 02 a possible month. But there is no 31st of February, and 2023 was not a leap year. The function `dmy()` from `lubridate`, which we met in the chapter on dates, knows the calendar, and returns `NA` for both. On the other hand, `dmy()` reads `4/7/2024` correctly, which the `DMY` pattern rejects for lack of leading zeros.

::: callout-tip
## Audit tip

Use regex to find dates written in the wrong *format*, and a date parser like `dmy()` to find dates that do not *exist*. A voucher dated 31 February, or a bill dated after the date of its payment, is worth a closer look. Impossible dates are sometimes typing errors, and sometimes a sign that the document was prepared later, in a hurry.
:::

## A suggested workflow

1.  **Write down the official format.** Start from the rule that defines the identifier, such as PAN, IFSC or GSTIN, in plain words. For a PAN, it is five letters, then four digits, then one letter.
2.  **Build the pattern piece by piece.** Join the parts with `%R%`, using `ascii_upper()`, `ascii_digit()`, `char_class()` and `repeated()`. Keep alternatives together with `or()`.
3.  **Anchor the whole value.** Wrap the pattern in `exactly()`. Otherwise a longer or shorter value may pass just because a part of it matches.
4.  **Print the pattern.** Print the `rebus` object to see the regular expression it builds. Check that it means what you intended, for example that lower case letters are not allowed by mistake.
5.  **Test on known values.** Try the pattern with `str_detect()` on a few valid and invalid samples before running it on the whole beneficiary or vendor list.
6.  **List the failures.** Run the pattern over the column received from the auditee and pick out the records that fail. Each one is a data quality issue to be explained.
7.  **Check dates twice.** Use `DMY` to find dates written in the wrong format, and `dmy()` to find dates that do not exist, like 31 February. Look closely at impossible dates in vouchers and bills.

::: callout-important
## Remember

A pattern checks only the shape of a value, not whether it is true. A date like 31/02/2024 passes `DMY`, and a GSTIN in the right format may have a wrong check character. Use `dmy()` for dates and `gstin_valid()` for GSTINs, and confirm doubtful cases from the records.
:::

## Summary of functions used

| Function | Package | What it does |
|-----------------------|----------------|--------------------------------|
| `%R%`, `or()`, `%|%` | rebus | Join patterns, and alternatives |
| `exactly()`, `START`, `END` | rebus | Anchor a pattern to the whole string, or to its start or end |
| `char_class()`, `negated_char_class()` | rebus | Sets of characters to match, or not to match |
| `ascii_upper()`, `ascii_digit()`, `digit()` and others | rebus | Built-in character classes, with optional repetition |
| `one_or_more()`, `zero_or_more()`, `optional()`, `repeated()` | rebus | Quantifiers |
| `whole_word()`, `BOUNDARY` | rebus | Word boundaries |
| `capture()`, `group()`, `REF1` | rebus | Capturing and non-capturing groups, and back references |
| `lookahead()`, `lookbehind()` and their negative forms | rebus | Look-arounds |
| `ISO_DATE`, `DMY`, `roman()` | rebus | Ready patterns for dates and Roman numerals |
| `dmy()` | lubridate | Converts day-month-year text to dates, `NA` for impossible dates |

: Functions used in this chapter {#tbl-rebusfunctions}
