```{r}
#| include: false
source("_common.R")
```

# Factors {#sec-factors}

Audit data is full of categories. Expenditure is booked under heads from an approved list. Payments are made in districts, by offices, to types of vendors. Works are graded by quality inspectors as Poor, Average or Good. Units are rated as Low, Medium or High risk. Bills are grouped by how long they took to pay.

Such categories cause three kinds of trouble. First, a value may be typed wrongly, or may not be in the approved list at all. A head called *Hospitality* in a list that has no such head needs an explanation. Second, many categories have a natural order that is not alphabetical. Sorted as text, *High* risk comes before *Low*, and *April* comes after *August*. Third, charts drawn from categories come out in alphabetical order, with too many bars, which hides the point we want to make.

R has a special class for such data, called `factor`. A factor stores a category, along with the fixed list of values it may take, called its **levels**, in an order that we choose. We met this data type briefly in @sec-fctrss. In this chapter we learn it in detail. We create factors in base R, and then use the **forcats** package to check, order, rename and lump their levels. All along we use audit data. At the end, from @sec-fct-audit, we use factors to catch unapproved heads, filter units by risk and draw a clean chart of the top vendors.

**Prerequisites**

```{r}
#| message: false
#| warning: false
library(tidyverse)   # includes forcats
library(patchwork)
library(conflicted)
conflicts_prefer(dplyr::filter)
```

## Factors in base R

Factors in R are built on the `integer` data type. Base R has two main functions to create factors from character vectors.

- `factor()`
- `as.factor()`

Of these, `factor()` gives full control, as it is the basic function to create factors in base R. Let us discuss both.

### Creating factors from scratch

The quality control wing of a Public Works Department inspects completed works. It grades each work on a five-point scale, with the codes `P`, `A`, `G`, `VG` and `E`. These stand for Poor, Average, Good, Very good and Excellent. Let us keep the allowed codes, in their proper order, in a vector.

```{r}
grade_codes <- c("P", "A", "G", "VG", "E")
```

Now let us say the grades of 10 works were typed in from inspection reports.

```{r}
# Grades of 10 works, as typed in
# Notice one grade typed as small case g, and one as EX, which is not a valid code
grades <- c("G", "g", "A", "VG", "EX", "G", "P", "P", "A", "A")

# Let us create a factor, with the allowed codes as levels
grade_f <- factor(grades, levels = grade_codes)
grade_f
```

In the output above, R has put `NA` silently in place of the values that are not in the allowed codes (the levels). The works graded `g` and `EX` have lost their grades without any warning.

Now check its type and class.

```{r}
typeof(grade_f)
class(grade_f)
```

The `factor` class is built on the `integer` type. Underneath, each value is stored as a number, the position of its level. What we see in the output are the labels, which are the levels by default.

```{r}
unclass(grade_f)
```

We can show meaningful labels without changing the levels, by giving a vector to the `labels` argument. Let us create the grades of 100 works, and label them in full.

```{r}
set.seed(123)
# Grades of 100 works
grades2 <- sample(grade_codes, 100, replace = TRUE)
grades2_f <- factor(grades2, levels = grade_codes,
                    labels = c("Poor", "Average", "Good", "Very good", "Excellent"))
# Let us view some works
head(grades2_f)
# Check its levels
levels(grades2_f)
```

The function `summary()` gives the count of each level of a factor.

```{r}
summary(grades2_f)
```

### Coercing objects to factor

So far, we created a factor from a set of allowed values (the `levels`), and showed them with meaningful `labels`. The function `is.factor()` checks whether a vector is a factor. The function `as.factor()` converts a character vector to a factor. It uses all the distinct values in the vector as levels, but sorts them alphabetically.

```{r}
is.factor(grades2)
is.factor(grades2_f)
grades2_coerced <- as.factor(grades2)
head(grades2_coerced)
levels(grades2_coerced)
# Check summary too
summary(grades2_coerced)
```

The levels are now in alphabetical order, `A`, `E`, `G`, `P`, `VG`. That order means nothing. *Excellent* sits between *Average* and *Good*. Our factor created from scratch did not have this problem. There we gave the levels ourselves, in a meaningful order.

### Order in factors {#sec-ordfact}

Even with the levels in the right order, R does not know that *Good* is better than *Average*. To tell it, we create an *ordered factor*. An ordered factor can be created in two ways.

- Use the argument `ordered = TRUE` in the function `factor()`.
- Convert an existing factor to an ordered factor, with the function `ordered()`.

The second way orders the levels as they already are in the factor.

```{r}
grades2_ordered <- factor(
  grades2,
  levels = grade_codes,
  labels = c("Poor", "Average", "Good", "Very good", "Excellent"),
  ordered = TRUE
)
head(grades2_ordered)
```

The levels are now printed with `<` between them. An ordered factor has another benefit. We can compare its values. Suppose we want to know how many works were graded *Good* or better.

```{r}
#| warning: true
# How many works are graded Good or better?
sum(grades2_ordered >= "Good")
# But an unordered factor gives only a warning, and NA
sum(grades2_f >= "Good")
```

::: callout-warning
## Caution

Notice that the unordered factor did not stop with an error. It gave a warning, and the answer `NA`. In a long script, such a warning is easily missed, and the `NA` silently flows into later results. So always read the warnings, and use ordered factors wherever the order matters.
:::

We can check whether a factor is ordered with the function `is.ordered()`.

```{r}
is.ordered(grades2_ordered)
is.ordered(grades2_f)
```

### Functions returning factors as output

In @sec-cutt we learnt the function `cut()`, which divides numbers into categories and returns a factor. In the next sections we learn more functions that work with factors, as input, as output or both.

## Factors in `forcats`

The package `forcats`, part of the core tidyverse, gives safer and more useful ways to create and handle factors. In `forcats` we have the function `fct()` to create a factor. It stops with an error if any value is not in the given levels. This catches typing errors that `factor()` would hide. For example, here a month in a return has been typed wrongly.

```{r error=TRUE}
months_31 <- c("Jan", "Mar", "May", "Jul", "Aug", "Oxt", "Dec")
fct(months_31, levels = month.abb)
```

## Inspecting factors

### Summarising factors

The `summary()` method gives the count of each level.

```{r}
grades2_fct <- fct(grades2)
summary(grades2_fct)
```

Like `count()` in `dplyr`, forcats has `fct_count()`, which gives the count of each level, and optionally the proportions. The difference from `summary()` is in the type of output. `fct_count()` returns a `tibble`.

```{r}
fct_count(grades2_fct)
# Sort in decreasing counts
fct_count(grades2_fct, sort = TRUE)
# Include proportions also
fct_count(grades2_fct, sort = TRUE, prop = TRUE)
```

### Unique levels only

The function `fct_unique()` returns a factor with the unique values, removing duplicates.

```{r}
fct_unique(grades2_fct)
```

## Order in factors

### Default ordering in factors

A factor created with `fct()` has its levels in the order given in the `levels` argument. If that argument is not given, the levels are in the order of first appearance. This is different from `factor()`, which sorts them alphabetically. We saw this in the output above, where `G` came first because the first work in the data was graded `G`.

To learn more functions of `forcats`, we now use a larger dataset. A state health mission has given us its bills for the financial year 2024-25. There are bills from 18 districts, under five heads. Each bill has a month, an amount, and the number of days taken to pay it. The mission's report groups the bills by delay, with *Not paid* listed first. We have planted two features in the data. The bills of one district are much larger than the rest. Equipment bills rush up in February and March.

```{r}
#| code-fold: true
#| code-summary: "Code to create the bills of the health mission"
set.seed(2024)
districts <- c("Anandpur", "Badalgarh", "Chandanpur", "Devgarh", "Gopalganj", "Haripur",
               "Jaitpur", "Kishangarh", "Lalpur", "Madhopur", "Nayagaon", "Pratapgarh",
               "Rampur", "Sitapur", "Sundernagar", "Tilakpur", "Udaypur", "Vijaynagar")
heads     <- c("Civil works", "Equipment", "Training", "Office expenses", "Publicity")
fy_months <- seq(as.Date("2024-04-01"), by = "month", length.out = 12)
n <- 4000

bills <- tibble(
  bill_no     = sprintf("HM/24-25/%05d", 1:n),
  district    = sample(districts, n, TRUE, prob = runif(18, 0.3, 2)),
  head        = sample(heads, n, TRUE, prob = c(4, 2, 2, 3, 1)),
  month       = sample(fy_months, n, TRUE),
  amount      = round(rlnorm(n, log(60000), 0.8)),
  days_to_pay = round(rexp(n, 1 / 25))
)

# Planted. Bills of Haripur are about three times the usual size
bills <- bills |>
  mutate(amount = if_else(district == "Haripur", amount * 3, amount))
# Planted. Equipment bills are four times larger in February and March
bills <- bills |>
  mutate(amount = if_else(head == "Equipment" & month >= as.Date("2025-02-01"),
                          amount * 4, amount))
# Some bills are not paid yet
bills$days_to_pay[sample(n, 240)] <- NA

bills <- bills |>
  mutate(delay = case_when(is.na(days_to_pay) ~ "Not paid",
                           days_to_pay <= 30  ~ "Within 30 days",
                           days_to_pay <= 60  ~ "31 to 60 days",
                           days_to_pay <= 90  ~ "61 to 90 days",
                           TRUE               ~ "Over 90 days"),
         delay = factor(delay, levels = c("Not paid", "Within 30 days", "31 to 60 days",
                                          "61 to 90 days", "Over 90 days")))
```

```{r}
bills
```

### Reordering factors

Let us total the bills of each district, in lakh, and draw them as bars.

```{r fig-reord, fig.show='hold', fig.align='center', fig.cap="Districts in the default, alphabetical order"}
bills |>
  summarise(amount_lakh = sum(amount) / 1e5, .by = district) |>
  ggplot(aes(district, amount_lakh)) +
  geom_col() +
  coord_flip() +
  labs(x = "District", y = "Total of bills (Rs lakh)")
```

In @fig-reord the districts are in alphabetical order. To compare them, we would rather sort the bars by the amount. The function `fct_reorder()` does this. It orders the levels of a factor by the values of another, numeric, variable.

```
fct_reorder(
  .f,
  .x,
  .fun = median,
  ...,
  .na_rm = NULL,
  .default = Inf,
  .desc = FALSE
)
```

The main arguments are these.

- `.f` is the factor to be sorted.
- `.x` is the numeric variable by which `.f` is sorted.
- `.fun` is the function used when a level of `.f` has several values of `.x`. The default is `median`.

So we can reorder the districts in the chart above with this function. See @fig-reord2.

```{r fig-reord2, fig.show='hold', fig.align='center', fig.cap="Districts sorted by the total of their bills"}
bills |>
  summarise(amount_lakh = sum(amount) / 1e5, .by = district) |>
  ggplot(aes(fct_reorder(district, amount_lakh), amount_lakh)) +
  geom_col() +
  coord_flip() +
  labs(x = "District", y = "Total of bills (Rs lakh)")
```

```{r}
#| include: false
dist_tot <- bills |> summarise(amt = sum(amount), .by = district) |> arrange(desc(amt))
```

Now the order is clear at a glance. `r dist_tot$district[1]` has the largest total, at ₹`r round(dist_tot$amt[1] / 1e7, 2)` crore. A large total may simply mean many bills, so we also need to look at the size of each bill.

`fct_reorder()` is also useful in sorting box plots. Let us draw the bill amounts of each district as boxes, first in the default order (@fig-reord3).

```{r fig-reord3, fig.show='hold', fig.align='center', fig.cap="Unsorted boxes with unordered levels"}
bills |>
  ggplot(aes(x = amount / 1e5, y = district)) +
  geom_boxplot() +
  labs(x = "Amount of bill (Rs lakh)", y = NULL) +
  ggtitle("Unsorted boxes")
```

```{r fig-reord4, fig.show='hold', fig.align='center', fig.cap="Boxes sorted on median by reordering factor levels"}
bills |>
  mutate(district = fct_reorder(district, amount)) |>
  ggplot(aes(x = amount / 1e5, y = district)) +
  geom_boxplot() +
  labs(x = "Amount of bill (Rs lakh)", y = NULL) +
  ggtitle("Boxes sorted on median")
```

```{r}
#| include: false
dist_med <- bills |> summarise(med = median(amount), .by = district) |> arrange(desc(med))
```

In @fig-reord4 the boxes are sorted by the median bill. One district stands apart at the top. The median bill of `r dist_med$district[1]` is ₹`r round(dist_med$med[1] / 1e5, 2)` lakh, against ₹`r format(round(dist_med$med[2]), big.mark = ",")` for the next district. That is `r round(dist_med$med[1] / median(bills$amount[bills$district != dist_med$district[1]]), 1)` times the median bill of all the other districts. The auditor would ask why. Higher rates, bills that combine several jobs, or inflated claims are all possible.

### Reordering factors with two other variables

Sometimes a factor needs to be sorted by the last (or first) values of two other variables. A common case is a line chart with one line for each group. The legend is easiest to read when it lists the lines in the same order as their ends on the right. Here `fct_reorder2()` is useful. Compared with `fct_reorder()`, it takes an extra argument `.y`. Its syntax is

```
fct_reorder2(
  .f,
  .x,
  .y,
  .fun = last2,
  ...,
  .na_rm = NULL,
  .default = -Inf,
  .desc = TRUE
)
```

The default function `last2` sorts the levels of `.f` by the last value of `.y`, when `.y` is plotted against `.x`. Let us draw the monthly total of each head. See @fig-reord5.

```{r fig-reord5, fig.show='hold', fig.align='center', fig.width=10, fig.cap="Factors reordered on two criteria"}
monthly_heads <- bills |>
  summarise(amount_lakh = sum(amount) / 1e5, .by = c(month, head))

default <- monthly_heads |>
  ggplot(aes(month, amount_lakh)) +
  geom_line(aes(group = head, color = head)) +
  labs(x = NULL, y = "Rs lakh") +
  ggtitle("Default legend")

aligned <- monthly_heads |>
  mutate(head = fct_reorder2(head, month, amount_lakh)) |>
  ggplot(aes(month, amount_lakh)) +
  geom_line(aes(group = head, color = head)) +
  labs(x = NULL, y = "Rs lakh") +
  ggtitle("Legend aligned with \nlast values of each line")

default + aligned
```

On the right, the legend starts with *Equipment*, whose line ends highest in March. This is the rush of equipment bills that we planted. Purchases bunched at the end of the year, to use up the budget, are a common audit finding.

### Changing the order of a few levels only

Sometimes the levels of a factor are already in a meaningful order. The delay bands of our bills are an example. Look at @fig-reord6.

```{r fig-reord6, fig.show='hold', fig.align='center', fig.cap="Default order of delay bands"}
bills |>
  ggplot(aes(delay)) +
  geom_bar() +
  labs(x = "Time taken to pay", y = "Number of bills")
```

The bands run from the shortest to the longest delay. But *Not paid* comes first, because the mission's report listed it first. It would read better at the end, after *Over 90 days*. In such cases we may use `fct_relevel()`. It takes a factor, followed by the levels we want to move. By default it moves them to the *front*. The argument `after` says where to put them instead. `after = Inf` moves them to the end. See @fig-reord7.

```{r fig-reord7, fig.show='hold', fig.align='center', fig.cap="Modifying levels manually"}
bills |>
  ggplot(aes(fct_relevel(delay, "Not paid", after = Inf))) +
  geom_bar() +
  labs(x = "Time taken to pay", y = "Number of bills")
```

### Ordering bar charts in order of frequency {#sec-infreq}

The function `fct_infreq()` sorts the levels of a factor in decreasing order of frequency. So it can be used to sort bar charts. Let us count the bills under each head (@fig-fct16).

```{r fig-fct16, fig.align='center', fig.show='hold', fig.width=10, fig.cap="Default order (left) and decreasing order of frequency (right)"}
default <- bills |>
  ggplot(aes(head)) +
  geom_bar() +
  labs(x = NULL, y = "Number of bills")

decreasing <- bills |>
  ggplot(aes(fct_infreq(head))) +
  geom_bar() +
  labs(x = NULL, y = "Number of bills")
default + decreasing
```

### Reversing the factor levels

With `fct_rev()` we can reverse the order of the levels of any factor. See @fig-reord8.

```{r fig-reord8, fig.show='hold', fig.align='center', fig.cap="Reversing levels"}
bills |>
  mutate(head = fct_rev(fct_infreq(head))) |>
  ggplot(aes(head)) +
  geom_bar() +
  labs(x = NULL, y = "Number of bills")
```

### Other reordering

There are two more functions to reorder the levels of a factor.

- `fct_inorder()` orders them by the order in which they first appear.
- `fct_inseq()` orders them by the numeric value of the level.

Readers may explore these functions by themselves.

### More on ordering factors

In @sec-ordfact we saw how an unordered factor can be turned into an ordered factor. This has a side effect when plotting with ggplot2. An ordered factor gets the *viridis* colour scale by default, which runs from dark to light. An unordered factor gets a set of distinct colours. Let us colour the bills of each head by their delay band, first as an ordered factor (@fig-moreord), and then as an unordered one (@fig-moreord2).

```{r fig-moreord, fig.cap="Use of colour scale in ordered and unordered factors"}
bills |>
  mutate(delay = ordered(fct_relevel(delay, "Not paid", after = Inf))) |>
  ggplot(aes(head, fill = delay)) +
  geom_bar() +
  labs(x = NULL, y = "Number of bills") +
  ggtitle("Ordered factor")
```

```{r fig-moreord2, fig.cap="Use of colour scale in ordered and unordered factors"}
bills |>
  mutate(delay = fct_relevel(delay, "Not paid", after = Inf)) |>
  ggplot(aes(head, fill = delay)) +
  geom_bar() +
  labs(x = NULL, y = "Number of bills") +
  ggtitle("Unordered factor")
```

The shades of the ordered factor suit delay bands, since each band is worse than the one before. Distinct colours suit categories with no order, like heads or districts.

## Levels in factors

### Modifying factor levels by applying a function

The function `fct_relabel()` in `forcats` applies a function `.fun` to each level of the factor `.f` given to it. For example, let us change the district names to upper case (@fig-reord9). The order of the levels does not change.

```{r fig-reord9, fig.show='hold', fig.align='center', fig.cap="Applying a function to all labels"}
bills |>
  mutate(district = fct_relabel(fct_reorder(district, amount), str_to_upper)) |>
  ggplot(aes(x = amount / 1e5, y = district)) +
  geom_boxplot() +
  labs(x = "Amount of bill (Rs lakh)", y = NULL)
```

### Modifying factor levels manually

With the function `fct_recode()` we can change the levels of a factor by hand. We give the new levels as a series of named values. The name gives the new level, and the value gives the old level. Levels not mentioned are left as they are. Levels can be removed by naming them `NULL`.

Say the same head has been entered under two names in a voucher register.

```{r}
x <- factor(c("Repairs", "Repair & maintenance", "Stationery", "Supplies", "Repairs"))
fct_recode(x, Repairs = "Repair & maintenance")
```

To collapse several levels into one, we can use its cousin `fct_collapse()`. Here we put *Stationery* under *Supplies*, and both spellings of repairs under *Repairs*.

```{r}
fct_collapse(x,
             Supplies = c("Stationery", "Supplies"),
             Repairs  = c("Repairs", "Repair & maintenance"))
```

### Lump uncommon factor levels into `other`

Package `forcats` has a family of four functions that lump together the levels meeting some condition.

- `fct_lump_min()` lumps levels that appear fewer than `min` times.
- `fct_lump_prop()` lumps levels that appear in fewer than (or equal to) `prop` * `n` times.
- `fct_lump_n()` lumps all levels except the `n` most frequent (or least frequent if `n` < 0).
- `fct_lump_lowfreq()` lumps together the least frequent levels, ensuring that `"Other"` is still the smallest level.

Apart from the factor `f`, these functions take one or more of these arguments.

- `n`. A positive `n` keeps the `n` most common values. A negative `n` keeps the `-n` least common values.
- `prop`. A positive `prop` lumps values that do not appear at least `prop` of the time. A negative `prop` lumps values that do not appear at most `-prop` of the time.
- `min`. Keeps levels that appear at least `min` times.
- `w`. An optional numeric vector giving weights for the frequency of each value (not level) in `f`.
- `other_level`. The name of the level used for the lumped values. The default is `"Other"`. It is always placed at the end of the levels.

Let us see some examples. First, the number of bills of each district.

```{r fig-fctlump, fig.show='hold', fig.align='center', fig.cap="Lumping Factors"}
# Number of bills of each district
bills |>
  count(district)

# Let us keep the 5 districts with the most bills only
bills |>
  ggplot(aes(fct_lump_n(district, n = 5))) +
  geom_bar() +
  labs(x = NULL, y = "Number of bills")
```

In @fig-fctlump we see five districts, plus the `"Other"` level placed last. We may also notice that the bars are not sorted yet.

We may also use the `w` argument, if we already have our factor and the counts of its levels in another vector. See @fig-fctlump2.

```{r fig-fctlump2, fig.show='hold', fig.align='center', fig.cap="Lumping Factors by making use of w argument"}
bills |>
  count(district) |>
  # Making use of `w` argument
  mutate(district = fct_lump_n(district, n = 5, w = n)) |>
  ggplot(aes(district, n)) +
  geom_col() +
  labs(x = NULL, y = "Number of bills")
```

`geom_col()` draws bars whose heights are already in the data, here the counts `n`. It is the same as `geom_bar(stat = "identity")`. To sort the levels in decreasing order of frequency, we can use the function `fct_infreq()`, which we learnt in @sec-infreq. It can also take the `w` argument, if our levels are already counted. So, to sort the levels in @fig-fctlump2, we may use this function one step before lumping. See @fig-fctlump3.

```{r fig-fctlump3, fig.show='hold', fig.align='center', fig.cap="Sorting and Lumping Factors"}
bills |>
  count(district) |>
  # Sorting making use of `w` argument
  mutate(district = fct_infreq(district, w = n),
         district = fct_lump_n(district, n = 5, w = n)) |>
  ggplot(aes(district, n)) +
  geom_col() +
  labs(x = NULL, y = "Number of bills")
```

::: callout-note
The `"Other"` bar is the tallest here, because it joins thirteen districts. That is expected. It does not mean that "Other" is a large district. In @sec-fct-vendors we see how to move such a bar to the end, so that it does not look like the biggest item.
:::

## Audit use case, checking heads against the approved list {#sec-fct-audit}

Expenditure is booked under heads from an approved list. A head typed wrongly, or invented, is a data quality problem. Sometimes it is a way of hiding an expense. `factor()` turns any value not in its `levels` into `NA` silently, as we saw with the grades of works. `fct()` refuses, and tells us which values are not allowed. Both behaviours are useful. Let us take a few vouchers.

```{r}
approved_heads <- c("Office expenses", "Repairs", "Supplies", "Professional fees", "Travel")

vouchers <- tibble(
  voucher_no = sprintf("PV-%03d", 1:8),
  head = c("Repairs", "Supplies", "Ofice expenses", "Travel", "Hospitality",
           "Repairs", "supplies", "Professional fees")
)
```

```{r error=TRUE}
fct(vouchers$head, levels = approved_heads)
```

`fct()` stops with an error, listing the values which are not approved. To find the vouchers concerned, we let `factor()` turn them into `NA`, and pick those rows.

```{r}
vouchers |>
  mutate(head_f = factor(head, levels = approved_heads)) |>
  filter(is.na(head_f))
```

Three vouchers have heads outside the approved list. *Ofice expenses* is a typing error, and *supplies* differs only in case. But *Hospitality* is a head which does not exist in the approved list at all. It needs an explanation. The auditor would ask the office under what sanction the hospitality expense was incurred, and which approved head it should have been charged to.

::: callout-tip
## Audit tip

Do not correct the heads in the original column. Keep the auditee's data as received, and put any corrected head in a new column. Then the list of exceptions, and how each was resolved, can go into the working papers.
:::

## Audit use case, risk ratings as an ordered factor {#sec-fct-risk}

Audit planning often rates units as Low, Medium or High risk. These ratings have a natural order. As an ordered factor, they can be compared and filtered directly.

```{r}
units <- tibble(
  unit = c("Division A", "Division B", "Division C", "Division D", "Division E"),
  risk = factor(c("Medium", "Low", "High", "Very high", "Low"),
                levels = c("Low", "Medium", "High", "Very high"), ordered = TRUE)
)

# Units of Medium risk or above, to be covered this year
units |>
  filter(risk >= "Medium") |>
  arrange(desc(risk))
```

Without the ordered factor, `"Medium"` and `"High"` would be compared as text, alphabetically, and `"High"` would wrongly come before `"Low"`. The list above goes straight into the annual audit plan, with the highest risk units first.

## Audit use case, top vendors and the rest {#sec-fct-vendors}

Charts of expenditure by vendor usually have too many bars. A common audit chart shows the top few vendors, and puts the rest together as *All others*. Since we are interested in *money*, not in the number of payments, we lump by the amount, using the `w` argument (@fig-fctvendor). The code below creates a register of 2,000 payments to 40 vendors. Click on it to see the code.

```{r}
#| code-fold: true
#| code-summary: "Code to create the vendor payments"
set.seed(124)
payments <- tibble(
  vendor = sample(sprintf("Vendor %02d", 1:40), 2000, TRUE,
                  prob = c(10, 8, 6, rep(1, 37))),
  amount = round(rlnorm(2000, log(25000), 1.1))
)
head(payments)
```

```{r}
#| label: fig-fctvendor
#| fig-cap: "Payments to the top 8 vendors, and all others together"
payments |>
  mutate(vendor = fct_lump_n(vendor, n = 8, w = amount, other_level = "All others"),
         vendor = fct_infreq(vendor, w = amount) |> fct_relevel("All others", after = Inf) |>
           fct_rev()) |>
  summarise(amount_lakh = sum(amount) / 1e5, .by = vendor) |>
  ggplot(aes(amount_lakh, vendor)) +
  geom_col(fill = "steelblue") +
  labs(x = "Amount paid (Rs lakh)", y = NULL) +
  theme_bw()
```

`fct_lump_n()` keeps the 8 vendors with the most money and lumps the rest. `fct_infreq()` with `w = amount` sorts the vendors by the amount paid. And `fct_relevel(..., after = Inf)` moves *All others* to the end, so that it does not look like the biggest vendor. Horizontal bars, with the vendor on the y-axis, keep long names readable. But ggplot2 draws the first level at the *bottom* of the y-axis. So we reverse the order with `fct_rev()`, to bring the biggest vendor to the top.

```{r}
#| include: false
top3_share <- payments |>
  summarise(amt = sum(amount), .by = vendor) |>
  arrange(desc(amt)) |>
  summarise(share = 100 * sum(amt[1:3]) / sum(amt)) |>
  pull(share)
```

The chart shows that the three largest vendors received `r round(top3_share)`% of all the money. The next step is to see how these vendors were selected, and whether their purchases went through open tenders.

## A suggested workflow

1.  **Get the approved list.** Obtain the approved list of codes from the auditee or from the rules, such as expenditure heads, vendor types or districts. Keep it as a character vector, in a meaningful order.
2.  **Test the codes with `fct()`.** Convert the column with `fct()`, giving the approved list as `levels`. It stops with an error and names every value that is not allowed.
3.  **Pick out the offending records.** Convert the column with `factor()` instead, so that unapproved values become `NA`, and filter those rows. Separate simple typing errors from heads that do not exist at all.
4.  **Count each category.** Use `fct_count()` to see how many vouchers fall under each head. A head with very few or very many entries may need a closer look.
5.  **Use ordered factors for grades and ratings.** Store quality grades or risk ratings as an ordered factor, with `ordered = TRUE`. A filter like `risk >= "Medium"` then picks the right units, which a text comparison would not.
6.  **Tidy the level names.** Correct or combine levels with `fct_recode()` and `fct_collapse()`, for example where the same head is spelt in two ways.
7.  **Sort the levels by what matters.** Use `fct_reorder()` to sort districts by amount or median bill. The unusual district then sits at the top of the chart.
8.  **Prepare the levels for charts.** Lump the smaller vendors into *All others* with `fct_lump_n()` and `w = amount`. Sort the bars with `fct_infreq()`, and keep *All others* at the end with `fct_relevel()`.

::: callout-important
## Remember

`factor()` turns any value not in `levels` into `NA` silently. Such records can drop out of totals without anyone noticing. Use `fct()` when the list of allowed values is known, and always read the warnings R gives.
:::

## Summary of functions used

| Function | Package | What it does |
|-----------------------|----------------|--------------------------------|
| `factor()`, `as.factor()`, `levels()` | base R | Create factors and see their levels. Values not in `levels` become `NA` |
| `is.factor()`, `unclass()` | base R | Check whether a vector is a factor, and see the integers underneath |
| `factor(ordered = TRUE)`, `ordered()`, `is.ordered()` | base R | Ordered factors, which can be compared with `<` and `>=` |
| `fct()` | forcats | Create a factor, with an error for values not in `levels` |
| `fct_count()`, `fct_unique()` | forcats | Counts and unique values of a factor |
| `fct_reorder()`, `fct_reorder2()` | forcats | Order levels by another variable |
| `fct_infreq()`, `fct_rev()`, `fct_relevel()`, `fct_inorder()`, `fct_inseq()` | forcats | Order levels by frequency, reverse them, or move some by hand |
| `fct_relabel()`, `fct_recode()`, `fct_collapse()` | forcats | Change or combine level names |
| `fct_lump_n()`, `fct_lump_prop()`, `fct_lump_min()`, `fct_lump_lowfreq()` | forcats | Lump rare levels into *Other* |

: Functions used in this chapter {#tbl-fctfunctions}
