```{r}
#| include: false
source("_common.R")
```

# Factors {#sec-factors}
We often have a requirement, where we need to have our character variable representing categorical data that should take values from a fixed and known set of finite values (or categories).  Additionally sometimes these categories need to be sorted in a specific order which may not be alphabetical.  Categorical data like this plays an important role in data analytics. 

To deal with such categorical data, we have a special class `factor` in R. We met this data type briefly in @sec-fctrss. Audit data is full of categories, like expenditure heads, types of vendors, districts, risk ratings and months. So let us discuss factors in a little more detail in this chapter, and apply them to audit data at the end, in @sec-fct-audit.

**Prerequisites**

```{r}
#| message: false
#| warning: false
library(tidyverse)   # includes forcats
library(patchwork)
library(conflicted)
conflicts_prefer(dplyr::filter)
```

## Factors in base R
Factors in R are objects built over atomic data-type `integer`.  We have two primary functions to create (and coerce) factors from character vectors.

- `factor`
- `as.factor()`

Of these, `factor` provides us with full customisation as it is the basic function to create factor objects in base R.  Let us discuss both.

### Creating factors from scratch
Now, let us create factor from a character vector having values from `shirt_sizes`.

```{r}
shirt_sizes <- c("S", "M", "L", "XL", "XXL")
```

Let us say, we have 10 shirts from these sizes randomly.

```{r}
# 10 shirts with following sizes
# Notice one shirt with size small case l and one XXXL
shirts <- c("S", "s", "L", "XL", "XXXL", "S", "M", "M", "L", "L")

# Let us create a factor of shirt_sizes
shirt_f <- factor(shirts, levels = shirt_sizes)
shirt_f
```
In the output above we get `NA`s silently in place of values that were not available in our allowed values (read levels).  

Now check its type and class.

```{r}
typeof(shirt_f)
class(shirt_f)
```
We can see that `factor` class is actually built upon the `integer` class underneath and labels (taking values from levels by default) shown in the output.

```{r}
unclass(shirt_f)
```

We can, however, modify labels without modifying the levels by providing values (in a vector) to the `labels` argument.  Let us now create another factor with labels different than levels.
```{r}
set.seed(123)
# 100 Shirts
shirts2 <- sample(shirt_sizes, 100, replace = TRUE)
shirts2_f <- factor(shirts2, levels = shirt_sizes, labels = c("Small", "Medium", "Large", "Extra Large", "Extra Extra large"))
# Let's view some shirts
head(shirts2_f)
# Check its levels
levels(shirts2_f)
```

Function `summary` gives us count for each level in the given factor.

```{r}
summary(shirts2_f)
```

### Coercing objects to factor
Up to now, we created a factor from set of allowable finite values (read `levels`) and displayed them using meaningful `labels`.  Function `is.factor()` will check whether the given vector is a factor or not. On the other hand, function `as.factor` will coerce the existing character vector to a factor variable by using all the distinct values available therein as levels, but sorted alphabetically.

```{r}
is.factor(iris$Species)
shirts2_coerced <- as.factor(shirts2)
head(shirts2_coerced)
levels(shirts2_coerced)
# Check summary too
summary(shirts2_coerced)
```

This problem was not there in our earlier factor created from scratch, because there it took levels from the provided vector which we sorted meaningfully ourselves, and thus having more control.  

### Order in factors {#sec-ordfact}
To sort the factors in a meaningful way, we can actually create ordered factor.  Ordered factor can be created either by 

- using `ordered = TRUE` argument in function `factor`; or by 
- coercing a given factor to an ordered factor using function `ordered()`.  

Clearly, latter method will again order the given factor as per the levels present in the factor.

```{r}
shirts2_ordered <- factor(
  shirts2,
  levels = shirt_sizes,
  labels = c("Small", "Medium", "Large", 
             "Extra Large", "Extra Extra large"),
  ordered = TRUE
)

```

Ordering a factor in R has another benefit, we can actually perform calculations on the ordered factor.
Suppose if we want to find how many shirts do we have of sizes "Large" or greater.


```{r}
#| warning: true
# How many shirts are there of sizes L or greater?
sum(shirts2_ordered >= "Large")
# But an unordered factor gives only a warning, and NA
sum(shirts2_f >= "Large")
```

::: callout-warning
## Caution

Notice that the unordered factor did not stop with an error. It gave a warning, and the answer `NA`. In a long script, such a warning is easily missed, and the `NA` silently flows into later results. So always read the warnings, and use ordered factors wherever the order matters.
:::

We can check the given factor is an ordered factor or not using function `is.ordered()`.  E.g.
```{r}
is.ordered(shirts2_ordered)
is.ordered(shirts2_f)
```

### Functions returning factors as output
Readers may remember that in an earlier chapter we learnt about a function `cut()` which returns a factor variable as output.  In the next sections we will learn some functions which will be useful while working with factor variables, either as input or output or both.

## Factors in `forcats`
Package `forcats` which is part of core tidyverse provides us with more robust and useful ways to create and deal with factor variables.  In `forcats` we have function `fct` for creating factor variables. It will produce errors if any value is not available in the given levels, to avoid bugs/typographical errors in the code.  E.g.
```{r error=TRUE}
months_31 <- c("Jan", "Mar", "May", "Jul", "Aug", "Oxt", "Dec")
fct(months_31, levels = month.abb)
```

## Inspecting Factors
### Summarising factors
The `summary()` method works well to give the counts for each level.
```{r}
shirts2_fct <- fct(shirts2)
summary(shirts2_fct)
```

Like `count` in `dplyr`, here we have `fct_count()` to give us level wise counts and/or proportions.  The difference from `summary()` is however in output type.  Function `fct_count()` returns a `tibble` instead.  E.g.
```{r}
fct_count(shirts2_fct)
# Sort in decreasing counts
fct_count(shirts2_fct, sort = TRUE)
# Include proportions also
fct_count(shirts2_fct, sort = TRUE, prop = TRUE)
```
### Unique levels only
Function `fct_unique()` from the package, returns a factor with unique values, removing duplicates.  E.g.
```{r}
fct_unique(shirts2_fct)
```

## Order in Factors
### Default ordering in factors
Orders created in factor variables using `fct` are sorted as per the levels given in the level argument. If the argument is not supplied then it is sorted on the basis of first appearance (as against alphabetically in the `factor`), as we observed in output of above example.

To learn more functions from `forcats` we will use `gss_cat` data frame which is part of the `forcats` package itself and GSS here stands for General Social Survey.  It actually consists of many factor variables.  For some other use cases, we will also use `economics_long` data which is part of `tidyr` package.

### Reordering factors
If we analyse the (mean) number of hours spent per day on TV watching across different religions in `gss_cat`, we can see that -

```{r fig-reord, fig.show='hold', fig.align='center', fig.cap="Factors with default order"}
gss_cat |>
  summarise(tv = mean(tvhours, na.rm = TRUE), .by = relig) |>
  ggplot(aes(relig, tv)) +
  geom_col() +
  coord_flip()
```

In above case (@fig-reord), the order of `relig` factor is meaningful when sorted on the basis of summarised values of another numerical variable present in the data.  Function `fct_reorder` is helpful in these scenarios.

```
fct_reorder(
  .f,
  .x,
  .fun = median,
  ...,
  .na_rm = NULL,
  .default = Inf,
  .desc = FALSE
)
```
Where - 

- `.f` is the factor variable to be sorted.
- `.x` is the numerical variable based on which `.f` is to be sorted.
- `.fun` is the optional function (default is `median`) to be used when there are multiple values of .x for any of the level in `.f`.

So, in above example, we can re-order the levels using this function.  See @fig-reord2 -

```{r fig-reord2, fig.show='hold', fig.align='center', fig.cap="Factors with reordered levels"}
gss_cat |>
  summarise(tv = mean(tvhours, na.rm = TRUE), .by = relig) |>
  ggplot(aes(fct_reorder(relig, tv), tv)) +
  geom_col() +
  coord_flip() +
  labs(x = "Religion")
```

This function is also useful in sorting box-plots.  As an example refer @fig-reord3.
```{r fig-reord3, fig.show='hold', fig.align='center', fig.cap="Unsorted boxes with unordered levels"}
economics_long |> 
  ggplot(aes(x = value01, y = variable)) +
  geom_boxplot() +
  ggtitle("Unsorted Boxes")
```
```{r fig-reord4, fig.show='hold', fig.align='center', fig.cap="Boxes sorted on Median reordering factor levels"}
economics_long |> 
  mutate(variable = fct_reorder(variable, value01)) |> 
  ggplot(aes(x = value01, y = variable)) +
  geom_boxplot() +
  ggtitle("Boxes sorted on Median")
```

### Reordering factors with two other variables.
Sometimes, a factor variable needs to be sorted on the basis of first (or last) values of two other variables.  In such `fct_reorder2()` is useful.  As compared to `fct_reorder()` it takes an extra argument `.y` and is having syntax like

```
fct_reorder2(
  .f,
  .x,
  .y,
  .fun = last2,
  ...,
  .na_rm = NULL,
  .default = -Inf,
  .desc = TRUE
)
```
Here default function is `last2` which simply means that levels of `.f` are sorted on the basis of last values of `.y` when plotted against `.x` as in grouped line charts.  E.g. See @fig-reord5.
```{r fig-reord5, fig.show='hold', fig.align='center', fig.cap="Factors reordered on two criteria"}
default <- economics_long |> 
  filter(date < dmy("31122003"), date >= dmy("01011995")) |> 
  ggplot(aes(date, value01)) +
  geom_line(aes(group = variable, color = variable)) +
  ggtitle("Default legend")

aligned <- economics_long |> 
  filter(date < dmy("31122003"), date >= dmy("01011995")) |> 
  mutate(variable = fct_reorder2(variable, date, value01)) |> 
  ggplot(aes(date, value01)) +
  geom_line(aes(group = variable, color = variable)) +
  ggtitle("Legend aligned with \nlast values of each line")

default + aligned
```


### Changing orders of few factor labels only
Sometimes, we may have a factor whose levels are already meaningfully sorted.  E.g. income levels in `gss_cat`.  Check the plot in @fig-reord6.

```{r fig-reord6, fig.show='hold', fig.align='center', fig.cap="Default Income levels"}
gss_cat |> 
  ggplot(aes(rincome)) +
  geom_bar() +
  coord_flip()
```

The levels of income are already sorted in a meaningful way.  However, sometimes we may want to change order of a few levels only.  E.g. `Not applicable` in the @fig-reord6 which if re-levelled in the end may be more meaningful.  In such cases, we may use `fct_relevel`.  The function takes a factor variable and thereafter we may pass all those levels as arguments which we want to move in the end.  After rearranging the bars (levels), the plot will look like as in @fig-reord7

```{r fig-reord7, fig.show='hold', fig.align='center', fig.cap="Modifying levels manually"}
gss_cat |> 
  ggplot(aes(fct_relevel(rincome, "Not applicable"))) +
  geom_bar() +
  coord_flip() +
  labs(x= "Income levels")
```

### Ordering bar charts in order of frequency {#sec-infreq}
Function `fct_infreq()` sorts the levels of a factor in decreasing order of frequency, so it can be used to sort bar charts (Refer charts in @fig-fct16).

```{r fig-fct16, fig.align='center', fig.show='hold', fig.cap="Default order (left) and decreasing order of frequency (right)"}

default <- mpg |> 
  ggplot(aes(trans)) +
  geom_bar()

increasing <- mpg |> 
  ggplot(aes(fct_infreq(trans))) +
  geom_bar()
default + increasing
```


### Reversing the factor levels
Using `fct_rev()` we can reverse the order of levels in any factor. E.g. @fig-reord8.
```{r fig-reord8, fig.show='hold', fig.align='center', fig.cap="Reversing levels"}
mpg |> 
  mutate(trans = fct_rev(fct_infreq(trans))) |> 
  ggplot(aes(trans)) +
  geom_bar()
```

### Other reordering
There are two more functions which can be used to reorder factor levels -

- `fct_inorder()`: by the order in which they first appear.
- `fct_inseq()`: by numeric value of level.

Readers may explore these functions by themselves.

### More on ordering factors
In @sec-ordfact we saw how an unordered factor can be turned into an ordered factor.  This ordering can cause one side effect while plotting in ggplot2.  Ordered factor use `scale_color_viridis` by default whereas unordered factor doesn't.  See following example (Notice how the color scale has been @fig-moreord to @fig-moreord2).

```{r fig-moreord, fig.cap="Use of color scale in ordered and unordered factors"}
mtcars |> 
  mutate(cyl = ordered(factor(cyl))) |> 
  ggplot(aes(wt, mpg)) +
  geom_point(size = 5, aes(color = cyl)) +
  ggtitle("Ordered Factor")
```

```{r fig-moreord2, fig.cap="Use of color scale in ordered and unordered factors"}
mtcars |> 
  mutate(cyl = factor(cyl)) |> 
  ggplot(aes(wt, mpg)) +
  geom_point(size = 5, aes(color = cyl)) +
  ggtitle("Unordered Factor")
```


## Levels in Factors

### Modifying factor levels by applying a function

Function `fct_relabel` in `forcats` powerhouse applies a function `.fun` to each of the level in `.f` factor supplied to it.  E.g. Change the case of each of the variable name in `economics_long` (@fig-reord9).

```{r fig-reord9, fig.show='hold', fig.align='center', fig.cap="Applying a function to all labels"}
economics_long |> 
  mutate(variable = fct_relabel(variable, str_to_upper)) |> 
  ggplot(aes(x = value01, y = variable)) +
  geom_boxplot() 
```

### Modifying factor levels manually
Using function `fct_recode()` we can change the levels from the given factor manually.  We have to provide new levels manually through a sequence of named character vectors where the name gives the new level, and the value gives the old level. Levels not otherwise mentioned will be left as is. Levels can be removed by naming them `NULL`.  See Example
```{r}
x <- factor(c("apple", "bear", "banana", "dear"))
fct_recode(x, fruit = "apple", fruit = "banana")
```
To collapse multiple levels (lumping) into one we can use its cousin `fct_collapse()`.  Example
```{r}
x <- factor(c("apple", "bear", "banana", "dear"))
fct_collapse(x, fruit = c("apple", "banana"))
```
### Lump uncommon factor levels into `other`
Package `forcats` provides us a family of 4 functions that are useful in lumping together the levels meeting some given criteria.  These are

- `fct_lump_min()`: lumps levels that appear fewer than min times.
- `fct_lump_prop()`: lumps levels that appear in fewer than (or equal to) `prop` * `n` times.
- `fct_lump_n()` lumps all levels except for the n most frequent (or least frequent if `n` < 0)
- `fct_lump_lowfreq()` lumps together the least frequent levels, ensuring that `"other"` is still the smallest level.

These all functions, apart from taking factors `f` as argument, also take one or more argument, which fits into the case-

- `n` Positive `n` preserves the most common `n` values. Negative `n` preserves the least common `-n` values.
- `prop` Positive `prop` lumps values which do not appear at least `prop` of the time. Negative prop lumps values that do not appear at most `-prop` of the time.
- `min` Preserve levels that appear at least `min` number of times.
- `w` An optional numeric vector giving weights for frequency of each value (not level) in `f`.
- `other_level`: Value of level used for `"other"`(default) values. Always placed at end of levels.

Some examples-

```{r fig-fctlump, fig.show='hold', fig.align='center', fig.cap="Lumping Factors"}
# We can check the religion in `gss_cat` and followers count
gss_cat |> 
  count(relig)

# Let's restrict these to 5 religions only
gss_cat |> 
  ggplot(aes(fct_lump_n(relig, n = 5))) +
  geom_bar()
```

In @fig-fctlump we can see that there are five religions plus `"other"` category placed in last.  We may also notice that bars are not sorted yet.

We may also make use of `w` argument, if we already have our factor and its counts of levels in another vector.  See @fig-fctlump2 -

```{r fig-fctlump2, fig.show='hold', fig.align='center', fig.cap="Lumping Factors by making use of w argument"}
gss_cat |> 
  count(relig) |> 
  # Making use of `w` argument
  mutate(relig = fct_lump_n(relig, n = 5, w = n)) |> 
  ggplot(aes(relig, n)) +
  geom_col()
```

`geom_col()` draws bars whose heights are already in the data, here the counts `n`. It is the same as `geom_bar(stat = "identity")`. To sort the levels in decreasing order of frequency, we can use the function `fct_infreq()`, which we learnt in @sec-infreq. It may also take `w` argument (optionally, of course) if we have our factor levels already counted.  So, to sort the levels in @fig-fctlump2, we may make use of this function one step before lumping.  See @fig-fctlump3.

```{r fig-fctlump3, fig.show='hold', fig.align='center', fig.cap="Sorting and Lumping Factors"}
gss_cat |> 
  count(relig) |> 
  # Sorting making use of `w` argument
  mutate(relig = fct_infreq(relig, w = n),
         relig = fct_lump_n(relig, n = 5, w = n)) |> 
  ggplot(aes(relig, n)) +
  geom_col()
```

## Audit use cases {#sec-fct-audit}

### Catching codes that are not in the approved list

Expenditure is booked under heads from an approved list. A head typed wrongly, or invented, is a data quality problem, and sometimes a way of hiding an expense. `factor()` turns any value not in its `levels` into `NA` silently, as we saw with the shirt sizes. `fct()` refuses, and tells us which values are not allowed. Both behaviours are useful. Let us take a few vouchers.

```{r}
approved_heads <- c("Office expenses", "Repairs", "Supplies", "Professional fees", "Travel")

vouchers <- tibble(
  voucher_no = sprintf("PV-%03d", 1:8),
  head = c("Repairs", "Supplies", "Ofice expenses", "Travel", "Hospitality",
           "Repairs", "supplies", "Professional fees")
)
```

```{r error=TRUE}
fct(vouchers$head, levels = approved_heads)
```

`fct()` stops with an error, listing the values which are not approved. To find the vouchers concerned, we let `factor()` turn them into `NA`, and pick those rows.

```{r}
vouchers |>
  mutate(head_f = factor(head, levels = approved_heads)) |>
  filter(is.na(head_f))
```

Three vouchers have heads outside the approved list. *Ofice expenses* is a typing error and *supplies* differs only in case, but *Hospitality* is a head which does not exist in the approved list at all. It needs an explanation.

::: callout-tip
## Audit tip

Do not correct the heads in the original column. Keep the auditee's data as received, and put any corrected head in a new column. Then the list of exceptions, and how each was resolved, can go into the working papers.
:::

### Risk ratings, an ordered factor

Audit planning often rates units as Low, Medium or High risk. These ratings have a natural order. As an ordered factor, they can be compared and filtered directly.

```{r}
units <- tibble(
  unit = c("Division A", "Division B", "Division C", "Division D", "Division E"),
  risk = factor(c("Medium", "Low", "High", "Very high", "Low"),
                levels = c("Low", "Medium", "High", "Very high"), ordered = TRUE)
)

# Units of Medium risk or above, to be covered this year
units |>
  filter(risk >= "Medium") |>
  arrange(desc(risk))
```

Without the ordered factor, `"Medium"` and `"High"` would be compared as text, alphabetically, and `"High"` would wrongly come before `"Low"`.

### Top vendors and the rest

Charts of expenditure by vendor usually have too many bars. A common audit chart shows the top few vendors, and puts the rest together as *Other*. Since we are interested in *money*, not in the number of payments, we lump by the amount, using the `w` argument (@fig-fctvendor). The code below creates a register of 2,000 payments to 40 vendors. Click on it to see the code.

```{r}
#| code-fold: true
#| code-summary: "Code to create the vendor payments"
set.seed(124)
payments <- tibble(
  vendor = sample(sprintf("Vendor %02d", 1:40), 2000, TRUE,
                  prob = c(10, 8, 6, rep(1, 37))),
  amount = round(rlnorm(2000, log(25000), 1.1))
)
head(payments)
```

```{r}
#| label: fig-fctvendor
#| fig-cap: "Payments to the top 8 vendors, and all others together"
payments |>
  mutate(vendor = fct_lump_n(vendor, n = 8, w = amount, other_level = "All others"),
         vendor = fct_infreq(vendor, w = amount) |> fct_relevel("All others", after = Inf) |>
           fct_rev()) |>
  summarise(amount_lakh = sum(amount) / 1e5, .by = vendor) |>
  ggplot(aes(amount_lakh, vendor)) +
  geom_col(fill = "steelblue") +
  labs(x = "Amount paid (Rs lakh)", y = NULL) +
  theme_bw()
```

`fct_lump_n()` keeps the 8 vendors with the most money and lumps the rest. `fct_infreq()` with `w = amount` sorts the vendors by the amount paid. And `fct_relevel(..., after = Inf)` moves *All others* to the end, so that it does not look like the biggest vendor. Horizontal bars, with the vendor on the y-axis, keep long names readable. But ggplot2 draws the first level at the *bottom* of the y-axis, so we reverse the order with `fct_rev()`, to bring the biggest vendor to the top.

## A suggested workflow

1.  **Get the approved list.** Obtain the approved list of codes from the auditee or from the rules, such as expenditure heads, vendor types or districts. Keep it as a character vector, in a meaningful order.
2.  **Test the codes with `fct()`.** Convert the column with `fct()`, giving the approved list as `levels`. It stops with an error and names every value that is not allowed.
3.  **Pick out the offending records.** Convert the column with `factor()` instead, so that unapproved values become `NA`, and filter those rows. Separate simple typing errors from heads that do not exist at all.
4.  **Count each category.** Use `fct_count()` to see how many vouchers fall under each head. A head with very few or very many entries may need a closer look.
5.  **Use ordered factors for ratings.** Store risk ratings or grades as an ordered factor, with `ordered = TRUE`. A filter like `risk >= "Medium"` then picks the right units, which a text comparison would not.
6.  **Tidy the level names.** Correct or combine levels with `fct_recode()` and `fct_collapse()`, for example where the same head is spelt in two ways.
7.  **Prepare the levels for charts.** Lump the smaller vendors into *All others* with `fct_lump_n()` and `w = amount`. Sort the bars with `fct_infreq()` or `fct_reorder()`, and keep *All others* at the end with `fct_relevel()`.

::: callout-important
## Remember

`factor()` turns any value not in `levels` into `NA` silently. Such records can drop out of totals without anyone noticing. Use `fct()` when the list of allowed values is known, and always read the warnings R gives.
:::

## Summary of functions used

| Function | Package | What it does |
|-----------------------|----------------|--------------------------------|
| `factor()`, `as.factor()`, `levels()` | base R | Create factors and see their levels. Values not in `levels` become `NA` |
| `factor(ordered = TRUE)`, `ordered()`, `is.ordered()` | base R | Ordered factors, which can be compared with `<` and `>=` |
| `fct()` | forcats | Create a factor, with an error for values not in `levels` |
| `fct_count()`, `fct_unique()` | forcats | Counts and unique values of a factor |
| `fct_reorder()`, `fct_reorder2()` | forcats | Order levels by another variable |
| `fct_infreq()`, `fct_rev()`, `fct_relevel()`, `fct_inorder()` | forcats | Order levels by frequency, reverse them, or move some by hand |
| `fct_relabel()`, `fct_recode()`, `fct_collapse()` | forcats | Change or combine level names |
| `fct_lump_n()`, `fct_lump_prop()`, `fct_lump_min()`, `fct_lump_lowfreq()` | forcats | Lump rare levels into *Other* |

: Functions used in this chapter {#tbl-fctfunctions}
