```{r}
#| include: false
source("_common.R")
```

# Finding string similarity {#sec-strsim}

Comparing two numbers is easy. We can subtract one from the other, or use any of several statistical measures. Comparing two pieces of text is not so simple. Is *Geeta* the same as *Gita*? Is *M/s Sharma Traders* the same as *Sharma Trader's*? Is *Mohd. Iqbal* the same person as *Mohammad Iqbal*? A computer, checking for exact equality, will say "no" to all three.

**String similarity** methods solve this problem. They take two strings and return a score that tells how alike they are. A good method should give a high score to *Geeta* and *Gita*, but not to *Apple* and *Microsoft*. Matching strings in this approximate way is often called **fuzzy matching**.

Over several decades, many methods have been developed for this. They have different strengths and weaknesses. They fall into three broad groups.

-   **Edit-based (lexical) methods** count how many small changes, like adding, removing or swapping letters, turn one string into the other. They are good at catching typing and spelling mistakes.
-   **Token-based methods** break strings into small pieces, such as groups of letters or whole words, and compare the pieces. They are less affected by the order of words.
-   **Phonetic methods** compare how strings *sound*. They are good at catching different spellings of the same name.

## Why does an auditor need string similarity?

Names and addresses are typed by different people, at different times, in different systems. So the same person or firm is rarely written the same way everywhere. Exact matching misses such cases. Fuzzy matching finds them. Typical audit uses include these.

-   **Duplicate beneficiaries or vendors**, registered twice with slightly different names, as we saw in @sec-dup-fuzzy.
-   **Conflict of interest.** Matching the names of vendors' proprietors, partners or directors against the list of employees.
-   **Ineligible beneficiaries.** Matching a list of beneficiaries against a list of deceased persons, government employees or income-tax payers, when there is no common ID.
-   **Ghost employees.** Matching the pay roll against attendance or posting records.
-   **Linking two databases** that were never designed to be linked, such as a scheme database and a land record.

## Prerequisites

We will mainly use the `stringdist` package. For phonetic codes, we also use `phonics` and `cagmetaphone`.

```{r}
#| message: false
library(tidyverse)
library(stringdist)
```

## First, clean the strings

Before measuring similarity, we should remove differences that do not matter. Otherwise, the methods waste their effort on them. For Indian names and firm names, common steps are these.

-   **Case.** Convert everything to the same case. Most methods treat *A* and *a* as different letters.
-   **Punctuation and extra spaces.** Remove full stops, apostrophes and commas, and squeeze multiple spaces into one.
-   **Titles and common words.** Remove *Dr*, *Shri*, *Smt*, *Mr*, *Ms*, *M/s*, and words like *and sons* or *pvt ltd*, when they do not help to tell entities apart.
-   **Common short forms.** Make *Mohd* and *Md* into *Mohammad*, and similar.

Let us write a small function for this, using `stringr` functions that we learnt in @sec-strng.

```{r}
clean_name <- function(x) {
  x |>
    str_to_lower() |>
    str_remove_all("['’]") |>
    str_replace_all("[[:punct:]]", " ") |>
    str_replace_all("\\b(dr|shri|smt|mr|mrs|ms|m s|kumari|late)\\b", " ") |>
    str_replace_all("\\b(mohd|md)\\b", "mohammad") |>
    str_squish()
}

clean_name(c("Dr. Suchitra", "M/s Sharma Trader's", "Mohd. Iqbal", "Shri  R.K.  Gupta"))
```

We first remove apostrophes, so that *Trader's* becomes *traders* and not *trader s*. Then `[[:punct:]]`, which matches any other punctuation mark, is replaced by a space. `\\b` marks the boundary of a word, so that *dr* is removed only as a whole word, and not from the middle of *Indra*. We discussed such patterns in @sec-regex.

::: callout-tip
## Audit tip

Keep the original names in one column and the cleaned names in another. Match on the cleaned names, but always report the original names. The auditee will recognise *M/s Sharma Trader's*, not *sharma trader*.
:::

## Distance and similarity

`stringdist()` returns a **distance**. Zero means the strings are identical, and larger numbers mean more different. Its companion `stringsim()` returns a **similarity** between 0 and 1. One means identical, and zero means completely different. The two carry the same information, scaled differently.

```{r}
stringdist("helo", "hello", method = "lv")
stringsim("helo", "hello", method = "lv")
```

Both functions are vectorised. They compare the first element of one vector with the first of the other, the second with the second, and so on. Their main arguments are these.

-   `a` and `b` are the two strings, or vectors of strings, to be compared.
-   `method` is the method to use. The default is `osa`, explained below. Others include `lv`, `dl`, `hamming`, `lcs`, `qgram`, `cosine`, `jaccard`, `jw` and `soundex`.
-   `q` is the size of the letter groups, for the token-based methods.
-   `p` is the Winkler prefix factor, for the `jw` method.

Similarity is often easier to work with, because it is on the same 0 to 1 scale for every method, whatever the length of the strings.

## Edit-based methods

Edit-based methods count the smallest number of changes needed to turn one string into another. They match strings based on a simple model of typing errors.

### Levenshtein distance

It is named after *Vladimir Levenshtein*, who considered this distance in 1965. The Levenshtein distance between two words is the minimum number of single-character edits, that is, insertions, deletions or substitutions, required to change one word into the other. It is often simply called the *edit distance*.

For the two words *helo* and *hello*, there is a missing letter *l*. To change *helo* into *hello*, we need to insert just that one letter. So the distance is `1`, as we saw above.

### Optimal String Alignment and Damerau-Levenshtein

Consider *atharv* and its typo *ahtarv*. The letters *th* have been swapped to *ht*. Such swaps, called **transpositions**, are among the most common typing errors. Levenshtein counts a swap as two substitutions. **Optimal String Alignment** (`osa`) and **Damerau-Levenshtein** (`dl`) count it as a single edit.

```{r}
c(lv  = stringdist("atharv", "ahtarv", method = "lv"),
  osa = stringdist("atharv", "ahtarv", method = "osa"),
  dl  = stringdist("atharv", "ahtarv", method = "dl"))
```

This is why `osa` is the default method of `stringdist()`. For most typing errors, `osa` and `dl` give the same answer.

### Hamming distance

This distance is computed by placing one string over the other, and counting the positions where the letters differ. It was designed for strings of the **same length**, such as codes. For strings of different lengths, `stringdist()` returns `Inf`.

```{r}
stringdist(c("ABCPS1234K", "ABCPS1234K"),
           c("ABCPS1243K", "ABCPS123K"), method = "hamming")
```

The first pair, two PAN-like codes with two digits swapped, differs in 2 positions. The second pair has different lengths, so the distance is infinite. Hamming distance is useful for fixed-length identifiers like PAN, IFSC, vehicle numbers or account numbers of a known length.

### Longest common subsequence

The `lcs` method looks for the longest **subsequence**, that is, letters appearing in the same order, though not necessarily next to each other, that both strings share. The distance is the number of letters left out of it, in both strings together.

### Jaro and Jaro-Winkler

The **Jaro** method gives a high score to two strings if

1.  they contain the same characters, not too far from each other, and
2.  the matching characters appear in the same order.

To be exact, two characters count as matching only if they are no further apart than one less than half the length of the longer string. So if the longer string has 6 letters, a letter of the first string must be found within 2 positions of the same place in the second.

**Jaro-Winkler** (`jw`) adds a bonus when the strings *begin* alike, controlled by the argument `p`, usually set to 0.1. The idea is that people rarely make mistakes in the first few letters of a name. Without `p`, the `jw` method gives the plain Jaro distance.

```{r}
c(jaro          = stringdist("martha", "marhta", method = "jw"),
  jaro_winkler  = stringdist("martha", "marhta", method = "jw", p = 0.1))
```

Jaro-Winkler distance always lies between 0 and 1, whatever the length of the strings. This makes it easy to use one cut-off for all names, which is why it is very popular for matching names.

## Token-based methods

Token-based methods break each string into small pieces and compare the sets of pieces.

### Q-grams

A **q-gram** is a group of `q` consecutive characters. To illustrate, take `q = 3`. Groups of three letters are also called *trigrams*. For *atharv* and its possible typo *ahtarv*, the trigrams are as follows.

-   For *atharv*, `{ath, tha, har, arv}`
-   For *ahtarv*, `{aht, hta, tar, arv}`

Only one trigram, *arv*, is common to both. Two measures can be built from these sets.

-   The **q-gram distance** (`qgram`) counts the trigrams that are present in one string but not in the other. Here 3 of each string's 4 trigrams are unmatched, so the distance is 6.
-   The **Jaccard** distance (`jaccard`) divides the number of *shared* trigrams by the number of *distinct* trigrams in the two strings together. Together, there are 7 distinct trigrams, and only 1 is shared. So the Jaccard similarity is 1/7, about 14%, and the Jaccard distance is 6/7.

```{r}
c(qgram   = stringdist("atharv", "ahtarv", method = "qgram", q = 3),
  jaccard = stringdist("atharv", "ahtarv", method = "jaccard", q = 3),
  jaccard_similarity = stringsim("atharv", "ahtarv", method = "jaccard", q = 3))
```

We can see that q-grams are not very effective for transpositions. A single swap of two letters breaks several trigrams at once. The **cosine** method (`cosine`) is a close cousin, which compares how often each q-gram occurs.

### Word order

Token-based methods have one big advantage. They are not much affected by the **order** of words. In Indian records, the same person may be written as *Rajesh Kumar Sharma* in one place and *Sharma Rajesh Kumar* in another.

```{r}
pair <- c("rajesh kumar sharma", "sharma rajesh kumar")

c(osa     = stringsim(pair[1], pair[2], method = "osa"),
  jw      = stringsim(pair[1], pair[2], method = "jw", p = 0.1),
  qgram_2 = stringsim(pair[1], pair[2], method = "qgram", q = 2),
  cosine  = stringsim(pair[1], pair[2], method = "cosine", q = 2))
```

OSA, a pure edit-based measure, sees these as very different. Jaro-Winkler is in between. The token-based measures see them as nearly the same. Another simple trick, which works with any method, is to **sort the words** of each name before comparing.

```{r}
sort_words <- function(x) {
  map_chr(str_split(x, " "), \(w) paste(sort(w), collapse = " "))
}

sort_words(pair)
```

After sorting, both names become *kumar rajesh sharma*, and every method sees them as identical.

## Phonetic methods

*Phonetic matching methods* compare strings by how they sound. Consider *Geeta* and *Gita*. They sound alike, so one person may write *Geeta* and another *Gita*. Neither is a misspelling of the other. They simply sound the same.

### Soundex

Created by *Robert Russell* and *Margaret King Odell* in 1918, this method was meant to match names and surnames based on the basic rules of English pronunciation. Similar-sounding names get the same code. The code keeps the first letter, followed by three digits for the next few consonant sounds. `stringdist()` gives `0` when two strings have the same Soundex code, and `1` otherwise.

```{r}
stringdist(c("geeta", "susheel", "ram"), c("gita", "sushil", "rama"), method = "soundex")
```

Because Soundex looks at only the first few sounds, long names that merely *begin* alike also match. We saw this problem with firm names in @sec-dup-fuzzy.

### Metaphone

Developed by *Lawrence Philips* in 1990, Metaphone is more accurate than Soundex, because it takes groups of letters into account, such as *ph* sounding like *f*. Its disadvantage shows up with names that are not English, as it is based on the rules of English pronunciation. The `phonics` package calculates it.

```{r}
library(phonics)

metaphone(c("geeta", "gita", "susheel", "sushil", "laxmi", "lakshmi"))
```

*Geeta* and *Gita* get the same code, and so do *Susheel* and *Sushil*. But *Laxmi* and *Lakshmi* do not. Metaphone codes *x* as *KS* and *sh* as *X*, so the two names get *LKSM* and *LKXM*. Such variations, very common in Indian names, are better handled during cleaning, for example by replacing *ksh* with *x* in both lists before matching.

`metaphone()` works on single words. It cannot handle spaces or other special characters, so we apply it to cleaned single words, or to each word of a name separately.

### Double Metaphone

Following Metaphone, *Philips* also designed the **Double Metaphone**. As its name suggests, it returns two codes, a *primary* and a *secondary* one, so there are more chances to match. At the same time, it means a higher chance of a false match. There are three matching levels.

-   Primary code against primary code is the **strongest** match.
-   Secondary code against primary code is a **normal** match.
-   Secondary code against secondary code is the **weakest** match.

*Philips* later refined it further, as **Metaphone 3**. That algorithm is, however, proprietary and not open-source.

Double Metaphone can be calculated with the function `DoubleMetaphone()` from the `PGRdup` package. The function is vectorised. It returns a list of two vectors, `primary` and `alternate`.

```{r}
suppressPackageStartupMessages(library(PGRdup))

DoubleMetaphone(c("geeta", "gita", "susheel", "sushil"))
```

::: callout-warning
## `PGRdup` keeps only four characters

`DoubleMetaphone()` of `PGRdup` cuts every code to its first **four** characters. This was the default in the original design of the algorithm. It works well for short single names like the ones above. But longer names, and names of firms, get the same code merely because they *begin* alike.

```{r}
DoubleMetaphone(c("Bhattacharya", "Bhattacharjee", "Sharma Traders", "Sharma Transport"))$primary
```

Both surnames get `PTKR`. Both firms get `XRMT`. In audit work, where we compare full names of beneficiaries or vendors, this can produce many false matches.
:::

### Full length Double Metaphone with `cagmetaphone`

To overcome this limitation, an R package `cagmetaphone` was developed by *Atharv Tyagi* during an internship under the author's guidance. It is a port of the Python reference implementation of Double Metaphone. It keeps codes of up to 32 characters instead of four. The package is not on CRAN yet, and can be installed from GitHub.

```{r}
#| eval: false
# install.packages("remotes")
remotes::install_github("AtharvTyagi1805/cagmetaphone")
```

It has two functions. `double_metaphone()` takes a single string and returns its two codes. `double_metaphone_vec()` takes a vector of strings and returns a data frame with columns `primary` and `secondary`.

```{r}
library(cagmetaphone)

double_metaphone("Guptha")

double_metaphone_vec(c("Bhattacharya", "Bhattacharjee", "Sharma Traders", "Sharma Transport"))
```

Now the four names get four different primary codes. For short names, the codes are the same as those from `PGRdup`.

Let us apply the three matching levels to a few pairs of names. We write a small function, so that we can use it again later.

```{r}
dm_match <- function(a, b) {
  dm_a <- double_metaphone_vec(a)
  dm_b <- double_metaphone_vec(b)
  case_when(
    dm_a$primary == dm_b$primary ~ "strong",
    dm_a$secondary == dm_b$primary | dm_a$primary == dm_b$secondary ~ "normal",
    dm_a$secondary == dm_b$secondary ~ "weak",
    TRUE ~ "no match"
  )
}

nameset1 <- c("Geeta", "Susheel", "Ram", "Dr. Suchitra")
nameset2 <- c("Gita", "Sushil", "Rama", "Suchitra")

tibble(nameset1, nameset2,
       raw     = dm_match(nameset1, nameset2),
       cleaned = dm_match(clean_name(nameset1), clean_name(nameset2)))
```

On the raw names, *Dr. Suchitra* does not match *Suchitra*. The title "Dr." changes the beginning of the code, and phonetic codes depend heavily on how a name begins. After cleaning with our `clean_name()` function, the title is gone and the two match. This shows, once again, why cleaning comes first.

## Comparing the methods

Let us now put all the methods side by side, on pairs of names that show typical Indian variations. We use similarities, so that all methods are on the same 0 to 1 scale.

```{r}
pairs <- tribble(
  ~name_1,                ~name_2,                 ~variation,
  "geeta",                "gita",                  "Spelling (ee/i)",
  "lakshmi devi",         "laxmi devi",            "Spelling (ksh/x)",
  "atharv",               "ahtarv",                "Typing (swap)",
  "suresh chandra",       "suresh chander",        "Spelling (ending)",
  "rajesh kumar sharma",  "sharma rajesh kumar",   "Word order",
  "r k gupta",            "ram kishan gupta",      "Initials",
  "anil kumar",           "sunil kumar",           "Different person"
)

pairs |>
  mutate(osa    = stringsim(name_1, name_2, method = "osa"),
         jw     = stringsim(name_1, name_2, method = "jw", p = 0.1),
         qgram  = stringsim(name_1, name_2, method = "qgram", q = 2),
         dmeta  = dm_match(name_1, name_2)) |>
  mutate(across(where(is.double), \(x) round(x, 2))) |>
  knitr::kable()
```

Some lessons stand out from this table.

-   **No single method wins everywhere.** Jaro-Winkler handles spelling and typing well. Q-grams handle word order. The phonetic code handles *Geeta* and *Gita*, but not *Lakshmi* and *Laxmi*, which need a cleaning rule.
-   **Initials are hard for every method.** *R K Gupta* and *Ram Kishan Gupta* may well be the same person, but no method sees them as very similar. Matching on the surname, and checking initials separately, works better.
-   **Similar is not the same.** *Anil Kumar* and *Sunil Kumar* get high scores from all the lexical methods, yet they are different people. Here the phonetic code correctly says "no match", which shows the value of using two kinds of method together. A high score is a lead, not proof.

@tbl-strsimguide gives a rough guide.

| Situation | Suggested method |
|-------------------------------|-------------------------------|
| Typing mistakes in names | Jaro-Winkler (`jw`, with `p = 0.1`), or OSA |
| Codes of fixed length (PAN, IFSC) | Hamming, or OSA |
| Words in a different order | Q-grams or cosine, or sort the words first |
| Different spellings of the same sounding name | Double Metaphone (`cagmetaphone`) |
| Long text, like addresses | Q-grams or cosine |

: Choosing a string similarity method {#tbl-strsimguide}

## Matching two lists, an audit use case

So far we have compared names in pairs. In audit, we usually have two **lists**, and want to find which names in one list resemble names in the other. Let us take a common example, a check for **conflict of interest**. We have a list of employees of an office, and a list of proprietors of firms that supplied goods to it. Both lists are made up.

```{r}
employees <- tibble(
  emp_id   = c("E01", "E02", "E03", "E04", "E05", "E06", "E07"),
  emp_name = c("Rakesh Kumar Verma", "Smt. Sunita Devi", "Mohd. Iqbal Khan",
               "Pradeep Singh", "Kavita Sharma", "Anand Prakash", "Neha Gupta")
)

proprietors <- tibble(
  firm      = c("Verma Enterprises", "Sunrise Traders", "Iqbal & Sons",
                "Singh Suppliers", "Kavya Stationers", "Gupta Agencies"),
  proprietor = c("Rakesh Kr. Verma", "Arvind Mehta", "Mohammad Iqbal Khan",
                 "Pradip Singh", "Kavya Sharma", "Neeraj Gupta")
)
```

### Method 1, compare every pair

With small lists, we can simply compare every employee with every proprietor. `cross_join()`, which we used in @sec-dup-fuzzy, makes all the pairs. We compare the *cleaned* names, keep the pairs above a chosen similarity, and add the Double Metaphone match level as a second opinion.

```{r}
candidates <- employees |>
  cross_join(proprietors) |>
  mutate(jw = stringsim(clean_name(emp_name), clean_name(proprietor),
                        method = "jw", p = 0.1),
         sound = dm_match(clean_name(emp_name), clean_name(proprietor))) |>
  filter(jw >= 0.85) |>
  arrange(desc(jw))

candidates |>
  mutate(jw = round(jw, 2)) |>
  select(emp_name, proprietor, firm, jw, sound)
```

Let us read the result.

-   *Rakesh Kumar Verma* and *Rakesh Kr. Verma* are almost certainly the same person, written with a short form. The phonetic code misses this pair, because *Kr* sounds nothing like *Kumar*. Adding a rule to our cleaning function, to expand *kr* to *kumar*, would fix it. So a "no match" from one method does not prove that two names are different.
-   *Mohd. Iqbal Khan* and *Mohammad Iqbal Khan* become identical after cleaning, since our `clean_name()` function expands *Mohd*.
-   *Pradeep Singh* and *Pradip Singh* are very close, and sound the same. A strong lead.
-   *Kavita Sharma* and *Kavya Sharma*, and *Neha Gupta* and *Neeraj Gupta*, are close in spelling, but are probably different people. Note that the phonetic code does not match them. Such pairs show why every match must be checked against other details.

### Method 2, `amatch()`

`stringdist` has a ready function for finding, for each string, the closest match in another list. `amatch(x, table, maxDist)` returns, for each element of `x`, the position of the closest element in `table`, if it is within `maxDist`. Otherwise it returns `NA`.

```{r}
pos <- amatch(clean_name(employees$emp_name), clean_name(proprietors$proprietor),
              method = "jw", p = 0.1, maxDist = 0.15)

employees |>
  mutate(matched_proprietor = proprietors$proprietor[pos],
         matched_firm       = proprietors$firm[pos])
```

A Jaro-Winkler distance of 0.15 is the same as a similarity of 0.85, the cut-off we used above. `amatch()` is fast and convenient, but it returns only the *single* closest match for each name. When there may be more than one candidate, the pairwise method is safer.

::: callout-tip
## Audit tip

A name match alone is weak evidence, especially for common names. For every candidate pair, compare a second detail, such as father's name, date of birth, address, mobile number, PAN or bank account. In a conflict-of-interest check, a matching *bank account* or *address* between an employee and a vendor is often stronger evidence than the name itself.
:::

## Choosing a cut-off

Every fuzzy match needs a cut-off, and every cut-off is a trade-off.

-   A **strict** cut-off, say a Jaro-Winkler similarity of 0.95, finds few false matches, but misses many true ones.
-   A **loose** cut-off, say 0.80, finds most true matches, but also many false ones, which must then be checked by hand.

A practical approach is to start loose, sort the pairs by score, and read down the list. You will usually see that above a certain score most pairs are real, and below another score most are not. Set the cut-off between the two, keeping in mind how many pairs your team can actually check.

## Large lists

Comparing every pair becomes slow when the lists are large. Two lists of 10,000 names each give ten crore pairs. Three ideas help.

-   **Blocking.** Compare names only within smaller groups that must match anyway, such as the same district, the same year of birth, or the same first letter of the surname. Blocks can also be formed by phonetic codes.
-   **Inequality joins.** When comparing a list with itself, compare each pair only once, as we did with `inner_join(..., join_by(id < id))` in @sec-dup-fuzzy.
-   **`stringdistmatrix()`.** It calculates all distances between two vectors in one call, and uses several processor cores. The argument `nthread` controls how many.

::: callout-warning
## Privacy

Names, addresses and identity numbers are personal data. Do fuzzy matching only on secure systems approved for such data. Never paste such lists into online tools or AI chatbots.
:::

## A suggested workflow

1.  **Define the question.** Which two lists, or which one list, and what would a match mean?
2.  **Clean the names.** Case, punctuation, titles, common short forms and, where useful, word order.
3.  **Choose a method** suited to the kind of variation you expect, and a second method as a check.
4.  **Block** if the lists are large.
5.  **Compare and score**, with `stringsim()`, `cross_join()` or `amatch()`.
6.  **Choose a cut-off** by reading down the sorted list of pairs.
7.  **Confirm each candidate** with a second identifier, like date of birth, address or bank account.
8.  **Report the original names**, with the evidence for each match.

::: callout-important
## Remember

Fuzzy matching finds names that *look* or *sound* alike. It cannot tell whether they belong to the same person. Two different people can have very similar names, and one person can be written in very different ways. Treat every match as a question to be answered from other evidence.
:::

## Summary of functions used

| Function | Package | What it does |
|-----------------------|----------------|--------------------------------|
| `str_to_lower()`, `str_replace_all()`, `str_squish()` | stringr | Clean names before matching |
| `stringdist(a, b, method)` | stringdist | Distance between strings, 0 for identical |
| `stringsim(a, b, method)` | stringdist | Similarity between strings, from 0 to 1 |
| `amatch(x, table, maxDist)` | stringdist | Closest match for each string in another list |
| `stringdistmatrix()` | stringdist | All distances between two vectors at once |
| `metaphone()` | phonics | Metaphone code of a single word |
| `DoubleMetaphone()` | PGRdup | Double Metaphone codes, cut to four characters |
| `double_metaphone_vec()` | cagmetaphone | Full length Double Metaphone codes |
| `cross_join()` | dplyr | All pairs of rows from two tables |

: Functions used in this chapter {#tbl-strsimfunctions}
