```{r}
#| include: false
source("_common.R")
```

# Reverse geocoding {#sec-revgeocoding}

## Introduction

**Geocoding**, which we learnt in @sec-geocoding, is the process of converting a human-readable address into geographic coordinates, such as latitude and longitude. It lets us place addresses on a map.

**Reverse geocoding**, on the other hand, is the process of converting geographic coordinates (latitude and longitude) into a human-readable address, such as a street, village or city, district, state and country. In other words, it determines the address of a place from its coordinates.

Both are important in geospatial data analysis. They allow us to link spatial information with non-spatial information, such as demographic data or land records, and to see patterns and relationships based on location.

Reverse geocoding is especially useful in forensic data analysis. Mobile phones and handheld devices now record the coordinates of almost every activity. A photo of a newly built house, an attendance marked by a field worker, a ration distributed through a point-of-sale machine, a payment made at an ATM. Each comes with a latitude and a longitude. Reverse geocoding turns these numbers into places that an auditor can read and check, such as "is this photo really from the village where the work was claimed?".

## Widely used databases

**OpenStreetMap** (OSM) is one of the most popular and widely used databases for geocoding and reverse geocoding. OSM is a free and open map of the world, created and maintained by a community of volunteers. It has a rich and detailed set of spatial data, including street names, addresses and other points of interest. Its free service, called *Nominatim*, allows only light use, about one request per second.

Other popular services include **Google Maps**, **ArcGIS** and several Indian map providers. They provide geocoding and reverse geocoding through paid APIs, and are widely used in web and mobile applications.

::: callout-warning
## Privacy

Coordinates of a beneficiary's home are personal data, just like the address. Do not send such coordinates to public services. Use only services approved by your organisation. In this chapter, we use the coordinates of well-known monuments only.
:::

## Example in R

### Prerequisites

```{r}
#| message: false
#| warning: false
library(tidyverse)
library(tidygeocoder)
```

### Function to be used

We will use the `reverse_geo()` function from `tidygeocoder`. Its main arguments are these.

```
reverse_geo(
  lat,
  long,
  method = "osm",
  address = "address",
  full_results = FALSE,
  ...
)
```

-   `lat` and `long` are, as the names suggest, the latitude and longitude of the place whose address is to be found.
-   `method` is the service to be used. The default is `osm`. Other services available include `arcgis`, `geocodio` and `google`.
-   `address` is the name of the output column holding the address.
-   `full_results` is `FALSE` by default. If set to `TRUE`, it returns all available details from the service, such as the city, district, state and postcode in separate columns.
-   For other arguments, please check the help of the function.

### A single place

Let us find the address of the Taj Mahal, from its coordinates.

```{r}
#| message: false
reverse_geo(lat = 27.1751, long = 78.0421, method = "osm")
```

::: callout-tip
## Coordinates need care

Each digit of a coordinate matters. One degree of latitude is about 111 km, so the third decimal place is about 100 metres. If the longitude above is mistyped as 78.421 instead of 78.0421, the "Taj Mahal" turns out to be a village in Firozabad district, about 40 km away. When coordinates are typed by hand, always check a few of them on a map.
:::

### Several places at once

`reverse_geo()` is vectorised. We can give it vectors of latitudes and longitudes, and it returns one row for each. Let us build a sample dataset of four places of interest in India.

```{r}
sample_places <- tibble(
  lat  = c(27.1751, 24.8318, 27.1795, 28.5245),
  long = c(78.0421, 79.9199, 78.0211, 77.1855)
)

knitr::kable(sample_places)
```

When the coordinates are already in a data frame, the companion function **`reverse_geocode()`** is even more convenient. We simply name the latitude and longitude columns, and it adds the address as a new column.

```{r}
#| message: false
results <- sample_places |>
  reverse_geocode(lat = lat, long = long, method = "osm", full_results = TRUE)

results |>
  select(lat, long, address, any_of(c("state_district", "state", "postcode"))) |>
  knitr::kable()
```

With `full_results = TRUE`, the result has many more columns, one for each part of the address. Which columns appear depends on the place. A monument in a city and a spot in a village give different parts. `any_of()` selects only those of the listed columns which actually exist, so that the code does not fail if one is missing.

## Audit use case 1, checking geotagged photos

Many government schemes ask field staff to upload a geotagged photo of each completed work, such as a house, a pond or a road. The work is claimed to be at a particular place. Reverse geocoding the photo's coordinates tells us where the photo was *actually* taken.

Let us use our four coordinates as the geotags of four such photos. Each comes with the place where the work was *claimed* to be. One of the claims is wrong.

```{r}
photos <- sample_places |>
  mutate(photo_id      = c("PH-101", "PH-102", "PH-103", "PH-104"),
         claimed_place = c("Agra", "Khajuraho", "Mathura", "Delhi"),
         .before = 1)

photos
```

We reverse geocode the photos, and check whether the address returned contains the claimed place.

```{r}
#| message: false
photos_checked <- photos |>
  reverse_geocode(lat = lat, long = long, method = "osm") |>
  mutate(matches_claim = str_detect(str_to_lower(address), str_to_lower(claimed_place)))

photos_checked |>
  select(photo_id, claimed_place, address, matches_claim)
```

Photo PH-103 was claimed to show a work in Mathura. But its geotag is in Agra, more than 50 km away. Such a photo needs explanation. It may be a photo of a different work, or of no work at all.

::: callout-warning
## Caution

Comparing text in this way is a rough first check. Place names are spelt in many ways, such as *Gurgaon* and *Gurugram*, and the address returned by the service may use a different name, or a larger or smaller unit, than the claim. Treat a mismatch as a question to be checked, not as a finding. For large data, it is better to compare codes, such as district codes, or to use the map-based checks of @sec-maps, where we measured the distance between a geotag and the site directly.
:::

## Audit use case 2, checking geotags without sending them anywhere {#sec-revgeo-offline}

Reverse geocoding needs an online service, and we cannot send the coordinates of beneficiaries' homes or works to a public one. But many problems in geotags can be found without any service at all, using only the numbers themselves. Let us take the geotags of 1,000 works completed under a scheme in a district of Rajasthan. We have planted five common problems.

-   15 geotags with **latitude and longitude swapped**, a frequent error when data passes between systems.
-   8 geotags of **0, 0**, which is what some devices record when the GPS has failed.
-   20 geotags with only **two decimal places**, which suggests they were typed by hand, not captured by the phone.
-   3 geotags **outside India** altogether.
-   40 works in different villages, all geotagged at **exactly the same spot**. This is the block office, from where the photos were actually uploaded.

```{r}
#| code-fold: true
#| code-summary: "Code to create the geotags"
set.seed(7)
n <- 1000
geotags <- tibble(work_id = sprintf("WK%04d", 1:n),
                  village = sprintf("Village-%02d", sample(1:60, n, TRUE)),
                  lat  = round(runif(n, 26.60, 27.20), 6),
                  long = round(runif(n, 75.50, 76.10), 6))

problem_rows <- sample(1:n, 86)
swapped <- problem_rows[1:15]
zeros   <- problem_rows[16:23]
typed   <- problem_rows[24:43]
office  <- problem_rows[44:83]
abroad  <- problem_rows[84:86]

geotags <- geotags |>
  mutate(
    tmp  = lat,
    lat  = if_else(row_number() %in% swapped, long, lat),
    long = if_else(row_number() %in% swapped, tmp, long),
    lat  = if_else(row_number() %in% zeros, 0, lat),
    long = if_else(row_number() %in% zeros, 0, long),
    lat  = if_else(row_number() %in% typed, round(lat, 2), lat),
    long = if_else(row_number() %in% typed, round(long, 2), long),
    lat  = if_else(row_number() %in% office, 26.912437, lat),
    long = if_else(row_number() %in% office, 75.787271, long),
    lat  = if_else(row_number() %in% abroad, lat - 30, lat)
  ) |>
  select(-tmp)
```

Let us add one column for each check. India lies roughly between latitudes 6° and 38° north, and longitudes 68° and 98° east. A swapped point has its latitude in the range of India's longitudes, and the other way round.

```{r}
count_decimals <- function(x) {
  str_length(coalesce(str_extract(as.character(x), "(?<=\\.)\\d+"), ""))
}

geotag_checks <- geotags |>
  add_count(lat, long, name = "same_spot") |>
  mutate(
    zero_point    = lat == 0 & long == 0,
    swapped       = between(lat, 68, 98) & between(long, 6, 38),
    outside_india = !zero_point & !swapped &
                    !(between(lat, 6, 38) & between(long, 68, 98)),
    low_precision = !zero_point & (count_decimals(lat) <= 2 | count_decimals(long) <= 2),
    shared_spot   = same_spot >= 5 & !zero_point
  )

geotag_checks |>
  summarise(across(zero_point:shared_spot, sum))
```

`count_decimals()` uses a look-behind, `(?<=\\.)`, which we learnt in @sec-regex, to take the digits after the decimal point and count them. A phone usually gives five or six decimals. Two decimals mean an accuracy of only about a kilometre.

Every planted problem has been found. The most serious is the shared spot. Let us look at it.

```{r}
geotag_checks |>
  filter(shared_spot) |>
  summarise(works = n(), villages = n_distinct(village), .by = c(lat, long))
```

```{r}
#| include: false
spot <- geotag_checks |> filter(shared_spot) |> summarise(works = n(), villages = n_distinct(village))
```

`r spot$works` works, in `r spot$villages` different villages, have *exactly* the same geotag, to the sixth decimal place. No two houses or ponds can be at the same spot. These photos were taken, or at least uploaded, at one place. This one location, unlike a beneficiary's home, is an office, so it can be reverse geocoded to confirm what it is.

::: callout-tip
## Audit tip

Run these checks first, on the whole data, before any mapping or reverse geocoding. They need no internet, raise no privacy concern, and often find the most important problems. Swapped and zero coordinates are data errors to be corrected. Typed coordinates and shared spots suggest that the photo was not taken at the site, and need verification in the field.
:::

## A suggested workflow

1.  **Use an approved service** for any coordinates that relate to people.
2.  **Check the coordinates offline first**, for swapped, zero, typed, outside-India and shared values, as in the second use case.
3.  **Reverse geocode each unique location once**, using `reverse_geocode()` on a data frame, and save the results.
4.  **Compare** the returned place with the claimed place, by name or preferably by code.
5.  **Measure the distance** between the geotag and the claimed site, where the site's coordinates are known.
6.  **Verify** every mismatch with records or a field visit.

::: callout-important
## Remember

A geotag tells us where the *phone* was, not necessarily where the *work* is. And geotags can be faked. A correct geotag supports a claim, but does not prove it. A wrong one is a strong reason to look closer.
:::

## Summary of functions used

| Function | Package | What it does |
|-----------------------|----------------|--------------------------------|
| `reverse_geo(lat, long, method)` | tidygeocoder | Address for vectors of coordinates |
| `reverse_geocode(.tbl, lat, long)` | tidygeocoder | Adds the address to a data frame of coordinates |
| `full_results = TRUE` | tidygeocoder | Returns each part of the address in a separate column |
| `any_of()` | dplyr | Selects the listed columns which exist |
| `str_detect()`, `str_to_lower()` | stringr | Checks whether the claimed place appears in the address |
| `between()`, `add_count()` | dplyr | Range checks on coordinates, and counting records at the same spot |

: Functions used in this chapter {#tbl-revgeofunctions}
