```{r}
#| include: false
source("_common.R")
```

# Maps in R {#sec-maps}

Almost everything that government does happens *somewhere*. A hospital, a school, a road, a ration shop, a beneficiary's house. Records usually carry this location, as a district name, a pin code, or increasingly as exact coordinates captured by a mobile phone. When we put such data on a map, patterns appear that no table can show. Which areas are well covered, which are left out, and which records do not fit where they claim to be.

R can work with maps as easily as with tables. In this chapter, we will learn to read map data, draw maps, join data to locations, and measure distances. As always, we will apply each skill to an audit question.

## Where can maps help an auditor?

-   **Coverage and equity.** Are benefits reaching every district, or are some areas left out? A map of beneficiaries per lakh population shows this at a glance.
-   **Verifying geotagged evidence.** Many schemes now require geotagged photos, of houses built, assets created, or inspections done. A photo whose location does not match the claimed site is a red flag.
-   **Distance checks.** Patients travelling unusually far to a hospital, students enrolled in a school far from their homes, or travel claims for distances that do not match the map.
-   **Location mismatches.** A beneficiary recorded in one district but living in another, or a vendor whose address is far from where supplies were delivered.
-   **Spatial clusters.** Many claims, complaints or failures concentrated in one area.

## Basic ideas

### Types of map data

Map data comes in two main kinds. **Vector** data describes features by their shape.

-   **Points**, such as the location of a hospital, a school or a photo.
-   **Lines**, such as roads, rivers or canals.
-   **Polygons**, which are closed shapes, such as the boundary of a district or a plot of land.

**Raster** data describes the earth as a grid of cells, like a photograph. Satellite images and land-use maps are raster data. In this chapter we will work only with vector data, which is what most audit work needs.

### File formats

Vector map data is commonly stored in these formats.

-   **Shapefile.** The oldest and most common format. A single "shapefile" is really a set of files with the same name and different extensions, such as `.shp` (the shapes), `.shx` (an index), `.dbf` (the attribute table) and `.prj` (the coordinate system). All of them must be kept together. If the `.shp` file is missing, the shapes are lost.
-   **GeoPackage** (`.gpkg`). A modern, single-file format, which can hold several layers.
-   **GeoJSON** (`.geojson`). A text format, popular on the web.

### Coordinate reference systems

Every location is described by a pair of numbers. But numbers mean nothing unless we know the system they belong to. This system is called the **coordinate reference system** or **CRS**.

-   **Geographic** systems use **latitude and longitude**, in degrees. The most common one is called **WGS 84**, used by GPS and mobile phones. Its code is `EPSG:4326`.
-   **Projected** systems flatten the round earth onto a plane, and use **metres or kilometres**. Distances and areas are easier to measure in them.

We will learn more about CRS, with real latitudes and longitudes, in the chapter on geocoding. In this chapter, our map uses a simple flat grid in kilometres.

::: callout-warning
## Maps of India in publications

When publishing a map that shows the boundaries of India or its states, use boundaries from an authoritative source, such as those published by the Survey of India, and follow the applicable guidelines. Many freely available world maps show India's boundaries incorrectly. That is one reason why the examples in this chapter use an imaginary state.
:::

## Prerequisites

The package **`sf`**, short for *simple features*, is the standard tool for vector map data in R. An `sf` object is simply a data frame with a special column, usually called `geometry`, that holds the shape of each row. So everything we know about `dplyr` works on it, and `ggplot2` can draw it directly.

```{r}
#| message: false
library(tidyverse)
library(sf)
```

## Our data, a map of the imaginary state

We continue with the imaginary *Sarvajan Swasthya Yojana* (SSY) of the earlier chapters. The scheme runs in an imaginary state, *Sarvajan Pradesh*, with eight districts. All shapes and locations are made up, and are created by `tools/make_ssy_geo_data.R` in the book's repository.

The map is stored in a GeoPackage file. `st_layers()` lists the layers in it.

```{r}
st_layers("Data/ssy_map.gpkg")
```

There are two layers, the outline of the state and its districts. `st_read()` reads a layer.

```{r}
state     <- st_read("Data/ssy_map.gpkg", layer = "state", quiet = TRUE)
districts <- st_read("Data/ssy_map.gpkg", layer = "districts", quiet = TRUE)

districts
```

The printout shows that this is an `sf` object with 8 *features* (rows), each a polygon. Besides the `geom` column, it has the district name and its population in lakh. Note the line *Projected CRS*. Let us check the units of our coordinate system.

```{r}
st_crs(districts)$units
```

So all coordinates and distances are in kilometres.

## Drawing maps

The quickest way to see an `sf` object is the base R `plot()` function. It draws one small map for each attribute column. Here we draw just the geometry, with `st_geometry()`.

```{r}
#| label: fig-mapbase
#| fig-cap: "Districts of the imaginary state, drawn with plot()"
plot(st_geometry(districts))
```

For better maps, we use `ggplot2`. The geom **`geom_sf()`** draws `sf` objects. It needs no `x` or `y` aesthetic, because the geometry column already holds the shapes. `geom_sf_text()` places a label at the centre of each shape (@fig-mapdist).

```{r}
#| label: fig-mapdist
#| fig-cap: "The eight districts of Sarvajan Pradesh"
ggplot(districts) +
  geom_sf(aes(fill = district), colour = "white", show.legend = FALSE) +
  geom_sf_text(aes(label = district), size = 3.5) +
  theme_void()
```

`theme_void()` removes the axes and gridlines, which are rarely needed on a map.

## Adding points from a table

Locations often come as plain columns of numbers in a table. Our hospital locations are in a CSV file, with two columns, `x_km` and `y_km`.

```{r}
hospitals <- read_csv("Data/ssy_hospitals.csv", show_col_types = FALSE)

hosp_locations <- read_csv("Data/ssy_hospital_locations.csv", show_col_types = FALSE)
head(hosp_locations, 3)
```

To make them map points, we use `st_as_sf()`. We tell it which columns hold the coordinates, and which CRS they are in. Since these coordinates are on the same grid as our districts, we use the district map's CRS.

```{r}
hosp_sf <- hosp_locations |>
  left_join(hospitals, by = "hospital_id") |>
  st_as_sf(coords = c("x_km", "y_km"), crs = st_crs(districts))

hosp_sf
```

Now we can draw the hospitals on the map, as another layer (@fig-maphosp).

```{r}
#| label: fig-maphosp
#| fig-cap: "Empanelled hospitals of the scheme"
ggplot() +
  geom_sf(data = districts, fill = "grey95", colour = "grey60") +
  geom_sf(data = hosp_sf, aes(colour = type), size = 1.8) +
  labs(colour = "Hospital type") +
  theme_void()
```

## Joining by location

In ordinary joins, rows are matched by a common value, like a hospital ID. In a **spatial join**, rows are matched by *location*. `st_join()` finds, for each point, the polygon it lies in, and attaches that polygon's columns.

Our hospital table already has a `district` column, entered when the hospital was empanelled. A spatial join lets us check whether each hospital really lies in the district recorded against it.

```{r}
hosp_check <- hosp_sf |>
  st_join(districts |> select(district_on_map = district))

hosp_check |>
  st_drop_geometry() |>
  count(match = district == district_on_map)
```

`st_drop_geometry()` turns the `sf` object back into an ordinary data frame, which prints more neatly. Here every hospital lies in its recorded district. In real data, such a check often finds mismatches. Some are errors in entering the district. Others may be beneficiaries or units recorded in one district to claim a benefit available only there.

## A map of claims per lakh population

A **choropleth** is a map where each area is coloured according to a value. Let us colour each district by the number of claims raised by its hospitals, per lakh population.

First we count claims by the district of the hospital, and join the counts to the district map with an ordinary `left_join()`. The result is still an `sf` object, so it can be drawn.

```{r}
claims <- read_csv("Data/ssy_claims.csv", show_col_types = FALSE)

claims_by_district <- claims |>
  left_join(hospitals |> select(hospital_id, district), by = "hospital_id") |>
  count(district, name = "claims")

district_map <- districts |>
  left_join(claims_by_district, by = "district") |>
  mutate(claims_per_lakh = claims / population_lakh)

district_map |>
  st_drop_geometry() |>
  arrange(desc(claims_per_lakh))
```

```{r}
#| label: fig-mapchoro
#| fig-cap: "Claims per lakh population, by district of the hospital"
ggplot(district_map) +
  geom_sf(aes(fill = claims_per_lakh), colour = "white") +
  geom_sf_text(aes(label = district), size = 3) +
  scale_fill_distiller(palette = "YlOrRd", direction = 1) +
  labs(fill = "Claims per\nlakh population") +
  theme_void()
```

The darker districts in @fig-mapchoro have many more claims for their population. This alone is not a finding. A district may have more claims because it has more hospitals, or because patients come to it from other districts. But it shows the auditor where to start asking questions.

::: callout-tip
## Audit tip

Always map **rates**, like claims per lakh population or beneficiaries per thousand households, rather than raw counts. A map of raw counts mostly shows where more people live. We met the same point while comparing groups in @sec-sentiment.
:::

## Audit use case 1, how far do patients travel?

Most patients go to a hospital near their home. A hospital whose patients come mostly from far away deserves a closer look. Patients may be brought in by agents, or the claims may be for patients who never visited at all.

For a sample of claims, we have the location of the patient's home. Let us turn these into points.

```{r}
homes <- read_csv("Data/ssy_patient_homes.csv", show_col_types = FALSE) |>
  st_as_sf(coords = c("home_x_km", "home_y_km"), crs = st_crs(districts))

nrow(homes)
```

`st_distance()` measures the distance between features. With `by_element = TRUE`, it measures between the first home and the first hospital, the second home and the second hospital, and so on, instead of between every pair. So we first line up, for each home, the location of its hospital.

```{r}
hospital_of_home <- hosp_sf[match(homes$hospital_id, hosp_sf$hospital_id), ]

homes <- homes |>
  mutate(distance_km = as.numeric(st_distance(homes, hospital_of_home, by_element = TRUE)))

summary(homes$distance_km)
```

`st_distance()` returns the distances with their units attached. `as.numeric()` removes the units, so that we can use them like ordinary numbers. Half of the patients live within about `r round(median(homes$distance_km))` km of their hospital. Now let us see, for each hospital, what share of its patients come from more than 50 km away.

```{r}
travel <- homes |>
  st_drop_geometry() |>
  summarise(patients = n(),
            median_km = median(distance_km),
            pct_over_50km = 100 * mean(distance_km > 50),
            .by = hospital_id) |>
  filter(patients >= 15) |>
  arrange(desc(pct_over_50km))

travel |>
  head(8) |>
  mutate(across(where(is.double), \(x) round(x, 1)))
```

We kept only hospitals with at least 15 patients in the sample, to avoid judging a hospital on very few cases. The result is striking. For most hospitals, very few patients come from beyond 50 km. But for a group of six hospitals, more than half do. These are the same six "camp" hospitals that stood out in the chapters on PCA, clustering, association rules and text analytics.

A picture makes the point even more clearly. Let us draw lines from each patient's home to the hospital, for one of these hospitals and for an ordinary hospital (@fig-maptravel).

```{r}
#| label: fig-maptravel
#| fig-cap: "Where patients come from, for a flagged hospital and an ordinary one"
#| fig-height: 4
two_hospitals <- c(travel$hospital_id[1],
                   travel$hospital_id[travel$pct_over_50km == 0][1])

lines_df <- homes |>
  filter(hospital_id %in% two_hospitals) |>
  mutate(home_x = st_coordinates(geometry)[, 1],
         home_y = st_coordinates(geometry)[, 2]) |>
  st_drop_geometry() |>
  left_join(hosp_locations, by = "hospital_id") |>
  mutate(hospital_id = factor(hospital_id, levels = two_hospitals))

ggplot() +
  geom_sf(data = districts, fill = "grey95", colour = "grey70") +
  geom_segment(data = lines_df,
               aes(x = x_km, y = y_km, xend = home_x, yend = home_y),
               colour = "steelblue", alpha = 0.4) +
  geom_point(data = lines_df |> distinct(hospital_id, x_km, y_km),
             aes(x_km, y_km), colour = "firebrick", size = 3) +
  facet_wrap(~hospital_id) +
  theme_void()
```

`st_coordinates()` extracts the x and y values of points, so that we can use them with ordinary `ggplot2` geoms like `geom_segment()`. The hospital on the left draws its patients from across the state. The one on the right serves its own neighbourhood, as one would expect.

## Audit use case 2, were the inspections really done?

Each empanelled hospital must be inspected twice a year. The inspector uploads a photo from the hospital, and the phone records the location where the photo was taken. This location is called the **geotag**. If the photo was really taken at the hospital, its geotag should be within a few hundred metres of the hospital.

```{r}
inspections <- read_csv("Data/ssy_inspections.csv", show_col_types = FALSE) |>
  st_as_sf(coords = c("photo_x_km", "photo_y_km"), crs = st_crs(districts))

head(inspections, 3)
```

As before, we measure the distance from each photo to its hospital.

```{r}
hospital_of_photo <- hosp_sf[match(inspections$hospital_id, hosp_sf$hospital_id), ]

inspections <- inspections |>
  mutate(distance_km = as.numeric(st_distance(inspections, hospital_of_photo,
                                              by_element = TRUE)))

inspections |>
  st_drop_geometry() |>
  summarise(photos = n(),
            median_metres = 1000 * median(distance_km),
            over_1km = sum(distance_km > 1))
```

Most photos were taken within about `r round(1000 * median(inspections$distance_km))` metres of the hospital. But some were taken more than a kilometre away. Let us list them.

```{r}
inspections |>
  st_drop_geometry() |>
  filter(distance_km > 1) |>
  arrange(desc(distance_km)) |>
  mutate(distance_km = round(distance_km, 1))
```

These photos were taken tens of kilometres from the hospitals they claim to show. The inspections were, most likely, reported without anyone visiting the hospital. The table also shows the inspector and the date of each, which is what an audit observation needs. Two things stand out. One inspector, INS-03, accounts for three of the seven cases, which suggests a pattern in that inspector's work rather than an odd mistake. And one of the hospitals, H066, is again among the six flagged hospitals of the earlier chapters.

::: callout-warning
## Caution

GPS locations on phones can be off by a few metres, sometimes by a few hundred metres in crowded areas or indoors. So never flag a photo for being 50 metres away. Use a sensible threshold, like one kilometre in our case, and look at how far beyond it each case lies. Also remember that geotags can be faked with special apps. A correct geotag does not prove a visit, but a wrong one is a strong sign of trouble.
:::

## Buffers

A **buffer** is the area within a given distance of a feature. `st_buffer()` draws one. For example, the area within 25 km of every public hospital shows which parts of the state are served by public hospitals (@fig-mapbuffer).

```{r}
#| label: fig-mapbuffer
#| fig-cap: "Areas within 25 km of a public hospital"
public_cover <- hosp_sf |>
  filter(type == "Public") |>
  st_buffer(dist = 25) |>
  st_union()

ggplot() +
  geom_sf(data = districts, fill = "grey95", colour = "grey60") +
  geom_sf(data = public_cover, fill = "seagreen", alpha = 0.3, colour = NA) +
  geom_sf(data = hosp_sf |> filter(type == "Public"), colour = "seagreen", size = 1) +
  theme_void()
```

`st_union()` merges the overlapping circles into one shape. Areas outside the green shading are more than 25 km from any public hospital. In a performance audit of health services, such a map can show gaps in coverage.

## A suggested workflow

1.  **Get the map layers** you need, from an authoritative source, and check that all parts of each file are present.
2.  **Check the CRS** of every layer. Use a projected CRS in metres or kilometres for distances and areas.
3.  **Turn location columns into points** with `st_as_sf()`.
4.  **Check locations against recorded areas** with `st_join()`.
5.  **Map rates, not counts**, with `geom_sf()`.
6.  **Measure distances** with `st_distance()`, and look for records that are unusually far from where they should be.
7.  **Set sensible thresholds**, allowing for GPS error.
8.  **Verify** each flagged case with records and, where needed, a field visit.

::: callout-important
## Remember

A map is a powerful way to show a finding, but it is easy to mislead with it. Large areas catch the eye even when few people live there. Colours can exaggerate small differences. Make sure the map shows exactly the point you want to make, and that the numbers behind it are right.
:::

## Summary of functions used

| Function | Package | What it does |
|-----------------------|----------------|--------------------------------|
| `st_layers()`, `st_read()` | sf | Lists and reads layers from a map file |
| `st_crs()` | sf | Shows the coordinate reference system |
| `st_as_sf(coords = , crs = )` | sf | Turns x and y columns into map points |
| `geom_sf()`, `geom_sf_text()` | ggplot2 | Draws `sf` objects and labels them |
| `st_join()` | sf | Joins by location, such as point in polygon |
| `st_drop_geometry()` | sf | Turns an `sf` object back into a plain data frame |
| `st_distance(by_element = TRUE)` | sf | Distance between matching pairs of features |
| `st_coordinates()` | sf | Extracts x and y values of points |
| `st_buffer()`, `st_union()` | sf | Area within a distance, and merging shapes |
| `scale_fill_distiller()` | ggplot2 | Colour scale for choropleth maps |

: Functions used in this chapter {#tbl-mapfunctions}
