```{r}
#| include: false
source("_common.R")
```

# Rule-Based Validation {#sec-validate}

Much of audit is checking whether things were done *by the rules*. Pay must be within the scale of the post. Dearness allowance must be at the rate in force. Nobody may be paid after retirement. A bill must add up. Every one of these is a **rule** that can be written down precisely, and every rule that can be written down precisely can be checked by R, on every record, in seconds.

We have already checked many such rules in this book, one at a time, with `filter()`. That works, but as the number of rules grows, a pile of separate `filter()` statements becomes hard to keep track of. Which rules were checked? On which data? How many records failed each? Was anything missed?

In this chapter we learn to keep the rules together as a **rule book**, apply the whole rule book to the data in one step, and get a summary of every rule with the failing records listed. The package **validate** does this. The rule book can be kept in a spreadsheet, maintained by the audit team, with the source of each rule recorded against it. At the end we also meet **pointblank**, which produces a ready report of the results.

**Prerequisites**

```{r}
#| message: false
#| warning: false
library(tidyverse)
library(validate)
```

Both packages are on CRAN, and are installed with `install.packages(c("validate", "pointblank"))`.

## Our data, a month's pay bill {#sec-validate-data}

We audit the pay bill of a department for March 2025. There are two tables.

- `master`, the employee master: date of birth, date of joining, pay level, city class of the place of posting, pension scheme, PAN, bank IFSC and account number.
- `bill`, the pay bill: for each employee paid, the basic pay, dearness allowance (DA), house rent allowance (HRA), transport allowance (TA), gross pay, deductions and net pay.

The rules of the game, simplified from the 7th Central Pay Commission rules, are these.

- Basic pay must lie within the range of the employee's level in the pay matrix.
- DA is 53% of basic pay, the rate in force for our example.
- HRA is 30%, 20% or 10% of basic pay, for cities of class X, Y and Z.
- Gross pay is basic pay plus DA, HRA and TA, and net pay is gross pay less deductions.
- An employee retires on the last day of the month in which he or she attains 60 years. (A person attains an age on the day *before* the birthday, so someone born on the 1st of a month retires at the end of the previous month.)
- Employees who joined on or after 1 January 2004 are in the National Pension System (NPS); earlier entrants are in the General Provident Fund (GPF) scheme.
- PAN must be in the valid format (@sec-regex-validate), and IFSC must be present and valid.
- No two employees may be paid into the same bank account.
- Everyone in the bill must be in the employee master.

The code creates the two tables, and plants errors of each kind.

```{r}
#| code-fold: true
#| code-summary: "Code to create the employee master and the pay bill"
set.seed(2025)
pay_matrix <- tibble(
  level   = 1:14,
  min_pay = c(18000, 19900, 21700, 25500, 29200, 35400, 44900, 47600, 53100, 56100,
              67700, 78800, 123100, 144200),
  max_pay = c(56900, 63200, 69100, 81100, 92300, 112400, 142400, 151100, 167800, 177500,
              208700, 209200, 215900, 218200)
)
n <- 1200
make_pan <- function(k) {
  paste0(map_chr(1:k, \(i) paste(sample(LETTERS, 3, TRUE), collapse = "")), "P",
         sample(LETTERS, k, TRUE), sprintf("%04d", sample(1:9999, k, TRUE)), sample(LETTERS, k, TRUE))
}
master <- tibble(
  emp_id = sprintf("E%05d", 1:n),
  dob    = as.Date("1966-01-01") + sample(0:(365 * 37), n, TRUE)
) |>
  mutate(doj = pmin(dob + round(365.25 * runif(n, 21, 35)), as.Date("2024-12-01")),
         level = sample(1:14, n, TRUE, prob = c(8, 8, 7, 9, 8, 9, 10, 8, 7, 8, 6, 5, 4, 3)),
         city_class = sample(c("X", "Y", "Z"), n, TRUE, prob = c(0.3, 0.4, 0.3)),
         pension_scheme = if_else(doj >= as.Date("2004-01-01"), "NPS", "GPF"),
         pan = make_pan(n),
         ifsc = paste0(sample(c("SBIN", "PUNB", "CNRB", "UBIN", "BARB"), n, TRUE), "0",
                       sprintf("%06d", sample(1:999999, n, TRUE))),
         account_no = sprintf("%011.0f", runif(n, 1e10, 9.99e10)))

# Planted. Four employees who retired in 2024 are still in the master
retired <- sample(n, 4)
master$dob[retired] <- as.Date(c("1964-03-15", "1964-07-01", "1964-09-20", "1964-11-05"))

bill <- master |>
  left_join(pay_matrix, by = "level") |>
  mutate(basic_pay = round(min_pay + (max_pay - min_pay) * runif(n()) ^ 1.5, -2))

pick <- function(k, from = seq_len(nrow(bill))) sample(from, k)
# Planted. Basic pay above the maximum, or below the minimum, of the level
r_hi <- pick(3); bill$basic_pay[r_hi] <- bill$max_pay[r_hi] + c(5000, 12000, 30000)
r_lo <- pick(1); bill$basic_pay[r_lo] <- bill$min_pay[r_lo] - 2000

bill <- bill |>
  mutate(da  = round(0.53 * basic_pay),
         hra = round(c(X = 0.30, Y = 0.20, Z = 0.10)[city_class] * basic_pay),
         ta  = case_when(level >= 9 ~ 7200, level >= 3 ~ 3600, TRUE ~ 1800))

# Planted. DA paid at 55% instead of 53%
r_da <- pick(6); bill$da[r_da] <- round(0.55 * bill$basic_pay[r_da])
# Planted. HRA at the X-class rate for employees posted in Z-class cities
r_hra <- pick(8, which(bill$city_class == "Z")); bill$hra[r_hra] <- round(0.30 * bill$basic_pay[r_hra])

bill <- bill |>
  mutate(gross = basic_pay + da + hra + ta,
         pension_contrib = if_else(pension_scheme == "NPS", round(0.10 * (basic_pay + da)),
                                   round(0.08 * basic_pay)),
         income_tax = round(pmax(0, gross - 60000) * 0.12),
         net_pay = gross - pension_contrib - income_tax)

# Planted. Gross pay with TA added twice
r_gross <- pick(2); bill$gross[r_gross] <- bill$gross[r_gross] + bill$ta[r_gross]
bill$net_pay[r_gross] <- bill$net_pay[r_gross] + bill$ta[r_gross]
# Planted. Post-2004 entrants shown under GPF in the master
r_gpf <- pick(3, which(master$pension_scheme == "NPS")); master$pension_scheme[r_gpf] <- "GPF"
# Planted. PAN in an invalid format, and IFSC missing
r_pan <- pick(5); master$pan[r_pan] <- c("ABCDE1234", "AAAPX1234", "ABCP12345K", "abcpk1234q", "AB CPK1234Q")
r_ifsc <- pick(2); master$ifsc[r_ifsc] <- NA
# Planted. Two pairs of employees paid into the same account
r_acc <- pick(4); master$account_no[r_acc[3:4]] <- master$account_no[r_acc[1:2]]

bill <- bill |>
  select(emp_id, level, city_class, basic_pay, da, hra, ta, gross,
         pension_contrib, income_tax, net_pay)

# Planted. Three employees in the bill who are not in the master
strangers <- bill |> slice_sample(n = 3) |> mutate(emp_id = sprintf("E9%04d", 1:3))
bill <- bind_rows(bill, strangers)
```

```{r}
master
bill
```

## The rules, one `filter()` at a time {#sec-validate-filter}

Before using a package, let us see the plain way, which we already know. To find employees whose DA is not at 53%, allowing a rupee for rounding:

```{r}
bill |> filter(abs(da - 0.53 * basic_pay) > 1) |> select(emp_id, basic_pay, da)
```

This works perfectly for one rule. But we have more than a dozen rules, on two tables. Writing and re-running a dozen such statements, and keeping a record of which passed and which failed, quickly gets untidy. And a rule that silently returns no rows looks the same as a rule that was never run, or one that failed because a column name was mistyped. What we want is to treat the rules as *data*, kept together, applied together, and reported together.

## A rule book with `validator()` {#sec-validate-rules}

First, we bring together the columns the rules need. The bill has the pay figures, the master has the dates and identifiers, and the pay matrix has the limits of each level. We also work out each employee's date of retirement, as described above, with lubridate (@sec-lubridate).

```{r}
pay_check <- bill |>
  left_join(master |> select(-level, -city_class), by = "emp_id") |>
  left_join(pay_matrix, by = "level") |>
  mutate(retirement_date = rollforward(dob %m+% years(60) - days(1)))
```

`rollforward()` moves a date to the last day of its month. So the date of retirement is the last day of the month in which the employee attains 60, that is, the day before the 60th birthday.

Now the rule book. `validator()` takes the rules, each written as an R expression that is `TRUE` when the rule is met. Each rule gets a short name.

```{r}
pay_rules <- validator(
  # identity and completeness
  id_unique      = is_unique(emp_id),
  in_master      = !is.na(dob),
  ifsc_present   = !is.na(ifsc),
  # pay within the level, and the allowances at the right rates
  pay_in_range   = basic_pay >= min_pay & basic_pay <= max_pay,
  da_rate        = abs(da - 0.53 * basic_pay) <= 1,
  hra_rate       = abs(hra - ifelse(city_class == "X", 0.30,
                                    ifelse(city_class == "Y", 0.20, 0.10)) * basic_pay) <= 1,
  # the bill adds up
  gross_total    = gross == basic_pay + da + hra + ta,
  net_total      = net_pay == gross - pension_contrib - income_tax,
  # service conditions
  not_retired    = retirement_date >= as.Date("2025-03-31"),
  pension_scheme = (doj >= as.Date("2004-01-01") & pension_scheme == "NPS") |
                   (doj <  as.Date("2004-01-01") & pension_scheme == "GPF"),
  # identifiers
  pan_format     = grepl("^[A-Z]{3}[PCHFATBLJG][A-Z][0-9]{4}[A-Z]$", pan),
  account_unique = is_unique(account_no)
)
pay_rules
```

A few points about the rules.

- `is_unique()` comes from validate. It checks that each value occurs only once in the column.
- Employees not in the master get `NA` for every master column after the left join, so `!is.na(dob)` is `TRUE` only for those found in the master.
- Amounts computed with percentages can differ by a rupee because of rounding, so the rate rules allow a difference of up to ₹1 (@sec-basics warned against comparing computed decimals with `==`).
- The PAN pattern requires the fourth character to be one of the holder types (P for an individual, C for a company, and so on), as in @sec-regex-validate.
- validate accepts only certain kinds of expressions as rules: comparisons, logical operators like `&` and `|`, and functions returning `TRUE` or `FALSE`. A rule written with `ifelse()` is ignored with a warning, so the pension scheme rule is written with `&` and `|`: "joined on or after 2004 *and* in NPS, *or* joined before 2004 *and* in GPF".

## Confronting the data with the rules {#sec-validate-confront}

`confront()` applies every rule to the data, and `summary()` reports the result of each.

```{r}
results <- confront(pay_check, pay_rules)
summary(results)
```

```{r}
#| include: false
res_sum <- summary(results)
fails_of <- function(rule) res_sum$fails[res_sum$name == rule]
nna_of   <- function(rule) res_sum$nNA[res_sum$name == rule]
```

Each row is one rule. `items` is the number of records checked, `passes` and `fails` the number that met or broke the rule, and `nNA` the number where the rule could not be decided because a value was missing. `error` and `warning` say whether R could evaluate the rule at all, which catches mistyped column names. `expression` shows the rule as validate understood it. Notice that validate has rewritten `gross == ...` with a tiny tolerance, since exact equality of decimals is unsafe.

This one table is the backbone of the audit working paper. Every rule is listed, run or not, passed or not. Let us read it.

- `in_master` fails for `r fails_of("in_master")` records, the employees paid but not in the master.
- `pay_in_range` fails for `r fails_of("pay_in_range")`, and `da_rate` and `hra_rate` for `r fails_of("da_rate")` and `r fails_of("hra_rate")`.
- `not_retired` fails for `r fails_of("not_retired")`, employees paid after their retirement, and `pension_scheme` for `r fails_of("pension_scheme")`.
- `not_retired` and `pension_scheme` also show `nNA` of `r nna_of("not_retired")`. These are the employees missing from the master, for whom the dates are empty, so the rules could not be decided.
- `ifsc_present` and `pan_format` fail for `r fails_of("ifsc_present")` and `r fails_of("pan_format")` records, *including* those same missing employees. `!is.na()` and `grepl()` return `FALSE`, not `NA`, for a missing value, so these rules count a missing value as a failure. How a rule treats missing values depends on how it is written, so read the counts with that in mind.

`plot()` shows the same as a chart, which is useful for a quick view of many rules.

```{r}
#| fig-cap: "Results of each rule"
#| fig-height: 5
plot(results)
```

::: callout-caution
A rule with `nNA` greater than zero has *not* been passed by those records, only left undecided. Never count them as passes. Find out why the values are missing, and decide whether that is a finding in itself, as it is here.
:::

## The failing records {#sec-validate-violating}

To examine the failures, `violating()` returns the rows of the data that break any of the given rules. Let us take the DA and HRA rules.

```{r}
violating(pay_check, results[c("da_rate", "hra_rate")]) |>
  select(emp_id, city_class, basic_pay, da, hra) |>
  mutate(da_pct = round(100 * da / basic_pay, 1), hra_pct = round(100 * hra / basic_pay, 1))
```

The DA of six employees has been drawn at 55%, and eight employees in Z-class cities have been given HRA at the X-class rate of 30%. For a full picture, `values()` gives a `TRUE`/`FALSE` matrix with one column per rule and one row per record. Adding it to the data gives each employee's result on every rule, from which we can count the rules each employee breaks.

```{r}
rule_matrix <- values(results)
dim(rule_matrix)

flags <- pay_check |>
  select(emp_id) |>
  bind_cols(as_tibble(!rule_matrix[, colnames(rule_matrix) != "id_unique" &
                                     colnames(rule_matrix) != "account_unique"]))
flags |>
  mutate(rules_broken = rowSums(across(-emp_id), na.rm = TRUE)) |>
  filter(rules_broken > 0) |>
  count(rules_broken)
```

(The two `is_unique()` rules are left out of the matrix here, since validate records them in a slightly different way.) For the duplicate bank accounts, we look directly.

```{r}
pay_check |>
  filter(!is.na(account_no), account_no %in% account_no[duplicated(account_no)]) |>
  select(emp_id, level, account_no, net_pay) |>
  arrange(account_no)
```

## Valuing the findings {#sec-validate-value}

Each failure is checked against the records before it becomes an observation. For the money involved, we compute the excess paid under each rule.

```{r}
excess <- pay_check |>
  mutate(
    hra_due = round(c(X = 0.30, Y = 0.20, Z = 0.10)[city_class] * basic_pay),
    excess_da  = pmax(0, da - round(0.53 * basic_pay)),
    excess_hra = pmax(0, hra - hra_due),
    paid_after_retirement = if_else(!is.na(retirement_date) & retirement_date < as.Date("2025-03-31"),
                                    net_pay, 0),
    paid_not_in_master = if_else(is.na(dob), net_pay, 0)
  ) |>
  summarise(across(c(excess_da, excess_hra, paid_after_retirement, paid_not_in_master), sum))
excess
```

```{r}
#| include: false
excess_total <- sum(unlist(excess))
```

In one month's pay bill, about ₹`r round(excess_total / 1e5, 1)` lakh is at stake, before counting the pay above the scale, the doubtful accounts and the PAN errors. Since pay bills repeat every month, the annual effect of the same errors would be about twelve times as much, which is why such checks are worth running every month and not once a year.

## Keeping the rule book in a spreadsheet {#sec-validate-spreadsheet}

Rules written in R code are fine for the analyst who wrote them. An audit team, however, needs rules that anyone can read, review and keep up to date, along with the *source* of each rule, so that each failure can be traced to the order it breaks. validate can read the rules from a data frame, with columns `rule`, `name`, `label` and `description`. Such a table can be kept in Excel and read with `readxl::read_excel()` (@sec-read).

```{r}
rule_book <- tribble(
  ~name,          ~label,                              ~rule,                                                  ~description,
  "da_rate",      "DA at the rate in force",           "abs(da - 0.53 * basic_pay) <= 1",                       "DA order in force for March 2025: 53% of basic pay",
  "pay_in_range", "Basic pay within the level",        "basic_pay >= min_pay & basic_pay <= max_pay",           "Pay matrix, 7th CPC",
  "not_retired",  "Not paid after superannuation",     "retirement_date >= as.Date('2025-03-31')",              "Superannuation at the end of the month of attaining 60",
  "in_master",    "Paid employee exists in master",    "!is.na(dob)",                                           "Internal control: payment only to verified employees",
  "pan_format",   "PAN in valid format",               "grepl('^[A-Z]{3}[PCHFATBLJG][A-Z][0-9]{4}[A-Z]$', pan)", "Income Tax Act; needed for TDS"
)

book_rules <- validator(.data = rule_book)
book_rules
```

The labels and descriptions travel with the rules, so the summary can be joined back to the rule book to produce a readable table for the working papers.

```{r}
summary(confront(pay_check, book_rules)) |>
  left_join(rule_book |> select(name, label, description), by = "name") |>
  select(label, items, passes, fails, nNA, description)
```

::: callout-tip
## Audit tip

Build one rule book for each recurring audit area, such as pay bills, pension, contracts or a scheme, and keep it with the audit programme. Each year, update the rates and limits, add the rules suggested by last year's findings, and re-run the whole book on the new data. Over time, the rule book becomes the team's institutional memory, and new members can see at once what is checked and why.
:::

## A ready report with pointblank {#sec-validate-pointblank}

The package **pointblank** does a similar job with a different style. Instead of a rule book, we build an *agent*, give it a series of validation steps, and let it *interrogate* the data. Its strength is the report it produces, a formatted table of every step with the share of records that passed, which can be shared directly.

```{r}
#| message: false
#| warning: false
library(pointblank)

agent <- create_agent(tbl = pay_check, tbl_name = "Pay bill, March 2025",
                      label = "Pay bill checks") |>
  rows_distinct(columns = vars(emp_id)) |>
  col_vals_not_null(columns = vars(ifsc)) |>
  col_vals_between(columns = vars(basic_pay), left = vars(min_pay), right = vars(max_pay)) |>
  col_vals_expr(expr = expr(abs(da - 0.53 * basic_pay) <= 1)) |>
  col_vals_regex(columns = vars(pan), regex = "^[A-Z]{3}[PCHFATBLJG][A-Z][0-9]{4}[A-Z]$") |>
  col_vals_gte(columns = vars(retirement_date), value = as.Date("2025-03-31")) |>
  interrogate()
```

Printing the agent gives the report.

```{r}
agent
```

Each row is a step, with the number of records tested, and the number and share that passed and failed. Missing values count as failures in pointblank by default (the argument `na_pass` changes this), which is why the steps involving master columns show a few more failures than validate did. `get_sundered_data()` splits the data into the rows that passed every step and those that failed at least one.

```{r}
failed_rows <- get_sundered_data(agent, type = "fail")
nrow(failed_rows)
```

pointblank can also set *action levels*, such as a warning if more than 1% of records fail a step, and stopping a process if more than 5% do. This is useful when the same checks are run every month on data received from a department, as an early warning that something has gone wrong with the data itself.

::: callout-note
Which package to use? **validate** suits audit work where the rule book itself is the main product: rules kept as data, with labels and sources, applied to many data sets. **pointblank** suits repeated checks on incoming data with a polished report. Both are good, and they can be used together, as here.
:::

## A suggested workflow

1. **Collect the rules** from the acts, rules, orders and manuals governing the area, and note the source of each.
2. **Write them as a rule book**, in a spreadsheet, with a name, label, rule and description for each.
3. **Prepare the data**: join the tables the rules need, and compute derived values like the date of retirement.
4. **Confront** the data with the rule book, and read the summary, rule by rule. Check rules with errors, warnings or missing values first.
5. **List the failing records** with `violating()`, and examine them against the documents.
6. **Value the findings**, and project the effect over the year where the data repeats.
7. **Report** the summary table with the rule sources, or a pointblank report.
8. **Maintain the rule book**: update rates, and add rules from new findings.

::: callout-important
## Remember

- Many audit checks are rules that can be written exactly. Keep them together as a rule book, not as scattered filters.
- `validator()` holds the rules, `confront()` applies them, `summary()` reports every rule, and `violating()` lists the failing records.
- Records with missing values are undecided, not passed.
- Keep the rule book in a spreadsheet, with the source of each rule, so that the team can review and update it.
- A rule failure is a lead to be verified against documents, like every other analytic finding.
:::

## Summary of functions used

| Function | Package | What it does |
|-----------------------|----------------|--------------------------------|
| `validator()`, `validator(.data = )` | validate | Create a rule book, from expressions or from a data frame of rules |
| `confront()` | validate | Apply the rules to the data |
| `summary()`, `plot()` | validate | Passes, fails and missing values for each rule |
| `violating()` | validate | Rows that break the rules |
| `values()` | validate | Result of every rule for every record, as a matrix |
| `is_unique()` | validate | Rule that each value of a column occurs only once |
| `rollforward()`, `%m+%` | lubridate | Last day of the month, and adding years to dates safely |
| `create_agent()`, `interrogate()` | pointblank | Set up and run validation steps |
| `rows_distinct()`, `col_vals_not_null()`, `col_vals_between()`, `col_vals_expr()`, `col_vals_regex()`, `col_vals_gte()` | pointblank | Validation steps |
| `get_sundered_data()` | pointblank | Split the data into passing and failing rows |

: Functions used in this chapter {#tbl-validatefunctions}
