```{r}
#| include: false
source("_common.R")
```

# Anomaly Detection {#sec-anomaly}

```{r an100, echo=FALSE, warning=FALSE, message=FALSE}
library(tidyverse)
library(magrittr)
```

Anomalies play a significant role in the field of audit. An anomaly is something that does not fit. It may be a payment far larger than usual, a claim that breaks a pattern, or a month when expenditure suddenly jumps. As auditors examine data, whether financial statements, transactions or other records, such anomalies give valuable leads.

Anomalies often point to problems like these.

-   **Fraud.** Fictitious transactions, inflated bills, unauthorised access or misappropriation of assets.
-   **Errors and misstatements.** Wrong entries, keying mistakes and wrong classification.
-   **Weak internal controls.** Transactions that should have been stopped by a control, but were not.
-   **Non-compliance.** Payments that break rules, limits or entitlements.
-   **Inefficiency.** Processes that take unusually long or cost unusually more.

Detecting and investigating anomalies is therefore a core skill for an auditor. In this chapter, we will learn several methods to detect them, from simple boxplots to machine learning methods. @tbl-anomethods gives a road map.

| Method | Works on | Good at finding | Section |
|-------------------|------------------|----------------------|------------|
| Boxplot (IQR rule) | One numeric column | Values far from the middle of the data | @sec-ano-inspect |
| Grubbs' test | One numeric column, normally distributed | Whether the single most extreme value is an outlier | @sec-ano-grubbs |
| Robust z-score, within peer groups | One numeric column, or a ratio | Values extreme for their own peer group, without masking | @sec-ano-robustz |
| Robust Mahalanobis distance | A few related numeric columns | Combinations that break the usual relationship | @sec-ano-mahal |
| Seasonal Hybrid ESD | One time series | Anomalies after allowing for seasonality | @sec-ano-shesd |
| k-Nearest Neighbour distance | Many numeric columns | Records far away from all others (global anomalies) | @sec-ano-knn |
| Local Outlier Factor (LOF) | Many numeric columns | Records unlike their own neighbours (local anomalies) | @sec-ano-lof |
| Isolation Forest | Many columns, numeric and categorical | Records which are easy to separate from the rest | @sec-ano-iforest |
| LOF with Gower distance | Many columns, numeric and categorical | Rare combinations of categories and values | @sec-ano-gower |
| STL decomposition (`timetk`) | One time series | Odd periods after allowing for trend and season | @sec-ano-ts |
| Change point detection (`changepoint`) | One time series | The date of a lasting shift in level | @sec-ano-changepoint |
| Combining methods by rank | Scores from several methods | One prioritised list, using the strengths of each | @sec-ano-combine |

: Anomaly detection methods covered in this chapter {#tbl-anomethods}

## Definition and types of anomalies

Anomalies are patterns or data points that deviate significantly from the expected or normal behaviour within a dataset. They are also known as *outliers*, *novelties* or *deviations*. There are three types of anomalies.

-   **Point anomalies.** A single record which is unusual when compared with the rest of the data. For example, a stationery bill of ₹ 2,00,000 in an office where such bills are usually below ₹ 10,000.
-   **Contextual anomalies.** A record which is unusual only in a particular context. For example, an electricity bill of ₹ 3 lakh may be normal for an office in May, when air conditioners run all day, but unusual for the same office in December. The amount is not odd by itself. The *month* makes it odd.
-   **Collective anomalies.** A group of records which is unusual when seen together, though each record may look normal by itself. For example, twenty cash withdrawals of ₹ 49,000 each from one account within a week. No single withdrawal is unusual. Together, they suggest that someone is avoiding a reporting limit of ₹ 50,000.

## Anomaly detection by inspection {#sec-ano-inspect}

Many times, anomalies or outliers can be spotted just by looking at the data. The `summary()` function prints the *minimum*, *maximum*, *lower* and *upper quartiles*, the *mean* and the *median*. It gives a sense of how far an extreme point lies from the rest of the data.

As an example, let us take the mean annual temperatures in degrees Fahrenheit in New Haven, Connecticut, from 1912 to 1971. This data is available in base R as `nhtemp`.

```{r an101}
summary(nhtemp)
```

The easiest way to see how unusual a value is, is a graphical summary like a **boxplot**. In base R, it is drawn with the `boxplot()` function. It shows the distribution of values as a box with whiskers, and draws extreme values as separate points. This makes them easy to spot. We can also draw boxplots using `ggplot2`. Examples of both are shown in @fig-an1.

::: {#fig-an1}

```{r fig.show='hold', fig.align='center', out.width="45%"}
boxplot(nhtemp, main = "Box plot of nhtemp data")
ggplot(iris, aes(Species, Petal.Length, color = Species )) +
  geom_boxplot(outlier.colour = "red", outlier.size = 2) +
  theme_bw() +
  labs(title = "Petal Lengths in Iris Data")
```

Outliers through Boxplot
:::

The separate points in a boxplot are decided by a simple rule, often called the **IQR rule**. The box covers the middle 50% of the data, from the lower quartile (Q1) to the upper quartile (Q3). The difference Q3 - Q1 is called the *inter-quartile range* or IQR. Any value more than 1.5 × IQR below Q1, or above Q3, is drawn as a separate point.

We do not have to read these values from the plot. `boxplot.stats()` returns them in its `out` element.

```{r}
boxplot.stats(nhtemp)$out
```

::: callout-tip
## Audit tip

The boxplot is the quickest first check on any amount column. Draw it *by group* rather than for the whole data. For example, draw a boxplot of bill amounts for each vendor or each head of account. A bill which is normal for the whole office may be very unusual for its own vendor. We saw this idea in the RSF test in @sec-nigrini.
:::

It is important to note that a **point anomaly** need not always be an extreme value. A point anomaly can also arise as an unusual *combination* of values across several columns. We will see such an example in @sec-ano-gower.

A **collective anomaly**, on the other hand, is a group of records which is unusual together. In @fig-an2, a period of five consecutive days with high temperatures is shown by red points. These daily temperatures are unusual because they occur together, and are likely caused by the same weather event.

```{r fig-an2, echo=FALSE, fig.align='center', fig.show='hold', fig.cap="Collective Anomalies", out.height="27%"}
set.seed(567)
c(rnorm(15, 30, 2), rnorm(5, 40, 2.1), rnorm(10, 30, 2)) %>% ts() -> myts
dates <- 1:30
plot(dates, myts, type = 'l',ylab = "Temp in Celsius", xlab = 'Dates',
     main = "Collective Anomalies through a dummy data")
points(16:20, myts[16:20], col = 'red', pch = 16)
points(1:15, myts[1:15], col = 'black', pch = 16)
points(21:30, myts[21:30], col = 'black', pch = 16)
```

## Grubbs' test {#sec-ano-grubbs}

Visual inspection works well when most data points are grouped together and a few lie clearly apart. But it is a matter of judgement. Statistics offers formal tests to decide whether a value is an outlier. **Grubbs' test**, named after Frank Grubbs, is one of these.

Grubbs' test checks whether the single value lying farthest from the mean could be an outlier. This value will be either the largest or the smallest value in the data. The test assumes that the data is *normally distributed*. So, before using it, we must check that this assumption is reasonable. A histogram is a quick way to check this. (For other ways of checking normality, refer to the chapter on linear regression.)

Let us run the test on the `nhtemp` data, which we saw in @fig-an1. First, the histogram.

```{r fig-an3, fig.align='center', fig.show='hold', fig.cap="Histogram for Grubbs' test", out.height="27%"}
hist(nhtemp, breaks = 20)
```

In @fig-an3 we can see that the data is roughly bell shaped, so the assumption is nearly satisfied. Let us run the test, using `grubbs.test()` from the `outliers` package.

```{r an}
library(outliers)
grubbs.test(nhtemp)
```

How do we read this? The *null hypothesis* of the test is that there is **no** outlier. The output shows the alternative hypothesis, that the highest value, 54.6, is an outlier. The p-value is about `0.15`, which is more than the usual cut-off of `0.05`. So we do not have enough evidence to reject the null hypothesis. In simple words, we **cannot conclude** that 54.6 is an outlier, though it looked separate in the boxplot.

::: callout-warning
## Caution

Grubbs' test checks only *one* value at a time, the most extreme one. If there are several outliers, they can hide each other. For testing several suspected outliers together, a related test called Rosner's test can be used. It is available as `rosnerTest()` in the `EnvStats` package. Also remember that most financial data, like bill amounts, is *not* normally distributed. It usually has a long tail of large values. For such data, Grubbs' test is not suitable unless the data is first transformed, for example by taking logs.
:::

## Robust z-scores and peer groups {#sec-ano-robustz}

The best-known way to measure how unusual a value is, is the **z-score**: how many standard deviations it lies from the mean. A value with a z-score beyond 3, or below -3, is commonly called an outlier. But the plain z-score has two weaknesses which matter a great deal in audit data.

**It is not robust.** The mean and the standard deviation are themselves pulled by extreme values. A few large outliers raise the mean and inflate the standard deviation, and so make *themselves* look less unusual. This is called *masking*. See this small example.

```{r}
x <- c(10, 11, 12, 10, 11, 12, 11, 95, 98, 99)
round((x - mean(x)) / sd(x), 2)
```

The three obviously odd values, 95, 98 and 99, have z-scores below 1.5. They have hidden one another. The remedy is to use measures which outliers cannot pull: the **median** in place of the mean, and the **median absolute deviation** (MAD) in place of the standard deviation. The MAD is the median of the distances of the values from their median. R's `mad()` scales it so that, for normal data, it matches the standard deviation. The result is the **robust z-score**.

```{r}
round((x - median(x)) / mad(x), 2)
```

Now the three odd values stand out at once. A common rule, suggested by Iglewicz and Hoaglin, is to flag robust z-scores beyond 3.5. For money amounts, which are skewed, compute the score on the logarithm of the amount, as we did in @sec-ano-ts.

**It compares everything with everything.** The second weakness matters even more. A district hospital spends far more per patient than a primary health centre. Compared with all facilities together, every district hospital looks "unusual", and a primary health centre spending three times as much as other PHCs looks normal. The fair comparison is with the **peer group**: PHCs with PHCs, district hospitals with district hospitals.

Let us see both points on the medicine expenditure of 300 health facilities of a State.

```{r}
#| code-fold: true
#| code-summary: "Code to create the facility data"
set.seed(2028)
types <- c(PHC = 200, CHC = 70, DH = 30)   # primary, community health centres, district hospitals
facilities <- tibble(type = rep(names(types), types)) |>
  mutate(facility = paste0(type, "-", sprintf("%03d", row_number())), .by = type) |>
  mutate(beds = round(case_when(type == "PHC" ~ runif(n(), 6, 10),
                                type == "CHC" ~ runif(n(), 30, 50),
                                TRUE          ~ runif(n(), 100, 300))),
         patients = round(beds * rlnorm(n(), log(140), 0.25)),          # patients per month
         cost_per_patient = round(case_when(type == "PHC" ~ 120, type == "CHC" ~ 330,
                                            TRUE ~ 850) * rlnorm(n(), 0, 0.18)),
         # Planted. Medicine spending far above the facility's own peers
         cost_per_patient = case_when(facility == "PHC-037" ~ 410,
                                      facility == "CHC-012" ~ 1250,
                                      TRUE ~ cost_per_patient),
         medicine_lakh = round(patients * cost_per_patient / 1e5, 2))
```

```{r}
facilities |>
  summarise(facilities = n(), median_cost = median(cost_per_patient), .by = type)
```

First, the plain z-score of the cost per patient, over all 300 facilities.

```{r}
facilities |>
  mutate(z = (cost_per_patient - mean(cost_per_patient)) / sd(cost_per_patient)) |>
  filter(abs(z) > 3) |>
  count(type)
```

```{r}
#| include: false
plain_flags <- facilities |> mutate(z = (cost_per_patient - mean(cost_per_patient)) / sd(cost_per_patient)) |> filter(abs(z) > 3)
```

Of the `r nrow(plain_flags)` facilities flagged, `r sum(plain_flags$type == "DH")` are district hospitals, flagged simply for being district hospitals. Now the robust z-score of the log cost, computed *within each type* with `.by`.

```{r}
robust_z <- function(x) (x - median(x)) / mad(x)

facilities <- facilities |>
  mutate(rz_peer = robust_z(log(cost_per_patient)), .by = type)

facilities |>
  filter(abs(rz_peer) > 3.5) |>
  select(facility, type, cost_per_patient, rz_peer) |>
  mutate(rz_peer = round(rz_peer, 1))
```

Only the two planted facilities are flagged. PHC-037 spends ₹410 per patient, ordinary for a CHC but more than three times what other PHCs spend, which the plain z-score over all facilities could not see. CHC-012 spends far more than other CHCs.

::: callout-tip
## Audit tip

Robust z-scores within peer groups are the simplest anomaly method that an audit team can explain to anyone, and often the most useful. Choose peer groups that *should* behave alike: facilities of one type, schools of one level, police stations of one category, DDOs of one department, vendors of one commodity. Use a ratio that removes size, like cost per patient or expenditure per student. And make sure each group has enough members, at least 20 or 30, for the median and MAD to be reliable.
:::

## Unusual combinations, robust Mahalanobis distance {#sec-ano-mahal}

Robust z-scores look at one measure at a time. Some anomalies show only in a *combination*. When a natural ratio exists, like cost per patient, the ratio captures the combination, and the z-score of the ratio is the best tool. But often the measures are related in a way no single ratio captures.

Take the primary health centres of a district. Larger PHCs see more patients in the outpatient department (OPD) and also conduct more deliveries, so the two numbers rise together. Each delivery in a public facility earns an incentive under the Janani Suraksha Yojana, so inflated delivery figures are a known risk. Now suppose one PHC reports a fairly *low* OPD but a fairly *high* number of deliveries. Neither number is unusual by itself. Together, they are.

```{r}
#| code-fold: true
#| code-summary: "Code to create the PHC data"
set.seed(12)
n_phc <- 200
size  <- rlnorm(n_phc, 0, 0.35)              # some PHCs serve larger populations
phc <- tibble(phc        = sprintf("PHC-%03d", 1:n_phc),
              opd        = round(1100 * size * rlnorm(n_phc, 0, 0.12)),   # patients per month
              deliveries = round(30 * size * rlnorm(n_phc, 0, 0.15)))     # deliveries per month
# Planted. Low OPD (15th percentile) but high deliveries (88th percentile)
phc$opd[77]        <- round(quantile(phc$opd, 0.15))
phc$deliveries[77] <- round(quantile(phc$deliveries, 0.88))
```

```{r}
phc <- phc |> mutate(z_opd = robust_z(log(opd)), z_deliveries = robust_z(log(deliveries)))
phc |> filter(phc == "PHC-077") |> mutate(across(starts_with("z_"), \(z) round(z, 2)))
```

Neither robust z-score of PHC-077 comes anywhere near 3.5. A scatter plot shows the problem at once.

```{r}
#| fig-cap: "OPD and deliveries rise together; PHC-077 breaks the pattern"
#| fig-height: 4
ggplot(phc, aes(opd, deliveries)) +
  geom_point(alpha = 0.6) +
  geom_point(data = filter(phc, phc == "PHC-077"), colour = "red", size = 3) +
  scale_x_log10() + scale_y_log10() +
  labs(x = "OPD patients per month (log scale)", y = "Deliveries per month (log scale)") +
  theme_minimal()
```

The **Mahalanobis distance** measures how far a point lies from the centre of the data, *taking account of how the measures vary together*. A point along the main band of the scatter, even far out, is not unusual. A point *across* the band, like PHC-077, is. It works for any number of measures, where scatter plots are no longer possible.

Like the z-score, it needs a centre and a measure of spread, and like the z-score, these are pulled by outliers. So we use robust estimates, from `cov.rob()` in the package `MASS`, which comes with R. Its "minimum covariance determinant" method finds the centre and spread of the most tightly packed half of the data, ignoring the rest. We work on the logarithms, since the counts are skewed.

```{r}
X <- cbind(log(phc$opd), log(phc$deliveries))
robust_fit <- MASS::cov.rob(X, method = "mcd")
phc$mahal <- mahalanobis(X, center = robust_fit$center, cov = robust_fit$cov)

phc |>
  arrange(desc(mahal)) |>
  select(phc, opd, deliveries, mahal) |>
  mutate(mahal = round(mahal, 1)) |>
  head(5)
```

PHC-077 stands far above the rest. How large is "large"? For normal data, the squared Mahalanobis distance, which is what `mahalanobis()` returns, follows a *chi-square* distribution, with as many degrees of freedom as there are measures. So `qchisq(0.999, df = 2)` gives a cut-off that a normal point exceeds only once in a thousand times.

```{r}
cutoff <- qchisq(0.999, df = ncol(X))
round(cutoff, 1)
phc |> filter(mahal > cutoff) |> pull(phc)
```

Only PHC-077 crosses the cut-off. The next step would be to compare its delivery register with its OPD register, the beds used, and the incentive payments made.

## Seasonal Hybrid ESD algorithm {#sec-ano-shesd}

Grubbs' test has two limitations. It needs normally distributed data, and it looks only at the extreme values of the whole data. Neither suits a time series with seasonal ups and downs, like monthly sales or quarterly revenue. A value may be normal for one season and unusual for another.

The **Seasonal Hybrid ESD** (S-H-ESD) algorithm was developed at Twitter for exactly this purpose. It first removes the seasonal pattern, and then looks for unusual values in what remains. It is available in the `AnomalyDetection` package, which can be installed from GitHub using `remotes::install_github("twitter/AnomalyDetection")`.

As an example, let us use the `JohnsonJohnson` data available in base R, which has quarterly earnings of the company. Since the data is quarterly, the seasonal pattern repeats every 4 records, so we use `period = 4`.

```{r fig-an4, fig.show='hold', fig.align='center', fig.cap="Seasonal Hybrid ESD Algorithm", message=FALSE, warning=FALSE, out.height="28%"}
#remotes::install_github("twitter/AnomalyDetection")
library(AnomalyDetection)
AnomalyDetectionVec(as.vector(JohnsonJohnson),
                    period = 4, # Seasonality
                    plot = TRUE, # set FALSE when plot not needed
                    direction = 'both' # set 'pos' for higher values
                                       # or 'neg' for lower values
                    )
```

In @fig-an4, the anomalies are marked with blue circles. They are also listed in the `$anoms` element of the output, with their positions and values.

Notice that all the flagged points are near the end of the series. Johnson & Johnson's earnings grew very fast, and the seasonal ups and downs grew wider with them. So the later quarters look unusual compared with the earlier ones, though they follow the same pattern. This is a common trap with growing financial series. We will see how to deal with it, by working on the log scale, in @sec-ano-ts.

## k-Nearest Neighbour distance score {#sec-ano-knn}

One big limitation of the methods so far is that they work on a single column, or a single series. In real audit work, data is rarely so simple. A vendor may be described by its total payments, number of bills, average bill and share of round amounts. An anomaly may show up only when all these are seen together. The **k-Nearest Neighbour** (kNN) method works on many columns at once.

The idea of kNN is that records which are close to each other are similar. For each record, we find its `k` nearest neighbours, and calculate the average distance to them. A record lying far away from even its nearest neighbours is likely to be an anomaly.

Distance between records is usually measured as the straight-line *Euclidean distance*. For this to be fair, all columns must be on the same scale. Otherwise a column measured in lakhs will swamp a column measured in single digits. So we must always **scale** the data first.

**Problem Statement-1:** Let us try to identify outliers in the flower measurements of the `virginica` species in the `iris` data. For an auditor, it may help to think of each flower as a vendor, and its four measurements as four indicators of the vendor's behaviour. We will use the `scale()` function of base R to normalise the data. Remember that `scale()` returns a matrix, so we may need to convert it into a data frame for `ggplot2`.

```{r an102}
# Sample data
df <- iris[101:150, 1:4]
#  Scale the data
df <- scale(df)
```

Let us visualise the data, first in two dimensions only. As the data is already scaled, we can also see all four dimensions together in a boxplot. Both are shown in @fig-an5. The points are numbered by their row numbers, so that we can refer to them later.

::: {#fig-an5}

```{r warning=FALSE, message=FALSE, fig.align='center', fig.show='hold', out.width="45%"}

df %>%
  as.data.frame() %>%
  mutate(row = row_number()) %>%
  ggplot(aes(Sepal.Length, Sepal.Width)) +
  geom_point(color = "seagreen") +
  ggrepel::geom_text_repel(aes(label = row), arrow = grid::arrow()) +
  theme_bw()

# Helper Function
find_outlier <- function(x) {
  return(x < quantile(x, .25) - 1.5*IQR(x) | x > quantile(x, .75) + 1.5*IQR(x))
}

df %>%
  as.data.frame() %>%
  mutate(row = row_number()) %>%
  pivot_longer(-row, names_to = "Dimension", values_to = "Values") %>%
  group_by(Dimension) %>%
  mutate(outlier = ifelse(find_outlier(Values), row, NA)) %>%
  ggplot(aes(Dimension, Values)) +
  geom_boxplot(outlier.colour = "red") +
  geom_text(aes(label=outlier), na.rm=TRUE, hjust=-.5) +
  theme_bw()
```

Sepal Length Vs. Widths in Virginica
:::

Now, let us find the outliers in R. We will use the `get.knn()` function from the `FNN` package. We must choose the value of `k`, the number of neighbours, beforehand. Let us take `k = 5`.

```{r an6, warning=FALSE, message=FALSE}
# Load the library
library(FNN)
# get kNN object, using k = 5, on first two columns only
virginica_knn <- FNN::get.knn(df[, 1:2], 5)
```

The object created above has two matrices, each with `k` columns. `nn.index` holds the row numbers of the nearest neighbours of each record. `nn.dist` holds the distances to them. Let us view the top 6 rows of each.

```{r an7}
head(virginica_knn$nn.index)
head(virginica_knn$nn.dist)
```

Using `rowMeans()`, we can calculate the average distance of each record from its neighbours. This is the **kNN score**. The bigger the score, the more likely the record is an outlier. Let us store the score and show it as the size of the points. In @fig-an8, the points lying far away are bigger, because their chance of being outliers is higher.

```{r fig-an8, warning=FALSE, message=FALSE, fig.align='center', fig.show='hold', fig.cap="k-Nearest Neighbour Distance", out.height="28%"}
score <- rowMeans(virginica_knn$nn.dist)
df %>%
  as.data.frame() %>%
  mutate(row = row_number(),
         score = score) %>%
  ggplot(aes(Sepal.Length, Sepal.Width)) +
  geom_point(aes(size = score),color = "seagreen") +
  ggrepel::geom_text_repel(aes(label = row), arrow = grid::arrow()) +
  theme_bw() +
  labs(title = "KNN method of outlier")

```

Now, we can run the algorithm on all four columns of the data.

```{r an88}
virginica_knn <- FNN::get.knn(df, 5)
score <- rowMeans(virginica_knn$nn.dist)
# Which point is farthest-Outlier
which.max(score)
```

::: callout-note
## How to choose `k`?

There is no single correct value of `k`. A very small `k`, like 1 or 2, makes the score sensitive to chance. Two odd records lying close to each other will look normal, because each is the other's nearest neighbour. A very large `k` blurs the picture. A value between 5 and 20 works well in most cases. It is a good practice to try two or three values and see whether the same records come out on top.
:::

## Local Outlier Factor {#sec-ano-lof}

kNN is good at finding records which are far from *all* others. These are sometimes called **global anomalies**. But it can miss records which lie close to a group, yet are not like their immediate neighbours. These are called **local anomalies**.

To understand the difference, see the dummy data in @fig-an9. Think of it as sales and profit of branches of a company.

```{r fig-an9, echo=FALSE, warning=FALSE, message=FALSE, fig.align='center', fig.show='hold', fig.cap="Local Outlier factor", out.height="28%"}
set.seed(123)
dummy <- data.frame(
  sales = rnorm(25, 500, 100),
  profit = rnorm(25, 75, 10)
)
dummy2 <- data.frame(
  sales = rnorm(25, 1000, 5),
  profit = rnorm(25, 75, 2)
)
dummy3 <- data.frame(
  sales = rnorm(5, 500, 100),
  profit = rnorm(5, 120, 10)
)
dummy4 <- data.frame(
  sales = rnorm(5, 1000, 10),
  profit = rnorm(5, 90, 2)
)
dummy <- rbind(dummy, dummy2, dummy3, dummy4)

plot(profit ~ sales, data = dummy, xlim = c(0, 1100), ylim = c(0, 200))
points(profit ~ sales, data = dummy[51:55,], xlim = c(0, 1100), ylim = c(0, 200), col = 'red', pch = 20)
points(profit ~ sales, data = dummy[56:60,], xlim = c(0, 1100), ylim = c(0, 200), col = 'blue', pch = 20)
```

The red points are global outliers. They are far from everything else. The blue points are not far from the tight group on the right. But that group is very tightly packed, and the blue points do not share its tightness. So they are local anomalies. kNN may miss them, because their distance from the group is small in absolute terms.

The **Local Outlier Factor** (LOF) solves this. Instead of distance, it looks at *density*, that is, how crowded the neighbourhood of a record is. It compares the density around each record with the density around its neighbours.

-   A score close to **1** means the record is as crowded as its neighbours. It is normal.
-   A score clearly **above 1** means the record is in a much emptier area than its neighbours. It may be an anomaly.

We will use the `lof()` function from the `dbscan` package. Its second argument is the number of neighbours to consider, like `k` in kNN. Let us apply it to the same `virginica` data, using all four columns.

::: {#fig-an10}

```{r warning=FALSE, message=FALSE, fig.align='center', fig.show='hold', out.width="48%"}
library(dbscan)
lof_score <- lof(df, 5)
df %>%
  as.data.frame() %>%
  mutate(row = row_number(),
         score = lof_score) %>%
  ggplot(aes(Sepal.Length, Sepal.Width)) +
  geom_point(aes(size = score),color = "seagreen") +
  ggrepel::geom_text_repel(aes(label = row), arrow = grid::arrow()) +
  scale_size_continuous(guide = "none") +
  theme_bw() +
  labs(title = "LOF method of outlier")

df_prcomp <- prcomp(df, scale. = TRUE)

df %>%
  as.data.frame() %>%
  mutate(row = row_number(),
         score = lof_score,
         PC1 = df_prcomp$x[,1],
         PC2 = df_prcomp$x[,2]) %>%
  ggplot(aes(PC1, PC2)) +
  geom_point(aes(size = score),color = "seagreen") +
  ggrepel::geom_text_repel(aes(label = row), arrow = grid::arrow()) +
  scale_size_continuous(guide = "none") +
  theme_bw() +
  labs(title = "LOF method of outlier\nVisualised on principal components")
```

LOF
:::

In @fig-an10 (left), we have plotted the data in two dimensions only, though LOF was calculated on all four. So some big points appear *inside* the cloud of points. These are local anomalies, which are unusual in the other two dimensions.

To see all four dimensions in one picture, we can use the first two *principal components* (@fig-an10, right). Principal components combine many columns into a few new columns which keep most of the information. We learnt about them in @sec-pca. For now, it is enough to know that they give a fair two-dimensional picture of four-dimensional data.

Let us also compare the biggest anomaly according to each method.

```{r an11}
# Biggest Anomaly - kNN
which.max(score)

# Biggest Anomaly - LOF
which.max(lof_score)
```

The two methods do not agree on the top record. kNN picks record `r which.max(score)`, which is far from all others. LOF picks record `r which.max(lof_score)`, which is unusual compared to its own neighbourhood. Both are useful, and they answer slightly different questions.

Histograms of both `kNN` and `LOF` scores can also be drawn, as in @fig-an12. In both, most records have low scores, and a few stand apart on the right.

::: {#fig-an12}

```{r warning=FALSE, message=FALSE, fig.align='center', fig.show='hold', out.width="48%"}
hist(score, breaks = 40,
     main = "Histogram of KNN scores in IRIS-Virginica data")
hist(lof_score, breaks = 40,
     main = "Histogram of LOF scores in IRIS-Virginica data")
```

Histogram - KNN(Left) and LOF (Right)
:::

::: callout-tip
## Audit tip

kNN and LOF give *scores*, not a yes or no answer. Sort the records by score and examine them from the top, until the cases stop being interesting. The histogram of scores helps decide where to stop. Look for the point where the long tail on the right begins.
:::

## Isolation Forest {#sec-ano-iforest}

Isolation Forest takes a completely different approach. It does not measure distance or density. Instead, it asks a simple question. *How easy is it to separate this record from all the others?*

Imagine we pick a column at random, and a random cut-point within its range. This splits the data into two parts. We repeat this on each part, again and again, until every record sits alone. Such a series of random splits is called an **isolation tree**. A normal record, sitting in a crowd, needs many splits before it is alone. An anomaly, sitting far from the crowd, gets separated in very few splits. So the **number of splits** (the path length in the tree) tells us how unusual a record is.

One tree is based on random choices, so its result can vary. A **forest** is a collection of many such trees. Averaging over many trees gives a stable score.

We will use the `isotree` package, which is available on CRAN. The main function is `isolation.forest()`. Its argument `ntrees` sets the number of trees. `sample_size` sets how many records are drawn to build each tree. We also set `ndim = 1`, which gives the classic isolation forest that splits on one column at a time, and a `seed` so that the results can be reproduced.

::: callout-note
## A note on packages

Earlier versions of this book used the `isofor` package, which is available only on GitHub and is no longer maintained. `isotree` does the same job, is faster, and is regularly updated on CRAN. It can be installed with `install.packages("isotree")`.
:::

Let us work on the same problem statement as above. First, let us build a single tree.

```{r an13, message=FALSE, warning=FALSE}
library(isotree)
# Generate a single tree
virginica_1 <- isolation.forest(df, ntrees = 1, sample_size = 50,
                                ndim = 1, seed = 1, nthreads = 1)

# Generate isolation score
iso_score_1 <- predict(virginica_1, df)

# View first 10 scores
iso_score_1[1:10]
```

*Score interpretation*: The scores lie between 0 and 1. The closer a score is to 1, the more likely the record is an anomaly. Scores below 0.5 usually mean normal records.

Let us visualise the scores on the principal component plot again (@fig-an24).

::: {#fig-an24}

```{r message=FALSE, warning=FALSE, fig.align='center', fig.show='hold', out.width="48%"}

df %>%
  as.data.frame() %>%
  mutate(row = row_number(),
         score = iso_score_1,
         PC1 = df_prcomp$x[,1],
         PC2 = df_prcomp$x[,2]) %>%
  ggplot(aes(PC1, PC2)) +
  geom_point(aes(size = score),color = "seagreen") +
  ggrepel::geom_text_repel(aes(label = row), arrow = grid::arrow()) +
  scale_size_continuous(guide = "none") +
  theme_bw() +
  labs(title = "Isolation Tree based outlier\nVisualised on principal components")

df %>%
  as.data.frame() %>%
  mutate(row = row_number(),
         score = iso_score_1,
         PC1 = df_prcomp$x[,1],
         PC2 = df_prcomp$x[,2],
         is_outlier = factor(ifelse(ntile(iso_score_1, 10) >= 10, "O", "N"))) %>%
  ggplot(aes(PC1, PC2)) +
  geom_point(aes(size = score, color = is_outlier)) +
  ggrepel::geom_text_repel(aes(label = row), arrow = grid::arrow()) +
  scale_color_manual(values = c(O = "red", N = "black"), guide = "none") +
  scale_size_continuous(guide = "none") +
  theme_bw() +
  labs(title = "Isolation Tree based outlier\nTop-10%")
```

Isolation Tree Method
:::

In the right panel, the top 10% of records by score are shown in red. `ntile(x, 10)` divides the records into ten equal groups by score, so group 10 is the top 10%.

Let us now build a forest of 100 trees (@fig-an14). The right panel compares the scores of one tree with those of the forest. If both gave the same scores, all points would lie on the diagonal line. They do not, because a single tree depends heavily on its random splits. As the number of trees increases, the scores become stable.

::: {#fig-an14}

```{r warning=FALSE, message=FALSE, fig.align='center', fig.show='hold', out.width="48%"}
iso_100 <- isolation.forest(df, ntrees = 100, sample_size = 50,
                            ndim = 1, seed = 1, nthreads = 1)

# Generate scores
iso_score_100 <- predict(iso_100, df)

# View Results
df %>%
  as.data.frame() %>%
  mutate(row = row_number(),
         score = iso_score_100,
         PC1 = df_prcomp$x[,1],
         PC2 = df_prcomp$x[,2],
         is_outlier = factor(ifelse(ntile(iso_score_100, 10) >= 10, "O", "N"))) %>%
  ggplot(aes(PC1, PC2)) +
  geom_point(aes(size = score, color = is_outlier)) +
  ggrepel::geom_text_repel(aes(label = row), arrow = grid::arrow()) +
  scale_color_manual(values = c(O = "red", N = "black"), guide = "none") +
  scale_size_continuous(guide = "none") +
  theme_bw() +
  labs(title = "Isolation Tree based outlier\nNumber of Trees = 100")

plot(iso_score_1, iso_score_100, xlim = c(0, 1), ylim = c(0, 1),
     main = "Comparison of Tree Vs. Forest Method")
abline(a = 0, b = 1)

```

100 Isolation Trees (Left) Comparison of 1 vs. 100 trees (Right)
:::

**Contour plots:** We can also show the anomaly scores of a whole area as a contour plot, like the height lines of a map. We build a fine grid of points covering the first two principal components, score every grid point, and draw lines joining equal scores. See @fig-an16. Records falling in the high-score regions are anomalies.

```{r fig-an16, warning=FALSE, message=FALSE, fig.align='center', fig.show='hold', fig.cap="Contour Plot", out.width="90%"}
# Create PRCOMP data
df_grid <- data.frame(
  PC1 = df_prcomp$x[,1],
  PC2 = df_prcomp$x[,2]
)

# Create Sequences
pc1_seq <- seq(min(df_prcomp$x[,1]), max(df_prcomp$x[,1]), length.out = 25)
pc2_seq <- seq(min(df_prcomp$x[,2]), max(df_prcomp$x[,2]), length.out = 25)
# Create Grid
my_grid <- expand.grid(PC1 = pc1_seq, PC2 = pc2_seq)

# Create model for Pr comp
iso_model <- isolation.forest(df_grid, ntrees = 100, sample_size = 50,
                              ndim = 1, seed = 1, nthreads = 1)
# append scores
my_grid$scores <- predict(iso_model, my_grid)

# Draw Plot
library(lattice)
contourplot(scores ~ PC1 + PC2, data = my_grid, region = TRUE)
```

**Including categorical variables.** One benefit of the isolation forest is that it can use categorical columns directly. The only condition is that they should be of `factor` type.

**Problem Statement-2:** Let us now find anomalies in the full `iris` data, including the `Species` column. We can check the column types before proceeding.

```{r an15}
sapply(iris, class)
```

Our condition is met. So we can build the forest directly. Refer @fig-an18.

```{r fig-an18, warning=FALSE, message=FALSE, fig.align='center', fig.show='hold', fig.cap="Including categorical variable", out.width="98%"}
# New isolation forest model, including Species
iso_model_new <- isolation.forest(iris, ntrees = 100, ndim = 1,
                                  seed = 1, nthreads = 1)
new_scores <- predict(iso_model_new, iris)
head(new_scores)

# Full PRCOMP for Visual
iris_pr <- prcomp(iris[, 1:4])

data.frame(
  PC1 = iris_pr$x[,1],
  PC2 = iris_pr$x[,2]
) %>%
  mutate(score = new_scores,
         Species = iris$Species) %>%
  ggplot(aes(PC1, PC2, color = Species)) +
  geom_point(aes(size = score)) +
  guides(size = "none") +
  theme_bw() +
  labs(title = "Isolation Forest\nVisualised on Principal Components")
```

::: callout-tip
## Audit tip

Isolation forest is a good default for large audit data. It is fast, needs no scaling, handles many columns, and accepts categorical columns. A common practice is to run it on a table of indicators built for each vendor, employee or beneficiary. Then examine the top 1% or top 50 records by score.
:::

## Including categorical variables in LOF {#sec-ano-gower}

We can also include categorical variables in the Local Outlier Factor, using the **Gower distance**. Gower distance can measure the distance between records which have both numeric and categorical columns, even when the categories have no natural order. It is calculated with the function `daisy()` from the `cluster` package. Let us calculate LOF scores on the full `iris` data, including `Species`. The result is shown in @fig-an20.

```{r fig-an20, warning=FALSE, message=FALSE, fig.align='center', fig.show='hold', fig.cap="Including categorical variable", out.width="90%"}
library(cluster)
iris_dist <- daisy(iris, metric = "gower")
iris_lof <- lof(iris_dist, minPts = 5)

data.frame(
  PC1 = iris_pr$x[,1],
  PC2 = iris_pr$x[,2]
) %>%
  mutate(score = iris_lof,
         Species = iris$Species) %>%
  ggplot(aes(PC1, PC2, color = Species)) +
  geom_point(aes(size = score)) +
  guides(size = "none") +
  theme_bw() +
  labs(title = "LOF Scores\nVisualised on Principal Components")
```

### How does Gower distance work?

kNN and LOF need a *distance* between every pair of records. For numbers this is easy. But what is the distance between "Air" and "Sleeper", or between "Group A" and "Group C"? Gower distance answers this in a simple way.

-   For a **numeric** column, it takes the difference between the two values and divides it by the range (maximum minus minimum) of that column. So the result is always between 0 and 1.
-   For a **categorical** column, the distance is 0 if the two values are the same, and 1 if they are different.
-   The final distance is the average of these column-wise distances.

So every column contributes equally, and the final distance is always between 0 and 1. This is why we did not have to scale the data before using `daisy()`.

### Why categories matter, an audit example

Let us see why including categorical columns can make a real difference. Consider travel allowance (TA) claims of employees. Each claim records the grade of the employee, the mode of travel, the distance travelled and the amount claimed. Suppose the entitlement rules are as below.

| Grade | Entitled to travel by |
|---------|----------------------|
| Group A | Air or AC-2 |
| Group B | AC-2 or AC-3 |
| Group C | AC-3 or Sleeper |

: Travel entitlement in our example {#tbl-taentitle}

We have about 300 claims, created in the folded code below. Three problem claims have been planted. Two Group C employees have claimed *air* fare. Their amounts are quite normal *for an air journey* of that distance. The third claim is an AC-3 journey with a hugely inflated amount.

```{r}
#| code-fold: true
#| code-summary: "Code to create the TA claims data"
set.seed(2611)

entitled <- list(`Group A` = c("Air", "AC-2"),
                 `Group B` = c("AC-2", "AC-3"),
                 `Group C` = c("AC-3", "Sleeper"))
rate_per_km <- c(Air = 6, `AC-2` = 2.6, `AC-3` = 1.8, Sleeper = 0.7)

n <- 300
claims <- tibble(grade = sample(names(entitled), n, replace = TRUE,
                                prob = c(0.25, 0.35, 0.4))) |>
  mutate(mode        = map_chr(grade, \(g) sample(entitled[[g]], 1)),
         distance_km = round(runif(n, 200, 1800)),
         amount      = round(distance_km * rate_per_km[mode] * 2 *
                               runif(n, 0.9, 1.15) + 500))

planted <- tibble(grade       = c("Group C", "Group C", "Group B"),
                  mode        = c("Air", "Air", "AC-3"),
                  distance_km = c(1150, 1420, 900),
                  amount      = c(14300, 17600, 41000))

claims <- bind_rows(claims, planted) |>
  mutate(claim_id = sprintf("TA-%03d", row_number()), .before = 1)
```

```{r}
tail(claims, 4)
```

**Step-1: LOF on numeric columns only.** First, let us do what we did earlier in this chapter. We use only the two numeric columns, distance and amount, scale them, and calculate LOF.

```{r}
claims$lof_numeric <- claims |>
  select(distance_km, amount) |>
  scale() |>
  lof(minPts = 10)

claims |>
  arrange(desc(lof_numeric)) |>
  head(5)
```

The inflated AC-3 claim, `TA-303`, is at the top, with a very high score. That is good. But the two Group C air claims are nowhere near the top. This is not surprising. Looking at distance and amount alone, these claims look exactly like the many genuine air claims of Group A officers.

**Step-2: LOF with categorical columns, using Gower distance.** Now we include grade and mode of travel as well. `daisy()` needs categorical columns as factors, so we convert them first.

```{r}
gower_dist <- claims |>
  select(grade, mode, distance_km, amount) |>
  mutate(across(c(grade, mode), as.factor)) |>
  daisy(metric = "gower")

claims$lof_gower <- lof(gower_dist, minPts = 10)

claims |>
  arrange(desc(lof_gower)) |>
  head(5)
```

Now all three planted claims are at the top. A Group C employee travelling by air is a rare *combination* in the data, and Gower distance makes that rarity visible to LOF.

Let us compare the rank of the three claims under both methods. A rank of 1 means the most anomalous claim.

```{r}
claims |>
  mutate(rank_numeric = rank(-lof_numeric),
         rank_gower   = rank(-lof_gower)) |>
  filter(claim_id %in% c("TA-301", "TA-302", "TA-303")) |>
  select(claim_id, grade, mode, amount, rank_numeric, rank_gower)
```

::: callout-tip
## Audit tip

If the rules are known and simple, as in our entitlement table, a direct rule-based check is better. Just join the claims with the entitlement table and list every claim where the mode is not allowed for the grade. Anomaly detection is useful when the rules are not known, are too many to code, or when we want to discover patterns that nobody has thought of. In practice, use both.
:::

## Time series anomalies {#sec-ano-ts}

In time series data, an anomaly is a value which does not follow the usual trend, seasonal pattern or cycle of the series, and is clearly different from the rest of the data.

To detect such anomalies in R, we will use the package `timetk`. This package works on a `tibble` (data frame) with a date column, instead of a `ts` object. So we may need to prepare our data accordingly.

**Problem Statement:** Let us find anomalies, if any, in the monthly sunspots data available in base R^[Monthly numbers of sunspots, as from the World Data Center, aka SIDC. This is the version of the data that will occasionally be updated when new counts become available.].

```{r fig-an17, warning=FALSE, message=FALSE, fig.align='center', fig.show='hold', fig.cap="Time Series Anomalies", out.width="90%"}
start_date <- as.Date("1749-01-01")
# one date per month, as many as there are values in the series
date_sequence <- seq(start_date, by = "months", length.out = length(sunspot.month))

sunspot_df <- sunspot.month %>%
  as.data.frame() %>%
  set_names("value") %>%
  mutate(date = date_sequence)

library(timetk)
sunspot_df %>%
  plot_anomaly_diagnostics(date, value,
                           .interactive = FALSE)

sunspot_df %>%
  tk_anomaly_diagnostics(date, value) %>%
  filter(anomaly == 'Yes')
```

We can see the anomalies highlighted in red in @fig-an17, and listed by the code above. We may also apply the Seasonal Hybrid ESD algorithm discussed in @sec-ano-shesd to the same data.

### How does this method work?

`tk_anomaly_diagnostics()` works in two steps.

1.  It first splits the series into three parts, using a method called STL decomposition. The **trend** is the long-term rise or fall. The **season** is the pattern which repeats every year, or every week. The **remainder** is whatever is left after removing the trend and season.
2.  It then looks for unusually large values in the remainder only. The usual rule of the boxplot is used, so a value far outside the middle 50% of the remainders is called an anomaly.

So a value is called an anomaly only if it cannot be explained by the trend *and* the usual seasonal pattern. We learnt about these components of a time series in @sec-ts.

### An audit example, monthly expenditure

Every auditor in government knows the *March rush*. Offices spend much more in March than in other months, to use up the budget before the financial year ends. So a high expenditure in March is expected. What an auditor wants to find is a month which is unusual *even after allowing* for this pattern.

Let us create six years of monthly expenditure of an office, from April 2019 to March 2025, in ₹ lakh. The expenditure grows a little every year, and March is always much higher than other months. We have planted one odd month, October 2023, when some unusual payments were made.

```{r}
#| code-fold: true
#| code-summary: "Code to create the monthly expenditure data"
set.seed(2611)

months <- seq(as.Date("2019-04-01"), as.Date("2025-03-01"), by = "month")

# Seasonal pattern from April to March. March is the rush month.
season <- c(0.8, 0.85, 0.9, 0.9, 0.95, 0.95, 1, 1, 1.05, 1.1, 1.3, 2.4)
position_in_fy <- ((month(months) + 8) %% 12) + 1   # April = 1, ..., March = 12

expenditure <- tibble(
  month = months,
  amount = round(40 * 1.06^(seq_along(months) / 12) * season[position_in_fy] *
                   runif(length(months), 0.92, 1.08), 2)
)

# Planted case. Unusual payments in October 2023.
odd <- expenditure$month == as.Date("2023-10-01")
expenditure$amount[odd] <- expenditure$amount[odd] * 2.3
```

```{r}
head(expenditure)
```

**Attempt 1. A simple boxplot rule.** Let us first try what we learnt at the start of this chapter, using `boxplot.stats()`.

```{r}
outliers <- boxplot.stats(expenditure$amount)$out

expenditure |>
  filter(amount %in% outliers)
```

Every single March has been flagged, along with October 2023. Even February 2025 is flagged, only because expenditure has grown over the years and recent months are naturally higher. This is not useful. The March peaks are the normal pattern of this office, and the one really odd month is lost in the crowd.

**Attempt 2. Allowing for the season.** Now let us use `tk_anomaly_diagnostics()`. We tell it that the season repeats every 12 months, with `.frequency = 12`.

```{r}
expenditure |>
  tk_anomaly_diagnostics(month, amount, .frequency = 12, .message = FALSE) |>
  filter(anomaly == "Yes") |>
  select(month, observed, season, trend, remainder)
```

This is much better. October 2023 is flagged. But so are a few early Marches. Why? Look at how the March peak behaves. It is not a fixed *amount* above other months. It is a fixed *proportion*, about 2.4 times a normal month. As total expenditure grows over the years, the March peak grows in rupees too. STL expects a season of the same size every year, so it gets the early Marches slightly wrong.

**Attempt 3. Working on the log scale.** The fix is simple. We take the logarithm of the amounts. On the log scale, "2.4 times" becomes a fixed *difference*, whatever the size of the expenditure. So the seasonal pattern becomes the same size every year.

```{r}
expenditure |>
  mutate(log_amount = log(amount)) |>
  tk_anomaly_diagnostics(month, log_amount, .frequency = 12, .message = FALSE) |>
  filter(anomaly == "Yes") |>
  select(month, observed, remainder)
```

Now only October 2023 is flagged. This is exactly the month we planted. The same result can be seen in @fig-expanom.

```{r}
#| label: fig-expanom
#| fig-cap: "Anomalies in monthly expenditure, on log scale"
#| message: false
expenditure |>
  mutate(log_amount = log(amount)) |>
  plot_anomaly_diagnostics(month, log_amount, .frequency = 12,
                           .interactive = FALSE, .message = FALSE,
                           .title = "Monthly expenditure (log scale)")
```

::: callout-tip
## Audit tip

Use the log scale whenever the ups and downs of a series grow along with its level. This is true of most financial data, like expenditure, revenue, sales or tax collections. A quick look at a line plot of the data will tell you. If the March peaks get taller over the years, use the log.
:::

::: callout-note
## The March rush is itself an audit issue

The seasonal method treats the March rush as *normal*, because it happens every year. But for an auditor, the rush itself may be a finding. Ministry of Finance guidelines on expenditure management require that expenditure in March should not exceed 15% of the budget for the year. We can check this directly, by finding the share of March in each financial year's expenditure.

```{r}
expenditure |>
  mutate(fin_year = if_else(month(month) >= 4, year(month), year(month) - 1)) |>
  summarise(total = sum(amount),
            march = sum(amount[month(month) == 3]),
            .by = fin_year) |>
  mutate(march_share_pct = round(100 * march / total, 1))
```

Here `fin_year` 2019 means the financial year 2019-20, and so on. In every year, March accounts for more than 15% of the annual expenditure.
:::

## When did behaviour change? Change points {#sec-ano-changepoint}

The time series methods of @sec-ano-ts find *single periods* that are unusual. Sometimes the question is different. *When* did a series shift to a new level, and stay there? A vendor's rate that steps up when a new officer takes charge, a sudden and lasting rise in a hospital's claims after it was empanelled with a new third-party administrator, or collections that drop after a change in the inspection system. Such a lasting shift is not one odd month, so methods looking for odd months may miss it. **Change point detection** looks for exactly these shifts.

Let us take the rate charged per bag of cement by one supplier to a public works division, month by month for four years.

```{r}
#| code-fold: true
#| code-summary: "Code to create the cement rates"
set.seed(7)
cement <- tibble(month = seq(as.Date("2021-04-01"), by = "month", length.out = 48),
                 # Planted. A lasting step up after the 30th month, and a one-off spike in month 12
                 rate  = round(c(rnorm(30, 365, 8), rnorm(18, 405, 8))))
cement$rate[12] <- 440
```

The package `changepoint` finds the points where the *mean* of a series changes, with `cpt.mean()`. The `PELT` method searches efficiently for any number of change points. One preparation is needed. The function measures changes in units of the series' natural month-to-month noise, and assumes that noise has a standard deviation of 1. So we first divide the series by a robust estimate of its noise, the MAD of the month-to-month differences divided by $\sqrt{2}$.

```{r}
#| message: false
#| warning: false
library(changepoint)
noise <- mad(diff(cement$rate)) / sqrt(2)
cp <- cpt.mean(cement$rate / noise, method = "PELT")
cpts(cp)
```

```{r}
#| include: false
cps <- cpts(cp)
```

`cpts()` gives the positions after which the mean changes: `r paste(cps, collapse = ", ")`. Let us read them with the dates. The changes after months `r cps[1]` and `r cps[2]` enclose a single month, `r format(cement$month[cps[2]], "%B %Y")`. That is the one-off spike, a *point* anomaly, which the method sees as a tiny segment of its own. The change after month `r cps[3]`, `r format(cement$month[cps[3]], "%B %Y")`, is the lasting shift. The average rate in each segment shows its size.

```{r}
cement |>
  mutate(segment = cut(row_number(), breaks = c(0, cps, nrow(cement)))) |>
  summarise(from = min(month), to = max(month), months = n(), mean_rate = round(mean(rate)),
            .by = segment)
```

```{r}
#| fig-cap: "Cement rates with the detected segments"
#| fig-height: 3.5
cement |>
  mutate(segment = cut(row_number(), breaks = c(0, cps, nrow(cement)))) |>
  mutate(segment_mean = mean(rate), .by = segment) |>
  ggplot(aes(month, rate)) +
  geom_line(colour = "grey50") +
  geom_point() +
  geom_step(aes(y = segment_mean), colour = "red", linewidth = 1, direction = "mid") +
  labs(x = NULL, y = "Rate per bag (₹)") +
  theme_minimal()
```

From `r format(cement$month[cps[3] + 1], "%B %Y")`, the supplier has charged about ₹`r round(mean(cement$rate[(cps[3] + 1):48]) - mean(cement$rate[setdiff(1:cps[3], cps[2])]))` more per bag, month after month. The audit questions follow naturally. Did the market rate rise at the same time? Did other suppliers' rates rise too? Was the rate contract revised, and by whom? Did anything else change that month, like the officer in charge?

::: callout-tip
## Audit tip

Run change point detection on many series at once, say the rate of every major item from every supplier, with `map()` (@sec-purrr), and list the series with a lasting shift, the date and the size of the shift. Then compare the dates with known events: rate revisions, transfers of officers, new contracts. A shift with no explanation is a strong lead. `cpt.meanvar()` looks for changes in both the mean and the variability, which can reveal, for example, a period when rates suddenly stopped varying, as if copied.
:::

## Combining several methods {#sec-ano-combine}

Each method in this chapter sees anomalies in its own way. A robust z-score sees one extreme measure. The Mahalanobis distance sees a broken relationship. An isolation forest sees records that are easy to separate from the rest, and LOF sees records unlike their own neighbours. A record flagged by several methods is a stronger lead than one flagged by a single method, and different methods catch different kinds of irregularity. So in practice we run several, and combine their verdicts into **one ranked list** for examination.

The scores of different methods are on different scales: a z-score of 4, a Mahalanobis distance of 25, an isolation score of 0.6. They cannot be added directly. The simplest fair way is to convert each method's scores into **ranks**, or rather percentile ranks between 0 and 1, and average them. A record which every method places near the top gets a combined score near 1.

Let us add a third measure to our PHC data, medicine expenditure, and plant two more cases of different kinds.

```{r}
#| code-fold: true
#| code-summary: "Code to add medicine expenditure and two more planted cases"
set.seed(21)
phc <- phc |>
  mutate(medicine_lakh = round(opd * 120 * rlnorm(n(), 0, 0.15) / 1e5, 2))
# Planted. PHC-140 spends far more on medicines than its OPD explains
phc$medicine_lakh[140] <- round(phc$medicine_lakh[140] * 2.6, 2)
# Planted. PHC-021 reports far less of everything than any other PHC
phc$opd[21]           <- round(min(phc$opd) * 0.45)
phc$deliveries[21]    <- max(1, round(min(phc$deliveries) * 0.45))
phc$medicine_lakh[21] <- round(min(phc$medicine_lakh) * 0.45, 2)
```

Now we score every PHC with four methods. All of them work on the logarithms of the three measures.

```{r}
#| message: false
library(isotree)
library(dbscan)

logs <- phc |> transmute(l_opd = log(opd), l_del = log(deliveries), l_med = log(medicine_lakh))

scores <- phc |>
  transmute(
    phc,
    # 1. largest robust z-score among the three measures, and medicine per OPD patient
    robust_z = pmax(abs(robust_z(logs$l_opd)), abs(robust_z(logs$l_del)),
                    abs(robust_z(logs$l_med)), abs(robust_z(logs$l_med - logs$l_opd))),
    # 2. robust Mahalanobis distance
    mahal = {
      fit <- MASS::cov.rob(as.matrix(logs), method = "mcd")
      mahalanobis(as.matrix(logs), fit$center, fit$cov)
    },
    # 3. isolation forest score
    iforest = {
      set.seed(1)
      predict(isolation.forest(logs, ntrees = 200), logs)
    },
    # 4. local outlier factor, on scaled values
    lof = lof(scale(logs), minPts = 10)
  )
```

Each column is a score where *higher means more unusual*. Now we turn each into a percentile rank with `percent_rank()`, and average.

```{r}
combined <- scores |>
  mutate(across(c(robust_z, mahal, iforest, lof), percent_rank, .names = "pr_{.col}")) |>
  mutate(combined = (pr_robust_z + pr_mahal + pr_iforest + pr_lof) / 4) |>
  arrange(desc(combined))

combined |>
  select(phc, starts_with("pr_"), combined) |>
  mutate(across(where(is.numeric), \(x) round(x, 3))) |>
  head(8)
```

```{r}
#| include: false
planted <- c("PHC-021", "PHC-077", "PHC-140")
avg_rank_077 <- which(combined$phc == "PHC-077")
```

PHC-021 and PHC-140 top the list. But where is PHC-077, the PHC whose deliveries were out of line with its OPD? It is at position `r avg_rank_077`. Let us see how each method ranked the three planted PHCs, out of `r nrow(phc)`.

```{r}
map(c(robust_z = "robust_z", mahalanobis = "mahal", isolation_forest = "iforest", lof = "lof"),
    \(col) rank(-combined[[col]])[match(planted, combined$phc)]) |>
  as_tibble() |>
  mutate(phc = planted, .before = 1)
```

Three methods placed PHC-077 near the top, but the robust z-score, which looks at one measure at a time, saw nothing unusual in it, exactly as we found in @sec-ano-mahal. Its low rank on that one method pulled the average down. This is the weakness of averaging: it *dilutes* an anomaly that only some methods can see, and different kinds of anomaly are, by their nature, seen by different methods.

A remedy that auditors find natural is a **vote**. Count, for each record, how many methods place it in their top 5%, and rank records first by the number of votes, then by the average.

```{r}
top_share <- 0.05
voted <- combined |>
  mutate(votes = (pr_robust_z >= 1 - top_share) + (pr_mahal >= 1 - top_share) +
                 (pr_iforest >= 1 - top_share) + (pr_lof >= 1 - top_share)) |>
  arrange(desc(votes), desc(combined))

voted |>
  select(phc, votes, combined) |>
  mutate(combined = round(combined, 3)) |>
  head(8)
```

```{r}
#| include: false
vote_pos <- match(planted, voted$phc)
n_two_plus <- sum(voted$votes >= 2)
```

Now the three planted PHCs are at positions `r paste(sort(vote_pos), collapse = ", ")`. A record that most methods flag, for whatever reason each has, rises to the top, while a record that one method alone happens to place high stays lower. In all, `r n_two_plus` PHCs are flagged by two or more methods, a short and well-founded list for examination. Not every record on it is planted. The others with several votes are simply the smallest or largest PHCs in the data, and an examination would quickly clear them. That is normal: anomaly methods find the *unusual*, and only the records can say whether it is *wrong*.

::: callout-tip
## Audit tip

Keep the individual method scores alongside the combined score in your working papers. When a record is examined, they explain *why* it was selected: "medicine spending three times what its OPD explains" (robust z of the ratio), or "deliveries out of line with OPD" (Mahalanobis). An explained selection is much easier to defend than a single number.
:::

::: callout-note
Averaging and voting both give every method an equal say. If experience shows that one method finds more real issues in your data, give it a larger weight. And when outcomes of past examinations are available, consider the supervised methods of @sec-logistic and @sec-forest, which learn directly from what was found.
:::

## Choosing a method

With so many methods, which one should an auditor use? @tbl-anochoose gives a simple guide.

| Your data | Suggested method |
|----------------------------------|-------------------------------|
| One amount column, quick check | Boxplot, drawn group-wise |
| Entities that fall into natural peer groups | Robust z-score of a size-free ratio, within each group |
| A few related measures for each entity | Robust Mahalanobis distance |
| One column, roughly bell-shaped, one suspect value | Grubbs' test |
| A monthly, weekly or daily series | STL decomposition with `timetk`, on log scale for money |
| A series that may have shifted to a new level | Change point detection with `changepoint` |
| Several methods, one list to examine | Average of percentile ranks |
| A table of numeric indicators for each vendor, employee etc. | Isolation forest first, then LOF to look for local anomalies |
| Numeric and categorical columns together | LOF with Gower distance, or isolation forest with factors |
| Known rules, like entitlements or limits | A direct rule-based check, before any of the above |

: Choosing an anomaly detection method {#tbl-anochoose}

## A suggested workflow

1.  **Understand the data.** Know what each column means, and what "normal" looks like for the business.
2.  **Apply the known rules first.** Limits, entitlements and rates can be checked directly, without any algorithm.
3.  **Look at the data.** Draw boxplots and histograms, group-wise. Many anomalies are visible at this stage itself.
4.  **Prepare the data for the algorithm.** Scale numeric columns for kNN and LOF. Use the log for money amounts. Convert categorical columns to factors.
5.  **Compare like with like.** Use peer groups and size-free ratios, so that big entities are not flagged merely for being big.
6.  **Run more than one method, and combine them by rank.** A record flagged by several methods is a stronger lead.
7.  **Rank, do not label.** Sort records by score and examine them from the top.
8.  **Verify with documents.** Explain every flagged record from the records, before treating it as a finding.

::: callout-important
## Remember

An anomaly is only *unusual*. It is not necessarily *wrong*. Many anomalies have perfectly good explanations, like a one-time bulk purchase or a new scheme. Anomaly detection tells the auditor *where to look*. Only the examination of records can tell *what happened*.
:::

## Summary of functions used

| Function | Package | What it does |
|-----------------------|----------------|--------------------------------|
| `summary()`, `boxplot()` | base R | Quick view of spread and extreme values |
| `boxplot.stats()$out` | grDevices | Lists the values beyond the boxplot whiskers |
| `grubbs.test()` | outliers | Tests whether the most extreme value is an outlier |
| `median()`, `mad()` | stats | Robust centre and spread, for robust z-scores |
| `cov.rob(method = "mcd")`, `mahalanobis()` | MASS, stats | Robust centre and covariance, and Mahalanobis distances |
| `qchisq()` | stats | Cut-off for squared Mahalanobis distances |
| `AnomalyDetectionVec()` | AnomalyDetection | Seasonal Hybrid ESD on a single series |
| `scale()` | base R | Puts all numeric columns on the same scale |
| `get.knn()` | FNN | Finds the k nearest neighbours and their distances |
| `lof()` | dbscan | Local Outlier Factor scores |
| `isolation.forest()`, `predict()` | isotree | Builds an isolation forest and gives anomaly scores |
| `daisy(metric = "gower")` | cluster | Distance for mixed numeric and categorical data |
| `prcomp()` | stats | Principal components, used here for plotting |
| `tk_anomaly_diagnostics()`, `plot_anomaly_diagnostics()` | timetk | Time series anomalies after removing trend and season |
| `cpt.mean()`, `cpt.meanvar()`, `cpts()` | changepoint | Find change points in the mean (and variance) of a series |
| `percent_rank()` | dplyr | Percentile ranks, for combining scores of different methods |

: Functions used in this chapter {#tbl-anofunctions}
