```{r}
#| include: false
source("_common.R")
```

# Anomaly Detection
```{r an100, echo=FALSE, warning=FALSE, message=FALSE}
library(tidyverse)
library(magrittr)
```

Anomalies play a significant role in the field of audit. As auditors start examining data whether it be financial statements, or transactions, or other relevant data, the detection of anomalies can provide valuable insights and help identify potential issues or irregularities. Anomalies often indicate the presence of fraudulent activities.   Unusual or unexpected patterns in financial transactions or account balances may suggest potential fraud or misappropriation of assets. Auditors actively search for anomalies that may indicate fraudulent behavior, such as fictitious transactions, unauthorized access, or unusual changes in financial data.  

Anomalies in the audit context serve as indicators of potential issues, including fraud, material misstatements, control weaknesses, compliance violations, and process inefficiencies. Detecting and investigating anomalies is crucial for auditors to provide accurate and reliable financial information, enhance internal controls, and support informed decision-making by stakeholders.

## Definition and types of anomalies
Anomalies are patterns or data points that deviate significantly from the expected or normal behavior within a dataset. These are also known as outliers, novelties, or deviations and can provide valuable insights into unusual events or behavior in various domains.

Types of anomalies:

- **Point Anomalies:** Individual data points that are considered anomalous when compared to the rest of the dataset. For example, a temperature sensor reading that is significantly different from the expected range.

- **Contextual Anomalies:** Data points that are considered anomalous within a specific context or subset of the dataset. For instance, a sudden increase in obsolete website traffic.

- **Collective Anomalies:** Groups of data points that exhibit anomalous behavior when analyzed together but might appear normal when considered individually. An example is a sudden drop in sales for multiple related products.

**Anomaly detection, in R** 

## Anomaly detection - by inspection
Several times, anomalies or outliers are detectable by observation.  The `summary()` function prints the *maximum*, *minimum*, *upper* and *lower quartiles*, and the *mean* and *median*, and can give a sense for how far an extreme point lies from the rest of the data. E.g. Let's visualise the mean annual temperatures in degrees Fahrenheit in New Haven, Connecticut, from 1912 to 1971, through a boxplot (Refer @fig-an1.  This data is available in base R, as `nhtemp`.

```{r an101}
summary(nhtemp)
```

The easiest way to get a sense for how unusual a particular value is is by using a graphical summary like a boxplot. In R, this is created using the `boxplot` function. The `boxplot` function takes a column of values as an input argument, here illustrated with the temperature data, and produces a box and whiskers representation of the distribution of the values. The extreme values are represented as distinct points, making them easier to spot.  We can also make use of ggplot2.  Examples from both base R and ggplot2 are shown in @fig-an1.

::: {#fig-an1}

```{r fig.show='hold', fig.align='center', out.width="45%"}
boxplot(nhtemp, main = "Box plot of nhtemp data")
ggplot(iris, aes(Species, Petal.Length, color = Species )) +
  geom_boxplot(outlier.colour = "red", outlier.size = 2) +
  theme_bw() +
  labs(title = "Petal Lengths in Iris Data")
```

Outliers through Boxplot
:::

It's important to note that a **point anomaly** is not necessarily always extreme. A point anomaly can also arise as an unusual combination of values across several attributes.  

A **collective anomaly**, on the other hand, is a collection of similar data instances that can be considered anomalous together when compared to the rest of the data. For example, a consecutive 5 day period of high temperatures are shown by the red points in the plot. These daily temperatures are unusual because they occur together, and are likely caused by the same underlying weather event. Refer @fig-an2.

```{r fig-an2, echo=FALSE, fig.align='center', fig.show='hold', fig.cap="Collective Anomalies", out.height="27%"}
set.seed(567)
c(rnorm(15, 30, 2), rnorm(5, 40, 2.1), rnorm(10, 30, 2)) %>% ts() -> myts
dates <- 1:30
plot(dates, myts, type = 'l',ylab = "Temp in Celsius", xlab = 'Dates', 
     main = "Collective Anomalies through a dummy data")
points(16:20, myts[16:20], col = 'red', pch = 16)
points(1:15, myts[1:15], col = 'black', pch = 16)
points(21:30, myts[21:30], col = 'black', pch = 16)
```

## Grubb's Test
We saw that visual assessment of outliers works well when the majority of data points are grouped together and rest of them lie separate.  In statistics there are several tests to detect anomalies.  **Grubb's Test** is one of these. Grubb's test assesses whether the point that lies farthest from the mean in a dataset could be an outlier. This point will either be the largest or smallest value in the data.  This test is however based on assumption that data points are normally distributed.  So before proceeding for this test, we must be sure that there is a plausible explanation for this assumption. *(We can check normality of data points by plotting a histogram.  For other methods please refer to chapter on linear regression.)*

Let's run this test on `nhtemp` data example, which we have already seen in @fig-an1 (Left). So let's check its normality assumption.
```{r fig-an3, fig.align='center', fig.show='hold', fig.cap="Histogram for Grubb's test", out.height="27%"}
hist(nhtemp, breaks = 20)
```

In @fig-an3 we can see that our assumption is nearly satisfied.  Let's run the test.

```{r an}
library(outliers)
outliers::grubbs.test(nhtemp)
```

As p value is `0.15` we do not have strong evidence to reject null hypothesis that extreme maximum value is an outlier.

## Seasonal Hybrid ESD Algorithm

As we have seen that above test may not be appropriate for anomaly detection (normality assumption as well as detection of extreme values only), particularly detecting anomalies from a time series that may have seasonal variations.  We may install `AnomalyDetection` package development version from github using `devtools::install_github("twitter/AnomalyDetection")`.  Example: in `JohnsonJohnson` data having quarterly sales data, we can use the following syntax.

```{r fig-an4, fig.show='hold', fig.align='center', fig.cap="Seasonal Hybrid ESD Algorithm", message=FALSE, out.height="28%"}
#devtools::install_github("twitter/AnomalyDetection")
library(AnomalyDetection)
AnomalyDetectionVec(as.vector(JohnsonJohnson), 
                    period = 4, # Sesaonality
                    plot = TRUE, # set FALSE when plot not needed
                    direction = 'both' # set 'pos' for higher values
                                       # or 'neg' for lower values
                    )
```

In @fig-an4, we can see that in output `$anoms` containing anomalies are denoted in blue dots.

## k-Nearest Neighbour Distance Score
One of greatest limitation of above two methods was that, these were applicable on univariate data series, whereas, in real world, data analysis will rarely be univariate.  The `knn` technique works on multivariate data, but for visulaisation purposes, we will first visulaise the results on bivariate data only. 

**K-Nearest Neighbour** or **KNN** is a distance-based classifier, meaning that it implicitly assumes that the smaller the distance between two points, the more similar they are.  For bivariate data, we can understand the algorithm by plotting the data-points in a two dimensional scatter plot.  Now as the distance between any two points are usually calculated using *Euclidean Distance Metric*, we should ensure that the data is normalised/scaled before proceeding for distance calculation.

**Problem Statement-1:**  Let us try to identify outliers in `virginica` Species' flower measurements.  So let's prepare the data.   We will make use of `scale` function available in base R, to normalise the data.  Remember that `scale` returns a matrix, so we may need to convert it into data.frame while using `ggplot2`.

```{r an102}
# Sample data
df <- iris[101:150, 1:4]
#  Scale the data
df <- scale(df)
```

Let us visualise the data (in first two dimensions only). However, as our data is scaled already, we can try to visualise all the dimensions using a boxplot.  See both figures in @fig-an5.  The points/ouliers have been numbered (on the basis of row numbers) for interpretaion purposes.

::: {#fig-an5}

```{r warning=FALSE, message=FALSE, fig.align='center', fig.show='hold', out.width="45%"}

df %>% 
  as.data.frame() %>% 
  mutate(row = row_number()) %>% 
  ggplot(aes(Sepal.Length, Sepal.Width)) +
  geom_point(color = "seagreen") +
  ggrepel::geom_text_repel(aes(label = row), arrow = grid::arrow()) +
  theme_bw()

# Helper Function
find_outlier <- function(x) {
  return(x < quantile(x, .25) - 1.5*IQR(x) | x > quantile(x, .75) + 1.5*IQR(x))
}

df %>% 
  as.data.frame() %>% 
  mutate(row = row_number()) %>% 
  pivot_longer(-row, names_to = "Dimension", values_to = "Values") %>% 
  group_by(Dimension) %>% 
  mutate(outlier = ifelse(find_outlier(Values), row, NA)) %>% 
  ggplot(aes(Dimension, Values)) +
  geom_boxplot(outlier.colour = "red") +
  geom_text(aes(label=outlier), na.rm=TRUE, hjust=-.5) +
  theme_bw()
```

Sepal Length Vs. Widths in Virginica
:::

Now, let's proceed to identify outliers in R. We will make use of `get.knn` function from `FNN` package.  However, `k` parameter is required beforehand.

```{r an6, warning=FALSE, message=FALSE}
# Load the library
library(FNN)
# get kNN object, using k = 5
viginica_knn <- FNN::get.knn(df[, 1:2], 5)
```

The `knn` object created above will have two sub-objects (both matrices having columns equal to chosen `k`), one having nearest neighbors' indices and another having distances from those.  Let's view their top 6 rows.

```{r an7}
head(viginica_knn$nn.index)
head(viginica_knn$nn.dist)
```

Using `rowMeans` we can calculate mean distance for each data point.  The bigger this score is, the chances of that record being an outlier are relatively higher.  Let's also store that mean distance in a variable/column say `score` in main dataset, and visualise the results by setting the point-size with this mean distance (actually its square root).  In @fig-an8, we may notice that points lying far away are bigger becuase their chances of being outliers is high.

```{r fig-an8, warning=FALSE, message=FALSE, fig.align='center', fig.show='hold', fig.cap="k-Nearest Neighbour Distance", out.height="28%"}
score <- rowMeans(viginica_knn$nn.dist)
df %>% 
  as.data.frame() %>% 
  mutate(row = row_number(),
         score = score) %>% 
  ggplot(aes(Sepal.Length, Sepal.Width)) +
  geom_point(aes(size = score),color = "seagreen") +
  ggrepel::geom_text_repel(aes(label = row), arrow = grid::arrow()) +
  theme_bw() +
  labs(title = "KNN method of outlier")

```

Now, we can run the algorithm to find the outlier on the basis of all variables in the dataset.

```{r an88}
virginica_knn <- FNN::get.knn(df, 5)
score <- rowMeans(virginica_knn$nn.dist)
# Which point is farthest-Outlier
which.max(score)
```


## Local Outlier Factor
As against kNN which uses distances of k neighbors, this algorithm uses density of each data point vis-a-vis density of its nearest neighbors. kNN distance seems to be good at detecting points that are really far from their neighbors, sometimes called global anomalies, but sometimes fail to capture all of the points that might be considered anomalous like local anomalies.  Local Anomalies may lie near to a cluster still they won't be like their neighbors. To understand it better, see the plot in @fig-an9 consisting of dummy data.

```{r fig-an9, echo=FALSE, warning=FALSE, message=FALSE, fig.align='center', fig.show='hold', fig.cap="Local Outlier factor", out.height="28%"}
set.seed(123)
dummy <- data.frame(
  sales = rnorm(25, 500, 100),
  profit = rnorm(25, 75, 10)
)
dummy2 <- data.frame(
  sales = rnorm(25, 1000, 5),
  profit = rnorm(25, 75, 2)
)
dummy3 <- data.frame(
  sales = rnorm(5, 500, 100),
  profit = rnorm(5, 120, 10)
)
dummy4 <- data.frame(
  sales = rnorm(5, 1000, 10),
  profit = rnorm(5, 90, 2)
)
dummy <- rbind(dummy, dummy2, dummy3, dummy4)

plot(profit ~ sales, data = dummy, xlim = c(0, 1100), ylim = c(0, 200))
points(profit ~ sales, data = dummy[51:55,], xlim = c(0, 1100), ylim = c(0, 200), col = 'red', pch = 20)
points(profit ~ sales, data = dummy[56:60,], xlim = c(0, 1100), ylim = c(0, 200), col = 'blue', pch = 20)
```

The red points may be global outliers, being far from its immediate neigbors, yet the blue points may be local anomalies as these are not like their immediate neighbors and may be local anomalies.

As stated, LOF segregates the data points based on the ratio of density of that point with that of densities of its neighbors.  A score `> 1` thus indicates that the data point may be an anomaly.  Let's see that on same problem statement.

::: {#fig-an10}

```{r warning=FALSE, message=FALSE, fig.align='center', fig.show='hold', out.width="48%"}
library(dbscan)
lof_score <- lof(df, 5)
df %>% 
  as.data.frame() %>% 
  mutate(row = row_number(),
         score = lof_score) %>% 
  ggplot(aes(Sepal.Length, Sepal.Width)) +
  geom_point(aes(size = score),color = "seagreen") +
  ggrepel::geom_text_repel(aes(label = row), arrow = grid::arrow()) +
  scale_size_continuous(guide = FALSE) +  
  theme_bw() +
  labs(title = "LOF method of outlier")

df_prcomp <- prcomp(df, scale. = TRUE)

df %>% 
  as.data.frame() %>% 
  mutate(row = row_number(),
         score = lof_score,
         PC1 = df_prcomp$x[,1],
         PC2 = df_prcomp$x[,2]) %>% 
  ggplot(aes(PC1, PC2)) +
  geom_point(aes(size = score),color = "seagreen") +
  ggrepel::geom_text_repel(aes(label = row), arrow = grid::arrow()) +
  scale_size_continuous(guide = FALSE) +  
  theme_bw() +
  labs(title = "LOF method of outlier\nVisualised on principal components")
```

LOF
:::

Clearly, in @fig-an10 (left), where we have plotted the data in 2 dimensions only despite that we have attempted to find the LOF on the basis of all four dimensions. We can see presence of local anomalies, within the clustered data points. E.g. points nos. 31, 23, etc. were earlier given more weight instead of point nos.15, 10, etc. which are now given more weight.  However, if we want to visaulise it on first two principal components, we can do that (Refer @fig-an10 (Right)). We can also verify this.

```{r an11}
# Biggest Anomaly - kNN
which.max(score)

# Biggest Anomaly - LOF
which.max(lof_score)
```

Histograms of both `knn` and `LOF` scores can also be drawn, as in @fig-an12.

::: {#fig-an12}

```{r warning=FALSE, message=FALSE, fig.align='center', fig.show='hold', out.width="48%"}
hist(score, breaks = 40, 
     main = "Histogram of KNN scores in IRIS-Virginica data")
hist(lof_score, breaks = 40, 
     main = "Histogram of LOF scores in IRIS-Virginica data")
```

Histogram - KNN(Left) and LOF (Right)
:::

## Isolation Trees/Forest

Isolation Forest is an unsupervised tree based model, which actually works on path length instead of distance or density measures.  Tree based models use decision tree(s) to determine the value/class of an observation given its value. In other words, it determines or try to identify a path from the root node to a leaf node based on the value of the observation in question. Forest (or Random Forest) are actually collection of smaller trees and thus using ensemble learning methods to make decision, instead of a single complex tree.

Let us work on the same problem statement above.  Firstly, we will build a single decision tree.  WE will use a development version package `isofor` which can be downloaded using `remotes::install_github("Zelazny7/isofor")`.  To build forest/tree we will use function `iForest`.  Its argument `nt` will determine the number of trees to be built in ensemble.  Another important argument is `phi` which determines the number of samples to draw without replacement to construct each tree.  So let's use `nt = 1` and `phi = 100`.

```{r an13, message=FALSE, warning=FALSE}
# remotes::install_github("Zelazny7/isofor")
library(isofor)
# Generate a single tree
# Specify number of samples explicitly
viginica_1 <- iForest(df, nt = 1, phi = 100)

# Generate isolation score
iso_score_1 <- predict(viginica_1, df)

# View fisrt 10 scores
iso_score_1[1:10]
```
*Score interpretations*: The closer the score is to 1, the more likely the point is an anomaly.  However, if their scores are below 0.5, they are probably just normal points within the trend. 

Let's just visualise the scores on Principal Component plot again.

```{r fig-an24, message=FALSE, warning=FALSE, fig.align='center', fig.show='hold', fig.cap="Isolation Tree Method", out.width="48%"}

df %>% 
  as.data.frame() %>% 
  mutate(row = row_number(),
         score = iso_score_1,
         PC1 = df_prcomp$x[,1],
         PC2 = df_prcomp$x[,2]) %>% 
  ggplot(aes(PC1, PC2)) +
  geom_point(aes(size = score),color = "seagreen") +
  ggrepel::geom_text_repel(aes(label = row), arrow = grid::arrow()) +
  scale_size_continuous(guide = FALSE) +  
  theme_bw() +
  labs(title = "Isolation Tree based outlier\nVisualised on principal components")

df %>% 
  as.data.frame() %>% 
  mutate(row = row_number(),
         score = iso_score_1,
         PC1 = df_prcomp$x[,1],
         PC2 = df_prcomp$x[,2],
         is_outlier = factor(ifelse(ntile(iso_score_1, 10) >= 10, "O", "N"))) %>% 
  ggplot(aes(PC1, PC2)) +
  geom_point(aes(size = score, color = is_outlier)) +
  ggrepel::geom_text_repel(aes(label = row), arrow = grid::arrow()) +
  scale_color_manual(values = c(O = "red", N = "black"), guide = "none") +
  scale_size_continuous(guide = FALSE) +  
  theme_bw() +
  labs(title = "Isolation Tree based outlier\nTop-10%")
```
Let's also try a forest (ensemble) approach.  Refer @fig-an14, we can see that scores are modified using ensemble methods.  However, when we gradually increase number of trees, these scores will stabilise.

::: {#fig-an14}

```{r warning=FALSE, message=FALSE, fig.align='center', fig.show='hold', out.width="48%"}
iso_100 <- iForest(df, nt = 100, phi = 100)

# Generate scores

iso_score_100 <- predict(iso_100, df)

# View Results
df %>% 
  as.data.frame() %>% 
  mutate(row = row_number(),
         score = iso_score_100,
         PC1 = df_prcomp$x[,1],
         PC2 = df_prcomp$x[,2],
         is_outlier = factor(ifelse(ntile(iso_score_100, 10) >= 10, "O", "N"))) %>% 
  ggplot(aes(PC1, PC2)) +
  geom_point(aes(size = score, color = is_outlier)) +
  ggrepel::geom_text_repel(aes(label = row), arrow = grid::arrow()) +
  scale_color_manual(values = c(O = "red", N = "black"), guide = "none") +
  scale_size_continuous(guide = FALSE) +  
  theme_bw() +
  labs(title = "Isolation Tree based outlier\nNumber of Trees = 100")

plot(iso_score_1, iso_score_100, xlim = c(0, 1), ylim = c(0, 1), 
     main = "Comparision of Tree Vs. Forest Method")
abline(a = 0, b = 1)

```

100 Isolation Trees (Left) Comparison of 1 vs. 100 trees (Right)
:::

**Contour plots:**  We can also visualise the results of scores of anomaly detection, using contour plots.  See the plot in @fig-an16.

```{r fig-an16, warning=FALSE, message=FALSE, fig.align='center', fig.show='hold', fig.cap="Contour Plot", out.width="90%"}
# Create PRCOMP data
df_grid <- data.frame(
  PC1 = df_prcomp$x[,1],
  PC2 = df_prcomp$x[,2]
)

# Create Sequences
pc1_seq <- seq(min(df_prcomp$x[,1]), max(df_prcomp$x[,1]), length.out = 25)
pc2_seq <- seq(min(df_prcomp$x[,2]), max(df_prcomp$x[,2]), length.out = 25)
# Create Grid
my_grid <- expand.grid(PC1 = pc1_seq, PC2 = pc2_seq)

# Create model for Pr comp
iso_model <- iForest(df_grid, nt = 100, phi = 100)
# append scores
my_grid$scores <- predict(iso_model, my_grid)

# Draw Plot
library(lattice)
contourplot(scores ~ PC1 + PC2, data = my_grid, region = TRUE)
```

**Including categorical variables** One benefit of using forest tree method of anomaly detection, is that we can include categorical values also.  Only condition is that these should be of `factor` type.

**Problem Statement-2:** Let's now try to find out anomalies on full `iris` data.  We can check column types before proceeding.

```{r an15}
sapply(iris, class)
```

Our condition is met.  So we can proceed directly to build a decision tree/Forest.  For sake of simplicity, let's build a simple tree (one). Refer @fig-an18.

```{r fig-an18, warning=FALSE, message=FALSE, fig.align='center', fig.show='hold', fig.cap="Including categorical variable", out.width="98%"}
# New Iforest Model
iso_model_new <- iForest(iris, nt = 1, phi = 100)
new_scores <- predict(iso_model_new, iris)
head(new_scores)

# Full PRCOMP for Visual
iris_pr <- prcomp(iris[, 1:4])

data.frame(
  PC1 = iris_pr$x[,1],
  PC2 = iris_pr$x[,2]
) %>% 
  mutate(score = new_scores,
         Species = iris$Species) %>% 
  ggplot(aes(PC1, PC2, color = Species)) +
  geom_point(aes(size = score)) +
  guides(size = FALSE) +
  theme_bw() +
  labs(title = "Decision Tree\nVisualised on Principal Components")
```

## Including categorical variables in LOF
We can also include categorical variable in `Local outlier factor` using `gower distance` calculation method.  Gower method lets us calculate distance between categorical observations, most importantly when the categories are not ordered.  To calculate it we can use function `daisy` from library `cluster` in R. Let's see LOF score calculation on the above example. The plot generated can be seen in @fig-an20.

```{r fig-an20, warning=FALSE, message=FALSE, fig.align='center', fig.show='hold', fig.cap="Including categorical variable", out.width="90%"}
library(cluster)
iris_dist <- daisy(iris, metric = "gower")
iris_lof <- lof(iris_dist, minPts = 5)

data.frame(
  PC1 = iris_pr$x[,1],
  PC2 = iris_pr$x[,2]
) %>% 
  mutate(score = iris_lof,
         Species = iris$Species) %>% 
  ggplot(aes(PC1, PC2, color = Species)) +
  geom_point(aes(size = score)) +
  guides(size = "none") +
  theme_bw() +
  labs(title = "LOF Scores\nVisualised on Principal Components")
```

### How does Gower distance work?

kNN and LOF need a *distance* between every pair of records. For numbers this is easy. But what is the distance between "Air" and "Sleeper", or between "Group A" and "Group C"? Gower distance answers this in a simple way.

-   For a **numeric** column, it takes the difference between the two values and divides it by the range (maximum minus minimum) of that column. So the result is always between 0 and 1.
-   For a **categorical** column, the distance is 0 if the two values are the same, and 1 if they are different.
-   The final distance is the average of these column-wise distances.

So every column contributes equally, and the final distance is always between 0 and 1. This is why we did not have to scale the data before using `daisy()`.

### Why categories matter, an audit example

Let us see why including categorical columns can make a real difference. Consider travel allowance (TA) claims of employees. Each claim records the grade of the employee, the mode of travel, the distance travelled and the amount claimed. Suppose the entitlement rules are as below.

| Grade | Entitled to travel by |
|---------|----------------------|
| Group A | Air or AC-2 |
| Group B | AC-2 or AC-3 |
| Group C | AC-3 or Sleeper |

: Travel entitlement in our example {#tbl-taentitle}

We have about 300 claims, created in the folded code below. Three problem claims have been planted. Two Group C employees have claimed *air* fare. Their amounts are quite normal *for an air journey* of that distance. The third claim is an AC-3 journey with a hugely inflated amount.

```{r}
#| code-fold: true
#| code-summary: "Code to create the TA claims data"
set.seed(2611)

entitled <- list(`Group A` = c("Air", "AC-2"),
                 `Group B` = c("AC-2", "AC-3"),
                 `Group C` = c("AC-3", "Sleeper"))
rate_per_km <- c(Air = 6, `AC-2` = 2.6, `AC-3` = 1.8, Sleeper = 0.7)

n <- 300
claims <- tibble(grade = sample(names(entitled), n, replace = TRUE,
                                prob = c(0.25, 0.35, 0.4))) |>
  mutate(mode        = map_chr(grade, \(g) sample(entitled[[g]], 1)),
         distance_km = round(runif(n, 200, 1800)),
         amount      = round(distance_km * rate_per_km[mode] * 2 *
                               runif(n, 0.9, 1.15) + 500))

planted <- tibble(grade       = c("Group C", "Group C", "Group B"),
                  mode        = c("Air", "Air", "AC-3"),
                  distance_km = c(1150, 1420, 900),
                  amount      = c(14300, 17600, 41000))

claims <- bind_rows(claims, planted) |>
  mutate(claim_id = sprintf("TA-%03d", row_number()), .before = 1)
```

```{r}
tail(claims, 4)
```

**Step-1: LOF on numeric columns only.** First, let us do what we did earlier in this chapter. We use only the two numeric columns, distance and amount, scale them, and calculate LOF.

```{r}
claims$lof_numeric <- claims |>
  select(distance_km, amount) |>
  scale() |>
  lof(minPts = 10)

claims |>
  arrange(desc(lof_numeric)) |>
  head(5)
```

The inflated AC-3 claim, `TA-303`, is at the top, with a very high score. That is good. But the two Group C air claims are nowhere near the top. This is not surprising. Looking at distance and amount alone, these claims look exactly like the many genuine air claims of Group A officers.

**Step-2: LOF with categorical columns, using Gower distance.** Now we include grade and mode of travel as well. `daisy()` needs categorical columns as factors, so we convert them first.

```{r}
gower_dist <- claims |>
  select(grade, mode, distance_km, amount) |>
  mutate(across(c(grade, mode), as.factor)) |>
  daisy(metric = "gower")

claims$lof_gower <- lof(gower_dist, minPts = 10)

claims |>
  arrange(desc(lof_gower)) |>
  head(5)
```

Now all three planted claims are at the top. A Group C employee travelling by air is a rare *combination* in the data, and Gower distance makes that rarity visible to LOF.

Let us compare the rank of the three claims under both methods. A rank of 1 means the most anomalous claim.

```{r}
claims |>
  mutate(rank_numeric = rank(-lof_numeric),
         rank_gower   = rank(-lof_gower)) |>
  filter(claim_id %in% c("TA-301", "TA-302", "TA-303")) |>
  select(claim_id, grade, mode, amount, rank_numeric, rank_gower)
```

::: callout-tip
## Audit tip

If the rules are known and simple, as in our entitlement table, a direct rule-based check is better. Just join the claims with the entitlement table and list every claim where the mode is not allowed for the grade. Anomaly detection is useful when the rules are not known, are too many to code, or when we want to discover patterns that nobody has thought of. In practice, use both.
:::

## Time Series Anomalies
In time series data, an anomaly/outlier can be termed as a data point which is not following the common collective trend or seasonal or cyclic pattern of the entire data and is significantly distinct from rest of the data.

To calculate/detect anomalies, in R, we make use of package `timetk`.  The package works on `tibble` instead of `time series` data, so we may to prep our data/time series accordingly.

**Problem Statement** Let's find out anomalies, if any on Sunspots data available in base R^[Monthly numbers of sunspots, as from the World Data Center, aka SIDC. This is the version of the data that will occasionally be updated when new counts become available.].

```{r fig-an17, warning=FALSE, message=FALSE, fig.align='center', fig.show='hold', fig.cap="Time Series Anomalies", out.width="90%"}
start_date <- as.Date("1749-01-01")
# one date per month, as many as there are values in the series
date_sequence <- seq(start_date, by = "months", length.out = length(sunspot.month))

sunspot_df <- sunspot.month %>%
  as.data.frame() %>%
  set_names("value") %>%
  mutate(date = date_sequence)

library(timetk)
sunspot_df %>% 
  plot_anomaly_diagnostics(date, value,
                           .interactive = FALSE) 

sunspot_df %>% 
  tk_anomaly_diagnostics(date, value) %>% 
  filter(anomaly == 'Yes')
```

We can see anomalies highlighted in red, in @fig-an17 and anomalies filtered out in code above.  We may also implement Seasonal Hybrid ESD algorithm already discussed above.

### How does this method work?

`tk_anomaly_diagnostics()` works in two steps.

1.  It first splits the series into three parts, using a method called STL decomposition. The **trend** is the long-term rise or fall. The **season** is the pattern which repeats every year, or every week. The **remainder** is whatever is left after removing the trend and season.
2.  It then looks for unusually large values in the remainder only. The usual rule of the boxplot is used, so a value far outside the middle 50% of the remainders is called an anomaly.

So a value is called an anomaly only if it cannot be explained by the trend *and* the usual seasonal pattern. We learnt about these components of a time series in the chapter on time series analysis.

### An audit example, monthly expenditure

Every auditor in government knows the *March rush*. Offices spend much more in March than in other months, to use up the budget before the financial year ends. So a high expenditure in March is expected. What an auditor wants to find is a month which is unusual *even after allowing* for this pattern.

Let us create six years of monthly expenditure of an office, from April 2019 to March 2025, in ₹ lakh. The expenditure grows a little every year, and March is always much higher than other months. We have planted one odd month, October 2023, when some unusual payments were made.

```{r}
#| code-fold: true
#| code-summary: "Code to create the monthly expenditure data"
set.seed(2611)

months <- seq(as.Date("2019-04-01"), as.Date("2025-03-01"), by = "month")

# Seasonal pattern from April to March. March is the rush month.
season <- c(0.8, 0.85, 0.9, 0.9, 0.95, 0.95, 1, 1, 1.05, 1.1, 1.3, 2.4)
position_in_fy <- ((month(months) + 8) %% 12) + 1   # April = 1, ..., March = 12

expenditure <- tibble(
  month = months,
  amount = round(40 * 1.06^(seq_along(months) / 12) * season[position_in_fy] *
                   runif(length(months), 0.92, 1.08), 2)
)

# Planted case. Unusual payments in October 2023.
odd <- expenditure$month == as.Date("2023-10-01")
expenditure$amount[odd] <- expenditure$amount[odd] * 2.3
```

```{r}
head(expenditure)
```

**Attempt 1. A simple boxplot rule.** Let us first try what we learnt at the start of this chapter. `boxplot.stats()` returns, in its `out` element, the values which lie outside the whiskers of a boxplot.

```{r}
outliers <- boxplot.stats(expenditure$amount)$out

expenditure |>
  filter(amount %in% outliers)
```

Every single March has been flagged, along with October 2023. Even February 2025 is flagged, only because expenditure has grown over the years and recent months are naturally higher. This is not useful. The March peaks are the normal pattern of this office, and the one really odd month is lost in the crowd.

**Attempt 2. Allowing for the season.** Now let us use `tk_anomaly_diagnostics()`. We tell it that the season repeats every 12 months, with `.frequency = 12`.

```{r}
expenditure |>
  tk_anomaly_diagnostics(month, amount, .frequency = 12, .message = FALSE) |>
  filter(anomaly == "Yes") |>
  select(month, observed, season, trend, remainder)
```

This is much better. October 2023 is flagged. But so are a few early Marches. Why? Look at how the March peak behaves. It is not a fixed *amount* above other months. It is a fixed *proportion*, about 2.4 times a normal month. As total expenditure grows over the years, the March peak grows in rupees too. STL expects a season of the same size every year, so it gets the early Marches slightly wrong.

**Attempt 3. Working on the log scale.** The fix is simple. We take the logarithm of the amounts. On the log scale, "2.4 times" becomes a fixed *difference*, whatever the size of the expenditure. So the seasonal pattern becomes the same size every year.

```{r}
expenditure |>
  mutate(log_amount = log(amount)) |>
  tk_anomaly_diagnostics(month, log_amount, .frequency = 12, .message = FALSE) |>
  filter(anomaly == "Yes") |>
  select(month, observed, remainder)
```

Now only October 2023 is flagged. This is exactly the month we planted. The same result can be seen in @fig-expanom.

```{r}
#| label: fig-expanom
#| fig-cap: "Anomalies in monthly expenditure, on log scale"
#| message: false
expenditure |>
  mutate(log_amount = log(amount)) |>
  plot_anomaly_diagnostics(month, log_amount, .frequency = 12,
                           .interactive = FALSE, .message = FALSE,
                           .title = "Monthly expenditure (log scale)")
```

::: callout-tip
## Audit tip

Use the log scale whenever the ups and downs of a series grow along with its level. This is true of most financial data, like expenditure, revenue, sales or tax collections. A quick look at a line plot of the data will tell you. If the March peaks get taller over the years, use the log.
:::

::: callout-note
## The March rush is itself an audit issue

The seasonal method treats the March rush as *normal*, because it happens every year. But for an auditor, the rush itself may be a finding. Ministry of Finance guidelines on expenditure management require that expenditure in March should not exceed 15% of the budget for the year. We can check this directly, by finding the share of March in each financial year's expenditure.

```{r}
expenditure |>
  mutate(fin_year = if_else(month(month) >= 4, year(month), year(month) - 1)) |>
  summarise(total = sum(amount),
            march = sum(amount[month(month) == 3]),
            .by = fin_year) |>
  mutate(march_share_pct = round(100 * march / total, 1))
```

Here `fin_year` 2019 means the financial year 2019-20, and so on. In every year, March accounts for more than 15% of the annual expenditure.
:::

