# Duplicate Detection {#dup}

*The content is under development and is therefore not final*

## Finding duplicates
Duplicate detection in forensic audit is crucial for identifying fraud, ensuring financial integrity, and maintaining regulatory compliance. It helps uncover fraudulent activities such as double billing, expense reimbursement fraud, and payroll fraud, while also preventing financial misstatements that could mislead stakeholders. Many industries have strict compliance requirements, and duplicate transactions can lead to regulatory violations, fines, or legal consequences. Additionally, some duplicates arise from data entry errors or system glitches, impacting financial accuracy. In procurement and vendor transactions, detecting duplicate invoices helps prevent over-payments, fake vendors, and collusion. Moreover, duplicate transactions can result in tax over-claims, excessive deductions, or unnecessary cash outflows, affecting an organization’s financial health. By automating duplicate detection, audits become more efficient and data-driven, improving accuracy and reducing financial risk.

## Same-Same-Same test (Exact duplicates)


```{r}
library(janitor)
library(tidyverse)
```


## Same same Different test (Near Duplicates)

```{r}
iris %>% 
  filter(n_distinct(Species) > 1, .by = c(Sepal.Length, Sepal.Width)) %>% 
  get_dupes(Sepal.Length, Sepal.Width)
```

