```{r}
#| include: false
source("_common.R")
```

# Process Mining {#sec-process}

Most audit tests look at *transactions*: amounts, dates, payees. But every transaction is the end of a *process*. A purchase starts with an indent, needs a sanction, a tender, a purchase order, receipt and inspection of the goods, a bill, and only then a payment. The rules of the process exist for good reasons. Sanction must come before commitment. Goods must be received and inspected before they are paid for. The person who passes a bill must not be the person who makes the payment.

Today, most of these processes run through computer systems: e-procurement portals, IFMS, ERP systems, hospital management systems. Every system records each step as an **event**: which case it belongs to, what was done, when, and by whom. **Process mining** reads these event records and reconstructs the process *as it actually happened*, file by file. It answers questions that are very hard to answer by looking at transactions alone.

- What paths do the files actually follow, and how many follow the prescribed one?
- Which files skipped a step, or did steps in the wrong order?
- Where does the process wait, and for how long?
- Who did what, and did anyone do two things they should not both do?

In this chapter we use the **bupaR** family of packages: `bupaR` for event logs, `processmapR` for process maps, `edeaR` for statistics, and `processcheckR` for checking rules.

**Prerequisites**

```{r}
#| message: false
#| warning: false
library(tidyverse)
library(bupaR)          # event logs
library(processmapR)    # process maps and trace explorer
library(edeaR)          # process statistics and filters
library(processcheckR)  # rule checks
```

Install them once with `install.packages(c("bupaR", "processmapR", "edeaR", "processcheckR"))`.

## Event logs {#sec-process-eventlog}

Process mining needs an **event log**, a table with at least three columns.

- **Case**: the file or object that goes through the process, like a purchase file number, a pension case number, or a patient's admission number.
- **Activity**: what was done, like "Bill passed".
- **Timestamp**: when it was done.

A fourth column, the **resource**, the user or office who did it, is not essential but is very valuable in audit. Such logs are rarely handed over as such. They are usually extracted from the audit trail tables of a system, which record every action of every user. Asking the auditee's IT section for "the audit trail of the procurement module for the year, with case number, action, date-time and user" is often the first step.

## Our data, a year of purchase files {#sec-process-data}

We have the event log of 600 purchase files of a department for 2024-25, extracted from its procurement system. The prescribed process for a purchase is:

1. Indent raised (stores)
2. Sanction accorded (drawing and disbursing officer)
3. Tender published, and bids evaluated (purchase section), *except* for purchases up to ₹25,000, which in our example may be made without tender
4. Purchase order issued
5. Goods received (stores), and inspection done (technical section)
6. Bill submitted (accounts), bill passed (accounts officer), payment made (cashier)

Bills found defective are returned and resubmitted, which is a normal variation. Five kinds of irregularity are planted, which we will find.

```{r}
#| code-fold: true
#| code-summary: "Code to create the event log"
set.seed(2025)
n <- 600
steps <- c("Indent raised", "Sanction accorded", "Tender published", "Bids evaluated",
           "Purchase order issued", "Goods received", "Inspection done",
           "Bill submitted", "Bill passed", "Payment made")
# average days taken by each step after the previous one
mean_days <- c("Sanction accorded" = 3, "Tender published" = 4, "Bids evaluated" = 15,
               "Purchase order issued" = 5, "Goods received" = 20, "Inspection done" = 2,
               "Bill submitted" = 4, "Bill passed" = 6, "Payment made" = 5)
staff <- list("Indent raised" = sprintf("STORE%02d", 1:4), "Sanction accorded" = c("DDO01", "DDO02"),
              "Tender published" = sprintf("PUR%02d", 1:3), "Bids evaluated" = sprintf("PUR%02d", 1:3),
              "Purchase order issued" = sprintf("PUR%02d", 1:3), "Goods received" = sprintf("STORE%02d", 1:4),
              "Inspection done" = sprintf("TECH%02d", 1:3), "Bill submitted" = sprintf("ACC%02d", 1:4),
              "Bill passed" = c("AO01", "AO02"), "Payment made" = c("CASH01", "CASH02"))

files <- tibble(case_id = sprintf("PF-%04d", 1:n), value = round(rlnorm(n, log(80000), 1.2), -2)) |>
  mutate(start = as.POSIXct("2024-04-01 10:00", tz = "Asia/Kolkata") + sample(0:330, n, TRUE) * 86400)

make_file <- function(case_id, value, start) {
  acts <- if (value <= 25000) steps[-(3:4)] else steps     # small purchases skip the tender
  gaps <- c(0, pmax(0.1, rgamma(length(acts) - 1, 2, 2 / mean_days[acts[-1]])) * 86400)
  tibble(case_id, activity = acts, timestamp = start + cumsum(gaps),
         resource = map_chr(acts, \(a) sample(staff[[a]], 1)), value)
}
log <- pmap(files |> select(case_id, value, start), make_file) |> list_rbind()

# Normal variation. 50 bills returned once and resubmitted, which delays them
returned <- sample(files$case_id, 50)
resubmit <- log |> filter(case_id %in% returned, activity == "Bill submitted")
log <- log |>
  mutate(timestamp = if_else(case_id %in% returned & activity %in% c("Bill passed", "Payment made"),
                             timestamp + 6 * 86400, timestamp)) |>
  bind_rows(resubmit |> mutate(activity = "Bill returned", resource = "AO01",
                               timestamp = timestamp + 2 * 86400),
            resubmit |> mutate(timestamp = timestamp + 5 * 86400))

free <- setdiff(files$case_id, returned)
take <- function(k, pool) { chosen <- sample(pool, k); free <<- setdiff(free, chosen); chosen }
# Planted 1. Payment made before the goods were received and inspected
p_pay_first <- take(12, free)
log <- log |>
  mutate(last = max(timestamp), .by = case_id) |>
  mutate(timestamp = case_when(
    case_id %in% p_pay_first & activity == "Goods received"  ~ last + 10 * 86400,
    case_id %in% p_pay_first & activity == "Inspection done" ~ last + 12 * 86400,
    .default = timestamp)) |>
  select(-last)
# Planted 2. No sanction at all
p_no_sanction <- take(9, free)
log <- log |> filter(!(case_id %in% p_no_sanction & activity == "Sanction accorded"))
# Planted 3. Sanction given after the purchase order (post-facto sanction)
p_post_facto <- take(7, free)
log <- log |>
  mutate(po_time = timestamp[activity == "Purchase order issued"], .by = case_id) |>
  mutate(timestamp = if_else(case_id %in% p_post_facto & activity == "Sanction accorded",
                             po_time + 3 * 86400, timestamp)) |>
  select(-po_time)
# Planted 4. The same officer passed the bill and made the payment
p_same_user <- take(10, free)
log <- log |>
  mutate(passer = resource[activity == "Bill passed"][1], .by = case_id) |>
  mutate(resource = if_else(case_id %in% p_same_user & activity == "Payment made", passer, resource)) |>
  select(-passer)
# Planted 5. Purchases above Rs 25,000 made without tender
big <- files |> filter(value > 60000, case_id %in% free) |> pull(case_id)
p_no_tender <- take(6, big)
log <- log |> filter(!(case_id %in% p_no_tender & activity %in% c("Tender published", "Bids evaluated")))

log <- log |> arrange(case_id, timestamp)
```

```{r}
log
```

`eventlog()` turns this table into an event log, telling bupaR which column is which. It needs two more columns, which bupaR uses for logs that record the start and the end of each activity separately. In our log each event is recorded once, when it was completed, so we add an `activity_instance` number for each row and a `status` column that simply says `"complete"`.

```{r}
purchases <- log |>
  mutate(activity_instance = row_number(), status = "complete") |>
  eventlog(case_id = "case_id", activity_id = "activity",
           activity_instance_id = "activity_instance", lifecycle_id = "status",
           timestamp = "timestamp", resource_id = "resource")
purchases
```

The printout tells us the essentials: `r format(nrow(log), big.mark = ",")` events, `r n_cases(purchases)` cases, `r n_activities(purchases)` distinct activities, `r n_resources(purchases)` users, and the number of distinct *traces*, which we discuss next.

## What paths do the files take? {#sec-process-traces}

A **trace** is the sequence of activities of one case, in time order. If every file followed the prescribed process, there would be just two traces: one with a tender, and one for small purchases without. Let us see how many there are, and how common each is.

```{r}
n_traces(purchases)

purchases |>
  trace_coverage("trace") |>
  mutate(steps = str_count(trace, ",") + 1) |>
  select(steps, absolute, relative, cum_sum)
```

```{r}
#| include: false
tc <- purchases |> trace_coverage("trace")
```

There are `r n_traces(purchases)` traces. The two most common, the prescribed paths, cover `r round(100 * tc$cum_sum[2])`% of the files. Everything else is a *variant*, and every variant needs an explanation. The **trace explorer** draws them, one row per trace, with each activity as a coloured box, so that the differences can be seen.

```{r}
#| fig-cap: "The traces of the purchase files, most frequent first"
#| fig-height: 6
#| fig-width: 10
purchases |>
  trace_explorer(n_traces = 12, label_size = 2.5)
```

Read each row from left to right. The top two rows are the prescribed paths. Below them we can see bills returned and resubmitted, the normal variation. Then come the real exceptions: rows without a sanction box, rows with payment before receipt of goods, rows where the sanction comes after the purchase order, and rows without the tender boxes. Each is a small number of files.

## A process map {#sec-process-map}

A **process map** draws traces together as one diagram. Each activity is a box, and each arrow shows how often one activity was directly followed by another. Thick, dark arrows are the main road; thin ones are the side roads. `rankdir = "TB"` draws the map from top to bottom. Maps are drawn only 300 pixels high by default, which is too small for a process of ten steps, so we write a two-line function to set the height of a map (an *htmlwidget*, as interactive charts in R are called).

```{r}
set_height <- function(widget, pixels) {
  widget$height <- pixels
  widget
}
```

Drawn with every variant at once, a map soon looks like a plate of spaghetti, since every rare deviation adds arrows going back and forth. It is clearer to draw the main road first. `filter_trace_frequency(percentage = 0.9)` keeps the most frequent traces that together cover 90% of the cases.

```{r}
#| fig-cap: "Process map of the most frequent traces, with case counts"
#| out-width: "100%"
purchases |>
  filter_trace_frequency(percentage = 0.9) |>
  process_map(type = frequency("absolute-case"), rankdir = "TB") |>
  set_height(700)
```

This is the prescribed process, as most files follow it: the two routes, with and without a tender, joining at the purchase order, and the loop of bills returned and resubmitted. The map is interactive in the HTML version of this book: it can be zoomed and moved.

Now the opposite: only the files *outside* the main road, with `reverse = TRUE`. This is the map an auditor is most interested in.

```{r}
#| fig-cap: "Process map of the less frequent traces"
#| out-width: "100%"
purchases |>
  filter_trace_frequency(percentage = 0.9, reverse = TRUE) |>
  process_map(type = frequency("absolute-case"), rankdir = "TB") |>
  set_height(700)
```

Here the deviations stand out: arrows from *Indent raised* straight to *Tender published* or to *Purchase order issued*, skipping the sanction; arrows from *Purchase order issued* back to *Sanction accorded*; and arrows from *Payment made* to *Goods received*, payment before receipt. Each arrow is labelled with the number of files taking it.

The same map can show **time** instead of counts. With `performance()`, each arrow shows the median time between the two activities.

```{r}
#| fig-cap: "Process map with the median days taken between activities"
#| out-width: "100%"
purchases |>
  filter_trace_frequency(percentage = 0.9) |>
  process_map(type = performance(median, "days"), rankdir = "TB") |>
  set_height(700)
```

```{r}
#| include: false
tt <- purchases |> throughput_time("log", units = "days")
```

The longest waits are between issuing the purchase order and receiving the goods, and between publishing the tender and evaluating the bids. From indent to payment, a file takes a median of `r round(as.numeric(tt$median))` days, and the slowest took `r round(as.numeric(tt$max))`.

::: callout-note
Other filters help to focus a map too: `filter_case()` for particular files, `filter_activity()` to keep only some activities, and `filter_time_period()` for part of the year. A map of each part is usually clearer than one map of everything.
:::

## Checking the rules {#sec-process-rules}

Maps show the overall picture. For audit findings we need the exact list of cases breaking each rule. `processcheckR` expresses process rules declaratively. Some of its rule types are:

| Rule | Meaning |
|----------------|--------------------------------------------|
| `contains("A")` | A occurs in the case |
| `precedence("A", "B")` | Every B is preceded by an A |
| `response("A", "B")` | Every A is followed by a B |
| `starts("A")`, `ends("B")` | The case starts with A, or ends with B |
| `absent("A")` | A does not occur |
| `and("A", "B")`, `xor("A", "B")` | Both occur, or exactly one of them |

: Some rule types in processcheckR {#tbl-process-rules}

`check_rules()` checks several rules at once, adding a `TRUE`/`FALSE` column for each rule to every event. We keep one row per case.

```{r}
rule_results <- purchases |>
  check_rules(
    has_sanction      = contains("Sanction accorded"),
    sanction_before_po = precedence("Sanction accorded", "Purchase order issued"),
    goods_before_pay  = precedence("Goods received", "Payment made"),
    inspect_before_pay = precedence("Inspection done", "Payment made")
  ) |>
  as_tibble() |>
  distinct(case_id, has_sanction, sanction_before_po, goods_before_pay, inspect_before_pay)

rule_results |>
  summarise(across(-case_id, \(ok) sum(!ok))) |>
  pivot_longer(everything(), names_to = "rule", values_to = "cases_failing")
```

```{r}
#| include: false
rr <- rule_results |> summarise(across(-case_id, \(ok) sum(!ok)))
```

Note that `sanction_before_po` fails for `r rr$sanction_before_po` files. It includes the `r rr$has_sanction` files with no sanction at all, since a purchase order with *no* sanction before it also breaks the rule. The remaining `r rr$sanction_before_po - rr$has_sanction` had a sanction, but only *after* the order was placed, a post-facto sanction.

```{r}
rule_results |>
  filter(has_sanction, !sanction_before_po) |>
  select(case_id) |>
  left_join(log |> filter(activity %in% c("Sanction accorded", "Purchase order issued")),
            by = "case_id") |>
  select(case_id, activity, timestamp) |>
  head(6)
```

The rule about tenders depends on the value of the purchase, which `processcheckR` does not know about. Such rules are easiest with `dplyr` on the log itself: files above ₹25,000 that have no *Tender published* event.

```{r}
log |>
  summarise(value = first(value), tendered = any(activity == "Tender published"), .by = case_id) |>
  filter(value > 25000, !tendered) |>
  arrange(desc(value))
```

These purchases should have been made through a tender. They may be cases of splitting, of urgency, or of deliberate avoidance of competition, and the files must say which.

## Who did what? Segregation of duties {#sec-process-sod}

The resource column lets us check **segregation of duties**: certain pairs of actions on the same file must be done by different people. The officer who passes a bill must not also make the payment. With `pivot_wider()` (@sec-tidyr), we put the user of each activity side by side, one row per file.

```{r}
users_by_file <- log |>
  filter(activity %in% c("Bill passed", "Payment made")) |>
  summarise(user = first(resource), .by = c(case_id, activity)) |>
  pivot_wider(names_from = activity, values_from = user)

users_by_file |> filter(`Bill passed` == `Payment made`)
```

In these files, one officer passed the bill and then paid it, with no second person to check. The same approach checks any pair of incompatible duties, such as the person who raised the indent also receiving the goods.

A **resource map**, a process map of who hands the work to whom, shows the same at a glance. Each box is a user, and each arrow a handover.

```{r}
#| fig-cap: "Handover of work between users"
#| out-width: "100%"
purchases |>
  filter_activity(c("Bill passed", "Payment made")) |>
  resource_map(rankdir = "TB") |>
  set_height(450)
```

Arrows from AO01 or AO02 to CASH01 or CASH02 are the normal handovers. An arrow from an accounts officer to *themselves* is a file where the same person did both.

## How long do things take? {#sec-process-time}

The time between events is also an audit measure. `throughput_time()` gives the time from the first to the last event of each case.

```{r}
case_times <- purchases |> throughput_time("case", units = "days")
case_times |> arrange(desc(throughput_time)) |> head(5)
```

Bills that are returned and resubmitted take longer. Comparing the two groups shows the cost of defective bills, here a few days on each file.

```{r}
case_times |>
  mutate(returned = case_id %in% (log |> filter(activity == "Bill returned") |> pull(case_id))) |>
  summarise(files = n(), median_days = round(median(as.numeric(throughput_time))), .by = returned)
```

A **dotted chart** shows every event of every file against the calendar, one line per file. It reveals patterns over the year, like a rush of payments in March, or a period when files piled up.

```{r}
#| fig-cap: "Dotted chart of the purchase files over the year"
#| fig-height: 5
purchases |>
  filter_activity(c("Indent raised", "Purchase order issued", "Payment made")) |>
  dotted_chart(x = "absolute", sort = "start")
```

## Using process mining in audit {#sec-process-audit}

The same approach applies to any process run through a computer system.

| Process | Case | Typical checks |
|------------------------|----------------|--------------------------------|
| Procurement | Purchase file, tender | Sanction before order, tender above limits, goods before payment |
| Pension sanction | Pension case | Verification steps done, time from retirement to first payment |
| Hospital insurance claims | Admission | Pre-authorisation before treatment, discharge before claim |
| Licence or permit issue | Application | Inspection before approval, time taken against service guarantee |
| Grievance redressal | Grievance | Closed without action, reopened cases, delays |
| Revenue refund | Refund claim | Verification before sanction, same user sanctioning and paying |

: Process mining in different audit areas {#tbl-process-uses}

::: callout-tip
## Audit tip

Ask for the system's audit trail, not a report prepared from it. A report shows what the system's designers thought worth reporting; the audit trail shows everything that happened, including actions later undone. Check that it is complete: every case in the transaction tables should appear in the log, timestamps should be in one time zone (@sec-lubri-hours), and user IDs should match the list of authorised users.
:::

::: callout-caution
An event log shows what was *recorded* in the system, not necessarily what happened. A sanction given on paper but entered in the system late will appear post-facto. A payment entered under another officer's login will show that officer as the resource. Treat every deviation as a question for the auditee, and confirm it from the files, before reporting it.
:::

## A suggested workflow

1. **Get the event log**: the audit trail of the system, with case, activity, timestamp and user.
2. **Check the log**: completeness against the transaction tables, time zones, and spelling of activity names.
3. **Create the event log object** with `eventlog()`.
4. **Explore the traces** with `trace_coverage()` and `trace_explorer()`, and draw a process map.
5. **Write the rules** of the process and check them with `check_rules()`, adding `dplyr` checks for rules that depend on values.
6. **Check segregation of duties** from the resource column.
7. **Measure times** with `throughput_time()` and performance maps.
8. **Confirm deviations from the files** before reporting.

::: callout-important
## Remember

- An event log has, for each event, a case, an activity, a timestamp and, ideally, a user.
- A trace is the sequence of activities of one case; variants from the prescribed trace need explanation.
- Process maps show the actual process, with counts or times on the arrows.
- `processcheckR` checks rules like `contains()` and `precedence()` case by case.
- The user column allows segregation-of-duties checks.
- A deviation in the log is a question, not yet a finding.
:::

## Summary of functions used

| Function | Package | What it does |
|-----------------------|----------------|--------------------------------|
| `eventlog()` | bupaR | Create an event log from a table with case, activity, timestamp and resource |
| `n_cases()`, `n_activities()`, `n_resources()`, `n_traces()` | bupaR | Counts in the event log |
| `trace_coverage()` | edeaR | Frequency of each trace |
| `throughput_time()` | edeaR | Time from first to last event, for the log or each case |
| `filter_activity()`, `filter_trace_frequency()`, `filter_case()` | edeaR, bupaR | Keep part of the event log |
| `trace_explorer()` | processmapR | Draw the traces, most frequent first |
| `process_map(type = frequency(), performance())` | processmapR | Process map with counts or times |
| `resource_map()` | processmapR | Map of handovers between users |
| `dotted_chart()` | processmapR | Events of every case against time |
| `check_rules()`, `contains()`, `precedence()` | processcheckR | Check process rules, case by case |
| `pivot_wider()` | tidyr | Users of different activities side by side, for segregation of duties |

: Functions used in this chapter {#tbl-processfunctions}
