# Sentiment Analysis through Word-Cloud

## Step-1:Prepare data and load libraries

As an example we will create a word cloud with Budget Speech made by Finance Minister during her Budget speech^[Data Source: [Indian Budget Portal](https://www.indiabudget.gov.in/)] 2022-23.  All of the budget speech is available in file called `budget.txt`.

Load Libraries

```{r message=FALSE, warning=FALSE}
library(tidyverse)
library(tidytext) #install.packages("tidytext")
library(wordcloud) #install.packages("wordcloud")
library(ggtext)
library(ggalt)
library(ggthemes)
library(ggpubr)
```


Load data
```{r warning=FALSE,message=FALSE}
dat <- read.table('data/budget.txt', header = FALSE, fill = TRUE)
```

## Step-2: Reshape the .txt data frame into one column

Above steps will create one row per line.  Let's create a tidy data frame out of this data.

```{r echo=TRUE}
tidy_dat <- dat %>% 
  pivot_longer(everything(), values_to = 'word', names_to = NULL)
```

## Step-3: Tokenize the data/words
To tokenize the words we will use function `unnest_tokens()` from `tidytext` library.  As a further step we will have a count of each word, using `dplyr::count` which will create a column `n` against each word.
```{r echo=TRUE}
tokens <- tidy_dat %>% 
  unnest_tokens(word, word) %>% 
  count(word, sort = TRUE) 
```
## Step-4: Clean stop words
The library `tidytext` has a default database which can eliminate stop words from above data.  Let's load this default stop words data.
```{r echo=TRUE}
data("stop_words")
```
We may then remove stop words using `dplyr::anti_join`.
```{r echo=TRUE}
tokens_clean <- tokens %>%
  anti_join(stop_words, by='word') %>% 
  # remove numbers
  filter(!str_detect(word, "^[0-9]"))
```
We may remove additional stop words those specific to this data/input.  To have an idea of these stop words, we may at firt, skip this step altogether and proceed to generate word cloud in next step directly.  After having a first look, we can identify and then remove these additional stop words seen in first round(s).
```{r echo=TRUE}
uni_sw <- data.frame(word = c("cent", "pm", "crore", 
                              "lakh", "set",
                              "level", "sir"))

tokens_clean <- tokens_clean %>% 
  anti_join(uni_sw, by = "word")
```

## Step-5: Plot/generate word cloud
Output/Word cloud of following code can be seen in figure \@ref(fig:wordcloud).

```{r wordcloud, fig.cap= "Word Cloud of FM's Budget Speech 2022", fig.align='center',fig.align='center', message=FALSE, warning=FALSE}
pal <- RColorBrewer::brewer.pal(8,"Dark2")

# plot the 40 most common words
tokens_clean %>% 
  with(wordcloud(word, 
                 n, 
                 random.order = FALSE, 
                 max.words = 40, 
                 colors=pal,
                 scale=c(2.5, .5)))
```



