```{r}
#| include: false
source("_common.R")
```

# Various Datasets, in base R, used in this book
```{r echo=FALSE}
helpExtract <- function(Function, section = "Usage", type = "m_code", ...) {
  A <- deparse(substitute(Function))
  x <- capture.output(tools:::Rd2txt(utils:::.getHelpFile(help(A, ...)),
                                     options = list(sectionIndent = 0)))
  B <- grep("^_", x)                    ## section start lines
  x <- gsub("_\b", "", x, fixed = TRUE) ## remove "_\b"
  X <- rep(FALSE, length(x))
  X[B] <- 1
  out <- split(x, cumsum(X))
  out <- out[[which(sapply(out, function(x) 
    grepl(section, x[1], fixed = TRUE)))]][-c(1, 2)]
  while(TRUE) {
    out <- out[-length(out)]
    if (out[length(out)] != "") { break }
  } 

  switch(
    type,
    m_code = c("```r", out, "```"),
    s_code = c("<<>>=", out, "@"),
    m_text = paste("    ", out, collapse = "\n"),
    s_text = c("\\begin{verbatim}", out, "\\end{verbatim}"),
    stop("`type` must be either `m_code`, `s_code`, `m_text`, or `s_text`")
  )
}
```

- `Nile`:
```{r, echo=FALSE, results='asis'}
cat(helpExtract(Nile, section = "Desc", type = "m_text"))
```

- `lynx`: 
```{r, echo=FALSE, results='asis'}
cat(helpExtract(lynx, section = "Desc", type = "m_text"))
```

- `mtcars`: 
```{r, echo=FALSE, results='asis'}
cat(helpExtract(mtcars, section = "Desc", type = "m_text"))
```

- `sunspot.month`:
```{r, echo=FALSE, results='asis'}
cat(helpExtract(sunspot.month, section = "Desc", type = "m_text"))
```

- `nhtemp`:
```{r, echo=FALSE, results='asis'}
cat(helpExtract(nhtemp, section = "Desc", type = "m_text"))
```

- `airquality`:
```{r, echo=FALSE, results='asis'}
cat(helpExtract(airquality, section = "Desc", type = "m_text"))
```

- `JohnsonJohnson`:
```{r, echo=FALSE, results='asis'}
cat(helpExtract(JohnsonJohnson, section = "Desc", type = "m_text"))
```

- `iris`:
```{r, echo=FALSE, results='asis'}
cat(helpExtract(iris, section = "Desc", type = "m_text"))
```

- `attitude`:
```{r, echo=FALSE, results='asis'}
cat(helpExtract(attitude, section = "Desc", type = "m_text"))
```

- `longley`:
```{r, echo=FALSE, results='asis'}
cat(helpExtract(longley, section = "Desc", type = "m_text"))
```

- `USArrests`:
```{r, echo=FALSE, results='asis'}
cat(helpExtract(USArrests, section = "Desc", type = "m_text"))
```

- `state.x77`:
```{r, echo=FALSE, results='asis'}
cat(helpExtract(state.x77, section = "Desc", type = "m_text"))
```

## Datasets created for this book {#sec-ssydata}

Some chapters use simulated data, created only for this book. All names and numbers in them are made up, and do not relate to any real entity.

**Sarvajan Swasthya Yojana (SSY).** An imaginary state health insurance scheme, used in @sec-pca and the following chapters on clustering and association rules. The files are in the `Data` folder of the book's repository, and the code that creates them is in `tools/make_health_scheme_data.R`.

| File | One row per | Main columns |
|-------------------|----------------|----------------------------------------|
| `ssy_hospitals.csv` | Empanelled hospital | `hospital_id`, `hospital_name`, `type` (Public or Private), `district`, `beds` |
| `ssy_packages.csv` | Treatment package | `package_code`, `package_name`, `specialty`, `rate`, `surgical`, `day_care`, `typical_los` |
| `ssy_claims.csv` | Claim paid in 2024-25 | `claim_id`, `hospital_id`, `patient_id`, `gender`, `age`, `admission_date`, `discharge_date`, `packages` (codes separated by `;`, main package first), `claim_amount` |

: Files of the SSY dataset {#tbl-ssyfiles}

Other chapters, such as those on duplicates, Nigrini's tests and gaps, create their own small data sets within the chapter. The code is shown, folded, in each chapter.
