```{r}
#| include: false
source("_common.R")
```

# Datasets Used in This Book {#sec-datasets}

The book uses two kinds of data: datasets that come with R and its packages, which are convenient for showing how a function works, and data simulated for this book, which carry the audit examples.

## Datasets that come with R {#sec-builtin-data}

The following datasets are available in every installation of R. Their descriptions below are taken from R's own help pages.
```{r echo=FALSE}
helpExtract <- function(Function, section = "Usage", type = "m_code", ...) {
  A <- deparse(substitute(Function))
  x <- capture.output(tools:::Rd2txt(utils:::.getHelpFile(help(A, ...)),
                                     options = list(sectionIndent = 0)))
  B <- grep("^_", x)                    ## section start lines
  x <- gsub("_\b", "", x, fixed = TRUE) ## remove "_\b"
  X <- rep(FALSE, length(x))
  X[B] <- 1
  out <- split(x, cumsum(X))
  out <- out[[which(sapply(out, function(x) 
    grepl(section, x[1], fixed = TRUE)))]][-c(1, 2)]
  while(TRUE) {
    out <- out[-length(out)]
    if (out[length(out)] != "") { break }
  } 

  switch(
    type,
    m_code = c("```r", out, "```"),
    s_code = c("<<>>=", out, "@"),
    m_text = paste("    ", out, collapse = "\n"),
    s_text = c("\\begin{verbatim}", out, "\\end{verbatim}"),
    stop("`type` must be either `m_code`, `s_code`, `m_text`, or `s_text`")
  )
}
```

- `anscombe`:
```{r, echo=FALSE, results='asis'}
cat(helpExtract(anscombe, section = "Desc", type = "m_text"))
```

- `iris`:
```{r, echo=FALSE, results='asis'}
cat(helpExtract(iris, section = "Desc", type = "m_text"))
```

- `nhtemp`:
```{r, echo=FALSE, results='asis'}
cat(helpExtract(nhtemp, section = "Desc", type = "m_text"))
```

- `Nile`:
```{r, echo=FALSE, results='asis'}
cat(helpExtract(Nile, section = "Desc", type = "m_text"))
```

The appendix on colours also uses `mpg`, a dataset of car models that comes with the package `ggplot2`, for a quick demonstration of palettes.

## Data simulated for this book {#sec-ssydata}

Most audit examples use data simulated in R, because real audit data cannot be published. The data is made to look like what auditors receive, and irregularities are *planted* in it for the analysis to find. All names of persons, firms, offices and places are made up, and do not relate to any real entity.

In most chapters, the code that creates the data is shown, folded, at the start of the example, so that readers can see exactly what was planted, change it, and run it again. The table lists these datasets, chapter by chapter, in the order of the book. Chapters which use only a few values typed in by hand are left out.

| Chapter | Data | What is planted |
|----------------|--------------------------|--------------------------------------|
| @sec-basics | A bill register read from Excel | Amounts typed as text, one as "NIL"; an account number that lost its leading zero; round amounts; bills paid after more than 30 days |
| @sec-subset | A payment register of one division; a list of approved firms | Payments above ₹5 lakh; a missing and a negative amount; a firm not on the approved list |
| @sec-basefuns | A voucher register of four treasuries | Bills split just below ₹50,000; duplicate bill numbers; a negative amount; missing offices; cheques missing from the series |
| @sec-pipes | A payment register of four offices | Six bills of one firm, each just below ₹1 lakh, within a fortnight |
| @sec-fp | Vouchers of four offices; quarter-wise expenditure | Large payments at one office in the second half of March |
| @sec-read | Files received from treasuries, in CSV, text and Excel | Large payments to one firm; amounts with Indian commas or "-"; a voucher missing from a file |
| @sec-clean | Bills of a division; a messy office file of payments | A bill paid twice; title, blank and total rows; inconsistent names; amounts stored as text; four date formats |
| @sec-pdfxl | A bank statement in PDF; sanction orders in PDF; a district statement in Excel | An altered balance; a duplicate order number; a sanction beyond delegated powers; a district spending over budget; a typed-over total; highlighted cells |
| @sec-merge | A pension workbook with a sheet for each district; district salary files; block returns and a district statement | A district missing a month; a return short of its last muster roll; a file with its own column headings; a file not received; a file saved twice |
| @sec-DPLYRR | A payment register and a vendor master; works and their running bills | Bills just below ₹1 lakh; GST at 28% instead of 18%; a negative TDS; works paid beyond the sanction |
| @sec-joins | Payment files of three districts and a vendor master; tour claims and leave records; returns due and received; bills and rate revisions | A vendor paid but not in the master; overlapping tour claims; a tour during leave; returns not received; bills charged at old rates after a revision |
| @sec-tidyr | A month-wise district statement | A March rush; months not reported |
| @sec-lubridate | A bill register; a log of e-payment approvals | Payments on Sundays and holidays; a bill paid before it was received; approvals at night and at weekends by one user |
| @sec-duckdb | 24 lakh treasury vouchers in 12 files | Duplicate payments; one account receiving the salaries of 14 employees |
| @sec-strng | Small tables of vendor names, bank narrations and a salary upload | Name variants of one vendor; an invoice paid twice; employee codes that lost their leading zeros |
| @sec-regex | A vendor master; Aadhaar-like numbers | Invalid PAN, GSTIN, IFSC, mobile numbers and emails; vendors sharing one email; made-up and mistyped numbers |
| @sec-rebus | A vendor master; voucher dates | A short PAN; a PAN in lower case; IFSCs with a letter in place of zero; impossible dates |
| @sec-baseplot | A bill register of five offices; scheme spending in 40 blocks | A March rush; bills split just below ₹1 lakh; a block spending far beyond its beneficiaries |
| @sec-ggplot2 | A bill register of a scheme; monthly expenditure; district budgets | A March rush; bills split just below ₹2 lakh; large bills paid within a day; districts spending over budget |
| @sec-factors | Bills of a health mission; payments to vendors | One district's bills three times the usual size; equipment bills rising in February and March; mistyped and unapproved heads |
| @sec-descriptive | A payment ledger; expenditure of tehsil offices | Negative and zero amounts; missing vendors; dates outside the year; repeated voucher numbers; offices with unusual salary and fuel spending |
| @sec-sampling | A population of payments | Misstatements, for monetary unit sampling |
| @sec-tests | Medical reimbursement claims of six offices | One lax officer; a slow office; inflated claims |
| @sec-validate | A month's pay bill and employee master | Wrong DA and HRA; pay outside the scale; retired employees paid; wrong pension scheme; invalid PAN; shared accounts; employees not in the master |
| @sec-benford | Payment vouchers of a works division; payments of ten district offices | Bills just below ₹50,000; a genuine rate contract that distorts the digits; made-up bills in one office |
| @sec-nigrini | Bills of a department | A keying error; an inflated bill; bills below a limit; made-up round amounts |
| @sec-dup | Vendor master and payments | Invoices paid twice; altered invoice numbers; a bill raised again with a suffix; look-alike vendors; repeated rows |
| @sec-strsim | Employees and proprietors of suppliers | Names of the same persons spelt differently; similar names of different persons |
| @sec-linkage | An old-age pension list and a death register | Pensioners still paid after death; namesakes |
| @sec-gaps | Cheque issue registers; cash book closing dates | Missing cheque leaves; a wrongly entered cheque number; an out-of-sequence date; days without closing |
| @sec-process | The event log of 600 purchase files | Payment before receipt; no sanction; post-facto sanction; same officer passing and paying; tenders avoided |
| @sec-regression | Office expenses of block offices; road works; hospitals; salaries of sub-divisional offices | Inflated vehicle hire in two offices; ghost employees; a keying error in staff strength |
| @sec-logistic | Audited and pending GST refund claims | A hidden pattern of risk, mainly credit from suppliers who do not file returns |
| @sec-trees | Works contracts; scholarship decisions | Risky combinations of tender features; one district applying a looser income limit |
| @sec-forest | GST refund claims with many red flags | Risk factors of irregular refunds, mixed with features that do not matter |
| @sec-dea | Community health centres; transport depots | Inefficient units; inflated outputs |
| @sec-pca, @sec-kmeans | The SSY data, described below | Camp hospitals; hospitals upcoding to intensive care |
| @sec-apriori | Claims of the SSY data; bids for 400 tenders | Packages billed again as add-ons; a cartel of four firms |
| @sec-ts | Monthly expenditure of a department; yearly fines, royalty and excise; quarterly commercial tax; stamp duty receipts | A March rush; one March far above even the usual rush; a lasting drop in receipts from one month |
| @sec-anomaly | Stationery bills; health facilities; monthly receipts of a sub-registrar office and monthly expenditure; indicators of 60 vendors; TA claims; supplier rates | A bill far above the vendor's usual; outliers within peer groups; odd combinations; a spike and a shortfall in receipts; a vendor unlike all others; a supplier paid much faster than its peers; a service provider billing like a works contractor; air fares claimed by Group C staff; a lasting rate increase |
| @sec-text | Grievances of the SSY data; journal entries; tender specifications | Complaints about the camp hospitals; suspicious entries; brand-specific specifications |
| @sec-sentiment | Patient feedback of the SSY data; feedback on fair price shops | Negative feedback on the camp hospitals; feedback written by staff |
| @sec-graph | Contractors sharing details; tenders and bids | Two hidden groups of linked firms and a bridge between them; tenders where all bidders are linked |
| @sec-netaudit | Beneficiaries; bids; invoices; employee and vendor masters | A ring of shared IDs; an agent's mobile on many beneficiaries; a cartel; circular trading; employees linked to vendors |
| @sec-maps | Map, hospitals and inspections of the SSY data; housing beneficiaries | Patients of camp hospitals coming from far; inspection photos taken far from the hospital; beneficiaries of different villages photographed at one house |
| @sec-geocoding | Travel claims; works with the quarry and lead claimed | An inflated travel distance; a distant quarry; an inflated lead |
| @sec-revgeocoding | Geotags of completed works | Swapped coordinates; zero and hand-typed geotags; geotags outside India; many works at one spot |
| @sec-reports | Payments of a farmers' assistance scheme | Instalments paid twice; wrong amounts; a shared account |

: Data simulated for the chapters {#tbl-simdata}

### The Sarvajan Swasthya Yojana data {.unnumbered}

Several chapters share one larger dataset, for an imaginary State health insurance scheme, *Sarvajan Swasthya Yojana* (SSY). They are the chapters on principal components (@sec-pca), clustering (@sec-kmeans), association rules (@sec-apriori), text (@sec-text, @sec-sentiment and @sec-textviz) and maps (@sec-maps). Its files are in the `Data` folder of the book's repository. The code that creates them is in `tools/make_health_scheme_data.R`, `tools/make_ssy_text_data.R` and `tools/make_ssy_geo_data.R`.

| File | One row per | Main columns |
|-------------------|----------------|----------------------------------------|
| `ssy_hospitals.csv` | Empanelled hospital | `hospital_id`, `hospital_name`, `type` (Public or Private), `district`, `beds` |
| `ssy_packages.csv` | Treatment package | `package_code`, `package_name`, `specialty`, `rate`, `surgical`, `day_care`, `typical_los` |
| `ssy_claims.csv` | Claim paid in 2024-25 | `claim_id`, `hospital_id`, `patient_id`, `gender`, `age`, `admission_date`, `discharge_date`, `packages` (codes separated by `;`, main package first), `claim_amount` |
| `ssy_grievances.csv` | Grievance filed against a hospital | `grievance_id`, `received_on`, `district`, `hospital_id`, `category`, `text` |
| `ssy_feedback.csv` | Comment left by a patient | `feedback_id`, `hospital_id`, `comment`, `rating` |
| `ssy_map.gpkg` | Map layer | The state and its districts |
| `ssy_hospital_locations.csv` | Empanelled hospital | `hospital_id`, `x_km`, `y_km` |
| `ssy_patient_homes.csv` | Claim | `claim_id`, `hospital_id`, `home_x_km`, `home_y_km` |
| `ssy_inspections.csv` | Inspection of a hospital | `hospital_id`, `inspection_no`, `inspection_date`, `inspector`, `photo_x_km`, `photo_y_km` |

: Files of the SSY dataset {#tbl-ssyfiles}

Planted in it are six hospitals behaving like "camps" (high claims per bed, short stays, weekend admissions, repeated admissions, unbundled packages) and two hospitals upcoding to intensive care, which the chapters find in different ways. The text and map files carry the same story. Grievances and feedback complain about the camp hospitals, their patients come from far away, and some inspection photos were taken far from the hospital.
