```{r}
#| include: false
source("_common.R")
```

# Datasets Used in This Book {#sec-datasets}

The book uses two kinds of data: datasets that come with R and its packages, which are convenient for showing how a function works, and data simulated for this book, which carry the audit examples.

## Datasets that come with R {#sec-builtin-data}

The following datasets are available in every installation of R. Their descriptions below are taken from R's own help pages.
```{r echo=FALSE}
helpExtract <- function(Function, section = "Usage", type = "m_code", ...) {
  A <- deparse(substitute(Function))
  x <- capture.output(tools:::Rd2txt(utils:::.getHelpFile(help(A, ...)),
                                     options = list(sectionIndent = 0)))
  B <- grep("^_", x)                    ## section start lines
  x <- gsub("_\b", "", x, fixed = TRUE) ## remove "_\b"
  X <- rep(FALSE, length(x))
  X[B] <- 1
  out <- split(x, cumsum(X))
  out <- out[[which(sapply(out, function(x) 
    grepl(section, x[1], fixed = TRUE)))]][-c(1, 2)]
  while(TRUE) {
    out <- out[-length(out)]
    if (out[length(out)] != "") { break }
  } 

  switch(
    type,
    m_code = c("```r", out, "```"),
    s_code = c("<<>>=", out, "@"),
    m_text = paste("    ", out, collapse = "\n"),
    s_text = c("\\begin{verbatim}", out, "\\end{verbatim}"),
    stop("`type` must be either `m_code`, `s_code`, `m_text`, or `s_text`")
  )
}
```

- `Nile`:
```{r, echo=FALSE, results='asis'}
cat(helpExtract(Nile, section = "Desc", type = "m_text"))
```

- `lynx`: 
```{r, echo=FALSE, results='asis'}
cat(helpExtract(lynx, section = "Desc", type = "m_text"))
```

- `mtcars`: 
```{r, echo=FALSE, results='asis'}
cat(helpExtract(mtcars, section = "Desc", type = "m_text"))
```

- `sunspot.month`:
```{r, echo=FALSE, results='asis'}
cat(helpExtract(sunspot.month, section = "Desc", type = "m_text"))
```

- `nhtemp`:
```{r, echo=FALSE, results='asis'}
cat(helpExtract(nhtemp, section = "Desc", type = "m_text"))
```

- `airquality`:
```{r, echo=FALSE, results='asis'}
cat(helpExtract(airquality, section = "Desc", type = "m_text"))
```

- `JohnsonJohnson`:
```{r, echo=FALSE, results='asis'}
cat(helpExtract(JohnsonJohnson, section = "Desc", type = "m_text"))
```

- `iris`:
```{r, echo=FALSE, results='asis'}
cat(helpExtract(iris, section = "Desc", type = "m_text"))
```

- `attitude`:
```{r, echo=FALSE, results='asis'}
cat(helpExtract(attitude, section = "Desc", type = "m_text"))
```

- `longley`:
```{r, echo=FALSE, results='asis'}
cat(helpExtract(longley, section = "Desc", type = "m_text"))
```

- `USArrests`:
```{r, echo=FALSE, results='asis'}
cat(helpExtract(USArrests, section = "Desc", type = "m_text"))
```

- `state.x77`:
```{r, echo=FALSE, results='asis'}
cat(helpExtract(state.x77, section = "Desc", type = "m_text"))
```

## Data simulated for this book {#sec-ssydata}

Most audit examples use data simulated in R, because real audit data cannot be published. The data is made to look like what auditors receive, and irregularities are *planted* in it for the analysis to find. All names of persons, firms, offices and places are made up, and do not relate to any real entity.

In most chapters, the code that creates the data is shown, folded, at the start of the example, so that readers can see exactly what was planted, change it, and run it again. The table lists these datasets.

| Chapter | Data | What is planted |
|----------------|--------------------------|--------------------------------------|
| @sec-clean | A messy office file of payments | Inconsistent names, blanks, amounts stored as text |
| @sec-pdfxl | A bank statement in PDF; sanction orders in PDF; a district statement in Excel | An altered balance; a duplicate order number; a sanction beyond delegated powers; a typed-over total; highlighted cells |
| @sec-DPLYRR | Payments against sanctions | Payments in excess of the sanctioned amount |
| @sec-joins | Bills and rate revisions | Bills charged at old rates after a revision |
| @sec-tidyr | A month-wise district statement | A March rush; months not reported |
| @sec-lubridate | A bill register; a log of e-payment approvals | Payments on Sundays and holidays; approvals at night and at weekends by one user |
| @sec-duckdb | 24 lakh treasury vouchers in 12 files | Duplicate payments; one account receiving the salaries of 14 employees |
| @sec-baseplot, @sec-ggplot2 | Bills and district budgets | Bills just below a limit; districts spending over budget |
| @sec-descriptive | A payment ledger | Several data problems, to be found by profiling |
| @sec-sampling | A population of payments | Misstatements, for monetary unit sampling |
| @sec-tests | Medical reimbursement claims of six offices | One lax officer; a slow office; inflated claims |
| @sec-regex | A vendor master; Aadhaar-like numbers | Invalid identifiers; made-up and mistyped numbers |
| @sec-strsim | Employees and proprietors of suppliers | Names of the same persons spelt differently |
| @sec-validate | A month's pay bill and employee master | Wrong DA and HRA; pay outside the scale; retired employees paid; wrong pension scheme; invalid PAN; shared accounts; employees not in the master |
| @sec-nigrini | Bills of a department | A keying error; an inflated bill; bills below a limit; made-up round amounts |
| @sec-dup | Vendor master and payments | Invoices paid twice; altered invoice numbers; look-alike vendors |
| @sec-linkage | An old-age pension list and a death register | Pensioners still paid after death; namesakes |
| @sec-gaps | A cheque issue register | A wrongly entered cheque number; an out-of-sequence date |
| @sec-process | The event log of 600 purchase files | Payment before receipt; no sanction; post-facto sanction; same officer passing and paying; tenders avoided |
| @sec-regression | Salary expenditure of offices | Ghost employees; a keying error in staff strength |
| @sec-logistic, @sec-forest | GST refund claims | Risk factors of irregular refunds |
| @sec-trees | Works contracts; scholarship decisions | Risky combinations of tender features; one district applying a looser income limit |
| @sec-ts | Stamp duty receipts | A drop in receipts from one month |
| @sec-anomaly | Health facilities; PHC indicators; monthly expenditure; supplier rates | Outliers within peer groups; odd combinations; unusual months; a lasting rate increase |
| @sec-dea | Community health centres; transport depots | Inefficient units; inflated outputs |
| @sec-text, @sec-sentiment | Grievances, feedback and tender specifications | Suspicious entries; brand-specific specifications; feedback written by staff |
| @sec-netaudit | Beneficiaries, firms and invoices | A ring of shared IDs; an agent's mobile on many beneficiaries; circular trading |
| @sec-maps | Beneficiaries with photographed locations | Beneficiaries of different villages photographed at one house |
| @sec-reports | Payments of a farmers' assistance scheme | Instalments paid twice; wrong amounts; a shared account |

: Data simulated for the chapters {#tbl-simdata}

### The Sarvajan Swasthya Yojana data {.unnumbered}

The chapters on principal components, clustering, association rules and text analytics share one larger dataset, for an imaginary State health insurance scheme, *Sarvajan Swasthya Yojana* (SSY). Its files are in the `Data` folder of the book's repository, and the code that creates them is in `tools/make_health_scheme_data.R`.

| File | One row per | Main columns |
|-------------------|----------------|----------------------------------------|
| `ssy_hospitals.csv` | Empanelled hospital | `hospital_id`, `hospital_name`, `type` (Public or Private), `district`, `beds` |
| `ssy_packages.csv` | Treatment package | `package_code`, `package_name`, `specialty`, `rate`, `surgical`, `day_care`, `typical_los` |
| `ssy_claims.csv` | Claim paid in 2024-25 | `claim_id`, `hospital_id`, `patient_id`, `gender`, `age`, `admission_date`, `discharge_date`, `packages` (codes separated by `;`, main package first), `claim_amount` |

: Files of the SSY dataset {#tbl-ssyfiles}

Planted in it are six hospitals behaving like "camps" (high claims per bed, short stays, weekend admissions, repeated admissions, unbundled packages) and two hospitals upcoding to intensive care, which the chapters find in different ways.
