```{r}
#| include: false
source("_common.R")
```

# R Packages Used in This Book {#sec-packages}

R comes with a large set of functions, called *base R*. Most of the work in this book also uses *packages*, collections of functions written by others and shared freely, mostly through CRAN, the Comprehensive R Archive Network. @sec-packages-intro explains how packages are installed and loaded. This appendix lists all the packages used in the book, with a single command to install them.

## Packages by part {#sec-packages-list}

| Part | Packages | Used for |
|----------------|--------------------------|--------------------------------------|
| Throughout | `tidyverse` (including `dplyr`, `tidyr`, `readr`, `purrr`, `stringr`, `forcats`, `lubridate`, `ggplot2`, `tibble`) | Reading, transforming, reshaping and charting data |
| I. R Foundations | `purrr`, `magrittr`, `conflicted` | Functional programming, the older pipe, and handling clashing function names |
| II. Getting Data Ready | `readxl`, `writexl`, `openxlsx`, `janitor`, `pdftools`, `tidyxl`, `unpivotr`, `data.table`, `duckdb`, `dbplyr` | Excel and PDF files, cleaning names, large data |
| III. Working with Text Fields | `rebus` | Readable regular expressions |
| IV. Exploring and Sampling | `scales`, `ggrepel`, `patchwork`, `psych`, `skimr`, `GGally`, `PerformanceAnalytics`, `broom` | Charts, descriptive statistics and tidy test results |
| V. Forensic Tests | `validate`, `pointblank`, `benford.analysis`, `stringdist`, `phonics`, `PGRdup`, `fullmetaphone`, `bupaR`, `processmapR`, `edeaR`, `processcheckR` | Rule books, Benford's law, fuzzy matching and phonetic codes, process mining |
| VI. Modelling and Risk Scoring | `broom`, `performance`, `car`, `moderndive`, `nnet`, `rpart`, `rpart.plot`, `ranger`, `deaR` | Regression, classification and efficiency |
| VII. Finding Hidden Patterns | `cluster`, `factoextra`, `arules`, `forecast`, `timetk`, `zoo`, `outliers`, `AnomalyDetection`, `FNN`, `dbscan`, `isotree`, `changepoint` | Principal components, clustering, association rules, time series and anomalies |
| VIII. Beyond Tables | `tidytext`, `SnowballC`, `wordcloud`, `igraph`, `igraphdata`, `ggraph`, `visNetwork`, `sf`, `leaflet`, `tidygeocoder` | Text, networks, maps and geocoding |
| IX. Reporting | `gt`, `openxlsx`, `quarto` | Formatted tables, Excel working papers, rendering reports |
| Appendices | `RColorBrewer`, `colorspace`, `palmerpenguins` | Colour palettes, and an example dataset |

: Packages used in each part of the book {#tbl-packages}

`MASS`, `lattice` and `nnet`, also used in the book, come with every installation of R and need no separate installation.

## Installing all the packages {#sec-packages-install}

The following command installs every CRAN package used in the book. It needs an internet connection, and may take a while the first time. If your office network blocks downloads, see the note on installing packages on an office network in @sec-packages-intro.

```{r}
#| eval: false
install.packages(c(
  "tidyverse", "conflicted", "readxl", "writexl", "openxlsx", "janitor",
  "pdftools", "tidyxl", "unpivotr", "data.table", "duckdb", "dbplyr",
  "scales", "ggrepel", "patchwork", "psych", "skimr", "GGally", "PerformanceAnalytics", "broom",
  "rebus", "stringdist", "phonics", "PGRdup",
  "validate", "pointblank", "benford.analysis", "bupaR", "processmapR", "edeaR", "processcheckR",
  "performance", "car", "moderndive", "rpart", "rpart.plot", "ranger", "cluster", "factoextra",
  "arules", "forecast", "timetk", "zoo", "outliers", "FNN", "dbscan", "isotree", "changepoint", "deaR",
  "tidytext", "SnowballC", "wordcloud",
  "igraph", "igraphdata", "ggraph", "visNetwork",
  "sf", "leaflet", "tidygeocoder",
  "gt", "quarto", "RColorBrewer", "colorspace", "palmerpenguins", "remotes"
))
```

Two packages are not on CRAN, and are installed from GitHub with the package `remotes`.

```{r}
#| eval: false
# fullmetaphone, Double Metaphone with full-length codes, by Atharv Tyagi and the author
remotes::install_github("anilyayavar/dblmetaphone")
# Seasonal Hybrid ESD anomaly detection, by Twitter
remotes::install_github("twitter/AnomalyDetection")
```

The first of these, `fullmetaphone`, is a full implementation of the Double Metaphone algorithm keeping phonetic codes of up to 32 characters. It was developed by Atharv Tyagi and the author, and has been submitted to CRAN. It is used in @sec-strsim, @sec-dup and @sec-linkage.

::: callout-tip
## Keeping an analysis reproducible

Packages are updated from time to time, and occasionally a function changes its behaviour. For an audit that must be reproducible, record the versions used, as shown in @sec-TIDYVE with `sessionInfo()`. For larger projects, the package `renv` keeps a separate, fixed set of package versions for each project, so that an analysis written today runs the same way next year.
:::

## Versions used for this book {#sec-packages-versions}

This book was built with `r R.version.string`, and the following versions of the packages.

```{r}
#| echo: false
pkgs <- c("tidyverse", "dplyr", "tidyr", "readr", "purrr", "stringr", "lubridate", "ggplot2",
          "forcats", "tibble", "conflicted", "readxl", "writexl", "openxlsx", "janitor",
          "pdftools", "tidyxl", "unpivotr", "data.table", "duckdb", "dbplyr",
          "scales", "ggrepel", "patchwork", "psych", "skimr", "GGally", "PerformanceAnalytics",
          "broom", "rebus", "stringdist", "phonics", "PGRdup", "fullmetaphone",
          "validate", "pointblank", "benford.analysis", "bupaR", "processmapR", "edeaR",
          "processcheckR", "performance", "car", "moderndive", "nnet", "rpart", "rpart.plot",
          "ranger", "cluster", "factoextra", "arules", "forecast", "timetk", "zoo", "outliers",
          "AnomalyDetection", "FNN", "dbscan", "isotree", "changepoint", "deaR", "MASS",
          "tidytext", "SnowballC", "wordcloud", "igraph", "igraphdata", "ggraph", "visNetwork",
          "sf", "leaflet", "tidygeocoder", "gt", "quarto", "RColorBrewer", "colorspace",
          "palmerpenguins")
versions <- vapply(pkgs, \(p) tryCatch(as.character(utils::packageVersion(p)),
                                       error = \(e) "not installed"), character(1))
pkgs <- sort(pkgs)
versions <- versions[pkgs]
# lay out in three pairs of columns
k <- ceiling(length(pkgs) / 3)
pad <- function(x) c(x, rep("", 3 * k - length(x)))
tab <- data.frame(matrix(pad(pkgs), ncol = 3), matrix(pad(versions), ncol = 3))[, c(1, 4, 2, 5, 3, 6)]
knitr::kable(tab, col.names = rep(c("Package", "Version"), 3), row.names = FALSE)
```
